### 2024-08-28 1_get_data.qmd
first push

---
title: "Data getter for peekbank analysis"
format: html
---

# Load data

```{r}
library(here)
source(here("helper","common.R"))
source(here("helper","rt_helper.R"))
```

FIXME: we ideally want to figure out a more sustainable way to deal with this downstream dependency.

We load the frame-level data.

```{r}
load(file = here("..", "peekbank-method", "cached_intermediates", "1_d_aoi.Rds"))
```

# Trial level data frame

We have been worrying about windows, but let's just do two windows:

1. classic short: 200 - 2000
2. ICC maximizing long: 500 - 4000 - our analysis suggests this maximizes stability of individual differences (albeit not necessarily differences in language)

```{r}
d_trial <- d_aoi |>
  mutate(short_window = t_norm > 200 & t_norm <= 2000,
         long_window = t_norm > 500 & t_norm <= 4000) |>
  group_by(dataset_name, dataset_id, administration_id, 
           trial_id, age, target_label, distractor_label) |>
  summarise(short_window_accuracy = mean(correct[short_window], na.rm=TRUE),
            long_window_accuracy = mean(correct[long_window], na.rm=TRUE),
            short_window_prop_data = sum(!is.na(correct[short_window]), na.rm = TRUE) /
              length(correct[short_window]),
            long_window_prop_data = sum(!is.na(correct[long_window]), na.rm = TRUE) / 
              length(correct[long_window]),
            short_window_target = sum(correct[short_window], na.rm=TRUE),
            long_window_target = sum(correct[long_window], na.rm=TRUE),
            short_window_distractor = sum(!correct[short_window], na.rm=TRUE),
            long_window_distractor = sum(!correct[long_window], na.rm=TRUE)) |>
  mutate(short_window_elogit = log((short_window_target + .5) / (short_window_distractor + .5)),
         long_window_elogit = log((long_window_target + .5) / (long_window_distractor + .5)))
```

Compute RTs, relying on the RLE workflow from the shiny app. (We tested this extensively back in the `peekbank-shiny` repo).

```{r}
d_rle <- d_aoi |>
  filter(any(t_norm == 0), # must have data at 0
         t_norm >= 0) |> # only pass data after 0
  group_by(administration_id, trial_id, trial_order) |>
  reframe(lengths = rle(aoi)$lengths, 
            values = rle(aoi)$values)

d_rt <- d_rle %>%
  group_by(administration_id, trial_id, trial_order) %>%
  nest() %>%
  mutate(data = lapply(data, get_rt)) %>%
  unnest(cols = c(data))
```

Join this back into the main data frame. We keep raw RTs with all shifts, but `rt` will refer to only D-T shifts.

```{r}
d_trial_rt_unfiltered <- d_trial |>
  left_join(d_rt) |>
  mutate(raw_rt = rt,
         rt = ifelse(shift_type == "D-T", rt, NA),
         log_rt = log(rt)) 
```

# Filtering

Check out age distribution.

```{r}
ages <- d_trial_rt |>
  group_by(administration_id) |>
  summarise(age = mean(age))

ggplot(ages, aes(x = age)) + 
  geom_histogram()

sum(ages < 12)
```

We have NA and NaN (average of all NAs) throughout. 
- rows with zero data in both windows should be removed
- elogits for zero data in one window should be NA'd out. 
- NaN should be NA

Also, let's get rid of trials with very limited amounts of data. Classically we remove trials with less than 50% data.

We are also going to NA-out RTs if we have less than 50% data in the short window, just because we think that's going to mess up the computation.

Finally, it will make life easier for now if we filter kids >= 12months. We can revisit this.

```{r}
d_trial <- d_trial_rt_unfiltered |>
  filter(!(is.na(short_window_prop_data) & is.na(long_window_prop_data)), 
         !(short_window_prop_data < .5 & long_window_prop_data < .5), 
         age >= 12) |>
  mutate(across(contains("window"), ~ ifelse(is.na(.x), NA, .x))) |>
  mutate(short_window_accuracy = ifelse(short_window_prop_data >= .5, short_window_accuracy, NA),
         long_window_accuracy = ifelse(long_window_prop_data >= .5, long_window_accuracy, NA),
         rt = ifelse(short_window_prop_data >= .5, rt, NA),
         short_window_elogit = ifelse(short_window_prop_data >= .5, short_window_elogit, NA), 
         long_window_elogit = ifelse(long_window_prop_data >= .5, long_window_elogit, NA))

```

# Descriptives

Now let's look at a summary.

```{r}
d_trial |>
  ungroup() |>
  summarise(n = n(), 
            avg_acc_short = mean(short_window_accuracy, na.rm=TRUE),
            avg_acc_long = mean(long_window_accuracy, na.rm=TRUE),
            avg_rt = mean(rt, na.rm=TRUE),
            nas_in_acc_short = mean(is.na(short_window_accuracy)),
            nas_in_acc_long = mean(is.na(long_window_accuracy)),
            avg_prop_data_short = mean(short_window_prop_data, na.rm=TRUE),
            avg_prop_data_long = mean(long_window_prop_data, na.rm=TRUE),
            prop_rts = mean(!is.na(rt))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  kable(digits = 2)
```

And dataset-wise.

```{r}
d_trial |>
  group_by(dataset_name) |>
  summarise(n = n(), 
            avg_acc_short = mean(short_window_accuracy, na.rm=TRUE),
            avg_acc_long = mean(long_window_accuracy, na.rm=TRUE),
            avg_rt = mean(rt, na.rm=TRUE),
            nas_in_acc_short = mean(is.na(short_window_accuracy)),
            nas_in_acc_long = mean(is.na(long_window_accuracy)),
            average_prop_data_short = mean(short_window_prop_data, na.rm=TRUE),
            average_prop_data_long = mean(long_window_prop_data, na.rm=TRUE),
            prop_rts = mean(!is.na(rt))) |>
  arrange(desc(n)) |>
  kable(digits = 2)
```

Look at some histograms.

Short window.

```{r}
ggplot(d_trial, 
       aes(x = short_window_accuracy)) +
  geom_histogram() + 
  facet_wrap(~dataset_name, scales = "free_y")
```

Long window.

```{r}
ggplot(d_trial, 
       aes(x = long_window_accuracy)) +
  geom_histogram() + 
  facet_wrap(~dataset_name, scales = "free_y")
```
RTs.

```{r}
ggplot(d_trial, 
       aes(x = rt)) +
  geom_histogram() + 
  facet_wrap(~dataset_name, scales = "free_y")
```

# Create subject-level RDS as well

```{r}
d_sub <- d_trial |>
  group_by(dataset_name, administration_id, age) |>
  summarise(across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
```

# Save data

```{r}
saveRDS(d_trial, here("cached_intermediates","1_d_trial"))
saveRDS(d_sub, here("cached_intermediates","1_d_sub"))
```

### 2024-08-28 2_developmental_analysis.qmd
first push

---
title: "Trial-level developmental analysis"
format:
  html:
    toc: true
    toc-location: left
---

```{r}
library(here)
source(here("helper","common.R"))
d_trial <- readRDS(here("cached_intermediates","1_d_trial"))
d_sub <- readRDS(here("cached_intermediates","1_d_sub"))
```

Approaches:

- measures [long/short] accuracy, [log] rt, elogit
- across datasets vs. within
- trial level vs. participant level data

Let's not immediately elogit as that was just for models.

Let's also use only `mean(log(rt))` as that's correct if we think RTs are sampled from a log-normal distribution.

# Trial-level, across datasets

Across datasets. Short window looks less wiggly, for whatever reason.

```{r}
d_trial |>
  pivot_longer(contains("accuracy"), 
               values_to = "accuracy", names_to = "measure") |>
ggplot(aes(x = age, y = accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() + 
  facet_wrap(~measure)
```

RT

```{r}
d_trial |>
ggplot(aes(x = age, y = rt)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  scale_x_log10() +
  scale_y_log10()
```

# Trial-level, for each dataset

Only look at datasets with some (cross-sectional) variation.

```{r}
datasets_with_age_variation <- d_trial |>
  group_by(dataset_name) |>
  summarise(age_range = max(age) - min(age)) |>
  filter(age_range >= 4)

d_trial_agevar <- d_trial |>
  filter(dataset_name %in% datasets_with_age_variation$dataset_name)
```

```{r}
ggplot(d_trial_agevar, aes(x = age, y = short_window_accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm", col = "green", alpha = .5) +
  geom_smooth(method = "loess", alpha = .5) +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() + 
  facet_wrap(~dataset_name, scales = "free_x")
```

RT

```{r}
ggplot(d_trial_agevar,
       aes(x = age, y = rt)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm", col = "green", alpha = .5) +
  geom_smooth(method = "loess", alpha = .5) +
  scale_x_log10() +
  scale_y_log10() + 
  facet_wrap(~dataset_name, scales = "free_x")
```

# Participant-level, across datasets

Accuracy.

```{r}
d_sub |>
  pivot_longer(contains("accuracy"), 
               values_to = "accuracy", names_to = "measure") |>
ggplot(aes(x = age, y = accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() + 
  facet_wrap(~measure)
```

RT - but is it the mean of the log or the log of the mean? We choose log of the mean because this fits a log-normal distribution.

```{r}
d_sub |>
ggplot(aes(x = age, y = log_rt)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  scale_x_log10()  + 
  ylab("mean log RT")
```

# Participant-level, for each dataset

Within each dataset.

```{r}
d_sub_agevar <- d_sub |>
  filter(dataset_name %in% datasets_with_age_variation$dataset_name)
```

```{r}
ggplot(d_sub_agevar, aes(x = age, y = short_window_accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm", col = "green", alpha = .5) +
  geom_smooth(method = "loess", alpha = .5) +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() + 
  facet_wrap(~dataset_name, scales = "free_x")
```

elogits

```{r}
ggplot(d_sub_agevar, aes(x = age, y = short_window_elogit)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm", col = "green", alpha = .5) +
  geom_smooth(method = "loess", alpha = .5) +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() + 
  facet_wrap(~dataset_name, scales = "free_x")
```

RT - here we look at the mean of the log RTs

```{r}
ggplot(d_sub_agevar,
       aes(x = age, y = log_rt)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm", col = "green", alpha = .5) +
  geom_smooth(method = "loess", alpha = .5) +
  scale_x_log10() +
  scale_y_log10() + 
  facet_wrap(~dataset_name, scales = "free_x")
```

# Appendix: What if we remove zoner trials?

Remove floor and ceiling where there are no shifts?

The answer is that things look even better.

```{r}
d_trial |>
  pivot_longer(contains("accuracy"), 
               values_to = "accuracy", names_to = "measure") |>
  filter(accuracy > 0, accuracy < 1) |> 
ggplot(aes(x = age, y = accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() + 
  facet_wrap(~measure)
```

```{r}
d_sub_nozoner <- d_trial |>
  filter(short_window_accuracy > 0, short_window_accuracy < 1) |> 
  group_by(dataset_name, administration_id, age) |>
  summarise(across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
```

```{r}
d_sub_nozoner |>
  pivot_longer(contains("accuracy"), 
               values_to = "accuracy", names_to = "measure") |>
ggplot(aes(x = age, y = accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() + 
  facet_wrap(~measure)
```

### 2024-08-28 3_developmental_relationships.qmd
first push

---
title: "Subject-level developmental relationships"
format: html
---

```{r}
library(here)
source(here("helper","common.R"))
d_sub <- readRDS(here("cached_intermediates","1_d_sub"))
load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))
```

```{r}
d_cdi <- cdi_data |>
  select(administration_id, native_language, measure, language, 
         rawscore, percentile) |>
  filter(language == "English (American)", native_language == "eng") |>
  pivot_wider(names_from = "measure", names
```

```{r}
d_sub |>
  left_join(d_cdi)
```

# Relation between RT and accuracy

OK, let's take a look at these subject and word-level measures.

```{r}

acc_rt <- inner_join(ms_acc, ms_rt) |>
  mutate(age_group_years = cut(age, breaks = c(12,24,36,48,60)))

head(acc_rt)
```

Take a look at correlations to select variables.

```{r}
GGally::ggpairs(select(ungroup(acc_rt), -dataset_name, -age_group_years, -n_accuracy, -n_rt, -administration_id), 
        progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))

```
It turns out that the correlations between elogit(accuracy), log(age), and log(rt) are all stronger. So we'll adopt these.

```{r}
GGally::ggpairs(select(ungroup(acc_rt), elogit, log_rt, log_age), 
        progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))

```

So these are somewhat related to one another. Let's take a deeper look.

```{r}
ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = .5, lty = 2) + 
  xlab("Mean log RT") + 
  ylab("Mean elogit(accuracy)")
```

Broken out by age group.

```{r}
ggplot(acc_rt, aes(x = log_rt, y = elogit)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = .5, lty = 2) +
  facet_wrap(~age_group_years) + 
  xlab("Mean log RT") + 
  ylab("Mean elogit accuracy")

```

By dataset.

```{r}
ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = .5, lty = 2) +
  facet_wrap(~dataset_name) + 
  xlab("Mean log RT") + 
  ylab("Mean elogit accuracy")

```

## Dimensionality reduction

Let's do principal components analysis over the scaled variables.

```{r}
acc_rt_mat <- acc_rt |>
  ungroup() |>
  select(log_age, elogit, log_rt) |>
  mutate(log_age = scale(log_age)[,1], 
         log_rt = scale(log_rt)[,1],
         elogit = scale(elogit)[,1] )|>
  as.matrix()

acc_rt_prc <- prcomp(acc_rt_mat, 2)

acc_rt_prc
summary(acc_rt_prc)
ggbiplot::ggbiplot(acc_rt_prc, alpha = .1)

```

We see that there is a first component with ~60% of variance that is "faster, more accurate, older". Then we see a second component that relates to younger, faster, slightly less accurate. And the final piece is older and less accurate.

These would be interesting components to connect to CDI. In some sense, the claim of some of the Fernald processing corpus is that there is a second principal component here (namely, processing speed) that is meaningful and relates to later learning outcomes.

# Conclusions

Probably right now we need to try and connect this to CDIs...

### 2024-08-28 helper/common.R
copy helper

@@ -1,0 +1,28 @@
+# source me for all analyses 
+
+library(tidyverse)
+library(lme4)
+library(knitr) # for kable
+# library(ggrepel)
+library(ggthemes)
+# library(viridis)
+# library(cowplot)
+# remotes::install_github("jmgirard/agreement")
+# library(agreement)
+# library(tictoc)
+
+# Seed for random number generation
+set.seed(42)
+knitr::opts_chunk$set(cache.extra = knitr::rand_seed, cache = TRUE, 
+                      message=FALSE, warning=FALSE, error=FALSE)
+options(dplyr.summarise.inform = FALSE)
+
+
+.font <- "Source Sans Pro"
+theme_set(theme_bw(base_size = 14, base_family = .font))
+theme_update(panel.grid = element_blank(),
+             strip.background = element_blank(),
+             legend.key = element_blank(),
+             panel.border = element_blank(),
+             axis.line = element_line(),
+             strip.text = element_text(face = "bold"))

### 2024-08-28 3_developmental_relationships.qmd
3

@@
 
+This is broken due to some repeated fields.
+
 ```{r}
@@
   select(administration_id, native_language, measure, language, 
-         rawscore, percentile) |>
+         rawscore) |>
   filter(language == "English (American)", native_language == "eng") |>
-  pivot_wider(names_from = "measure", names
+  pivot_wider(names_from = "measure", values_from = "rawscore", 
+              values_fill = NA, values_fn = mean)
 ```
@@
 ```{r}
-d_sub |>
-  left_join(d_cdi)
+d_sub_cdi <- d_sub |>
+  left_join(d_cdi) |>
+  mutate(log_age = log(age)) |>
+  ungroup()
 ```
@@
 
-OK, let's take a look at these subject and word-level measures.
-
-```{r}
-
-acc_rt <- inner_join(ms_acc, ms_rt) |>
-  mutate(age_group_years = cut(age, breaks = c(12,24,36,48,60)))
-
-head(acc_rt)
-```
-
 Take a look at correlations to select variables.
@@
 ```{r}
-GGally::ggpairs(select(ungroup(acc_rt), -dataset_name, -age_group_years, -n_accuracy, -n_rt, -administration_id), 
-        progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
-
-```
-It turns out that the correlations between elogit(accuracy), log(age), and log(rt) are all stronger. So we'll adopt these.
-
-```{r}
-GGally::ggpairs(select(ungroup(acc_rt), elogit, log_rt, log_age), 
+GGally::ggpairs(select(d_sub_cdi, age, log_age, rt, log_rt, 
+                       short_window_accuracy, long_window_accuracy, 
+                       short_window_elogit, long_window_elogit, prod, comp), 
         progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
@@
 
-So these are somewhat related to one another. Let's take a deeper look.
+Let's select the best of these.
 
 ```{r}
-ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) +
-  geom_point(alpha = .3) +
-  geom_smooth(method = "lm") +
-  geom_hline(yintercept = .5, lty = 2) + 
-  xlab("Mean log RT") + 
-  ylab("Mean elogit(accuracy)")
-```
-
-Broken out by age group.
-
-```{r}
-ggplot(acc_rt, aes(x = log_rt, y = elogit)) +
-  geom_point(alpha = .3) +
-  geom_smooth(method = "lm") +
-  geom_hline(yintercept = .5, lty = 2) +
-  facet_wrap(~age_group_years) + 
-  xlab("Mean log RT") + 
-  ylab("Mean elogit accuracy")
+GGally::ggpairs(select(d_sub_cdi, log_age, rt,  
+                       long_window_accuracy, prod, comp), 
+        progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
 
@@
 
-By dataset.
+<!-- So these are somewhat related to one another. Let's take a deeper look.  -->
 
-```{r}
-ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) +
-  geom_point(alpha = .3) +
-  geom_smooth(method = "lm") +
-  geom_hline(yintercept = .5, lty = 2) +
-  facet_wrap(~dataset_name) + 
-  xlab("Mean log RT") + 
-  ylab("Mean elogit accuracy")
+<!-- ```{r} -->
+<!-- ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) + -->
+<!--   geom_point(alpha = .3) + -->
+<!--   geom_smooth(method = "lm") + -->
+<!--   geom_hline(yintercept = .5, lty = 2) +  -->
+<!--   xlab("Mean log RT") +  -->
+<!--   ylab("Mean elogit(accuracy)") -->
+<!-- ``` -->
 
-```
+<!-- Broken out by age group.  -->
+
+<!-- ```{r} -->
+<!-- ggplot(acc_rt, aes(x = log_rt, y = elogit)) + -->
+<!--   geom_point(alpha = .3) + -->
+<!--   geom_smooth(method = "lm") + -->
+<!--   geom_hline(yintercept = .5, lty = 2) + -->
+<!--   facet_wrap(~age_group_years) +  -->
+<!--   xlab("Mean log RT") +  -->
+<!--   ylab("Mean elogit accuracy") -->
+
+<!-- ``` -->
+
+<!-- By dataset.  -->
+
+<!-- ```{r} -->
+<!-- ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) + -->
+<!--   geom_point(alpha = .3) + -->
+<!--   geom_smooth(method = "lm") + -->
+<!--   geom_hline(yintercept = .5, lty = 2) + -->
+<!--   facet_wrap(~dataset_name) +  -->
+<!--   xlab("Mean log RT") +  -->
+<!--   ylab("Mean elogit accuracy") -->
+
+<!-- ``` -->
 
@@
 ```{r}
-acc_rt_mat <- acc_rt |>
+d_sub_cdi_mat <- d_sub_cdi |>
   ungroup() |>
-  select(log_age, elogit, log_rt) |>
+  select(log_age, rt,  long_window_accuracy, prod) |>
   mutate(log_age = scale(log_age)[,1], 
-         log_rt = scale(log_rt)[,1],
-         elogit = scale(elogit)[,1] )|>
+         rt = scale(rt)[,1],
+         long_window_accuracy = scale(long_window_accuracy)[,1], 
+         prod = scale(prod)[,1])
+
+d_sub_cdi_mat <- d_sub_cdi_mat |>
+  filter(complete.cases(d_sub_cdi_mat)) |>
   as.matrix()
 
-acc_rt_prc <- prcomp(acc_rt_mat, 2)
+acc_rt_prc <- prcomp(d_sub_cdi_mat, 2)
 

### 2024-08-28 2_developmental_analysis.qmd
update filenames

@@
 source(here("helper","common.R"))
-d_trial <- readRDS(here("cached_intermediates","1_d_trial"))
-d_sub <- readRDS(here("cached_intermediates","1_d_sub"))
+d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds"))
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
 ```

### 2024-08-28 3_developmental_relationships.qmd
update filenames

@@
 source(here("helper","common.R"))
-d_sub <- readRDS(here("cached_intermediates","1_d_sub"))
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
 load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))

### 2024-08-28 1_get_data.qmd
update filenames
fix df name

@@
 ```{r}
-ages <- d_trial_rt |>
+ages <- d_trial_rt_unfiltered |>
   group_by(administration_id) |>
@@
 ```{r}
-saveRDS(d_trial, here("cached_intermediates","1_d_trial"))
-saveRDS(d_sub, here("cached_intermediates","1_d_sub"))
+saveRDS(d_trial, here("cached_intermediates","1_d_trial.Rds"))
+saveRDS(d_sub, here("cached_intermediates","1_d_sub.Rds"))
 ```

### 2024-08-28 2_developmental_analysis.qmd
switch to percent scores
(rawscore interpretation is form-dependent)

--- 2024-08-28 2_developmental_analysis.qmd
@@
 ```{r}
-library(here)
+suppressPackageStartupMessages(library(here))
 source(here("helper","common.R"))
-d_trial <- readRDS(here("cached_intermediates","1_d_trial"))
-d_sub <- readRDS(here("cached_intermediates","1_d_sub"))
+d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds"))
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
 ```

### 2024-08-28 3_developmental_relationships.qmd
switch to percent scores
(rawscore interpretation is form-dependent)

--- 2024-08-28 3_developmental_relationships.qmd
@@
 ```{r}
-library(here)
-source(here("helper","common.R"))
-d_sub <- readRDS(here("cached_intermediates","1_d_sub"))
+suppressPackageStartupMessages(library(here))
+suppressPackageStartupMessages(source(here("helper","common.R")))
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
 load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))
@@
   select(administration_id, native_language, measure, language, 
-         rawscore) |>
+         CDI_percent) |> # rawscore is form-dependent
   filter(language == "English (American)", native_language == "eng") |>
-  pivot_wider(names_from = "measure", values_from = "rawscore", 
+  pivot_wider(names_from = "measure", values_from = "CDI_percent", 
               values_fill = NA, values_fn = mean)
@@
 
-```{r}
+```{r, fig.width=9, fig.height=9}
 GGally::ggpairs(select(d_sub_cdi, age, log_age, rt, log_rt, 
@@
 
-```{r}
+```{r, fig.width=9, fig.height=9}
 GGally::ggpairs(select(d_sub_cdi, log_age, rt,  
@@
 
+```{r control-for-age}
+d_sub_cdi_mat <- d_sub_cdi |>
+  ungroup() |>
+  select(age, rt,  long_window_accuracy, prod) |>
+  mutate(rt = scale(rt / age)[,1],
+         long_window_accuracy = scale(long_window_accuracy / age)[,1], 
+         prod = scale(prod / age)[,1]) |>
+  select(-age)
+
+d_sub_cdi_mat <- d_sub_cdi_mat |>
+  filter(complete.cases(d_sub_cdi_mat)) |>
+  as.matrix()
+
+acc_rt_prc <- prcomp(d_sub_cdi_mat, 2)
+
+acc_rt_prc
+summary(acc_rt_prc)
+ggbiplot::ggbiplot(acc_rt_prc, alpha = .1)
+```
+
+Controlling for age, the first component (~49% of variance) represents high production, fast RT, but low LWL accuracy.
+The second component (~31% of variance) captures slow RT, low accuracy, and moderate production, and the third component (~20%) captures fast RT, low accuracy, and low production).
+
+The claim of some of the Fernald processing corpus is that there is a second principal component here (namely, processing speed) that is meaningful and relates to later learning outcomes.
+
 ```{r}
@@
 ggbiplot::ggbiplot(acc_rt_prc, alpha = .1)
-
 ```
 
-We see that there is a first component with ~60% of variance that is "faster, more accurate, older".
-Then we see a second component that relates to younger, faster, slightly less accurate.
-And the final piece is older and less accurate.
-
-These would be interesting components to connect to CDI.
-In some sense, the claim of some of the Fernald processing corpus is that there is a second principal component here (namely, processing speed) that is meaningful and relates to later learning outcomes.
+There is a first component capturing ~50% of the variance that is low accuracy, low production, slower, and younger.
+A second component capturing ~30% of the variance is higher accuracy, low production, fast RT, and slightly young.
+A third component (~16%) is older, highly accurate, low production, and somewhat slow RT.
 
@@
 
-Probably right now we need to try and connect this to CDIs...
+The next thing to do is look at any longitudinal data we have, to see how predictive faster LWL at t0 is of higher CDIs at t1.

### 2024-08-29 1_get_data.qmd
extra changes from morning

@@
 
+Let's look at the RT distribution.
+
+```{r}
+ggplot(d_trial_rt_unfiltered, aes(x = rt)) + 
+  geom_histogram() + 
+  geom_vline(xintercept = 367, col = "red", lty =2 )
+```
+
 Check out age distribution.
@@
 
+Let's also remove RTs < 367 ms. These are unlikely to be real.
+
 ```{r}
@@
   mutate(across(contains("window"), ~ ifelse(is.na(.x), NA, .x))) |>
-  mutate(short_window_accuracy = ifelse(short_window_prop_data >= .5, short_window_accuracy, NA),
-         long_window_accuracy = ifelse(long_window_prop_data >= .5, long_window_accuracy, NA),
+  mutate(short_window_accuracy = ifelse(short_window_prop_data >= .5, 
+                                        short_window_accuracy, NA),
+         long_window_accuracy = ifelse(long_window_prop_data >= .5, 
+                                       long_window_accuracy, NA),
          rt = ifelse(short_window_prop_data >= .5, rt, NA),
-         short_window_elogit = ifelse(short_window_prop_data >= .5, short_window_elogit, NA), 
-         long_window_elogit = ifelse(long_window_prop_data >= .5, long_window_elogit, NA))
+         short_window_elogit = ifelse(short_window_prop_data >= .5, 
+                                      short_window_elogit, NA), 
+         long_window_elogit = ifelse(long_window_prop_data >= .5, 
+                                     long_window_elogit, NA), 
+         rt = ifelse(rt < 367, NA, rt), 
+         log_rt = ifelse(rt < 367, NA, log_rt))
 

### 2024-08-29 3_developmental_relationships.qmd
extra changes from morning

--- 2024-08-28 3_developmental_relationships.qmd
@@
   select(administration_id, native_language, measure, language, 
-         rawscore) |>
+         CDI_percent) |>
   filter(language == "English (American)", native_language == "eng") |>
-  pivot_wider(names_from = "measure", values_from = "rawscore", 
+  pivot_wider(names_from = "measure", values_from = "CDI_percent", 
               values_fill = NA, values_fn = mean)
@@
 GGally::ggpairs(select(d_sub_cdi, log_age, rt,  
-                       long_window_accuracy, prod, comp), 
+                       long_window_accuracy, short_window_accuracy, prod, comp), 
         progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
@@
 
-<!-- So these are somewhat related to one another. Let's take a deeper look.  -->
+# Individual relationships
 
-<!-- ```{r} -->
-<!-- ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) + -->
-<!--   geom_point(alpha = .3) + -->
-<!--   geom_smooth(method = "lm") + -->
-<!--   geom_hline(yintercept = .5, lty = 2) +  -->
-<!--   xlab("Mean log RT") +  -->
-<!--   ylab("Mean elogit(accuracy)") -->
-<!-- ``` -->
+```{r}
+m_rt <- mean(d_sub_cdi$rt, na.rm=TRUE)
+sd_rt <- sd(d_sub_cdi$rt, na.rm=TRUE)
+d_sub_cdi$factor_rt <- cut(d_sub_cdi$rt, 
+                           breaks = quantile(d_sub_cdi$rt, 
+                                             probs = c(0, .25, .5, .75, 1), 
+                                             na.rm= TRUE))
 
-<!-- Broken out by age group.  -->
+ggplot(filter(d_sub_cdi, !is.na(rt), !is.na(factor_rt)),
+       aes(x = log_age, 
+                      y = long_window_accuracy, 
+                      col = factor_rt)) +
+  geom_point(alpha = .3) +
+  geom_smooth() +
+  geom_hline(yintercept = .5, lty = 2)+
+  viridis::scale_color_viridis(discrete = TRUE, 
+                               name = "RT quantile")
 
-<!-- ```{r} -->
-<!-- ggplot(acc_rt, aes(x = log_rt, y = elogit)) + -->
-<!--   geom_point(alpha = .3) + -->
-<!--   geom_smooth(method = "lm") + -->
-<!--   geom_hline(yintercept = .5, lty = 2) + -->
-<!--   facet_wrap(~age_group_years) +  -->
-<!--   xlab("Mean log RT") +  -->
-<!--   ylab("Mean elogit accuracy") -->
+```
 
-<!-- ``` -->
+Broken out by age group.
+
+```{r}
+
+d_sub_cdi$age_group <- cut(d_sub_cdi$log_age, 
+                           breaks = log(c(12,24,36,48,60)), 
+                           labels = c("12-24","24-36","36-48","48-60"), 
+                           include_lowest = TRUE)
+
+ggplot(d_sub_cdi, 
+       aes(x = rt, 
+                      y = long_window_accuracy, 
+                      col = age_group)) +
+  geom_point(alpha = .3) +
+  geom_smooth() +
+  geom_hline(yintercept = .5, lty = 2)+
+  viridis::scale_color_viridis(discrete = TRUE, 
+                               name = "age group")
+```
 
@@
 
-Let's do principal components analysis over the scaled variables.
+Make into matrices.
 
@@
   as.matrix() 
+```
+
+Factor analysis
+
+```{r}
+# scree(d_sub_cdi_mat, pc=FALSE)  # Use pc=FALSE for factor analysis
+
+# install.packages("psych")
+mod1 <- fa(d_sub_cdi_mat, 1)
+mod2 <- fa(d_sub_cdi_mat, 2)
+
+```
+
+Let's do principal components analysis over the scaled variables.
+
+```{r}
 

### 2024-08-29 helper/common.R

@@ -10,6 +10,7 @@ library(ggthemes)
 # remotes::install_github("jmgirard/agreement")
 # library(agreement)
 # library(tictoc)
+library(GGally)
 
 # Seed for random number generation
 set.seed(42)

### 2024-08-29 1_get_data.qmd
add subject_id to cached data

@@
          long_window = t_norm > 500 & t_norm <= 4000) |>
-  group_by(dataset_name, dataset_id, administration_id, 
+  group_by(dataset_name, dataset_id, administration_id, subject_id,
            trial_id, age, target_label, distractor_label) |>
@@
 d_sub <- d_trial |>
-  group_by(dataset_name, administration_id, age) |>
+  group_by(dataset_name, subject_id, administration_id, age) |>
   summarise(across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))

### 2024-08-30 2_developmental_analysis.qmd
analyses

@@
 
+# Variability
+
+```{r}
+d_trial |>
+  group_by(administration_id, age) |>
+  summarise (sd_rt = sd(rt, na.rm=TRUE), 
+             sd_acc = sd(long_window_accuracy)) |>
+  pivot_wider(
+ggplot(aes(x = age, y = sd_rt)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  geom_hline(yintercept = .5, lty = 2) + 
+  scale_x_log10() 
+```
+
 # Appendix: What if we remove zoner trials?

### 2024-08-30 3_developmental_relationships.qmd
analyses

@@
 suppressPackageStartupMessages(library(here))
+suppressPackageStartupMessages(library(psych))
 suppressPackageStartupMessages(source(here("helper","common.R")))
@@
 
-This is broken due to some repeated fields.
+This is very slightly broken due to some repeated fields. We use `values_fn = mean` to suppress them.
 
@@
                                              probs = c(0, .25, .5, .75, 1), 
-                                             na.rm= TRUE))
+                                             na.rm= TRUE), 
+                           labels = c("0-25", "25-50", "50-75", "75-1"))
 
 ggplot(filter(d_sub_cdi, !is.na(rt), !is.na(factor_rt)),
-       aes(x = log_age, 
+       aes(x = age, 
            y = long_window_accuracy, 
@@
   geom_smooth() +
+  scale_x_log10() + 
   geom_hline(yintercept = .5, lty = 2)+
   viridis::scale_color_viridis(discrete = TRUE, 
-                               name = "RT quantile")
+                               name = "RT quantile") + 
+  xlab("Age (log months)") + 
+  ylab("Accuracy (long window)") + 
+  theme(legend.position = "bottom")
 
@@
 
-ggplot(d_sub_cdi, 
+ggplot(filter(d_sub_cdi, !is.na(age_group)), 
        aes(x = rt, 
@@
 
-<!-- By dataset.  -->
+By dataset.
 
-<!-- ```{r} -->
-<!-- ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) + -->
-<!--   geom_point(alpha = .3) + -->
-<!--   geom_smooth(method = "lm") + -->
-<!--   geom_hline(yintercept = .5, lty = 2) + -->
-<!--   facet_wrap(~dataset_name) +  -->
-<!--   xlab("Mean log RT") +  -->
-<!--   ylab("Mean elogit accuracy") -->
+```{r}
+datasets_with_age_variation <- d_trial |>
+  group_by(dataset_name) |>
+  summarise(age_range = max(age) - min(age)) |>
+  filter(age_range >= 6)
 
-<!-- ``` -->
+ggplot(filter(d_sub_cdi, 
+              !is.na(rt), !is.na(factor_rt), 
+              dataset_name %in% datasets_with_age_variation$dataset_name),
+       aes(x = age, 
+           y = long_window_accuracy, 
+           col = factor_rt)) +
+  geom_point(alpha = .3) +
+  geom_smooth(se= FALSE, method = "lm") +
+  scale_x_log10() + 
+  geom_hline(yintercept = .5, lty = 2)+
+  viridis::scale_color_viridis(discrete = TRUE, 
+                               name = "RT quantile") +
+  facet_wrap(~dataset_name) + 
+  xlab("Age (log months)") + 
+  ylab("Accuracy (long window)")
 
-## Dimensionality reduction
+```
 
-Make into matrices.
+# Dimensionality reduction
 
-```{r control-for-age}
+## Production
+
+Make into matrices and look at PCA
+
+```{r}
 d_sub_cdi_mat <- d_sub_cdi |>
   ungroup() |>
-  select(age, rt,  long_window_accuracy, prod) |>
-  mutate(rt = scale(rt / age)[,1],
-         long_window_accuracy = scale(long_window_accuracy / age)[,1], 
-         prod = scale(prod / age)[,1]) |>
-  select(-age)
+  select(log_age, rt,  long_window_accuracy, prod)
 
@@
 
-acc_rt_prc <- prcomp(d_sub_cdi_mat, 2)
+acc_rt_prc <- prcomp(x = d_sub_cdi_mat, scale = TRUE)
 
@@
 
-Controlling for age, the first component (~49% of variance) represents high production, fast RT, but low LWL accuracy.
-The second component (~31% of variance) captures slow RT, low accuracy, and moderate production, and the third component (~20%) captures fast RT, low accuracy, and low production).
-
 The claim of some of the Fernald processing corpus is that there is a second principal component here (namely, processing speed) that is meaningful and relates to later learning outcomes.
 
+Factor analysis
+
 ```{r}
-d_sub_cdi_mat <- d_sub_cdi |>
+mod1 <- fa(d_sub_cdi_mat, 1)
+mod2 <- fa(d_sub_cdi_mat, 2)
+```
+
+Scree plot
+
+```{r}
+fa.parallel(d_sub_cdi_mat)
+```
+
+## Comprehension
+
+```{r}
+d_sub_cdi_mat_comp <- d_sub_cdi |>
   ungroup() |>
-  select(log_age, rt,  long_window_accuracy, prod) |>
-  mutate(log_age = scale(log_age)[,1], 
-         rt = scale(rt)[,1],
-         long_window_accuracy = scale(long_window_accuracy)[,1], 
-         prod = scale(prod)[,1])
+  select(log_age, rt,  long_window_accuracy, comp)
 
-d_sub_cdi_mat <- d_sub_cdi_mat |>
-  filter(complete.cases(d_sub_cdi_mat)) |>
+d_sub_cdi_mat_comp <- d_sub_cdi_mat_comp |>
+  filter(complete.cases(d_sub_cdi_mat_comp)) |>
   as.matrix()
+
+acc_rt_prc_comp <- prcomp(x = d_sub_cdi_mat_comp, scale = TRUE)
+
+acc_rt_prc_comp
+summary(acc_rt_prc_comp)
+ggbiplot::ggbiplot(acc_rt_prc_comp, alpha = .1)
 ```
@@
 ```{r}
-# scree(d_sub_cdi_mat, pc=FALSE)  # Use pc=FALSE for factor analysis
+mod1_comp <- fa(d_sub_cdi_mat_comp, 1)
+mod2_comp <- fa(d_sub_cdi_mat_comp, 2)
+```
 
-# install.packages("psych")
-mod1 <- fa(d_sub_cdi_mat, 1)
-mod2 <- fa(d_sub_cdi_mat, 2)
+```{r}
+fa.parallel(d_sub_cdi_mat_comp)
+```
+
+# Longitudinal data
+
+```{r}
+longitudinal <- d_sub_cdi |>
+  group_by(dataset_name, subject_id) |>
+  count() |>
+  filter(n > 1)
+
+d_sub_long <- filter(d_sub_cdi, subject_id %in% longitudinal$subject_id)
+
+d_sub_wide <- d_sub_long |>
+  group_by(subject_id, 
+```
+
+```{r}
+library(lavaan)
 
 ```
-
-Let's do principal components analysis over the scaled variables.
-
-```{r}
-
-acc_rt_prc <- prcomp(d_sub_cdi_mat, 2)
-
-acc_rt_prc
-summary(acc_rt_prc)
-ggbiplot::ggbiplot(acc_rt_prc, alpha = .1)
-```
-
-There is a first component capturing ~50% of the variance that is low accuracy, low production, slower, and younger.
-A second component capturing ~30% of the variance is higher accuracy, low production, fast RT, and slightly young.
-A third component (~16%) is older, highly accurate, low production, and somewhat slow RT.
 

### 2024-08-30 1_get_data.qmd
updates from more hacking

@@
 
+Add CDIs.
+
+This is very slightly broken due to some repeated fields. We use `values_fn = mean` to suppress them.
+
+```{r}
+load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))
+
+d_cdi <- cdi_data |>
+  select(administration_id, native_language, measure, language, 
+         CDI_percent) |> # rawscore is form-dependent
+  filter(language == "English (American)", native_language == "eng") |>
+  pivot_wider(names_from = "measure", values_from = "CDI_percent", 
+              values_fill = NA, values_fn = mean)
+```
+
+```{r}
+d_sub <- d_sub |>
+  left_join(d_cdi) |>
+  mutate(log_age = log(age)) |>
+  ungroup()
+```
+
+Print descriptives.
+
+```{r}
+print(paste(length(unique(d_sub$subject_id)), "children,",
+            sum(!is.na(d_sub$prod)), "with CDI data"))
+print(paste(length(unique(d_sub$administration_id)), "administrations"))
+print(paste(length(unique(d_sub$dataset_name)), "datasets"))
+```
+
 # Save data

### 2024-08-30 2_developmental_analysis.qmd
updates from more hacking

--- 2024-08-28 2_developmental_analysis.qmd
@@
 d_trial |>
+  pivot_longer(c("rt","log_rt"), 
+               values_to = "rt", names_to = "measure") |>
 ggplot(aes(x = age, y = rt)) + 
@@
   scale_x_log10() +
-  scale_y_log10()
+  facet_wrap(~measure, scales = "free_y")
+```
+## Stats
+
+```{r}
+mods <- list(log_log = lm(log_rt ~ log_age, data = d_trial),
+             log_lin = lm(log_rt ~ age, data = d_trial),
+             lin_log = lm(rt ~ log_age, data = d_trial),
+             lin_lin = lm(rt ~ age, data = d_trial))
+
+models <- map_df(mods, ~broom::glance(.x)) |>
+  mutate(model = names(mods))
+
+kable(models, digits = 4)
 ```
@@
 ```
-
-RT - but is it the mean of the log or the log of the mean? We choose log of the mean because this fits a log-normal distribution.
+RTs
 
@@
 d_sub |>
-ggplot(aes(x = age, y = log_rt)) + 
+  pivot_longer(c("rt","log_rt"), 
+               values_to = "rt", names_to = "measure") |>
+ggplot(aes(x = age, y = rt)) + 
   geom_point(alpha = .05) + 
@@
   scale_x_log10()  + 
-  ylab("mean log RT")
+  ylab("mean log RT") + 
+  facet_wrap(~measure, scales = "free_y")
 ```
@@
 
+## Stats
+
+```{r}
+mods_sub <- list(log_log = lm(log_rt ~ log_age, data = d_sub),
+             log_lin = lm(log_rt ~ age, data = d_sub),
+             lin_log = lm(rt ~ log_age, data = d_sub),
+             lin_lin = lm(rt ~ age, data = d_sub))
+
+models_sub <- map_df(mods_sub, ~broom::glance(.x)) |>
+  mutate(model = names(mods_sub))
+
+kable(models_sub, digits = 4)
+```
+
+# Variability
+
+```{r}
+d_trial |>
+  group_by(administration_id, age) |>
+  summarise (sd_rt = sd(rt, na.rm=TRUE), 
+             sd_acc = sd(long_window_accuracy, na.rm=TRUE)) |>
+  pivot_longer(contains("sd"), 
+               values_to = "sd", names_to = "measure") |>
+ggplot(aes(x = age, y = sd)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  # geom_hline(yintercept = .5, lty = 2) + 
+  scale_x_log10() +
+  facet_wrap(~measure, scales = "free_y")
+```
+
 # Appendix: What if we remove zoner trials?

### 2024-08-30 3_developmental_relationships.qmd
updates from more hacking

@@
 d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
-load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))
-```
-
-This is very slightly broken due to some repeated fields. We use `values_fn = mean` to suppress them.
-
-```{r}
-d_cdi <- cdi_data |>
-  select(administration_id, native_language, measure, language, 
-         CDI_percent) |> # rawscore is form-dependent
-  filter(language == "English (American)", native_language == "eng") |>
-  pivot_wider(names_from = "measure", values_from = "CDI_percent", 
-              values_fill = NA, values_fn = mean)
-```
-
-```{r}
-d_sub_cdi <- d_sub |>
-  left_join(d_cdi) |>
-  mutate(log_age = log(age)) |>
-  ungroup()
 ```
@@
 ```{r, fig.width=9, fig.height=9}
-GGally::ggpairs(select(d_sub_cdi, age, log_age, rt, log_rt, 
+GGally::ggpairs(select(d_sub, age, log_age, rt, log_rt, 
                        short_window_accuracy, long_window_accuracy, 
@@
 ```{r, fig.width=9, fig.height=9}
-GGally::ggpairs(select(d_sub_cdi, log_age, rt,  
+GGally::ggpairs(select(d_sub, log_age, rt,  
                        long_window_accuracy, short_window_accuracy, prod, comp), 
@@
 ```{r}
-m_rt <- mean(d_sub_cdi$rt, na.rm=TRUE)
-sd_rt <- sd(d_sub_cdi$rt, na.rm=TRUE)
-d_sub_cdi$factor_rt <- cut(d_sub_cdi$rt, 
-                           breaks = quantile(d_sub_cdi$rt, 
+m_rt <- mean(d_sub$rt, na.rm=TRUE)
+sd_rt <- sd(d_sub$rt, na.rm=TRUE)
+d_sub$factor_rt <- cut(d_sub$rt, 
+                           breaks = quantile(d_sub$rt, 
                                              probs = c(0, .25, .5, .75, 1), 
@@
 
-ggplot(filter(d_sub_cdi, !is.na(rt), !is.na(factor_rt)),
+ggplot(filter(d_sub, !is.na(rt), !is.na(factor_rt)),
        aes(x = age, 
@@
 
-d_sub_cdi$age_group <- cut(d_sub_cdi$log_age, 
+d_sub$age_group <- cut(d_sub$log_age, 
                            breaks = log(c(12,24,36,48,60)), 
@@
 
-ggplot(filter(d_sub_cdi, !is.na(age_group)), 
+ggplot(filter(d_sub, !is.na(age_group)), 
        aes(x = rt, 
@@
 
-ggplot(filter(d_sub_cdi, 
+ggplot(filter(d_sub, 
               !is.na(rt), !is.na(factor_rt), 
@@
 ```{r}
-d_sub_cdi_mat <- d_sub_cdi |>
+d_sub_mat <- d_sub |>
   ungroup() |>
@@
 
-d_sub_cdi_mat <- d_sub_cdi_mat |>
-  filter(complete.cases(d_sub_cdi_mat)) |>
+d_sub_mat <- d_sub_mat |>
+  filter(complete.cases(d_sub_mat)) |>
   as.matrix()
 
-acc_rt_prc <- prcomp(x = d_sub_cdi_mat, scale = TRUE)
+acc_rt_prc <- prcomp(x = d_sub_mat, scale = TRUE)
 
@@
 ```{r}
-mod1 <- fa(d_sub_cdi_mat, 1)
-mod2 <- fa(d_sub_cdi_mat, 2)
+mod1 <- fa(d_sub_mat, 1)
+mod2 <- fa(d_sub_mat, 2)
 ```
@@
 ```{r}
-fa.parallel(d_sub_cdi_mat)
+fa.parallel(d_sub_mat)
 ```
@@
 ```{r}
-d_sub_cdi_mat_comp <- d_sub_cdi |>
+d_sub_mat_comp <- d_sub |>
   ungroup() |>
@@
 
-d_sub_cdi_mat_comp <- d_sub_cdi_mat_comp |>
-  filter(complete.cases(d_sub_cdi_mat_comp)) |>
+d_sub_mat_comp <- d_sub_mat_comp |>
+  filter(complete.cases(d_sub_mat_comp)) |>
   as.matrix()
 
-acc_rt_prc_comp <- prcomp(x = d_sub_cdi_mat_comp, scale = TRUE)
+acc_rt_prc_comp <- prcomp(x = d_sub_mat_comp, scale = TRUE)
 
@@
 ```{r}
-mod1_comp <- fa(d_sub_cdi_mat_comp, 1)
-mod2_comp <- fa(d_sub_cdi_mat_comp, 2)
+mod1_comp <- fa(d_sub_mat_comp, 1)
+mod2_comp <- fa(d_sub_mat_comp, 2)
 ```
@@
 ```{r}
-fa.parallel(d_sub_cdi_mat_comp)
+fa.parallel(d_sub_mat_comp)
 ```
@@
 ```{r}
-longitudinal <- d_sub_cdi |>
+longitudinal <- d_sub |>
   group_by(dataset_name, subject_id) |>
@@
 
-d_sub_long <- filter(d_sub_cdi, subject_id %in% longitudinal$subject_id)
+d_sub_long <- d_sub |>
+  filter(subject_id %in% longitudinal$subject_id, 
+                     !is.na(prod)) |>
+  group_by(subject_id) |>
+  arrange(administration_id) |>
+  mutate(admin_num = 1:n(), 
+         delta_t = c(0,diff(age)))
 
-d_sub_wide <- d_sub_long |>
-  group_by(subject_id, 
+ggplot(filter(d_sub_long, admin_num > 1), 
+       aes(x = delta_t)) +
+  geom_histogram()
+```
+
+```{r}
+
 ```

### 2024-09-15 1_get_data.qmd
more progress

@@
             average_prop_data_long = mean(long_window_prop_data, na.rm=TRUE),
-            prop_rts = mean(!is.na(rt))) |>
+            prop_rts = mean(!is.na(rt)), 
+            log_age = log(age)) |>
   arrange(desc(n)) |>

### 2024-09-15 2_developmental_analysis.qmd
more progress

--- 2024-08-28 2_developmental_analysis.qmd
@@
 suppressPackageStartupMessages(library(here))
+suppressPackageStartupMessages(library(broom))
+suppressPackageStartupMessages(library(broom.mixed))
+
 source(here("helper","common.R"))
-d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds"))
+d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
+  mutate(log_age = log(age))
 d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
@@
 
-# Trial-level, across datasets
+# Trial-level
+
+## Across datasets
 
@@
 d_trial |>
+  pivot_longer(c("rt","log_rt"), 
+               values_to = "rt", names_to = "measure") |>
   ggplot(aes(x = age, y = rt)) + 
@@
   scale_x_log10() +
-  scale_y_log10()
+  facet_wrap(~measure, scales = "free_y")
+```
+### Stats
+
+Accuracy models.
+
+```{r}
+acc_mods <- list(lwe_lin = lm(long_window_elogit ~ age, data = d_trial),
+                 lwe_log = lm(long_window_elogit ~ log_age, data = d_trial),
+                 swe_lin = lm(short_window_elogit ~ age, data = d_trial),
+                 swe_log = lm(short_window_elogit ~ log_age, data = d_trial),
+                 lwa_lin = lm(long_window_accuracy ~ age, data = d_trial),
+                 lwa_log = lm(long_window_accuracy ~ log_age, data = d_trial),
+                 swa_lin = lm(short_window_accuracy ~ age, data = d_trial),
+                 swa_log = lm(short_window_accuracy ~ log_age, data = d_trial))
+
+acc_mods_summary <- map_df(acc_mods, ~broom::glance(.x)) |>
+  mutate(model = names(acc_mods))
+
+kable(acc_mods_summary, digits = 4)
 ```
 
-# Trial-level, for each dataset
+Reaction time models.
+
+```{r}
+rt_mods <- list(log_log = lm(log_rt ~ log_age, data = d_trial),
+                log_lin = lm(log_rt ~ age, data = d_trial),
+                lin_log = lm(rt ~ log_age, data = d_trial),
+                lin_lin = lm(rt ~ age, data = d_trial))
+
+rt_mods_summary <- map_df(rt_mods, ~broom::glance(.x)) |>
+  mutate(model = names(rt_mods))
+
+kable(rt_mods_summary, digits = 4)
+```
+
+## Within dataset
 
@@
 ```
+
+Accuracy
 
@@
 
-# Participant-level, across datasets
+### Stats
 
-Accuracy.
+Accuracy models. This time LMERs.
 
 ```{r}
-d_sub |>
-  pivot_longer(contains("accuracy"), 
-               values_to = "accuracy", names_to = "measure") |>
-ggplot(aes(x = age, y = accuracy)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth() +
-  geom_smooth(method = "lm", col = "green") +
-  geom_hline(yintercept = .5, lty = 2) + 
-  scale_x_log10() + 
-  facet_wrap(~measure)
+d_trial$age_s <- scale(d_trial$age)[,1]
+d_trial$log_age_s <- scale(d_trial$log_age)[,1]
+
+acc_mods_lmer <- list(lwe_lin = lmer(long_window_elogit ~ 1 
+                                     + (age_s | dataset_name), data = d_trial),
+                      lwe_log = lmer(long_window_elogit ~ 1
+                                     + (log_age_s | dataset_name), data = d_trial),
+                      swe_lin = lmer(short_window_elogit ~ 1
+                                     + (age_s | dataset_name), data = d_trial),
+                      swe_log = lmer(short_window_elogit ~ 1
+                                     + (log_age_s | dataset_name), data = d_trial),
+                      lwa_lin = lmer(long_window_accuracy ~ 1
+                                     + (age_s | dataset_name), data = d_trial),
+                      lwa_log = lmer(long_window_accuracy ~ 1
+                                     + (log_age_s | dataset_name), data = d_trial),
+                      swa_lin = lmer(short_window_accuracy ~ 1
+                                     + (age_s | dataset_name), data = d_trial),
+                      swa_log = lmer(short_window_accuracy ~ 1
+                                     + (log_age_s | dataset_name), data = d_trial))
+
+acc_mods_lmer_summary <- map_df(acc_mods_lmer, ~broom.mixed::glance(.x)) |>
+  mutate(model = names(acc_mods_lmer), 
+         r2 = map_dbl(acc_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
+
+kable(acc_mods_lmer_summary, digits = 4)
 ```
 
-RT - but is it the mean of the log or the log of the mean? We choose log of the mean because this fits a log-normal distribution.
+Reaction time models.
 
 ```{r}
-d_sub |>
-ggplot(aes(x = age, y = log_rt)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth() +
-  geom_smooth(method = "lm", col = "green") +
-  scale_x_log10()  + 
-  ylab("mean log RT")
+rt_mods_lmer <- list(log_log = lmer(log_rt ~ 1 + 
+                                      (log_age_s | dataset_name), data = d_trial),
+                     log_lin = lmer(log_rt ~ 1 + 
+                                      (age_s | dataset_name), data = d_trial),
+                     lin_log = lmer(rt ~ 1 + 
+                                      (log_age_s | dataset_name), data = d_trial),
+                     lin_lin = lmer(rt ~ 1 + 
+                                      (age_s | dataset_name), data = d_trial))
+
+rt_mods_lmer_summary <- map_df(rt_mods_lmer, ~broom.mixed::glance(.x)) |>
+  mutate(model = names(rt_mods_lmer), 
+         r2 = map_dbl(rt_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
+
+kable(rt_mods_lmer_summary, digits = 4)
 ```
 
-# Participant-level, for each dataset
-
-Within each dataset.
+# Participant-level
 
@@
 ```
+
+## Across datasets
+
+Accuracy.
+
+```{r}
+d_sub_long <- d_sub |>
+  pivot_longer(contains("accuracy"), 
+               values_to = "accuracy", names_to = "measure") 
+d_sub_agevar_long <- d_sub_agevar |>
+  pivot_longer(contains("accuracy"), 
+               values_to = "accuracy", names_to = "measure")
+
+ggplot(d_sub_long, 
+       aes(x = age, y = accuracy)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  geom_smooth(data = d_sub_agevar_long,
+              aes(group = dataset_name, col = dataset_name), geom="line",
+              alpha = .1, method = "lm", size = .5, lty = 2) + 
+  geom_hline(yintercept = .5, lty = 2) + 
+  scale_x_log10() + 
+  facet_wrap(~measure) + 
+  theme(legend.position = "bottom") + 
+  ylab("Accuracy") + 
+  xlab("Age (months)")
+```
+
+Elogits
+
+```{r}
+d_sub_long_elogit <- d_sub |>
+  pivot_longer(contains("elogit"), 
+               values_to = "accuracy", names_to = "measure") 
+d_sub_agevar_long_elogit <- d_sub_agevar |>
+  pivot_longer(contains("elogit"), 
+               values_to = "accuracy", names_to = "measure")
+
+ggplot(d_sub_long_elogit, 
+       aes(x = age, y = accuracy)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  geom_smooth(data = d_sub_agevar_long_elogit,
+              aes(group = dataset_name, col = dataset_name), geom="line",
+              alpha = .1, method = "lm", size = .5, lty = 2) + 
+  geom_hline(yintercept = .5, lty = 2) + 
+  scale_x_log10() + 
+  facet_wrap(~measure) + 
+  theme(legend.position = "bottom") + 
+  ylab("Accuracy (elogit)") + 
+  xlab("Age (months)")
+```
+
+RTs
+
+```{r}
+d_sub_long_rt <- d_sub |>
+  pivot_longer(c("rt","log_rt"), 
+               values_to = "rt", names_to = "measure")
+
+d_sub_agevar_long_rt <- d_sub_agevar |>
+  pivot_longer(c("rt","log_rt"), 
+               values_to = "rt", names_to = "measure")
+
+ggplot(d_sub_long_rt, 
+       aes(x = age, y = rt)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  geom_smooth(data = d_sub_agevar_long_rt,
+              aes(group = dataset_name, col = dataset_name), geom="line",
+              alpha = .1, method = "lm", size = .5, lty = 2) + 
+  # geom_hline(yintercept = .5, lty = 2) + 
+  scale_x_log10() + 
+  facet_wrap(~measure, scales = "free_y") + 
+  theme(legend.position = "bottom") + 
+  ylab("mean log RT") + 
+  xlab("Age (months)")
+
+```
+
+### Stats
+
+Accuracy models.
+
+```{r}
+acc_mods_sub <- list(lwe_lin = lm(long_window_elogit ~ age, data = d_sub),
+                     lwe_log = lm(long_window_elogit ~ log_age, data = d_sub),
+                     swe_lin = lm(short_window_elogit ~ age, data = d_sub),
+                     swe_log = lm(short_window_elogit ~ log_age, data = d_sub),
+                     lwa_lin = lm(long_window_accuracy ~ age, data = d_sub),
+                     lwa_log = lm(long_window_accuracy ~ log_age, data = d_sub),
+                     swa_lin = lm(short_window_accuracy ~ age, data = d_sub),
+                     swa_log = lm(short_window_accuracy ~ log_age, data = d_sub))
+
+acc_mods_sub_summary <- map_df(acc_mods_sub, ~broom::glance(.x)) |>
+  mutate(model = names(acc_mods_sub))
+
+kable(acc_mods_sub_summary, digits = 4)
+```
+
+Reaction time models.
+
+```{r}
+mods_sub <- list(log_log = lm(log_rt ~ log_age, data = d_sub),
+                 log_lin = lm(log_rt ~ age, data = d_sub),
+                 lin_log = lm(rt ~ log_age, data = d_sub),
+                 lin_lin = lm(rt ~ age, data = d_sub))
+
+models_sub <- map_df(mods_sub, ~broom::glance(.x)) |>
+  mutate(model = names(mods_sub))
+
+kable(models_sub, digits = 4)
+```
+
+## Within dataset
+
+Within each dataset.
 
@@
 ```{r}
-ggplot(d_sub_agevar, aes(x = age, y = short_window_elogit)) + 
+ggplot(d_sub_agevar, aes(x = age, y = long_window_elogit)) +
   geom_point(alpha = .05) +
@@
   geom_hline(yintercept = .5, lty = 2) +
-  scale_x_log10() + 
+  # scale_x_log10() +
+  ylim(-2,3) + 
   facet_wrap(~dataset_name, scales = "free_x")
@@
   facet_wrap(~dataset_name, scales = "free_x")
+```
+
+### Stats
+
+Accuracy models.
+
+```{r}
+d_sub$age_s <- scale(d_sub$age)[,1]
+d_sub$log_age_s <- scale(d_sub$log_age)[,1]
+
+acc_mods_sub_lmer <- list(lwe_lin = lmer(long_window_elogit ~ 1 
+                                         + (age_s | dataset_name), data = d_sub),
+                          lwe_log = lmer(long_window_elogit ~ 1
+                                         + (log_age_s | dataset_name), data = d_sub),
+                          swe_lin = lmer(short_window_elogit ~ 1
+                                         + (age_s | dataset_name), data = d_sub),
+                          swe_log = lmer(short_window_elogit ~ 1
+                                         + (log_age_s | dataset_name), data = d_sub),
+                          lwa_lin = lmer(long_window_accuracy ~ 1
+                                         + (age_s | dataset_name), data = d_sub),
+                          lwa_log = lmer(long_window_accuracy ~ 1
+                                         + (log_age_s | dataset_name), data = d_sub),
+                          swa_lin = lmer(short_window_accuracy ~ 1
+                                         + (age_s | dataset_name), data = d_sub),
+                          swa_log = lmer(short_window_accuracy ~ 1
+                                         + (log_age_s | dataset_name), data = d_sub))
+
+acc_mods_sub_lmer_summary <- map_df(acc_mods_sub_lmer, ~broom.mixed::glance(.x)) |>
+  mutate(model = names(acc_mods_sub_lmer), 
+         r2 = map_dbl(acc_mods_sub_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
+
+kable(acc_mods_sub_lmer_summary, digits = 4)
+```
+
+Reaction time models.
+
+```{r}
+rt_mods_lmer <- list(log_log = lmer(log_rt ~ 1 + 
+                                      (log_age_s | dataset_name), data = d_sub),
+                     log_lin = lmer(log_rt ~ 1 + 
+                                      (age_s | dataset_name), data = d_sub),
+                     lin_log = lmer(rt ~ 1 + 
+                                      (log_age_s | dataset_name), data = d_sub),
+                     lin_lin = lmer(rt ~ 1 + 
+                                      (age_s | dataset_name), data = d_sub))
+
+rt_mods_lmer_summary <- map_df(rt_mods_lmer, ~broom.mixed::glance(.x)) |>
+  mutate(model = names(rt_mods_lmer), 
+         r2 = map_dbl(rt_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
+
+kable(rt_mods_lmer_summary, digits = 4)
+```
+
+# Variability
+
+```{r}
+d_var <- d_trial |>
+  group_by(dataset_name, administration_id, age) |>
+  summarise (`SD(log RT)` = sd(log_rt, na.rm=TRUE), 
+             `SD(accuracy)` = sd(long_window_elogit, na.rm=TRUE)) |>
+  pivot_longer(contains("sd"), 
+               values_to = "sd", names_to = "measure")
+
+v <- ggplot(d_var, aes(x = age, y = sd)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  geom_smooth(data = d_var |>
+  filter(dataset_name %in% datasets_with_age_variation$dataset_name),
+              aes(group = dataset_name, col = dataset_name), geom="line",
+              alpha = .1, method = "lm", size = .5, lty = 2) + 
+  scale_color_discrete(guide = FALSE) + 
+  scale_x_log10() +
+  facet_wrap(~measure, scales = "free_y")
+
+v 
+ggsave(v, filename = "figures/var.png", width = 8, height = 3.5)
+
+```
+# Pub stuff
+
+stats
+
+```{r}
+d_sub |>
+  group_by(dataset_name) |>
+  summarise(n = n(), 
+            `Mean Age` = mean(age), 
+            `Min Age` = min(age), 
+            `Max Age` = max(age), 
+            `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
+            `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
+  xtable(caption = "Characteristics of included datasets from Peekbank.")
+```
+
+```{r}
+a <- ggplot(d_sub_long_elogit |> filter(measure == "long_window_elogit"), 
+       aes(x = age, y = accuracy)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  geom_smooth(data = d_sub_agevar_long_elogit |> filter(measure == "long_window_elogit"),
+              aes(group = dataset_name, col = dataset_name), geom="line",
+              alpha = .1, method = "lm", size = .5, lty = 2) + 
+  geom_hline(yintercept = .5, lty = 2) + 
+  scale_x_log10() +
+  scale_color_discrete(guide = FALSE) + 
+  ylab("Accuracy (elogit)") + 
+  xlab("Age (months)")
+
+b <- ggplot(d_sub_long_rt |> filter(measure == "log_rt"), 
+       aes(x = age, y = rt)) + 
+  geom_point(alpha = .05) + 
+  geom_smooth() +
+  geom_smooth(method = "lm", col = "green") +
+  geom_smooth(data = d_sub_agevar_long_rt |> filter(measure == "log_rt"),
+              aes(group = dataset_name, col = dataset_name), geom="line",
+              alpha = .1, method = "lm", size = .5, lty = 2) + 
+  # geom_hline(yintercept = .5, lty = 2) + 
+  scale_x_log10() +
+  scale_color_discrete(guide = FALSE) + 
+  theme(legend.position = "bottom") + 
+  ylab("Reaction Time (log)") + 
+  xlab("Age (months)")
+
+f <- cowplot::plot_grid(a,b)
+ggsave(f, filename = "figures/acc_rt.png", width = 8, height = 3.5)
+
 ```

### 2024-09-15 3_developmental_relationships.qmd
more progress

@@
 
-# Relation between RT and accuracy
+# Relations between variables
+
+## Between datasets
+
+```{r}
+cor(select(d_sub, age, log_age, rt, log_rt, 
+                       short_window_accuracy, long_window_accuracy, 
+                       short_window_elogit, long_window_elogit, prod, comp), 
+    use = "pairwise.complete.obs") |>
+  round(2)
+```
 
@@
 
-Let's select the best of these.
+Let's select the most interesting of these.
 
 ```{r, fig.width=9, fig.height=9}
-GGally::ggpairs(select(d_sub, log_age, rt,  
-                       long_window_accuracy, short_window_accuracy, prod, comp), 
-                progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
+d_sub |>
+  select(log_age, log_rt,  
+         long_window_accuracy,  prod, comp) |>
+GGally::ggpairs(progress = FALSE, 
+                lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
 
+```
+
+## Within dataset
+
+All the variables.
+
+```{r}
+all_corrs <- d_sub |>
+select(dataset_name, age, log_age, rt, log_rt, 
+                       short_window_accuracy, long_window_accuracy, 
+                       short_window_elogit, long_window_elogit, prod, comp) |>
+  group_by(dataset_name) |>
+  nest() |>
+  mutate(corrs = map(data, ~cor(.x, use = "pairwise.complete.obs")),
+         long_corrs = map(corrs, ~pivot_longer(as_tibble(.x) |> 
+                                                mutate(variable1 = colnames(.x)), 
+                                               -variable1, names_to = "variable2", 
+                                               values_to = "r"))) |>
+  select(-data, -corrs) |>
+  unnest(cols = c(long_corrs))
+
+ggplot(all_corrs, aes(x = variable1, y = r, col = dataset_name)) + 
+  geom_point(alpha = .2) + 
+  stat_summary(fun.data = "mean_cl_boot", col = "black") +
+  coord_flip() + 
+  facet_wrap(~variable2)
+```
+
+Several variables.
+
+```{r}
+d_sub_agevar <- d_sub |>
+  filter(dataset_name %in% datasets_with_age_variation$dataset_name)
+
+all_corrs <- d_sub_agevar |>
+select(dataset_name, log_age, log_rt, rt,
+                       long_window_accuracy, long_window_elogit) |>
+  group_by(dataset_name) |>
+  nest() |>
+  mutate(corrs = map(data, ~cor(.x, use = "pairwise.complete.obs")),
+         long_corrs = map(corrs, ~pivot_longer(as_tibble(.x) |> 
+                                                mutate(variable1 = colnames(.x)), 
+                                               -variable1, names_to = "variable2", 
+                                               values_to = "r"))) |>
+  select(-data, -corrs) |>
+  unnest(cols = c(long_corrs)) |>
+  filter(variable1 != variable2, 
+         !(variable1 == "rt" & variable2 == "log_rt"), 
+         !(variable1 == "log_rt" & variable2 == "rt"),
+         !(variable1 == "long_window_accuracy" & variable2 == "long_window_elogit"),
+         !(variable1 == "long_window_elogit" & variable2 == "long_window_accuracy"))
+
+ggplot(all_corrs,
+       aes(x = variable1, y = r, col = dataset_name)) + 
+  geom_point(alpha = .5, pch = 20) +
+  stat_summary(fun.data = "mean_cl_boot", aes(group = 1)) +
+  geom_hline(yintercept = 0, lty = 2) + 
+  # coord_flip() +
+  facet_grid(.~variable2, scales = "free_x") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+
+Plot correlations of key variables now.  Annoying to figure out how to remove the upper diagonal...
+
+```{r}
+all_corrs <- d_sub_agevar |>
+select(dataset_name, log_age, log_rt,
+                       long_window_accuracy) |>
+  group_by(dataset_name) |>
+  nest() |>
+  mutate(corrs = map(data, ~cor(.x, use = "pairwise.complete.obs")),
+         long_corrs = map(corrs, ~pivot_longer(as_tibble(.x) |> 
+                                                mutate(variable1 = colnames(.x)), 
+                                               -variable1, names_to = "variable2", 
+                                               values_to = "r"))) |>
+  select(-data, -corrs) |>
+  unnest(cols = c(long_corrs)) |>
+  filter(variable1 != variable2)
+
+ggplot(all_corrs,
+       aes(x = interaction(variable1, variable2), y = r, col = dataset_name)) + 
+  geom_point(alpha = .5, pch = 20) +
+  stat_summary(fun.data = "mean_cl_boot", aes(group = 1)) +
+  geom_hline(yintercept = 0, lty = 2) + 
+  coord_flip() +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ```
@@
   summarise(age_range = max(age) - min(age)) |>
-  filter(age_range >= 6)
+  filter(age_range >= 4)
 
@@
 
+The claim of some of the Fernald processing corpus is that there is a second principal component here (namely, processing speed) that is meaningful and relates to later learning outcomes.
+
 ## Production
 
-Make into matrices and look at PCA
+Scree plot
 
@@
   ungroup() |>
-  select(log_age, rt,  long_window_accuracy, prod)
+  select(log_age, log_rt, long_window_elogit, prod)
 
@@
   as.matrix()
+
+fa.parallel(d_sub_mat)
+```
+
+Factor analysis first. This is probably more appropriate.
+
+```{r}
+mod1 <- fa(d_sub_mat, 1, rotate = "varimax")
+mod2 <- fa(d_sub_mat, 2, rotate = "varimax")
+
+summary(mod1)
+mod1
+```
+
+Make into matrices and look at PCA.
+
+```{r}
 
@@
 
-The claim of some of the Fernald processing corpus is that there is a second principal component here (namely, processing speed) that is meaningful and relates to later learning outcomes.
-
-Factor analysis
+## Production within individual datasets
 
 ```{r}
-mod1 <- fa(d_sub_mat, 1)
-mod2 <- fa(d_sub_mat, 2)
-```
+d_sub_mat_dataset <- d_sub |>
+  filter(dataset_name %in% datasets_with_age_variation$dataset_name) |>
+  group_by(dataset_name) |>
+  select(dataset_name, log_age, log_rt, long_window_elogit, prod)
 
-Scree plot
+d_sub_mat_grp <- d_sub_mat_dataset |>
+  nest() |>
+  mutate(complete_data = map(data, ~filter(.x[complete.cases(.x),])),
+    mat = as.matrix(complete_data), 
+    has_data = map_lgl(complete_data, ~nrow(.x)>0)) |>
+  filter(has_data) |>
+  mutate(factors = map(mat, ~fa(.x, 1, rotate = "varimax")), 
+         f_summary = map(factors, summary), 
+         loadings = map_df(factors, ~as.tibble(loadings(x))))
 
-```{r}
-fa.parallel(d_sub_mat)
+d_sub_mat_grp |>
+
 ```
@@
   ungroup() |>
-  select(log_age, rt,  long_window_accuracy, comp)
+  select(log_age, log_rt,  long_window_elogit, comp)
 
@@
 
+```
+
+Scree.
+
+```{r}
+fa.parallel(d_sub_mat_comp)
+```
+
+Factor analysis
+
+```{r}
+mod1_comp <- fa(d_sub_mat_comp, 1, rotate = "varimax")
+mod2_comp <- fa(d_sub_mat_comp, 2, rotate = "varimax")
+
+summary(mod1_comp)
+loadings(mod1_comp)
+```
+
+PCA.
+
+```{r}
 acc_rt_prc_comp <- prcomp(x = d_sub_mat_comp, scale = TRUE)
@@
 ggbiplot::ggbiplot(acc_rt_prc_comp, alpha = .1)
-```
-
-Factor analysis
-
-```{r}
-mod1_comp <- fa(d_sub_mat_comp, 1)
-mod2_comp <- fa(d_sub_mat_comp, 2)
-```
-
-```{r}
-fa.parallel(d_sub_mat_comp)
 ```

### 2024-09-16 3_developmental_relationships.qmd
add residualized age analysis and fix a couple bugs
- add residualized age correlations
- make sure trials are loaded
- move up dataset subsetting to those with age variation
- update rt_helper to most current version

@@
 d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
+d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds"))
 ```
@@
 GGally::ggpairs(progress = FALSE, 
-                lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
+                lower = list(continuous = GGally::wrap("points", alpha = 0.03)),
+                axisLabels = "show")
 
+```
+
+### Controlling for age (between dataset)
+
+```{r}
+d_residualized_by_age <- d_sub |>
+  select(subject_id, administration_id,log_age,log_rt, long_window_accuracy, prod, comp) |> 
+  pivot_longer(cols=c(log_rt,long_window_accuracy,  prod, comp),names_to="measure",values_to="value") |>
+  group_by(measure) |> 
+  nest() |> 
+  mutate(fits = map(data, \(d) lm(value ~ log_age, data = d) |>
+                      broom::augment(newdata = d))) |> 
+  unnest(fits) |> 
+  select(-data)
+
+d_residualized_by_age |>
+  select(administration_id, measure, .resid) |>
+  pivot_wider(names_from = measure, values_from = .resid) |>
+  select(-administration_id) |>
+  GGally::ggpairs(progress=FALSE,lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
 ```
@@
 ```{r}
+datasets_with_age_variation <- d_trial |>
+  group_by(dataset_name) |>
+  summarise(age_range = max(age) - min(age)) |>
+  filter(age_range >= 4)
+
 d_sub_agevar <- d_sub |>
@@
 
+Hmm, want to think more about  estimating an "average" correlation when the variation in age changes so dramatically between datasets
+
+```{r}
+#order by correlation value of log age and long window accuracy
+all_corrs_log_age_acc <- all_corrs |> 
+  ungroup() |> 
+  filter(variable1 == "log_age" & variable2 == "long_window_accuracy") |>
+  mutate(dataset_name_ordered= fct_reorder(dataset_name,r))
+
+d_sub_agevar <- d_sub_agevar |> 
+  left_join(all_corrs_log_age_acc)
+
+#quick peek at cross-dataset estimates, ranked by strength of correlation
+ggplot(d_sub_agevar,aes(log_age,long_window_accuracy))+geom_point()+geom_smooth(method="lm")+facet_wrap(~dataset_name_ordered)
+```
+
+### Controlling for age (within dataset)
+
+```{r}
+all_corrs_age_residualized <- d_sub_agevar |>
+  ungroup() |> 
+  #compute variables residualized by age
+select(dataset_name, administration_id,subject_id,log_age, log_rt, rt,
+                       long_window_accuracy, long_window_elogit,prod) |>
+  pivot_longer(cols=c(log_rt, rt,
+                       long_window_accuracy, long_window_elogit,  prod),names_to="measure",values_to="value") |>
+  group_by(dataset_name,measure) |> 
+  nest() |> 
+  mutate(fits = map(data, \(d) {if(sum(!is.na(d$value))>0) {lm(value ~ log_age, data = d) |> broom::augment(newdata = d)} })) |>
+  unnest(fits) |> 
+  select(dataset_name,subject_id,administration_id,measure, .resid) |> 
+  rename(resid = .resid) |> 
+  pivot_wider(names_from=measure,values_from=resid,names_prefix="resid_") |> 
+  #now group by dataset and compute pairwise correlations
+  select(-subject_id,-administration_id) |>
+  group_by(dataset_name) |>
+  nest() |>
+  mutate(corrs = map(data, ~cor(.x, use = "pairwise.complete.obs")),
+         long_corrs = map(corrs, ~pivot_longer(as_tibble(.x) |> 
+                                                mutate(variable1 = colnames(.x)), 
+                                               -variable1, names_to = "variable2", 
+                                               values_to = "r"))) |>
+  mutate(corrs = map(data, ~cor(.x, use = "pairwise.complete.obs")),
+         long_corrs = map(corrs, ~pivot_longer(as_tibble(.x) |> 
+                                                mutate(variable1 = colnames(.x)), 
+                                               -variable1, names_to = "variable2", 
+                                               values_to = "r"))) |>
+  select(-data, -corrs) |>
+  unnest(cols = c(long_corrs)) |>
+  filter(variable1 != variable2, 
+         !(variable1 == "resid_rt" & variable2 == "resid_log_rt"), 
+         !(variable1 == "resid_log_rt" & variable2 == "resid_rt"),
+         !(variable1 == "resid_long_window_accuracy" & variable2 == "resid_long_window_elogit"),
+         !(variable1 == "resid_long_window_elogit" & variable2 == "resid_long_window_accuracy"))
+
+ggplot(all_corrs_age_residualized,
+       aes(x = variable1, y = r, col = dataset_name)) + 
+  geom_point(alpha = .5, pch = 20) +
+  stat_summary(fun.data = "mean_cl_boot", aes(group = 1)) +
+  geom_hline(yintercept = 0, lty = 2) + 
+  # coord_flip() +
+  facet_grid(.~variable2, scales = "free_x") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+
 # Individual relationships
@@
 ```{r}
-datasets_with_age_variation <- d_trial |>
-  group_by(dataset_name) |>
-  summarise(age_range = max(age) - min(age)) |>
-  filter(age_range >= 4)
-
 ggplot(filter(d_sub, 

### 2024-09-26 1_get_data.qmd
more paper stuff, CFA

@@
   group_by(dataset_name, subject_id, administration_id, age) |>
-  summarise(across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
+  summarise(n_trials = sum(!is.na(long_window_accuracy)), 
+            n_trials_rt = sum(!is.na(rt)),
+            across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
 ```

### 2024-09-26 2_developmental_analysis.qmd
more paper stuff, CFA

@@
   group_by(dataset_name, administration_id, age) |>
-  summarise (`SD(log RT)` = sd(log_rt, na.rm=TRUE), 
-             `SD(accuracy)` = sd(long_window_elogit, na.rm=TRUE)) |>
-  pivot_longer(contains("sd"), 
+  summarise (`log(RT)` = sd(log_rt, na.rm=TRUE), 
+             `Accuracy` = sd(long_window_accuracy, na.rm=TRUE)) |>
+  pivot_longer(c("log(RT)","Accuracy"), 
                values_to = "sd", names_to = "measure")
@@
   scale_x_log10() +
-  facet_wrap(~measure, scales = "free_y")
+  facet_wrap(~measure, scales = "free_y") + 
+  ylab("Standard Deviation") + 
+  xlab("Age (months)")
 
@@
   group_by(dataset_name) |>
-  summarise(n = n(), 
+  summarise(`N subjects` = length(unique(subject_id)),
+            `N admins` = n(), 
             `Mean Age` = mean(age), 
@@
             `Max Age` = max(age), 
+            `Avg Trials` = mean(n_trials),
+            `Avg RT Trials` = mean(n_trials_rt),
             `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
             `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
-  xtable(caption = "Characteristics of included datasets from Peekbank.")
+  arrange(desc(`N subjects`)) |>
+  xtable(caption = "Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.")
 ```
@@
 ```{r}
-a <- ggplot(d_sub_long_elogit |> filter(measure == "long_window_elogit"), 
+a <- ggplot(d_sub_long |> filter(measure == "long_window_accuracy"), 
        aes(x = age, y = accuracy)) + 
@@
   geom_smooth(method = "lm", col = "green") +
-  geom_smooth(data = d_sub_agevar_long_elogit |> filter(measure == "long_window_elogit"),
+  geom_smooth(data = d_sub_agevar_long |> filter(measure == "long_window_accuracy"),
               aes(group = dataset_name, col = dataset_name), geom="line",
@@
   scale_color_discrete(guide = FALSE) + 
-  ylab("Accuracy (elogit)") + 
+  ylab("Accuracy") + 
   xlab("Age (months)")

### 2024-09-26 3_developmental_relationships.qmd
more paper stuff, CFA

--- 2024-09-15 3_developmental_relationships.qmd
@@
 d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
+library(lavaan)
+library(tidySEM)
+
 ```
@@
 
+## All data
+
+Scree plot
+
+```{r}
+d_sub_mat <- d_sub |>
+  ungroup() |>
+  select(log_age, log_rt, long_window_accuracy)
+
+d_sub_mat <- d_sub_mat |>
+  filter(complete.cases(d_sub_mat)) |>
+  as.matrix()
+
+fa.parallel(d_sub_mat)
+```
+```{r}
+mod1 <- fa(d_sub_mat, 1, rotate = "varimax")
+
+```
+
 ## Production
@@
   filter(has_data) |>
-  mutate(factors = map(mat, ~fa(.x, 1, rotate = "varimax")), 
-         f_summary = map(factors, summary), 
-         loadings = map_df(factors, ~as.tibble(loadings(x))))
+  mutate(fas = map(mat, ~fa(.x, 1, rotate = "varimax")), 
+         f_summary = map(fas, summary), 
+         loadings = map(fas, ~tibble(loadings = as.double(loadings(.x)), 
+                                     name = rownames(loadings(.x)))))
 
-d_sub_mat_grp |>
+ls <- d_sub_mat_grp |>
+  select(dataset_name, loadings) |>
+  unnest(cols = c(loadings))
+
+ggplot(ls, aes(x = name, y = loadings, col = dataset_name)) + 
+  geom_point(aes = .5) +
+  stat_summary(fun.data = "mean_cl_boot", aes(group = 1)) +
+  geom_hline(yintercept = 0, lty = 2) + 
+  coord_flip()
 
@@
 
+# CFA models
+
+```{r}
+
+d_cfa <- d_sub |>
+             rename(`accuracy` = long_window_accuracy) |>
+             mutate(accuracy = scale(accuracy), 
+                    log_age = scale(log_age), 
+                    log_rt = scale(log_rt)) |>
+  select(accuracy, log_age, log_rt)
+
+d_cfa <- d_cfa[complete.cases(d_cfa), ]
+
+fa_model <-  "F1  =~ log_age + log_rt + accuracy"
+fit <- cfa(fa_model, d_cfa, std.lv=TRUE)
+
+summary(fit)
+a <- graph_sem(model = fit)+ theme(panel.background = element_rect(fill = "white"))
+```
+
+```{r}
+d_cfa_prod <- d_sub |>
+             rename(`accuracy` = long_window_accuracy) |>
+             mutate(accuracy = scale(accuracy), 
+                    log_age = scale(log_age), 
+                    log_rt = scale(log_rt), 
+                    prod = scale(prod)) |>
+  select(accuracy, log_age, log_rt, prod)
+
+d_cfa_prod <- d_cfa_prod[complete.cases(d_cfa_prod), ]
+
+fa_model_prod <-  "F1  =~ log_age + log_rt + accuracy + prod"
+fit_prod <- cfa(fa_model_prod, d_cfa_prod, std.lv=TRUE)
+
+summary(fit_prod)
+b <- graph_sem(model = fit_prod) + 
+  theme(panel.background = element_rect(fill = "white"))
+b
+```
+```{r}
+d_cfa_comp <- d_sub |>
+             rename(`accuracy` = long_window_accuracy) |>
+             mutate(accuracy = scale(accuracy), 
+                    log_age = scale(log_age), 
+                    log_rt = scale(log_rt), 
+                    comp = scale(comp)) |>
+  select(accuracy, log_age, log_rt, comp)
+
+d_cfa_comp <- d_cfa_comp[complete.cases(d_cfa_comp), ]
+
+fa_model_comp <-  "F1  =~ log_age + log_rt + accuracy + comp"
+fit_comp <- cfa(fa_model_comp, d_cfa_comp, std.lv=TRUE)
+
+summary(fit_comp)
+c <- graph_sem(model = fit_comp,
+            layout = get_layout(x = fit_comp, layout_algorithm = "layout_as_star")) + 
+    theme(panel.background = element_rect(fill = "white"))
+c
+```
+
+```{r}
+plot_grid(a,b,c, nrow = 1, labels = c("A","B","C"))
+ggsave(here("figures","sem.png"), width = 25, height = 4)
+```
+
 # Longitudinal data
@@
 d_sub_long <- d_sub |>
-  filter(subject_id %in% longitudinal$subject_id, 
-                     !is.na(prod)) |>
+  filter(subject_id %in% longitudinal$subject_id) |>
   group_by(subject_id) |>
@@
 
-ggplot(filter(d_sub_long, admin_num > 1), 
-       aes(x = delta_t)) +
-  geom_histogram()
+ggplot(d_sub_long, 
+       aes(x = delta_t, fill = as.factor(admin_num))) +
+  geom_histogram() + 
+  facet_wrap(~dataset_name)
 ```
 
-```{r}
-
-```
+Longitudinal data with CDI is just FMW2013 right now.
 
 ```{r}
-library(lavaan)
+fmw <- filter(d_sub_long, dataset_name == "fmw_2013") |>
+  mutate(t = ifelse(age < 23, "t1","t2")) |>
+  group_by(subject_id, t) |>
+  summarise(age = mean(age), 
+            log_rt = mean(log_rt), 
+            prod = mean(prod), 
+            long_window_accuracy = mean(long_window_accuracy),
+            delta_t = mean(delta_t)) |>
+  pivot_wider(names_from = "t", 
+              values_from = c("age","log_rt","prod","long_window_accuracy", "delta_t"))
 
+summary(lm(prod_t2 ~ long_window_accuracy_t1 * log_rt_t1 + age_t2, 
+             data = fmw))
 ```

### 2024-09-26 3_developmental_relationships.qmd

--- 2024-09-16 3_developmental_relationships.qmd
@@
 suppressPackageStartupMessages(source(here("helper","common.R")))
+library(lavaan)
+library(tidySEM)
+
 d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))

### 2024-09-27 3_developmental_relationships.qmd
up to date for call

@@
 
-```{r, fig.width=9, fig.height=9}
+```{r}
 GGally::ggpairs(select(d_sub, age, log_age, rt, log_rt, 
@@
 ```
+
 ```{r}
 mod1 <- fa(d_sub_mat, 1, rotate = "varimax")
+mod1
 
@@
 ```
+Production.
 
@@
 ```
+
+Comprehension.
+
 ```{r}
@@
 
+Plot these as a grid.
+
 ```{r}
@@
 ggsave(here("figures","sem.png"), width = 25, height = 4)
+```
+
+## Measurement invariance across age?
+
+```{r}
+fa_model <-  "F1  =~ log_age + log_rt + accuracy"
+fit_younger <- cfa(fa_model, 
+                   filter(d_cfa, log_age <= median(log_age)),
+                   std.lv=TRUE)
+
+graph_sem(model = fit_younger)
 ```
@@
 
+## Without CDIs
+
+```{r}
+ggplot(d_sub_long, 
+       aes(x = age, y = long_window_accuracy, group = subject_id, col = dataset_name)) +
+  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
+  geom_smooth(aes(group = 1)) + 
+  scale_y_log10() + 
+  geom_hline(lty = 2, yintercept = .5)
+
+```
+
+```{r}
+ggplot(filter(d_sub_long, n_trials_rt > 3),
+       aes(x = age, y = log_rt, group = subject_id, col = dataset_name)) +
+  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
+  geom_smooth(aes(group = 1)) + 
+  scale_y_log10() 
+```
+
+## Model comparison of growth forms within kids
+
 # Conclusions
-
-The next thing to do is look at any longitudinal data we have, to see how predictive faster LWL at t0 is of higher CDIs at t1.

### 2024-10-04 1_get_data.qmd
refactor

--- 2024-09-15 1_get_data.qmd
@@
 source(here("helper","rt_helper.R"))
+
+# min number of trials
+ACC_MIN_TRIAL_CUTOFF <- 4
+RT_MIN_TRIAL_CUTOFF <- 2
 ```
@@
   group_by(dataset_name, subject_id, administration_id, age) |>
-  summarise(across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
+  summarise(n_trials = sum(!is.na(long_window_accuracy)), 
+            n_trials_rt = sum(!is.na(rt)),
+            rt_var = sd(rt, na.rm=TRUE), 
+            log_rt_var = sd(log_rt, na.rm=TRUE), 
+            short_window_acc_var = sd(short_window_accuracy, na.rm=TRUE),
+            long_window_acc_var = sd(long_window_accuracy, na.rm=TRUE),
+            across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
 ```
 
-Add CDIs.
+```{r}
+ggplot(d_sub, aes(x = n_trials, fill = n_trials < 4)) + 
+  geom_histogram(binwidth = 1) + 
+  facet_wrap(~dataset_name, scales = "free_y")
+```
+```{r}
+ggplot(d_sub, aes(x = n_trials_rt, fill = n_trials_rt < 2)) + 
+  geom_histogram(binwidth = 1) + 
+  facet_wrap(~dataset_name, scales = "free_y")
+```
+
+We impose the arbitrary cutoff that we need more than 4 accuracies and more than 2 RTs to analyze a participant.
+
+```{r}
+d_sub <- d_sub |>
+  mutate(across(contains("log_rt"), ~ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
+         across(contains("window"), ~ifelse(n_trials < ACC_MIN_TRIAL_CUTOFF, NA, .x)))
+
+```
+
+# Add CDI
 

### 2024-10-04 2_developmental_analysis.qmd
up to date for call
refactor

--- 2024-09-15 2_developmental_analysis.qmd
@@
 Let's also use only `mean(log(rt))` as that's correct if we think RTs are sampled from a log-normal distribution.
+
+TODO:
+- number of trials RT cleaning for slope comparison? (25% dataset? 2 RTs)
+- check factor invariance across age - Measurement invariance?!?
+- report on non-linearity model comparison
 
@@
 ```{r}
-ggplot(d_sub_agevar, aes(x = age, y = short_window_accuracy)) + 
+ggplot(d_sub_agevar, aes(x = age, y = long_window_accuracy)) + 
   geom_point(alpha = .05) + 
@@
 ```
+With fixed effects.
 
-Reaction time models.
+```{r}
+acc_mods_sub_lmer_fe <- list(lwe_lin = lmer(long_window_elogit ~ age_s  
+                                         + (age_s | dataset_name), data = d_sub),
+                          lwe_log = lmer(long_window_elogit ~ log_age_s 
+                                         + (log_age_s | dataset_name), data = d_sub),
+                          swe_lin = lmer(short_window_elogit ~ age_s 
+                                         + (age_s | dataset_name), data = d_sub),
+                          swe_log = lmer(short_window_elogit ~ log_age_s 
+                                         + (log_age_s | dataset_name), data = d_sub),
+                          lwa_lin = lmer(long_window_accuracy ~ age_s
+                                         + (age_s | dataset_name), data = d_sub),
+                          lwa_log = lmer(long_window_accuracy ~ log_age_s
+                                         + (log_age_s | dataset_name), data = d_sub),
+                          swa_lin = lmer(short_window_accuracy ~ age_s
+                                         + (age_s | dataset_name), data = d_sub),
+                          swa_log = lmer(short_window_accuracy ~ log_age_s
+                                         + (log_age_s | dataset_name), data = d_sub))
+
+acc_mods_sub_lmer_summary_fe <- map_df(acc_mods_sub_lmer_fe, ~broom.mixed::glance(.x)) |>
+  mutate(model = names(acc_mods_sub_lmer_fe), 
+         r2 = map_dbl(acc_mods_sub_lmer_fe, 
+                      ~performance::r2_nakagawa(.x)$R2_conditional))
+
+kable(acc_mods_sub_lmer_summary_fe, digits = 4)
+```
+
+### Reaction time models.
 
@@
 
-# Variability
+```{r}
+rt_mods_lmer_fe <- list(log_log = lmer(log_rt ~ log_age_s + 
+                                      (log_age_s | dataset_name), data = d_sub),
+                     log_lin = lmer(log_rt ~ age_s + 
+                                      (age_s | dataset_name), data = d_sub),
+                     lin_log = lmer(rt ~ log_age_s + 
+                                      (log_age_s | dataset_name), data = d_sub),
+                     lin_lin = lmer(rt ~ age_s + 
+                                      (age_s | dataset_name), data = d_sub))
 
-```{r}
-d_var <- d_trial |>
-  group_by(dataset_name, administration_id, age) |>
-  summarise (`SD(log RT)` = sd(log_rt, na.rm=TRUE), 
-             `SD(accuracy)` = sd(long_window_elogit, na.rm=TRUE)) |>
-  pivot_longer(contains("sd"), 
-               values_to = "sd", names_to = "measure")
+rt_mods_lmer_summary_fe <- map_df(rt_mods_lmer_fe, ~broom.mixed::glance(.x)) |>
+  mutate(model = names(rt_mods_lmer_fe), 
+         r2 = map_dbl(rt_mods_lmer_fe, ~performance::r2_nakagawa(.x)$R2_conditional))
 
-v <- ggplot(d_var, aes(x = age, y = sd)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth() +
-  geom_smooth(method = "lm", col = "green") +
-  geom_smooth(data = d_var |>
-  filter(dataset_name %in% datasets_with_age_variation$dataset_name),
-              aes(group = dataset_name, col = dataset_name), geom="line",
-              alpha = .1, method = "lm", size = .5, lty = 2) + 
-  scale_color_discrete(guide = FALSE) + 
-  scale_x_log10() +
-  facet_wrap(~measure, scales = "free_y")
-
-v 
-ggsave(v, filename = "figures/var.png", width = 8, height = 3.5)
-
-```
-# Pub stuff
-
-stats
-
-```{r}
-d_sub |>
-  group_by(dataset_name) |>
-  summarise(n = n(), 
-            `Mean Age` = mean(age), 
-            `Min Age` = min(age), 
-            `Max Age` = max(age), 
-            `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
-            `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
-  xtable(caption = "Characteristics of included datasets from Peekbank.")
-```
-
-```{r}
-a <- ggplot(d_sub_long_elogit |> filter(measure == "long_window_elogit"), 
-       aes(x = age, y = accuracy)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth() +
-  geom_smooth(method = "lm", col = "green") +
-  geom_smooth(data = d_sub_agevar_long_elogit |> filter(measure == "long_window_elogit"),
-              aes(group = dataset_name, col = dataset_name), geom="line",
-              alpha = .1, method = "lm", size = .5, lty = 2) + 
-  geom_hline(yintercept = .5, lty = 2) + 
-  scale_x_log10() +
-  scale_color_discrete(guide = FALSE) + 
-  ylab("Accuracy (elogit)") + 
-  xlab("Age (months)")
-
-b <- ggplot(d_sub_long_rt |> filter(measure == "log_rt"), 
-       aes(x = age, y = rt)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth() +
-  geom_smooth(method = "lm", col = "green") +
-  geom_smooth(data = d_sub_agevar_long_rt |> filter(measure == "log_rt"),
-              aes(group = dataset_name, col = dataset_name), geom="line",
-              alpha = .1, method = "lm", size = .5, lty = 2) + 
-  # geom_hline(yintercept = .5, lty = 2) + 
-  scale_x_log10() +
-  scale_color_discrete(guide = FALSE) + 
-  theme(legend.position = "bottom") + 
-  ylab("Reaction Time (log)") + 
-  xlab("Age (months)")
-
-f <- cowplot::plot_grid(a,b)
-ggsave(f, filename = "figures/acc_rt.png", width = 8, height = 3.5)
-
+kable(rt_mods_lmer_summary_fe, digits = 4)
 ```

### 2024-10-04 3_developmental_relationships.qmd
refactor

@@
 ```{r}
+AGEVAR_CUTOFF <- 6
+
 suppressPackageStartupMessages(library(here))
@@
   summarise(age_range = max(age) - min(age)) |>
-  filter(age_range >= 4)
+  filter(age_range >= AGEVAR_CUTOFF)
 
@@
   ungroup() |>
-  select(log_age, log_rt, long_window_accuracy)
+  select(log_age, rt, long_window_accuracy)
+
+d_sub_mat <- d_sub_mat |>
+  filter(complete.cases(d_sub_mat)) |>
+  as.matrix()
+
+fa.parallel(d_sub_mat)
+```
+
+Now with variance too.
+
+```{r}
+d_sub_mat <- d_sub |>
+  ungroup() |>
+  select(log_age, rt, long_window_accuracy, long_window_acc_var, rt_var)
 
@@
 mod1
-
 ```
@@
          log_age = scale(log_age), 
-                    log_rt = scale(log_rt)) |>
-  select(accuracy, log_age, log_rt)
+         rt = scale(rt)) |>
+  select(accuracy, log_age, rt)
 
@@
 
-fa_model <-  "F1  =~ log_age + log_rt + accuracy"
+fa_model <-  "F1  =~ log_age + rt + accuracy"
 fit <- cfa(fa_model, d_cfa, std.lv=TRUE)
@@
 summary(fit)
-a <- graph_sem(model = fit)+ theme(panel.background = element_rect(fill = "white"))
+graph_sem(model = fit)+ theme(panel.background = element_rect(fill = "white"))
 ```
@@
 
+## Within dataset
+
+```{r}
+
+d_cfa_dataset <- d_cfa |>
+  filter(dataset_name %in% datasets_with_age_variation$dataset_name) |>
+  group_by(dataset_name) |>
+  nest() |>
+  mutate(complete_data = map(data, ~filter(.x[complete.cases(.x),])),
+         mat = as.matrix(complete_data), 
+         has_data = map_lgl(complete_data, ~nrow(.x)>0)) |>
+  filter(has_data) |>
+  mutate(fa = map(mat, ~cfa(fa_model, .x, std.lv=TRUE)), 
+         loadings = map(fa, ~summary(.x)[[5]]))
+
+d_loadings <- d_cfa_dataset |>
+  select(dataset_name, loadings) |>
+  unnest(cols = c(loadings)) |>
+  filter(!is.na(se))
+
+ggplot(filter(d_loadings, lhs == "F1", rhs != "F1"),
+       aes(x = rhs, y = est, col = dataset_name)) + 
+  geom_point(alpha = .5) +
+  stat_summary(fun.data = "mean_cl_boot", aes(group = 1)) +
+  geom_hline(yintercept = 0, lty = 2) + 
+  coord_flip()+ 
+  ylab("Standardized loading") + 
+  xlab("Predictor") +
+  scale_color_discrete(name = "Dataset")
+```
+With production data.
+
+```{r}
+fa_model_prod <-  "F1  =~ log_age + rt + accuracy + prod"
+
+d_cfa_dataset_prod <-  d_sub |>
+  rename(`accuracy` = long_window_accuracy) |>
+  mutate(accuracy = scale(accuracy), 
+         log_age = scale(log_age), 
+         rt = scale(rt), 
+         prod = scale(prod)) |>
+  select(dataset_name, accuracy, log_age, rt, prod) |>
+  filter(dataset_name %in% datasets_with_age_variation$dataset_name) |>
+  group_by(dataset_name) |>
+  nest() |>
+  mutate(complete_data = map(data, ~filter(.x[complete.cases(.x),])),
+         mat = as.matrix(complete_data), 
+         has_data = map_lgl(complete_data, ~nrow(.x)>0)) |>
+  filter(has_data) |>
+  mutate(fa = map(mat, ~cfa(fa_model_prod, .x, std.lv=TRUE)), 
+         loadings = map(fa, ~summary(.x)[[5]]))
+
+d_loadings_prod <- d_cfa_dataset_prod |>
+  select(dataset_name, loadings) |>
+  unnest(cols = c(loadings)) |>
+  filter(!is.na(se))
+
+ggplot(filter(d_loadings_prod, lhs == "F1", rhs != "F1"),
+       aes(x = rhs, y = est, col = dataset_name)) + 
+  geom_point(alpha = .5) +
+  stat_summary(fun.data = "mean_cl_boot", aes(group = 1)) +
+  geom_hline(yintercept = 0, lty = 2) + 
+  coord_flip()+ 
+  ylab("Standardized loading") + 
+  xlab("Predictor") +
+  scale_color_discrete(name = "Dataset")
+```
 ## Measurement invariance across age?
 
+This doesn't work.
+
 ```{r}
-fa_model <-  "F1  =~ log_age + log_rt + accuracy"
+d_cfa_younger <- d_sub |>
+    filter(age < median(age)) |>
+  rename(`accuracy` = long_window_accuracy) |>
+  mutate(accuracy = scale(accuracy), 
+         log_age = scale(log_age), 
+         log_rt = scale(log_rt)) |>
+  select(accuracy, log_age, log_rt) 
+d_cfa_younger <- d_cfa_younger[complete.cases(d_cfa_younger), ]
+
+d_cfa_older <- d_sub |>
+  filter(age >= median(age)) |>
+  rename(`accuracy` = long_window_accuracy) |>
+  mutate(accuracy = scale(accuracy), 
+         log_age = scale(log_age), 
+         log_rt = scale(log_rt)) |>
+  select(accuracy, log_age, log_rt) 
+d_cfa_older <- d_cfa_older[complete.cases(d_cfa_older), ]
+
 fit_younger <- cfa(fa_model, 
-                   filter(d_cfa, log_age <= median(log_age)),
+                   data = d_cfa_younger,
+                   std.lv=TRUE)
+fit_older <- cfa(fa_model, 
+                 data = d_cfa_older,
                  std.lv=TRUE)
 
-graph_sem(model = fit_younger)
+younger_plot <- graph_sem(model = fit_younger)
+older_plot <- graph_sem(model = fit_older)
+plot_grid(younger_plot, older_plot)
 ```

### 2024-10-28 1_get_data.qmd
update

@@
 saveRDS(d_trial, here("cached_intermediates","1_d_trial.Rds"))
+write_csv(d_trial, here("cached_intermediates","1_d_trial.csv"), na = "")
 saveRDS(d_sub, here("cached_intermediates","1_d_sub.Rds"))
+write_csv(d_sub, here("cached_intermediates","1_d_sub.csv"), na = "")
 ```

### 2024-10-28 3_developmental_relationships.qmd
update

--- 2024-09-26 3_developmental_relationships.qmd
@@
 ```{r}
-cor(select(d_sub, age, log_age, rt, log_rt, 
-                       short_window_accuracy, long_window_accuracy, 
-                       short_window_elogit, long_window_elogit, prod, comp), 
+cor(select(d_sub, log_age, rt, 
+           long_window_accuracy, long_window_elogit, prod, comp), 
     use = "pairwise.complete.obs") |>
@@
 
-```{r, fig.width=9, fig.height=9}
-GGally::ggpairs(select(d_sub, age, log_age, rt, log_rt, 
-                       short_window_accuracy, long_window_accuracy, 
-                       short_window_elogit, long_window_elogit, prod, comp), 
+```{r}
+GGally::ggpairs(select(d_sub, log_age, rt, 
+           long_window_accuracy, long_window_elogit, prod, comp), 
                 progress = FALSE, lower = list(continuous = GGally::wrap("points", alpha = 0.03)))
-
-```
-
-Let's select the most interesting of these.
-
-```{r, fig.width=9, fig.height=9}
-d_sub |>
-  select(log_age, log_rt,  
-         long_window_accuracy,  prod, comp) |>
-GGally::ggpairs(progress = FALSE, 
-                lower = list(continuous = GGally::wrap("points", alpha = 0.03)),
-                axisLabels = "show")
 
@@
 d_residualized_by_age <- d_sub |>
-  select(subject_id, administration_id,log_age,log_rt, long_window_accuracy, prod, comp) |> 
-  pivot_longer(cols=c(log_rt,long_window_accuracy,  prod, comp),names_to="measure",values_to="value") |>
+  select(subject_id, administration_id, log_age, rt, long_window_accuracy, prod, comp) |> 
+  pivot_longer(cols=c(rt, long_window_accuracy,  prod, comp),names_to="measure",values_to="value") |>
   group_by(measure) |> 
@@
 ```
-
-# Dimensionality reduction
-
-The claim of some of the Fernald processing corpus is that there is a second principal component here (namely, processing speed) that is meaningful and relates to later learning outcomes.
-
-## All data
-
-Scree plot
-
-```{r}
-d_sub_mat <- d_sub |>
-  ungroup() |>
-  select(log_age, log_rt, long_window_accuracy)
-
-d_sub_mat <- d_sub_mat |>
-  filter(complete.cases(d_sub_mat)) |>
-  as.matrix()
-
-fa.parallel(d_sub_mat)
-```
-```{r}
-mod1 <- fa(d_sub_mat, 1, rotate = "varimax")
-
-```
-
-## Production
-
-Scree plot
-
-```{r}
-d_sub_mat <- d_sub |>
-  ungroup() |>
-  select(log_age, log_rt, long_window_elogit, prod)
-
-d_sub_mat <- d_sub_mat |>
-  filter(complete.cases(d_sub_mat)) |>
-  as.matrix()
-
-fa.parallel(d_sub_mat)
-```
-
-Factor analysis first. This is probably more appropriate.
-
-```{r}
-mod1 <- fa(d_sub_mat, 1, rotate = "varimax")
-mod2 <- fa(d_sub_mat, 2, rotate = "varimax")
-
-summary(mod1)
-mod1
-```
-
-Make into matrices and look at PCA.
-
-```{r}
-
-acc_rt_prc <- prcomp(x = d_sub_mat, scale = TRUE)
-
-acc_rt_prc
-summary(acc_rt_prc)
-ggbiplot::ggbiplot(acc_rt_prc, alpha = .1)
-```
-
-## Production within individual datasets
-
-```{r}
-d_sub_mat_dataset <- d_sub |>
-  filter(dataset_name %in% datasets_with_age_variation$dataset_name) |>
-  group_by(dataset_name) |>
-  select(dataset_name, log_age, log_rt, long_window_elogit, prod)
-
-d_sub_mat_grp <- d_sub_mat_dataset |>
-  nest() |>
-  mutate(complete_data = map(data, ~filter(.x[complete.cases(.x),])),
-    mat = as.matrix(complete_data), 
-    has_data = map_lgl(complete_data, ~nrow(.x)>0)) |>
-  filter(has_data) |>
-  mutate(fas = map(mat, ~fa(.x, 1, rotate = "varimax")), 
-         f_summary = map(fas, summary), 
-         loadings = map(fas, ~tibble(loadings = as.double(loadings(.x)), 
-                                     name = rownames(loadings(.x)))))
-
-ls <- d_sub_mat_grp |>
-  select(dataset_name, loadings) |>
-  unnest(cols = c(loadings))
-
-ggplot(ls, aes(x = name, y = loadings, col = dataset_name)) + 
-  geom_point(aes = .5) +
-  stat_summary(fun.data = "mean_cl_boot", aes(group = 1)) +
-  geom_hline(yintercept = 0, lty = 2) + 
-  coord_flip()
-
-```
-
-## Comprehension
-
-```{r}
-d_sub_mat_comp <- d_sub |>
-  ungroup() |>
-  select(log_age, log_rt,  long_window_elogit, comp)
-
-d_sub_mat_comp <- d_sub_mat_comp |>
-  filter(complete.cases(d_sub_mat_comp)) |>
-  as.matrix()
-
-```
-
-Scree.
-
-```{r}
-fa.parallel(d_sub_mat_comp)
-```
-
-Factor analysis
-
-```{r}
-mod1_comp <- fa(d_sub_mat_comp, 1, rotate = "varimax")
-mod2_comp <- fa(d_sub_mat_comp, 2, rotate = "varimax")
-
-summary(mod1_comp)
-loadings(mod1_comp)
-```
-
-PCA.
-
-```{r}
-acc_rt_prc_comp <- prcomp(x = d_sub_mat_comp, scale = TRUE)
-
-acc_rt_prc_comp
-summary(acc_rt_prc_comp)
-ggbiplot::ggbiplot(acc_rt_prc_comp, alpha = .1)
-```
-
-# CFA models
-
-```{r}
-
-d_cfa <- d_sub |>
-             rename(`accuracy` = long_window_accuracy) |>
-             mutate(accuracy = scale(accuracy), 
-                    log_age = scale(log_age), 
-                    log_rt = scale(log_rt)) |>
-  select(accuracy, log_age, log_rt)
-
-d_cfa <- d_cfa[complete.cases(d_cfa), ]
-
-fa_model <-  "F1  =~ log_age + log_rt + accuracy"
-fit <- cfa(fa_model, d_cfa, std.lv=TRUE)
-
-summary(fit)
-a <- graph_sem(model = fit)+ theme(panel.background = element_rect(fill = "white"))
-```
-
-```{r}
-d_cfa_prod <- d_sub |>
-             rename(`accuracy` = long_window_accuracy) |>
-             mutate(accuracy = scale(accuracy), 
-                    log_age = scale(log_age), 
-                    log_rt = scale(log_rt), 
-                    prod = scale(prod)) |>
-  select(accuracy, log_age, log_rt, prod)
-
-d_cfa_prod <- d_cfa_prod[complete.cases(d_cfa_prod), ]
-
-fa_model_prod <-  "F1  =~ log_age + log_rt + accuracy + prod"
-fit_prod <- cfa(fa_model_prod, d_cfa_prod, std.lv=TRUE)
-
-summary(fit_prod)
-b <- graph_sem(model = fit_prod) + 
-  theme(panel.background = element_rect(fill = "white"))
-b
-```
-```{r}
-d_cfa_comp <- d_sub |>
-             rename(`accuracy` = long_window_accuracy) |>
-             mutate(accuracy = scale(accuracy), 
-                    log_age = scale(log_age), 
-                    log_rt = scale(log_rt), 
-                    comp = scale(comp)) |>
-  select(accuracy, log_age, log_rt, comp)
-
-d_cfa_comp <- d_cfa_comp[complete.cases(d_cfa_comp), ]
-
-fa_model_comp <-  "F1  =~ log_age + log_rt + accuracy + comp"
-fit_comp <- cfa(fa_model_comp, d_cfa_comp, std.lv=TRUE)
-
-summary(fit_comp)
-c <- graph_sem(model = fit_comp,
-            layout = get_layout(x = fit_comp, layout_algorithm = "layout_as_star")) + 
-    theme(panel.background = element_rect(fill = "white"))
-c
-```
-
-```{r}
-plot_grid(a,b,c, nrow = 1, labels = c("A","B","C"))
-ggsave(here("figures","sem.png"), width = 25, height = 4)
-```
-
-# Longitudinal data
-
-```{r}
-longitudinal <- d_sub |>
-  group_by(dataset_name, subject_id) |>
-  count() |>
-  filter(n > 1)
-
-d_sub_long <- d_sub |>
-  filter(subject_id %in% longitudinal$subject_id) |>
-  group_by(subject_id) |>
-  arrange(administration_id) |>
-  mutate(admin_num = 1:n(), 
-         delta_t = c(0,diff(age)))
-
-ggplot(d_sub_long, 
-       aes(x = delta_t, fill = as.factor(admin_num))) +
-  geom_histogram() + 
-  facet_wrap(~dataset_name)
-```
-
-Longitudinal data with CDI is just FMW2013 right now.
-
-```{r}
-fmw <- filter(d_sub_long, dataset_name == "fmw_2013") |>
-  mutate(t = ifelse(age < 23, "t1","t2")) |>
-  group_by(subject_id, t) |>
-  summarise(age = mean(age), 
-            log_rt = mean(log_rt), 
-            prod = mean(prod), 
-            long_window_accuracy = mean(long_window_accuracy),
-            delta_t = mean(delta_t)) |>
-  pivot_wider(names_from = "t", 
-              values_from = c("age","log_rt","prod","long_window_accuracy", "delta_t"))
-
-summary(lm(prod_t2 ~ long_window_accuracy_t1 * log_rt_t1 + age_t2, 
-             data = fmw))
-```
-
-# Conclusions
-
-The next thing to do is look at any longitudinal data we have, to see how predictive faster LWL at t0 is of higher CDIs at t1.

### 2024-10-28 4_factor_analysis.qmd
update

---
title: "Factor analysis"
format: 
  html:
    toc: true
execute: 
  cache: true
---

```{r}
library(tidyverse)
library(lavaan)
library(tidySEM)
library(here)
library(psych)

d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds"))

```

Our goal here is to understand the dimensionality of the data.

Let's use the approach advocated by Nilam, namely test for a single dimension of latent variation.

# Full dataset

```{r}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                       ~ scale(.x)[,1])) 
```

## EFA

First we will fit EFA. You can do this over unscaled data, result is the same.

```{r}
fa.parallel(select(d_sub_mat, -dataset_name), fa = "fa", 
            use = "pairwise.complete.obs")
```

There is some variation on the second dimension, though.

```{r}
fa(select(d_sub_mat, -dataset_name), nfactor = 2,
            use = "pairwise", rotate = "varimax")
```

## CFA

We start with a one factor CFA.

```{r}
fa_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod + comp"
fit <- cfa(fa_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit, fit.measures=TRUE, standardize=TRUE)
```
r2:

```{r}
inspect(fit, 'r2')
```

Missing data coverage:

```{r}
inspect(fit, 'coverage')
```
Look at observed vs. fitted correlations.

```{r}
inspect(fit, what="cor.all")
lavCor(fit)
# resid(fit, "cor")

```

```{r}
graph_sem(model = fit, text_size = 2) + 
  theme(panel.background = element_rect(fill = "white"))
```

Extract F1.

```{r}
d_sub$f1 <- lavPredict(fit, newdata = d_sub_mat_s)
ggplot(d_sub, aes(x = log_age, y = f1)) + 
  geom_point(alpha = .1) + 
  geom_smooth() + 
  geom_smooth(method = "lm", col = "green")
```

This seems somewhat satisfactory (all loadings are in the same direction). But to be honest, the fit indices are not great. This is probably driven by some of the action in the second dimension in the scree plot. Not sure what is going on there, but part of it has to do with the relative lack of correlation between RT and comprehension. I wonder if that has to do with the relatively small amount of comprehension data we have.

# Alternative models

## Speed/accuracy separate

Compare to model with separate speed and accuracy. This one is better fitting.

```{r}

fa2_model <-  "accuracy =~ long_window_accuracy + long_window_acc_var + prod + comp
               speed =~ rt + rt_var"

fit2 <- cfa(fa2_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit2, fit.measures=TRUE, standardize=TRUE)
```

```{r}
layout = matrix(nrow=4, ncol = 4, 
                data = c(NA,NA,NA,"prod",
                         "rt_var",NA,NA,"comp",
                         NA,"speed","accuracy","long_window_accuracy",
                         "rt",NA,NA,"long_window_acc_var"), byrow = TRUE)

graph_sem(model = fit2, text_size = 2, layout = layout )

```

Big improvement.

```{r}
inspect(fit2, 'r2')
```

```{r}
anova(fit, fit2)
```

## Variation separated

Compare to model with vars separate but acc and RT grouped. This is definitely worse.

```{r}

fa2v_model <-  "language =~ rt + long_window_accuracy + prod + comp
               variability =~  long_window_acc_var + rt_var"

fit2v <- cfa(fa2v_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit2v, fit.measures=TRUE, standardize=TRUE)
```
```{r}
graph_sem(model = fit2v, text_size = 2) 
```

Model comparison. There is an improvement.

```{r}
anova(fit, fit2v)
```

Which of these is better?

```{r}
anova(fit2, fit2v)
```

The speed/accuracy model has vastly better fit.

## CDI vs. LWL

Compare to model with CDI variables separated. Note this is confounded by dataset.

```{r}

fa2t_model <-  "task =~ rt + long_window_accuracy + long_window_acc_var + rt_var
               CDI =~  prod + comp"

fit2t <- cfa(fa2t_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit2t, fit.measures=TRUE, standardize=TRUE)
```

```{r}
graph_sem(model = fit2t, text_size = 2) 
```

Model comparison. There is an improvement.

```{r}
anova(fit, fit2t)
```

Which of these is better?

```{r}
anova(fit2, fit2t)
```

## No comprehension

Removing comprehension doesn't seem to clean up the misfit issues.

```{r}
fa_prod_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod"
fit_prod <- cfa(fa_prod_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit_prod, fit.measures=TRUE, standardize=TRUE)
```

```{r}
graph_sem(model = fit_prod, text_size = 2)
```

## Without variances

Without variance.

```{r}

fa_model_novar <-  "F1  =~ rt + long_window_accuracy + prod + comp"
fit_novar <- cfa(fa_model_novar, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit_novar, fit.measures = TRUE, standardized =TRUE)
```

```{r}
graph_sem(model = fit_novar) + 
  theme(panel.background = element_rect(fill = "white"))
```

# Three-factor mode

Compare to model with separate speed and accuracy. This one is better fitting.

```{r}

fa3_model <- "vocab =~ prod + comp
              accuracy =~ long_window_accuracy + long_window_acc_var
              speed =~ rt + rt_var"

fit3 <- cfa(fa3_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit3, fit.measures=TRUE, standardize=TRUE)
```

```{r}
layout = matrix(nrow=2, ncol = 6,
                data = c("speed",NA,"accuracy",NA,"vocab",NA,
                         "rt_var","rt", "long_window_accuracy",
                         "long_window_acc_var", "prod","comp"), 
                byrow = TRUE)

graph_sem(model = fit3, text_size = 2, layout = layout)

```

# Fit models with dataset grouping

One big worry here is that it's between-group confounding that leads to model misfit, not actual factor structure. To deal with this, we fit the same model but with grouping by dataset.

We can't really handle the missing data here sadly, because there's too much missing prod and comp data.

```{r}
fa_model_reduced <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var"

fit_reduced <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
                   missing='fiml')
summary(fit_reduced, fit.measures=TRUE, standardize=TRUE)
```

Now the grouped dataset.

```{r}
fit_reduced_grouped <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
                   missing='fiml', 
                   group = "dataset_name")

summary(fit_reduced_grouped, fit.measures=TRUE, standardize=TRUE)
```

We can look at invariance.

```{r}
fit_reduced_grouped_loadings <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
                   missing='fiml', 
                   group = "dataset_name", 
                   group.equal = "loadings")

# summary(fit_reduced_grouped_loadings, fit.measures=TRUE, standardize=TRUE)

lavTestLRT(fit_reduced_grouped, fit_reduced_grouped_loadings)
```

This test suggests that we reject the null and the non-invariant model is better.

## Fernald and Marchman only

```{r}
fm_data <- d_sub_mat_s |>
  filter(dataset_name %in% c("adams_marchman_2018","fmw_2013"))

fa_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod + comp"

fm_fit <- cfa(fa_model, fm_data, std.lv=TRUE, missing='fiml')

summary(fm_fit, fit.measures=TRUE, standardize=TRUE)
```
```{r}
fa2_model <-  "accuracy =~ long_window_accuracy + long_window_acc_var + prod + comp
               speed =~ rt + rt_var"

fm_fit2 <- cfa(fa2_model, fm_data, std.lv=TRUE, missing='fiml')

summary(fm_fit2, fit.measures=TRUE, standardize=TRUE)
```

### 2024-10-28 5_longitudinal.qmd
update

---
title: "Longitudinal models"
format: html
---

Questions: 
(from Nilam)

1. Is the mean of the latent “language ability” factor scores different across 1st assessment, 2nd assessment, 3rd assessment, etc? This is basically an ANOVA on the latent factor scores – using “assessment number” as the categorical predictor. – an examination of “practice effects”.

Is the mean of the latent “language ability” factor scores increasing with age?
- This is basically a regression on the latent factor scores – and a growth model.
- In which case we also implicitly test research questions about the variances
- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?

- A more nuanced version of the “practice effects" examination might use delta_t. 
- Is the mean of the latent “language ability” factor different across retest interval?
- This is basically a regression on the latent factor scores.

- In which case we also implicitly test research questions about the variances
- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?

3. If we are interested in the covariances –
- Is the structure of the latent “language ability” factor changing with age?
- This again gets us into a multi-group SEM –
- Testing whether the factor loading structure differs across groups.
- This is most easily invoked with using categorical “time variable” – e.g, assessment number, age-group.
- However, it is also possible to do using a continuous time variable (e.g., age). In this case we basically use a moving window to obtain the covariances and then basically use age as a moderator of the factor loadings to examine the age x factor loading interaction.

# Descriptives

## Longitudinal data

```{r}
longitudinal <- d_sub |>
  group_by(dataset_name, subject_id) |>
  count() |>
  filter(n > 1)

d_sub_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)))

ggplot(d_sub_long, 
       aes(x = delta_t, fill = as.factor(admin_num))) +
  geom_histogram() + 
  facet_wrap(~dataset_name)
```
## Datasets with CDIs

Longitudinal data with CDI is just FMW2013 right now.

```{r}
fmw <- filter(d_sub_long, dataset_name == "fmw_2013") |>
  mutate(t = ifelse(age < 23, "t1","t2")) |>
  group_by(subject_id, t) |>
  summarise(age = mean(age), 
            log_rt = mean(log_rt), 
            prod = mean(prod), 
            long_window_accuracy = mean(long_window_accuracy),
            delta_t = mean(delta_t)) |>
  pivot_wider(names_from = "t", 
              values_from = c("age","log_rt","prod","long_window_accuracy", "delta_t"))

summary(lm(prod_t2 ~ long_window_accuracy_t1 * log_rt_t1 + age_t2, 
           data = fmw))
```

## Without CDIs

```{r}
ggplot(d_sub_long, 
       aes(x = age, y = long_window_accuracy, group = subject_id, col = dataset_name)) +
  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
  geom_smooth(aes(group = 1)) + 
  scale_y_log10() + 
  geom_hline(lty = 2, yintercept = .5)

```

```{r}
ggplot(filter(d_sub_long, n_trials_rt > 3),
       aes(x = age, y = log_rt, group = subject_id, col = dataset_name)) +
  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
  geom_smooth(aes(group = 1)) + 
  scale_y_log10() 
```
# Q1: age-related change in the latent score

Growth in latent language factor. This is the same as the previous factor analysis, but we only look within child and within dataset.

## 1 factor approach

This has the benefits of simplicity, even though we know that the 1F model is "wrong."

```{r}
fa_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod + comp"
fit <- cfa(fa_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

d_sub$f1 <- lavPredict(fit, newdata = d_sub_mat_s)

```

```{r}
ggplot(d_sub |>
         filter(subject_id %in% longitudinal$subject_id), 
       aes(x = age, y = f1, group = subject_id, col = dataset_name)) + 
  geom_point(alpha = .4) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = dataset_name), method = "lm")
```
Note that nesting age slopes doesn't converge. Could go Bayesian if we cared...

```{r}
long_mod <- lmer(f1 ~ age + (1 | subject_id) + (1 | dataset_name), 
                 data = d_sub)

summary(long_mod)
```

## 2 factor approach

Let's try looking at the speed/accuracy model across time.

```{r}
fa2_model <-  "accuracy =~ long_window_accuracy + long_window_acc_var + comp + prod
               speed =~ rt + rt_var"

fit2 <- cfa(fa2_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

preds <- predict(fit2)

d_sub$accuracy <- preds[,1]
d_sub$speed <- preds[,2]
```

Merge with longitudinal data.

```{r}
d_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)), 
         speed_0 = speed[1]) |>
  group_by(dataset_name) |>
  mutate(speed_0_high = speed_0 > median(speed_0, na.rm=TRUE))
```

Plot.

```{r}
ggplot(d_long |>
         pivot_longer(speed:accuracy, names_to = "variable", values_to = "value"), 
       aes(x = age, y = value, group = subject_id, col = dataset_name)) + 
  geom_point(alpha = .4) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = dataset_name), method = "lm") + 
  scale_x_log10() +
  facet_wrap(~variable) +
  theme(legend.position = "bottom") + 
  xlab("Age (months)") + 
  ylab("Latent score") 
```
Change in the latents over age.

```{r}
long_mod_acc <- lmer(accuracy ~ age + (1 | subject_id) + (1 | dataset_name), 
                 data = d_sub)
long_mod_speed <- lmer(speed ~ age + (1 | subject_id) + (1 | dataset_name), 
                 data = d_sub)

summary(long_mod_acc)
summary(long_mod_speed)
```

# Q2: relations between latents and other factors

Longitudinal model.

```{r}
lmod <- lmer(accuracy ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
     data = d_long)

summary(lmod)
```

```{r}
ggplot(d_long, 
       aes(x = age, y = accuracy, group = subject_id, col = dataset_name, lty = speed_0_high)) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = speed_0_high), method = "lm") + 
  facet_wrap(~dataset_name, scales="free_x")
```

### 2024-10-28 6_paper_analyses.qmd
update

---
title: "Analyses in the paper"
format: html
---

```{r}

suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(lavaan))
suppressPackageStartupMessages(library(tidySEM))

source(here("helper","common.R"))
d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
  mutate(log_age = log(age))
d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))

d_trial$age_s <- scale(d_trial$age)[,1]
d_trial$log_age_s <- scale(d_trial$log_age)[,1]

```

# Datasets table

```{r}
d_sub |>
  group_by(dataset_name) |>
  summarise(`N subjects` = length(unique(subject_id)),
            `N admins` = n(), 
            `Mean Age` = mean(age), 
            `Min Age` = min(age), 
            `Max Age` = max(age), 
            `Avg Trials` = mean(n_trials),
            `Avg RT Trials` = mean(n_trials_rt),
            `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
            `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
  arrange(desc(`N subjects`)) |>
  xtable(caption = "Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.")
```

# Speed and accuracy

Accuracy

```{r}
acc_mod = lmer(long_window_accuracy ~ log_age_s + (log_age_s | dataset_name) + 
                 (1 | subject_id), 
               data = d_trial)

summary(acc_mod)
```

Reaction time.

```{r}
rt_mod = lmer(rt ~ log_age_s + (log_age_s | dataset_name) + 
                 (1 | subject_id), 
               data = d_trial)

summary(rt_mod)
```

```{r}

datasets_with_age_variation <- d_trial |>
  group_by(dataset_name) |>
  summarise(age_range = max(age) - min(age)) |>
  filter(age_range >= AGEVAR_CUTOFF)

d_sub_agevar <- d_sub |>
  filter(dataset_name %in% datasets_with_age_variation$dataset_name)

a <- ggplot(d_sub,
       aes(x = age, y = long_window_accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_smooth(data = d_sub_agevar,
              aes(group = dataset_name, col = dataset_name), se=FALSE,
              alpha = .1, method = "lm", size = .5, lty = 2) + 
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() +
  scale_color_discrete(guide = FALSE) + 
  ylab("Accuracy") + 
  xlab("Age (months)")

b <- ggplot(d_sub,
       aes(x = age, y = rt)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_smooth(data = d_sub_agevar,
              aes(group = dataset_name, col = dataset_name), se=FALSE,
              alpha = .1, method = "lm", size = .5, lty = 2) + 
  # geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() +
  scale_color_discrete(guide = FALSE) + 
  theme(legend.position = "bottom") + 
  ylab("Reaction Time (log)") + 
  xlab("Age (months)")

f <- cowplot::plot_grid(a,b)
f
ggsave(f, filename = "figures/acc_rt.png", width = 8, height = 3.5)

```

# Variability

```{r}
d_var <- d_sub |>
  rename(RT = rt_var,
         Accuracy = long_window_acc_var) |>
  pivot_longer(c("RT","Accuracy"), 
               values_to = "sd", names_to = "measure")

v <- ggplot(d_var, aes(x = age, y = sd)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_smooth(data = d_var |>
  filter(dataset_name %in% datasets_with_age_variation$dataset_name),
              aes(group = dataset_name, col = dataset_name), 
              alpha = .1, method = "lm", size = .5, lty = 2, se=FALSE) + 
  scale_color_discrete(guide = FALSE) + 
  scale_x_log10() +
  facet_wrap(~measure, scales = "free_y") + 
  ylab("Standard Deviation") + 
  xlab("Age (months)")

v 
ggsave(v, filename = "figures/var.png", width = 8, height = 3.5)

```

Accuracy

```{r}
d_var$age_s <- scale(d_var$age)[,1]
d_var$log_age_s <- scale(d_var$log_age)[,1]

acc_var_mod <- lmer(sd ~ log_age_s + (log_age_s | dataset_name) + 
                 (1 | subject_id), 
               data = filter(d_var, measure == "Accuracy"))

summary(acc_mod)
```

Reaction time.

```{r}
rt_var_mod <- lmer(sd ~ log_age_s + (log_age_s | dataset_name) + 
                 (1 | subject_id), 
               data = filter(d_var, measure == "RT"))
summary(rt_var_mod)
```

# Factor analysis

```{r}
d_cfa <- d_sub |>
  rename(`accuracy` = long_window_accuracy) |>
  mutate(accuracy = scale(accuracy), 
         log_age = scale(log_age), 
         rt = scale(rt)) |>
  select(dataset_name, accuracy, log_age, rt)

d_cfa <- d_cfa[complete.cases(d_cfa), ]

fa_model <-  "F1  =~ log_age + rt + accuracy"
fit <- cfa(fa_model, d_cfa, std.lv=TRUE)

summary(fit)
graph_sem(model = fit)+ theme(panel.background = element_rect(fill = "white"))
```

### 2024-11-15 1_get_data.qmd
paper

@@
 ```{r}
-d_sub <- d_trial |>
+d_sub_nocdi <- d_trial |>
   group_by(dataset_name, subject_id, administration_id, age) |>
@@
 ```{r}
-ggplot(d_sub, aes(x = n_trials, fill = n_trials < 4)) + 
+ggplot(d_sub_nocdi, aes(x = n_trials, fill = n_trials < 4)) + 
   geom_histogram(binwidth = 1) + 
@@
 ```{r}
-ggplot(d_sub, aes(x = n_trials_rt, fill = n_trials_rt < 2)) + 
+ggplot(d_sub_nocdi, aes(x = n_trials_rt, fill = n_trials_rt < 2)) + 
   geom_histogram(binwidth = 1) + 
@@
 ```{r}
-d_sub <- d_sub |>
-  mutate(across(contains("log_rt"), ~ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
+d_sub_nocdi_exclusions <- d_sub_nocdi |>
+  mutate(across(c("rt", "rt_var"), ~ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
+         across(contains("log_rt"), ~ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
          across(contains("window"), ~ifelse(n_trials < ACC_MIN_TRIAL_CUTOFF, NA, .x)))
@@
 
-This is very slightly broken due to some repeated fields. We use `values_fn = mean` to suppress them.
+This join is very tricky and is done based on some careful work in `peekbank-method` to get the English CDIs and fuzzy join them.
 
@@
 
-d_cdi <- cdi_data |>
-  select(administration_id, native_language, measure, language, 
-         CDI_percent) |> # rawscore is form-dependent
-  filter(language == "English (American)", native_language == "eng") |>
-  pivot_wider(names_from = "measure", values_from = "CDI_percent", 
-              values_fill = NA, values_fn = mean)
-```
-
-```{r}
-d_sub <- d_sub |>
-  left_join(d_cdi) |>
+d_sub <- d_sub_nocdi_exclusions |>
+  left_join(select(cdi_data, -age, -dataset_name), 
+            by = c("subject_id","administration_id")) |>
   mutate(log_age = log(age)) |>
@@
 ```{r}
-print(paste(length(unique(d_sub$subject_id)), "children,",
-            sum(!is.na(d_sub$prod)), "with CDI data"))
+print(paste(length(unique(d_sub$subject_id)), "children"))
 print(paste(length(unique(d_sub$administration_id)), "administrations"))
+print(paste(sum(!is.na(d_sub$prod)| !is.na(d_sub$comp)), "administrations have CDI data"))
 print(paste(length(unique(d_sub$dataset_name)), "datasets"))

### 2024-11-15 4_factor_analysis.qmd
paper

@@
   ungroup() |>
-  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp)
+  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)
 
@@
                          "long_window_acc_var", "prod", "comp")), 
-                       ~ scale(.x)[,1])) 
+                       ~ age_scale(.x, age))) 
 ```
@@
 ```{r}
-fa.parallel(select(d_sub_mat, -dataset_name), fa = "fa", 
+fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
             use = "pairwise.complete.obs")
@@
             use = "pairwise", rotate = "varimax")
+```
+Here's what this looks like not forced to be orthogonal.
+
+```{r}
+fa(select(d_sub_mat, -dataset_name, -age), nfactor = 2,
+            use = "pairwise", rotate = "oblimin")
+
 ```
@@
 
-# Three-factor mode
+# Three-factor model
 
@@
 
+## Age regression
+
+```{r}
+
+fa3_age_model <- "
+# measurement model
+vocab =~ prod + comp
+accuracy =~ long_window_accuracy + long_window_acc_var
+speed =~ rt + rt_var
+
+# regressions
+vocab ~ age
+accuracy ~ age
+speed ~ age
+"
+
+fit3_age <- cfa(fa3_age_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
+
+summary(fit3_age, fit.measures=TRUE, standardize=TRUE)
+layout_age = matrix(nrow=5, ncol = 6,
+                data = c(NA,NA,"age",NA, NA, NA,
+                         NA, NA, NA, NA, NA, NA, 
+                         "speed",NA,"accuracy",NA,"vocab",NA,
+                         NA, NA, NA, NA, NA, NA, 
+                         "rt_var","rt", "long_window_accuracy",
+                         "long_window_acc_var", "prod","comp"), 
+                byrow = TRUE)
+
+graph_sem(model = fit3_age, text_size = 2, layout = t(layout_age))
+```
+
 # Fit models with dataset grouping
@@
 
-We can't really handle the missing data here sadly, because there's too much missing prod and comp data.
-
-```{r}
-fa_model_reduced <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var"
-
-fit_reduced <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
-                   missing='fiml')
-summary(fit_reduced, fit.measures=TRUE, standardize=TRUE)
-```
+These mdels generally do not work sadly.
 
@@
 ```{r}
-fit_reduced_grouped <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
-                   missing='fiml', 
-                   group = "dataset_name")
+fa3_model_l2 <- "
+level: 1
+  vocab =~ prod + comp
+  accuracy =~ long_window_accuracy + long_window_acc_var
+  speed =~ rt + rt_var
 
-summary(fit_reduced_grouped, fit.measures=TRUE, standardize=TRUE)
+level: 2
+  vocab =~ prod + comp
+  accuracy =~ long_window_accuracy + long_window_acc_var
+  speed =~ rt + rt_var"
+
+# optimization from https://lavaan.ugent.be/tutorial/multilevel.html
+fit3_grouped <- cfa(fa3_model_l2, d_sub_mat_s, std.lv=TRUE, missing='fiml', 
+                    cluster = "dataset_name", 
+                    optim.method = "em", 
+                    em.iter.max = 20000,
+                    em.fx.tol = 1e-08, em.dx.tol = 1e-04)
+
+summary(fit3_grouped, fit.measures=TRUE, standardize=TRUE)
 ```
 
-We can look at invariance.
+We can look at invariance. Not implemented due to convergence issues.
 
 ```{r}
-fit_reduced_grouped_loadings <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
-                   missing='fiml', 
-                   group = "dataset_name", 
-                   group.equal = "loadings")
+# fit_reduced_grouped_loadings <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
+#                    missing='fiml', 
+#                    group = "dataset_name", 
+#                    group.equal = "loadings")
 
@@
 
-lavTestLRT(fit_reduced_grouped, fit_reduced_grouped_loadings)
+# lavTestLRT(fit_reduced_grouped, fit_reduced_grouped_loadings)
 ```
@@
 This test suggests that we reject the null and the non-invariant model is better.
-
-## Fernald and Marchman only
-
-```{r}
-fm_data <- d_sub_mat_s |>
-  filter(dataset_name %in% c("adams_marchman_2018","fmw_2013"))
-
-fa_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod + comp"
-
-fm_fit <- cfa(fa_model, fm_data, std.lv=TRUE, missing='fiml')
-
-summary(fm_fit, fit.measures=TRUE, standardize=TRUE)
-```
-```{r}
-fa2_model <-  "accuracy =~ long_window_accuracy + long_window_acc_var + prod + comp
-               speed =~ rt + rt_var"
-
-fm_fit2 <- cfa(fa2_model, fm_data, std.lv=TRUE, missing='fiml')
-
-summary(fm_fit2, fit.measures=TRUE, standardize=TRUE)
-```

### 2024-11-15 5_longitudinal.qmd
paper

@@
 title: "Longitudinal models"
-format: html
+format: 
+  html:
+    toc: true
+execute: 
+  cache: true
 ---
 
-Questions: 
-(from Nilam)
+Notes from Nilam:
 
-1.
-Is the mean of the latent “language ability” factor scores different across 1st assessment, 2nd assessment, 3rd assessment, etc?
-This is basically an ANOVA on the latent factor scores – using “assessment number” as the categorical predictor.
-– an examination of “practice effects”.
+- don't z score, mean center + scale by the variance in one age group (e.g., youngest or oldest)
+- check data on Draft 2: latent change scores
+- propagate changes to D6 top model to bottom
+- consider adding shared parameter to ensure invariance across factors at t1 and t2 for D6 model 1
+- what is residual variance on each of the time points - if you say they are equal then it's measurement error, but if you let them vary, it's deviation
+- fit with shared error across all of the observed -->>
+Look at this part of the latent growth tutorial #observed residual variances (constrained to equality)- math2 ~~ sigma2_u*math2
 
-Is the mean of the latent “language ability” factor scores increasing with age?
-- This is basically a regression on the latent factor scores – and a growth model.
-- In which case we also implicitly test research questions about the variances
-- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?
+the two-wave model:
+change in inter-individual differences
+vs. differences in intra-individual change
 
-- A more nuanced version of the “practice effects" examination might use delta_t. 
-- Is the mean of the latent “language ability” factor different across retest interval?
-- This is basically a regression on the latent factor scores.
+This markdown documents attempts to do longitudinal models for the Peekbank data.
 
-- In which case we also implicitly test research questions about the variances
-- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?
+```{r}
+library(here)
+source(here("helper","common.R"))
+library(lavaan)
+library(tidySEM)
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
 
-3. If we are interested in the covariances –
-- Is the structure of the latent “language ability” factor changing with age?
-- This again gets us into a multi-group SEM –
-- Testing whether the factor loading structure differs across groups.
-- This is most easily invoked with using categorical “time variable” – e.g, assessment number, age-group.
-- However, it is also possible to do using a continuous time variable (e.g., age).
-In this case we basically use a moving window to obtain the covariances and then basically use age as a moderator of the factor loadings to examine the age x factor loading interaction.
+```
 
-# Descriptives
+```{r}
+d_sub_mat <- d_sub |>
+  ungroup() |>
+  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)
 
-## Longitudinal data
+d_sub_mat_s <- d_sub_mat |>
+  ungroup() |>
+  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
+                         "long_window_acc_var", "prod", "comp")), 
+                       ~ age_scale(.x, age))) 
+```
+
+# Descriptives of the longitudinal data
 
@@
        aes(x = delta_t, fill = as.factor(admin_num))) +
-  geom_histogram() + 
+  geom_histogram(binwidth = 1) + 
   facet_wrap(~dataset_name)
-```
-## Datasets with CDIs
-
-Longitudinal data with CDI is just FMW2013 right now.
-
-```{r}
-fmw <- filter(d_sub_long, dataset_name == "fmw_2013") |>
-  mutate(t = ifelse(age < 23, "t1","t2")) |>
-  group_by(subject_id, t) |>
-  summarise(age = mean(age), 
-            log_rt = mean(log_rt), 
-            prod = mean(prod), 
-            long_window_accuracy = mean(long_window_accuracy),
-            delta_t = mean(delta_t)) |>
-  pivot_wider(names_from = "t", 
-              values_from = c("age","log_rt","prod","long_window_accuracy", "delta_t"))
-
-summary(lm(prod_t2 ~ long_window_accuracy_t1 * log_rt_t1 + age_t2, 
-           data = fmw))
-```
-
-## Without CDIs
-
-```{r}
-ggplot(d_sub_long, 
-       aes(x = age, y = long_window_accuracy, group = subject_id, col = dataset_name)) +
-  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
-  geom_smooth(aes(group = 1)) + 
-  scale_y_log10() + 
-  geom_hline(lty = 2, yintercept = .5)
 
@@
 
+How much do we have production vs. comprehension vocab?
+
 ```{r}
-ggplot(filter(d_sub_long, n_trials_rt > 3),
-       aes(x = age, y = log_rt, group = subject_id, col = dataset_name)) +
-  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
-  geom_smooth(aes(group = 1)) + 
-  scale_y_log10() 
+d_long |>
+  group_by(dataset_name, subject_id) |>
+  summarise(n_prod = sum(!is.na(prod), na.rm=TRUE), 
+            n_comp = sum(!is.na(comp), na.rm=TRUE)) |>
+  summarise(n_prod = mean(n_prod, na.rm=TRUE), 
+            n_comp = mean(n_comp, na.rm=TRUE))
 ```
-# Q1: age-related change in the latent score
+# Q0: longitudinal stability of measures
 
-Growth in latent language factor. This is the same as the previous factor analysis, but we only look within child and within dataset.
+```{r}
+MONTH_CUTOFF <- 3
+
+d_reliability <- d_long |>
+  filter(delta_t <= MONTH_CUTOFF) |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "admin_num",
+              values_from = c("rt", "rt_var", "long_window_accuracy", 
+                              "long_window_acc_var", "prod", "comp"))
+
+d_reliability |>
+  summarise(rt_corr = cor(rt_1, rt_2, use = "pairwise.complete.obs"),
+            rt_var_corr = cor(rt_var_1, rt_var_2, use = "pairwise.complete.obs"),
+            acc_corr = cor(long_window_accuracy_1, 
+                          long_window_accuracy_2, use = "pairwise.complete.obs"),
+            acc_var_corr = cor(long_window_acc_var_1, 
+                              long_window_acc_var_2, use = "pairwise.complete.obs"),
+            prod_corr = cor(prod_1, prod_2, use = "pairwise.complete.obs"),
+            comp_corr = cor(comp_1, comp_2, use = "pairwise.complete.obs"))
+```
+
+# Q1: age-related changes in latent scores
+
+Let's look at growth in latent language factor.
+
+This is the same as the previous factor analysis, but we only look within child and within dataset.
+
+Summary - this is very interpretable and shows us that we see linear growth in the latents with the log of age.
+
+Downside is that we don't get a sense of the relations between the latents over time.
 
@@
          filter(subject_id %in% longitudinal$subject_id), 
-       aes(x = age, y = f1, group = subject_id, col = dataset_name)) + 
-  geom_point(alpha = .4) + 
-  geom_line(alpha = .4) + 
-  geom_smooth(aes(group = dataset_name), method = "lm")
+       aes(x = age, y = f1, group = subject_id)) + 
+  geom_point(alpha = .2) + 
+  geom_line(alpha = .2) + 
+  geom_smooth(aes(group = dataset_name), method = "lm") + 
+  facet_wrap(~dataset_name)
 ```
@@
 
-# Q2: relations between latents and other factors
+## 3 factor approach
 
-Longitudinal model.
+Let's try looking at the speed/accuracy/vocab model across time.
 
 ```{r}
-lmod <- lmer(accuracy ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
+fa3_model <- "vocab =~ prod + comp
+              accuracy =~ long_window_accuracy + long_window_acc_var
+              speed =~ rt + rt_var"
+
+fit3 <- cfa(fa3_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
+
+preds <- predict(fit3)
+
+d_sub$vocab <- preds[,1]
+d_sub$accuracy <- preds[,2]
+d_sub$speed <- preds[,3]
+```
+
+Merge with longitudinal data.
+
+```{r}
+d_long <- d_sub |>
+  filter(subject_id %in% longitudinal$subject_id) |>
+  group_by(subject_id) |>
+  arrange(administration_id) |>
+  mutate(admin_num = 1:n(), 
+         delta_t = c(0,diff(age)), 
+         speed_0 = speed[1]) |>
+  group_by(dataset_name) |>
+  mutate(speed_0_slow = speed_0 > median(speed_0, na.rm=TRUE))
+```
+
+Plot.
+
+```{r}
+ggplot(d_long |>
+         pivot_longer(vocab:accuracy, names_to = "variable", values_to = "value"), 
+       aes(x = age, y = value, group = subject_id, col = dataset_name)) + 
+  # geom_point(alpha = .1) + 
+  geom_line(alpha = .1) + 
+  geom_smooth(aes(group = dataset_name), method = "lm") + 
+  scale_x_log10() +
+  facet_wrap(~variable) +
+  theme(legend.position = "bottom") + 
+  xlab("Age (months)") + 
+  ylab("Latent score") 
+```
+Change in the latents over age.
+
+```{r}
+long_mod_acc <- lmer(accuracy ~ age + (1 | subject_id) + (1 | dataset_name), 
+                     data = d_sub)
+long_mod_speed <- lmer(speed ~ age + (1 | subject_id) + (1 | dataset_name), 
+                       data = d_sub)
+long_mod_vocab <- lmer(vocab ~ age + (1 | subject_id) + (1 | dataset_name), 
+                       data = d_sub)
+
+summary(long_mod_acc)
+summary(long_mod_speed)
+summary(long_mod_vocab)
+```
+
+# Q2: Does RT relate to growth in vocabulary?
+
+We see from the factor analysis that we are getting a coherent speed factor, but does it relate to anything else over time?
+
+## Draft 1: with LMERs
+
+Longitudinal model of change in accuracy. This is pretty crude, I just wanted to try it.
+
+```{r}
+lmod_acc <- lmer(accuracy ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
                  data = d_long)
 
-summary(lmod)
+summary(lmod_acc)
+```
+
+So speed at time zero does have an effect on accuracy. Low speed at t0 means accuracy is higher. And this interacts with delta_t, but isn't the coefficient in the wrong direction?
+
+Here's a visualization that doesn't quite capture this.
+
+```{r}
+ggplot(filter(d_long, !is.na(speed_0_slow)),
+       aes(x = age, y = accuracy, group = subject_id, 
+           col = dataset_name, 
+           lty = speed_0_slow)) + 
+  geom_line(alpha = .2) + 
+  geom_smooth(aes(group = speed_0_slow), method = "lm") + 
+  facet_wrap(~dataset_name, scales="free_x")
+```
+
+Longitudinal model of change in vocabulary now.
+
+```{r}
+lmod_vocab <- lmer(vocab ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
+                   data = d_long)
+
+summary(lmod_vocab)
+```
+We don't see differences in vocab over time (no interaction of delta_t).
+
+```{r}
+ggplot(d_long, 
+       aes(x = age, y = vocab, group = subject_id, 
+           col = dataset_name, 
+           lty = speed_0_slow)) + 
+  geom_line(alpha = .2) + 
+  geom_smooth(aes(group = speed_0_slow), method = "lm") + 
+  facet_wrap(~dataset_name, scales="free_x")
+```
+
+## Draft 2: latent change scores
+
+Idea behind this approach is to get a delta-vocabulary score for the longitudinal data and look at its relationship to vocab.
+
+```{r}
+d_long_s <- d_long |>
+  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
+                         "long_window_acc_var", "prod", "comp")), 
+                ~ age_scale(.x, age))) 
+```
+
+First fit the three-factor model to just the first obs of the longitudinal data.
+
+```{r}
+fa3l_model <- "vocab_l =~ prod + comp
+              accuracy_l =~ long_window_accuracy + long_window_acc_var
+              speed_l =~ rt + rt_var"
+
+fit3l <- cfa(fa3l_model, 
+             data = filter(d_long_s, admin_num == 1), 
+             std.lv=TRUE, missing='fiml')
+
+summary(fit3l, fit.measures=TRUE, standardize=TRUE)
+```
+```{r}
+layout_l = matrix(nrow=2, ncol = 6,
+                  data = c("speed_l",NA,"accuracy_l",NA,"vocab_l",NA,
+                           "rt_var","rt", "long_window_accuracy",
+                           "long_window_acc_var", "prod","comp"), 
+                  byrow = TRUE)
+
+graph_sem(model = fit3l, text_size = 2, layout = layout_l)
+
+```
+
+Now let's compare to the three-factor model again, with the fit to delta-prod and delta-comp.
+
+```{r}
+slopes <- d_long |>
+  group_by(dataset_name, subject_id) |>
+  nest() |>
+  mutate(delta_prod = map_dbl(data, 
+                              possibly(\(d) coef(lm(prod ~ delta_t, data = d))[2], 
+                                       otherwise = NA)), 
+         delta_comp = map_dbl(data, 
+                              possibly(\(d) coef(lm(comp ~ delta_t, data = d))[2], 
+                                       otherwise = NA))) |>
+  select(-data)
+
+d_long_s <- left_join(d_long_s, slopes)
+
+fa3l_delta_model <- "vocab_l =~ delta_prod + delta_comp
+              accuracy_l =~ long_window_accuracy + long_window_acc_var
+              speed_l =~ rt + rt_var"
+
+fit3l_delta <- cfa(fa3l_delta_model, 
+                   data = filter(d_long_s, admin_num == 1), 
+                   std.lv=TRUE, missing='fiml')
+
+summary(fit3l_delta, fit.measures=TRUE, standardize=TRUE)
 ```
@@
 ```{r}
-ggplot(d_long, 
-       aes(x = age, y = accuracy, group = subject_id, col = dataset_name, lty = speed_0_high)) + 
-  geom_line(alpha = .4) + 
-  geom_smooth(aes(group = speed_0_high), method = "lm") + 
-  facet_wrap(~dataset_name, scales="free_x")
+layout_ld = matrix(nrow=2, ncol = 6,
+                   data = c("speed_l",NA,"accuracy_l",NA,"vocab_l",NA,
+                            "rt_var","rt", "long_window_accuracy",
+                            "long_window_acc_var", "delta_prod","delta_comp"), 
+                   byrow = TRUE)
+graph_sem(model = fit3l_delta, text_size = 2, layout = layout_ld)
 ```
+
+## Draft 3: discrete variables
+
+Need to create discrete variables for each thing changing across time.
+
+Nice piece is that we can do this with all the data.
+
+```{r}
+d_sub_wide_mod_s <- d_sub |>
+  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
+                         "long_window_acc_var", "prod", "comp")),
+                ~ age_scale(.x, age)))
+
+wide_rts <- d_sub_wide_mod_s |>
+  select(dataset_name, subject_id, age, rt) |>
+  mutate(age = floor(age)) |>
+  pivot_wider(names_from = "age", values_from = "rt", 
+              names_prefix = "rt_", 
+              values_fn = mean)
+
+wide_accs <- d_sub_wide_mod_s |>
+  select(dataset_name, subject_id, age, long_window_accuracy) |>
+  mutate(age = floor(age)) |>
+  pivot_wider(names_from = "age", values_from = "long_window_accuracy", 
+              names_prefix = "acc_",
+              values_fn = mean)
+
+wide_prods <- d_sub_wide_mod_s |>
+  select(dataset_name, subject_id, age, prod) |>
+  mutate(age = floor(age)) |>
+  pivot_wider(names_from = "age", values_from = "prod", 
+              names_prefix = "prod_",
+              values_fn = mean)
+
+wide_comps <- d_sub_wide_mod_s |>
+  select(dataset_name, subject_id, age, comp) |>
+  mutate(age = floor(age)) |>
+  pivot_wider(names_from = "age", values_from = "comp", 
+              names_prefix = "comp_",
+              values_fn = mean)
+
+d_wide <- full_join(wide_rts, wide_accs) |>
+  full_join(wide_prods) |>
+  full_join(wide_comps)
+```
+
+```{r}
+ggplot(filter(d_sub_wide_mod_s, subject_id %in% longitudinal$subject_id),
+       aes(x = rt, y = long_window_accuracy, group = subject_id)) + 
+  geom_point(alpha = .2, aes(col = as.factor(subject_id)))+ 
+  geom_line(alpha = .2, aes(col = as.factor(subject_id))) +
+  geom_smooth(aes(group = 1), method = "lm") + 
+  scale_color_discrete(guide = FALSE)
+```
+
+```{r}
+coupled_change <- "accuracy_intercept =~ 1 * acc_13 
+accuracy_intercept =~ 1 * acc_14 
+accuracy_intercept =~ 1 * acc_15 
+accuracy_intercept =~ 1 * acc_16 
+accuracy_intercept =~ 1 * acc_17
+accuracy_intercept =~ 1 * acc_18 
+accuracy_intercept =~ 1 * acc_19
+accuracy_intercept =~ 1 * acc_20
+accuracy_intercept =~ 1 * acc_21
+accuracy_intercept =~ 1 * acc_22
+accuracy_intercept =~ 1 * acc_23
+accuracy_intercept =~ 1 * acc_24
+accuracy_intercept =~ 1 * acc_25
+accuracy_intercept =~ 1 * acc_26
+
+accuracy_slope =~ 0 * acc_13 
+accuracy_slope =~ 1 * acc_14 
+accuracy_slope =~ 2 * acc_15 
+accuracy_slope =~ 3 * acc_16 
+accuracy_slope =~ 4 * acc_17
+accuracy_slope =~ 5 * acc_18 
+accuracy_slope =~ 6 * acc_19
+accuracy_slope =~ 7 * acc_20
+accuracy_slope =~ 8 * acc_21
+accuracy_slope =~ 9 * acc_22
+accuracy_slope =~ 10 * acc_23
+accuracy_slope =~ 11 * acc_24
+accuracy_slope =~ 12 * acc_25
+accuracy_slope =~ 13 * acc_26
+
+rt_intercept =~ 1 * rt_13 
+rt_intercept =~ 1 * rt_14 
+rt_intercept =~ 1 * rt_15 
+rt_intercept =~ 1 * rt_16 
+rt_intercept =~ 1 * rt_17
+rt_intercept =~ 1 * rt_18 
+rt_intercept =~ 1 * rt_19
+rt_intercept =~ 1 * rt_20
+rt_intercept =~ 1 * rt_21
+rt_intercept =~ 1 * rt_22
+rt_intercept =~ 1 * rt_23
+rt_intercept =~ 1 * rt_24
+rt_intercept =~ 1 * rt_25
+rt_intercept =~ 1 * rt_26
+
+rt_slope =~ 0 * rt_13 
+rt_slope =~ 1 * rt_14 
+rt_slope =~ 2 * rt_15 
+rt_slope =~ 3 * rt_16 
+rt_slope =~ 4 * rt_17
+rt_slope =~ 5 * rt_18 
+rt_slope =~ 6 * rt_19
+rt_slope =~ 7 * rt_20
+rt_slope =~ 8 * rt_21
+rt_slope =~ 9 * rt_22
+rt_slope =~ 10 * rt_23
+rt_slope =~ 11 * rt_24
+rt_slope =~ 12 * rt_25
+rt_slope =~ 13 * rt_26
+
+# factor variances
+accuracy_intercept ~~ accuracy_intercept
+accuracy_slope ~~ accuracy_slope
+rt_intercept ~~ rt_intercept
+rt_slope ~~ rt_slope
+
+# covariances among factors 
+accuracy_intercept ~~ rt_intercept
+accuracy_slope ~~ rt_slope
+accuracy_intercept ~~ accuracy_slope
+rt_intercept ~~ rt_slope
+
+#residual covariances
+rt_13 ~~ sigma * acc_13
+rt_14 ~~ sigma * acc_14 
+rt_15 ~~ sigma * acc_15 
+rt_16 ~~ sigma * acc_16 
+rt_17 ~~ sigma * acc_17
+rt_18 ~~ sigma * acc_18 
+rt_19 ~~ sigma * acc_19
+rt_20 ~~ sigma * acc_20
+rt_21 ~~ sigma * acc_21
+rt_22 ~~ sigma * acc_22
+rt_23 ~~ sigma * acc_23
+rt_24 ~~ sigma * acc_24
+rt_25 ~~ sigma * acc_25
+rt_26 ~~ sigma * acc_26
+
+# means
+accuracy_intercept ~ 1
+accuracy_slope ~ 1
+rt_intercept ~ 1
+rt_slope ~ 1
+
+# manifest variances (made equivalent by naming theta)
+acc_13 ~~ theta * acc_13 
+acc_14 ~~ theta * acc_14 
+acc_15 ~~ theta * acc_15 
+acc_16 ~~ theta * acc_16 
+acc_17 ~~ theta * acc_17
+acc_18 ~~ theta * acc_18 
+acc_19 ~~ theta * acc_19
+acc_20 ~~ theta * acc_20
+acc_21 ~~ theta * acc_21
+acc_22 ~~ theta * acc_22
+acc_23 ~~ theta * acc_23
+acc_24 ~~ theta * acc_24
+acc_25 ~~ theta * acc_25
+acc_26 ~~ theta * acc_26
+
+rt_13 ~~ theta2 * rt_13
+rt_14 ~~ theta2 * rt_14 
+rt_15 ~~ theta2 * rt_15 
+rt_16 ~~ theta2 * rt_16 
+rt_17 ~~ theta2 * rt_17
+rt_18 ~~ theta2 * rt_18 
+rt_19 ~~ theta2 * rt_19
+rt_20 ~~ theta2 * rt_20
+rt_21 ~~ theta2 * rt_21
+rt_22 ~~ theta2 * rt_22
+rt_23 ~~ theta2 * rt_23
+rt_24 ~~ theta2 * rt_24
+rt_25 ~~ theta2 * rt_25
+rt_26 ~~ theta2 * rt_26
+
+# manifest means (fixed at zero)
+acc_13 ~ 0*1
+acc_14 ~ 0*1
+acc_15 ~ 0*1
+acc_16 ~ 0*1
+acc_17 ~ 0*1
+acc_18 ~ 0*1
+acc_19 ~ 0*1
+acc_20 ~ 0*1
+acc_21 ~ 0*1
+acc_22 ~ 0*1
+acc_23 ~ 0*1
+acc_24 ~ 0*1
+acc_25 ~ 0*1
+acc_26 ~ 0*1
+
+rt_13 ~ 0*1 
+rt_14 ~ 0*1 
+rt_15 ~ 0*1 
+rt_16 ~ 0*1 
+rt_17 ~ 0*1
+rt_18 ~ 0*1 
+rt_19 ~ 0*1
+rt_20 ~ 0*1
+rt_21 ~ 0*1
+rt_22 ~ 0*1
+rt_23 ~ 0*1
+rt_24 ~ 0*1
+rt_25 ~ 0*1
+rt_26 ~ 0*1"
+
+fit_cc <- lavaan(coupled_change, 
+                 data = d_wide,
+                 std.lv=TRUE, missing='fiml')
+
+summary(fit_cc)
+```
+
+```{r}
+layout_cc = matrix(
+  nrow=4, 
+  ncol = 16,
+  data = c("rt_intercept","rt_slope", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+           NA, NA, "rt_13","rt_14","rt_15","rt_16","rt_17","rt_18","rt_19","rt_20","rt_21","rt_22","rt_23","rt_24","rt_25","rt_26",
+           NA, NA, "acc_13","acc_14","acc_15","acc_16","acc_17","acc_18","acc_19","acc_20","acc_21","acc_22","acc_23","acc_24","acc_25","acc_26",
+           "accuracy_intercept","accuracy_slope", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA),
+  byrow = TRUE)
+
+graph_sem(model = fit_cc, text_size = 3, layout = layout_cc)
+```
+
+## Draft 4: Another version of coupled change
+
+This one doesn't have:
+
+* single residual variance
+* coupled covariances
+* pinned means
+
+```{r}
+coupled_change <- "
+accuracy_intercept =~ 1 * acc_13 + 1 * acc_15 + 1 * acc_16 + 1 * acc_17 + 1 * acc_18 + 1 * acc_19 + 1 * acc_20 + 1 * acc_21 + 1 * acc_22 + 1 * acc_23 + 1 * acc_24 + 1 * acc_25 + 1 * acc_26
+
+accuracy_slope =~ 0 * acc_13  + 1 * acc_14  + 2 * acc_15  + 3 * acc_16  + 4 * acc_17 + 5 * acc_18  + 6 * acc_19 + 7 * acc_20 + 8 * acc_21 + 9 * acc_22 + 10 * acc_23 + 11 * acc_24 + 12 * acc_25 + 13 * acc_26
+
+rt_intercept =~ 1 * rt_13 + 1 * rt_14  + 1 * rt_15  + 1 * rt_16  + 1 * rt_17 + 1 * rt_18  + 1 * rt_19+ 1 * rt_20 + 1 * rt_21 + 1 * rt_22 + 1 * rt_23 + 1 * rt_24 + 1 * rt_25 + 1 * rt_26
+
+rt_slope =~ 0 * rt_13 + 1 * rt_14 + 2 * rt_15 + 3 * rt_16 + 4 * rt_17+ 5 * rt_18 + 6 * rt_19 + 7 * rt_20 + 8 * rt_21 + 9 * rt_22 + 10 * rt_23 + 11 * rt_24 + 12 * rt_25 + 13 * rt_26"
+
+fit_cc <- growth(coupled_change, data = d_wide, missing='fiml')
+
+summary(fit_cc)
+```
+
+```{r}
+graph_sem(model = fit_cc, text_size = 3, layout = layout_cc)
+```
+
+## Draft 5: Adding vocab to the mix
+
+I was trying to figure out how to link the observed variables, but I think the observations are too sparse.
+
+```{r eval=FALSE}
+cc_vocab <- "
+accuracy_intercept =~ 1 * acc_13 + 1 * acc_15 + 1 * acc_16 + 1 * acc_17 + 1 * acc_18 + 1 * acc_19 + 1 * acc_20 + 1 * acc_21 + 1 * acc_22 + 1 * acc_23 + 1 * acc_24 + 1 * acc_25 + 1 * acc_26
+
+accuracy_slope =~ 0 * acc_13  + 1 * acc_14  + 2 * acc_15  + 3 * acc_16  + 4 * acc_17 + 5 * acc_18  + 6 * acc_19 + 7 * acc_20 + 8 * acc_21 + 9 * acc_22 + 10 * acc_23 + 11 * acc_24 + 12 * acc_25 + 13 * acc_26
+
+rt_intercept =~ 1 * rt_13 + 1 * rt_14  + 1 * rt_15  + 1 * rt_16  + 1 * rt_17 + 1 * rt_18  + 1 * rt_19+ 1 * rt_20 + 1 * rt_21 + 1 * rt_22 + 1 * rt_23 + 1 * rt_24 + 1 * rt_25 + 1 * rt_26
+
+rt_slope =~ 0 * rt_13 + 1 * rt_14 + 2 * rt_15 + 3 * rt_16 + 4 * rt_17+ 5 * rt_18 + 6 * rt_19 + 7 * rt_20 + 8 * rt_21 + 9 * rt_22 + 10 * rt_23 + 11 * rt_24 + 12 * rt_25 + 13 * rt_26
+
+prod_intercept =~ 1 * prod_13 + 1 * prod_14  + 1 * prod_15  + 1 * prod_16  + 1 * prod_17 + 1 * prod_18  + 1 * prod_19+ 1 * prod_20 + 1 * prod_21 + 1 * prod_22 + 1 * prod_23 + 1 * prod_24 + 1 * prod_25 + 1 * prod_26
+
+prod_slope =~ 0 * prod_13 + 1 * prod_14 + 2 * prod_15 + 3 * prod_16 + 4 * prod_17+ 5 * prod_18 + 6 * prod_19 + 7 * prod_20 + 8 * prod_21 + 9 * prod_22 + 10 * prod_23 + 11 * prod_24 + 12 * prod_25 + 13 * prod_26"
+
+fit_ccv <- growth(cc_vocab, data = d_wide, missing='fiml', 
+                  optim.method = "GN", 
+                  optim.gn.tol.x = .001, 
+                  optim.gn.iter.max = 500
+)
+
+summary(fit_ccv)
+```
+
+This model does not converge. Let's think about other ways to describe the same coupling.
+
+## Draft 6: Longitudinal common factor model
+
+Let's think about a two-point factor model, where we:
+* analyze all the data
+* average observations 0-2 months apart and then analyze 3-6 month growth
+
+First prep the data for this.
+
+```{r}
+d_sub_long <- d_sub |>
+  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
+                         "long_window_acc_var", "prod", "comp")),
+                ~ age_scale(.x,age)) |>
+  group_by(subject_id) |>
+  arrange(administration_id) |>
+  mutate(admin_num = 1:n(), 
+         delta_t = c(0,diff(age)))
+
+d_sub_long_t <- d_sub_long |> 
+  mutate(t = ifelse(delta_t > 2, "t2", "t1")) |>
+  group_by(subject_id, dataset_name, t) |>
+  summarise(across(all_of(c("rt", "rt_var", "long_window_accuracy",
+                            "long_window_acc_var", "prod", "comp")),
+                   ~ mean(.x, na.rm=TRUE))) |> 
+  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
+                         "long_window_acc_var", "prod", "comp")),
+                ~ ifelse(is.nan(.x), NA, .x)))
+
+d_sub_wide <- d_sub_long_t |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "t",
+              values_from = c("rt", "rt_var", "long_window_accuracy", 
+                              "long_window_acc_var", "prod", "comp"))
+```
+
+Big choice point here is whether to constrain the loadings, which makes something wrong but interpretable.
+
+```{r}
+fa3_model_long <- "
+# measurement model
+vocab_t1 =~ 1*prod_t1 + 1*comp_t1
+accuracy_t1 =~ 1*long_window_accuracy_t1 + 1*long_window_acc_var_t1
+speed_t1 =~ 1*rt_t1 + 1*rt_var_t1
+
+vocab_t2 =~ 1*prod_t2 + 1*comp_t2
+accuracy_t2 =~ 1*long_window_accuracy_t2 + 1*long_window_acc_var_t2
+speed_t2 =~ 1*rt_t2 + 1*rt_var_t2
+
+# longitudinal relationships
+vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+
+# factor covariances
+vocab_t1 ~~ accuracy_t1
+vocab_t1 ~~ speed_t1
+accuracy_t1 ~~ speed_t1
+
+vocab_t2 ~~ accuracy_t2
+vocab_t2 ~~ speed_t2
+accuracy_t2 ~~ speed_t2
+"
+
+fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(fit3_long, fit.measures=TRUE, standardize=TRUE)
+
+layout_long = matrix(nrow=5, ncol = 7,
+                     data = c("rt_var_t1","rt_t1", "long_window_accuracy_t1", "long_window_acc_var_t1", "prod_t1","comp_t1", NA,
+                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
+                              NA, NA, NA, NA, NA, NA,  NA,
+                               NA,"speed_t2",NA,"accuracy_t2",NA,"vocab_t2",NA,
+                               NA,"rt_var_t2","rt_t2", "long_window_accuracy_t2", "long_window_acc_var_t2", "prod_t2","comp_t2"), 
+                     byrow = TRUE)
+graph_sem(model = fit3_long, text_size = 3, layout = t(layout_long))
+
+```
+To test for fit:
+
+- Constrain regression coefficients to be zero and test for misfit (e.g., key fernald coefficient)
+- those are nested so chi2 test can compare between them 
+- maybe constrain ALL cross-influences to zero
+
+Alternative with no identification constraints:
+
+```{r}
+fa3_model_long <- "
+# measurement model
+vocab_t1 =~ 1*prod_t1 + comp_t1
+accuracy_t1 =~ 1*long_window_accuracy_t1 + long_window_acc_var_t1
+speed_t1 =~ 1*rt_t1 + rt_var_t1
+
+vocab_t2 =~ 1*prod_t2 + comp_t2
+accuracy_t2 =~ 1*long_window_accuracy_t2 + long_window_acc_var_t2
+speed_t2 =~ 1*rt_t2 + rt_var_t2
+
+# longitudinal relationships
+vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+
+# factor covariances
+vocab_t1 ~~ accuracy_t1
+vocab_t1 ~~ speed_t1
+accuracy_t1 ~~ speed_t1
+
+vocab_t2 ~~ accuracy_t2
+vocab_t2 ~~ speed_t2
+accuracy_t2 ~~ speed_t2
+
+# means for the latents
+vocab_t1 ~ 1
+vocab_t2 ~ 1
+accuracy_t1 ~ 1
+accuracy_t2 ~ 1
+speed_t1 ~ 1
+speed_t2 ~ 1
+
+# residual variance for the latents
+vocab_t1 ~~ NA*vocab_t1
+vocab_t2 ~~ NA*vocab_t2
+accuracy_t1 ~~ NA*accuracy_t1
+accuracy_t2 ~~ NA*accuracy_t2
+speed_t1 ~~ NA*speed_t1
+speed_t2 ~~ NA*speed_t2
+"
+
+fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(fit3_long, fit.measures=TRUE, standardize=TRUE)
+
+graph_sem(model = fit3_long, text_size = 3, layout = t(layout_long))
+```
+
+Now with baseline compariosns
+
+```{r}
+fa3_model_long_baseline <- "
+# measurement model
+vocab_t1 =~ 1*prod_t1 + comp_t1
+accuracy_t1 =~ 1*long_window_accuracy_t1 + long_window_acc_var_t1
+speed_t1 =~ 1*rt_t1 + rt_var_t1
+
+vocab_t2 =~ 1*prod_t2 + comp_t2
+accuracy_t2 =~ 1*long_window_accuracy_t2 + long_window_acc_var_t2
+speed_t2 =~ 1*rt_t2 + rt_var_t2
+
+# longitudinal relationships
+vocab_t2 ~ vocab_t1 + 0 * speed_t1 + 0 * accuracy_t1
+speed_t2 ~ 0 * vocab_t1 + speed_t1 + 0 * accuracy_t1
+accuracy_t2 ~ 0 * vocab_t1 + 0 * speed_t1 + accuracy_t1
+
+# factor covariances
+vocab_t1 ~~ accuracy_t1
+vocab_t1 ~~ speed_t1
+accuracy_t1 ~~ speed_t1
+
+vocab_t2 ~~ accuracy_t2
+vocab_t2 ~~ speed_t2
+accuracy_t2 ~~ speed_t2
+
+# means for the latents
+vocab_t1 ~ 1
+vocab_t2 ~ 1
+accuracy_t1 ~ 1
+accuracy_t2 ~ 1
+speed_t1 ~ 1
+speed_t2 ~ 1
+
+# residual variance for the latents
+vocab_t1 ~~ NA*vocab_t1
+vocab_t2 ~~ NA*vocab_t2
+accuracy_t1 ~~ NA*accuracy_t1
+accuracy_t2 ~~ NA*accuracy_t2
+speed_t1 ~~ NA*speed_t1
+speed_t2 ~~ NA*speed_t2
+"
+
+fit3_long_baseline <- sem(fa3_model_long_baseline, 
+                          d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(fit3_long_baseline, fit.measures=TRUE, standardize=TRUE)
+
+anova(fit3_long, fit3_long_baseline)
+```
+With a close baseline.
+
+```{r}
+fa3_model_long_closebaseline <- "
+# measurement model
+vocab_t1 =~ 1*prod_t1 + comp_t1
+accuracy_t1 =~ 1*long_window_accuracy_t1 + long_window_acc_var_t1
+speed_t1 =~ 1*rt_t1 + rt_var_t1
+
+vocab_t2 =~ 1*prod_t2 + comp_t2
+accuracy_t2 =~ 1*long_window_accuracy_t2 + long_window_acc_var_t2
+speed_t2 =~ 1*rt_t2 + rt_var_t2
+
+# longitudinal relationships
+vocab_t2 ~ vocab_t1 + 0* speed_t1 + accuracy_t1
+speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+
+# factor covariances
+vocab_t1 ~~ accuracy_t1
+vocab_t1 ~~ speed_t1
+accuracy_t1 ~~ speed_t1
+
+vocab_t2 ~~ accuracy_t2
+vocab_t2 ~~ speed_t2
+accuracy_t2 ~~ speed_t2
+
+# means for the latents
+vocab_t1 ~ 1
+vocab_t2 ~ 1
+accuracy_t1 ~ 1
+accuracy_t2 ~ 1
+speed_t1 ~ 1
+speed_t2 ~ 1
+
+# residual variance for the latents
+vocab_t1 ~~ NA*vocab_t1
+vocab_t2 ~~ NA*vocab_t2
+accuracy_t1 ~~ NA*accuracy_t1
+accuracy_t2 ~~ NA*accuracy_t2
+speed_t1 ~~ NA*speed_t1
+speed_t2 ~~ NA*speed_t2
+"
+
+fit3_long_closebaseline <- sem(fa3_model_long_closebaseline, 
+                               d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(fit3_long_closebaseline, fit.measures=TRUE, standardize=TRUE)
+
+anova(fit3_long, fit3_long_closebaseline)
+```
+
+## Draft 7: slope/intercept on delta-t
+
+Working from [https://thechangelab.stanford.edu/tutorials/growth-modeling/multivariate-latent-change-score-models/]()
+
+Right now I think we need to cut by 2 month intervals because of data sparsity.
+
+```{r}
+d_sub_long_dt <- d_sub |> 
+   mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
+                         "long_window_acc_var", "prod", "comp")), 
+                ~ age_scale(.x,age))) |>
+  group_by(subject_id) |>
+  arrange(administration_id) |>
+  mutate(admin_num = 1:n(), 
+         delta_t = c(0,diff(age))) |>
+  mutate(delta_t = cut(delta_t, 
+                       breaks = c(-10, 2, 4, 6, 30), 
+                       labels = c(0, 2, 4, 6))) |>
+  group_by(subject_id, dataset_name, delta_t) |>
+  summarise(across(all_of(c("rt", "rt_var", "long_window_accuracy",
+                            "long_window_acc_var", "prod", "comp")),
+                   ~ mean(.x, na.rm=TRUE))) |> 
+  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
+                         "long_window_acc_var", "prod", "comp")),
+                ~ ifelse(is.nan(.x), NA, .x)))
+
+d_sub_wide_dt <- d_sub_long_dt |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "delta_t",
+              values_from = c("rt", "rt_var", "long_window_accuracy", 
+                              "long_window_acc_var", "prod", "comp"))
+
+d_sub_wide_dt |>
+  ungroup() |>
+  summarise(across(rt_0:comp_4, 
+                   ~sum(!is.na(.x))))
+```
+
+Note there is no comp_6 and comp_4
+
+** Note propagate the intercept and also the fit variances from D6 ***
+
+```{r}
+dt_model <- "
+# measurement model
+vocab_int =~ 1*prod_0 + 1*comp_0 + 1*prod_2 + 1*comp_2 + 1*prod_4 + 1*prod_6 
+vocab_slope =~ 1*prod_0 + 1*comp_0 + 2*prod_2 + 2*comp_2 + 3*prod_4 + 4*prod_6
+
+accuracy_int =~ 1*long_window_accuracy_0 + 1*long_window_acc_var_0 + 1*long_window_accuracy_2 + 1*long_window_acc_var_2 + 1*long_window_accuracy_4 + 1*long_window_acc_var_4 + 1*long_window_accuracy_6 + 1*long_window_acc_var_6
+
+accuracy_slope =~ 1*long_window_accuracy_0 + 1*long_window_acc_var_0 + 2*long_window_accuracy_2 + 2*long_window_acc_var_2 + 3*long_window_accuracy_4 + 3*long_window_acc_var_4 + 4*long_window_accuracy_6 + 4*long_window_acc_var_6
+
+speed_int =~ 1*rt_var_0 + 1*rt_0 + 1*rt_var_2 + 1*rt_2 + 1*rt_var_4 + 1*rt_4 + 1*rt_var_6 + 1*rt_6 
+speed_slope =~ 1*rt_0 + 1*rt_var_0 + 2*rt_2 + 2*rt_var_2 + 3*rt_4 + 3*rt_var_4 + 4*rt_6 + 4*rt_var_6
+
+# covariances
+vocab_int ~~ vocab_slope
+accuracy_int ~~ accuracy_slope
+speed_int ~~ speed_slope
+
+# couplings
+# vocab_int ~~ accuracy_int
+# vocab_int ~~ speed_int
+# accuracy_int ~~ speed_int
+# vocab_slope ~~ accuracy_slope
+# vocab_slope ~~ speed_slope
+# accuracy_slope ~~ speed_slope
+
+# means for the latents
+vocab_int ~ 1
+vocab_slope ~ 1
+accuracy_int ~ 1
+accuracy_slope ~ 1
+speed_int ~ 1
+speed_slope ~ 1
+
+# residual variance for the latents
+vocab_int ~~ NA*vocab_int
+vocab_slope ~~ NA*vocab_slope
+accuracy_int ~~ NA*accuracy_int
+accuracy_slope ~~ NA*accuracy_slope
+speed_int ~~ NA*speed_int
+speed_slope ~~ NA*speed_slope
+
+# age regressions
+
+"
+
+fit_dt <- sem(dt_model, d_sub_wide_dt, std.lv=TRUE, missing='fiml')
+
+summary(fit_dt, fit.measures=TRUE, standardize=TRUE)
+
+layout_dt = matrix(nrow=11, ncol = 7,
+                   data = c("prod_0", "prod_2", "prod_4", "prod_6", "vocab_int", NA, NA,t
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "comp_0", "comp_2", NA, NA, "vocab_slope", NA,NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "long_window_accuracy_0", "long_window_accuracy_2", "long_window_accuracy_4", "long_window_accuracy_6", "accuracy_int", NA, NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "long_window_acc_var_0", "long_window_acc_var_2", "long_window_acc_var_4", "long_window_acc_var_6", "accuracy_slope", NA, NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "rt_0", "rt_2", "rt_4", "rt_6", "speed_int", NA, NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "rt_var_0", "rt_var_2", "rt_var_4", "rt_var_6", "speed_slope", NA, NA),
+                   byrow = TRUE)
+
+graph_sem(model = fit_dt, text_size = 3, layout = t(layout_dt[,c(7,6, 5,1:4)]))
+
+```

### 2024-11-15 6_paper_analyses.qmd
paper

@@
 suppressPackageStartupMessages(library(tidySEM))
+suppressPackageStartupMessages(library(xtable))
 
@@
 
+AGEVAR_CUTOFF <- 6
+
 ```
@@
 ```{r}
-d_sub |>
+dataset_characteristics <- d_sub |>
   group_by(dataset_name) |>
@@
             `Max Age` = max(age), 
-            `Avg Trials` = mean(n_trials),
-            `Avg RT Trials` = mean(n_trials_rt),
+            `Avg Trials` = mean(n_trials[!is.na(long_window_accuracy)]),
+            `Avg RT Trials` = mean(n_trials_rt[!is.na(rt)]),
             `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
             `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
-  arrange(desc(`N subjects`)) |>
+  arrange(desc(`N subjects`)) 
+knitr::kable(dataset_characteristics)
+```
+
+```{r}
+dataset_characteristics |>
   xtable(caption = "Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.")
@@
 b <- ggplot(d_sub,
-       aes(x = age, y = rt)) + 
+       aes(x = age, y = log_rt)) + 
   geom_point(alpha = .05) + 
@@
   # geom_hline(yintercept = .5, lty = 2) + 
-  scale_x_log10() +
+  # scale_x_log10() +
   scale_color_discrete(guide = FALSE) + 

### 2024-11-15 paper/paper.Rmd
paper

---
title: Continuous developmental changes in word recognition across early childhood

# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
author:
  - name: Alice Anonymous
    affiliation: a,1,2
  - name: Bob Security
    affiliation: a,b
address:
  - code: a
    address: Some Institute of Technology, Department, Street, City, State, Zip
  - code: b
    address: Another University Department, Street, City, State, Zip

corresponding_author:
  code: 2
  text: "To whom correspondence should be addressed. E-mail: bob@email.com"

# For footer text
lead_author_surname: Anonymous

author_contributions: |
  Please provide details of author contributions here.

abstract: |
  Being a fluent language user involves not just knowing how to combine words into meaningful sentences, but also being able to understand language as it unfolds in time. How does this skill develop over the course of early childhood?  And how does facility in word recognition relate to the growth of vocabulary?  We address these questions using data from Peekbank, an open database of eye-tracking experiments on children's early word recognition.  Combining 24 datasets with data from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development, both within and across studies. Factor analysis revealed a cross-sectional coupling of word recognition speed and accuracy to parent-reported vocabulary and a longitudinal relationship such that children with faster language processing showed larger vocabulary growth. Together, these findings are consistent with the view that language processing is –– among other things –– a skill, and language learning is a process of skill learning.

significance: |
  Early language development is a key part of children's development that also sets the stage for later academic success. The efficiency with which children recognize words provides a measurement of their real-time language processing, which has been argued to be a key part of the learning process. Here we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age. This trajectory is consistent with the idea that understanding language is a skill that improves gradually with practice.

acknowledgements: |
  This work was supported by a grant from the Jacobs Foundation.

keywords:
  - one
  - two
  - optional
  - optional
  - optional

## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
pnas_type: pnasresearcharticle

bibliography: peekbank.bib
csl: pnas.csl

## change to true to add optional line numbering
lineno: false

output: rticles::pnas_article
---

\dropcap{C}hildren acquiring a language are learning a body of knowledge -- a set of words and the ways they are combined -- but they are also learning to deploy this knowledge in the myriad complex, noisy, and fast-moving environments in which language is used. As children enter their second year, language explodes onto the scene; children's vocabulary and grammatical abilities grow rapidly in tandem [@bates1994developmental;@frank2021]. This growth in knowledge is also accompanied by changes in their language processing: they become quicker and more accurate in recognizing words and matching them with their referents [@fernald1998;@peter2019;@bergelson2020comprehension].

Yet unlike language production, which is manifest via overt behavior, evidence for word recognition is often more subtle. Very young children with incomplete knowledge may not be able to point to the correct referent of a word, but they may still have some representation of word meaning [@bergelson2012]. Eye tracking has thus emerged as a key method that allows the measurement of language comprehension with high temporal resolution: both adults and children reliably fixate the referent of a spoken word soon after it is spoken [@tanenhaus1995integration;@altmann1999incremental;@fernald1998;@fernald2008]. The relative timecourse of fixation then can provide an index of both the comprehender's ability and the properties of the stimulus.

The version of this method that is used with children goes by many names, including the "intermodal preferential looking paradigm" and the "looking while listening" paradigm (or LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008]. In these experiments, children are typically shown a series of trials in which two images are displayed side by side and they are asked to find one of them. For example, a ball and a book might be shown, and the child might hear "Look at the ball! Can you find it?" Accuracy is then computed as the proportion of time they fixate the correct image within a fixed window after the onset of the noun ("ball" in this case). Reaction time is typically computed only on trials in which the child is fixating the distractor image (the book); in these cases, the average time it takes to shift fixation to the target is used as an index of processing speed. 
The seminal work using this method showed that children's speed and accuracy both increase rapidly across the second year [@fernald1998;@reznick1990visual].

<!-- % MCF: we go back and forth now between word recognition and language processing. is this oK? -->
<!-- % MCF: I haven't adopted the fernald "efficiency" term, which groups speed and accuracy as one process... in part this seems to presuppose the answer to some bits of our factor analysis... -->

Language processing ability is hypothesized to play a key role in language learning. Each word that a child hears is an opportunity to learn; measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking]. These opportunities are likely helpful to different degrees, however [@kachergis2022toward;@mcmurray2007defusing]. Language is likely to be a better learning signal if it is focused on the child's own interest, grounded in the particulars of their current environment -- for example, in a game, routine, or object -- and uses vocabulary and syntax that they know [@tomasello2001perceiving;@roy2015predicting;@cartmill2013quality]. But processing is also prerequisite for learning. Consider a child hearing the utterance "can you put the ball in the box?" The faster and more accurately the child can recognize that the ball is a referent, the better they can use this evidence to help infer the speaker's intended meaning, perhaps making inferences about the meaning of "put" or "box" [@frank2009using]. Consistent with this idea, in one study, children's language processing skill mediated the relationship between home language input and vocabulary growth [@Weisleder2013].

<!-- % This method has provided an important window into how children process phonological [@???}, syntactic [@trueswell1999kindergarten}, and semantic [@mani} competition; how their lexical representations develop [@Swingley2002}; etc. -->

Speed of word recognition has been used as an index of individual differences in early childhood [@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow]. Word recognition speed at 18 months can predict children's later language ability over and above measures of vocabulary size [@marchman2008], though these predictions may be limited to particular ages or processing assessments [@peter2019]. Further, faster processing at age two is predictive of whether "late talkers" catch up to their peers or require further intervention [@fernald2012].  Critically, all of these assessments use words that children are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.

<!-- % MCF: is this question statement sharp enough?  -->

Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages. These samples provide neither the range of ages nor the number of participants to estimate the broader dynamics of how word recognition changes developmentally and how it connects with broader changes across the lifespan [@mcmurray2022slow;@colby2023efficiency]. Further, previous findings do not shed light on whether processing efficiency indexes a separate construct and might vary independently from vocabulary size, or whether it is tightly coupled to vocabulary such that children with larger vocabularies are also better language processors.

% suggest that speed of processing is related to concurrent and later language ability [@peter2019;@marchman2008]. This work is consistent with several explanations, however. First, it is possible that processing speed indexes a different construct than accuracy. On this kind of account, speedier processors may be differentiated from those children with a larger vocabulary; identifying these children thus helps to increase predictive accuracy. An alternative account, however, is that processing speed and accuracy reflect two aspects of how a single construct is translated into behavior (in this case, eye movements). On this second account, the increase in predictive accuracy by using reaction times is accounted for because each has independent measurement error -- the inclusion of both into a model thus increases precision. Thus, the factor structure of early language processing -- and in particular, whether there is evidence for orthogonal variation in early language related to processing speed -- is an open question.

```{r loading, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(psych))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(lavaan))
suppressPackageStartupMessages(library(tidySEM))
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(library(kableExtra))

source(here("helper","common.R"))
d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
  mutate(log_age = log(age))
d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))

d_trial$age_s <- scale(d_trial$age)[,1]
d_trial$log_age_s <- scale(d_trial$log_age)[,1]   
d_sub$age_s <- scale(d_sub$age)[,1]
d_sub$log_age_s <- scale(d_sub$log_age)[,1]

AGEVAR_CUTOFF <- 6

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
```

To address these questions, we made use of Peekbank, an open database of  looking while listening data from young children, stored in a harmonized format [@zettersten2023peekbank]. Aiming to map the development of language processing across ages 1 -- 6 years, we analyzed data from simple word recognition trials in which children were shown two pictures of concrete objects and heard an English word. (We focused on English purely for practical reasons -- we had very limited data from other languages). These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions). Table \ref{tab:datasets} shows the characteristics of individual datasets. The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allow us to take a more detailed look at early language processing than has previously been possible.

```{r dataset_table}
dataset_characteristics <- d_sub |>
  group_by(dataset_name) |>
  summarise(`N subjects` = length(unique(subject_id)),
            `N admins` = n(), 
            `Mean Age` = mean(age), 
            `Min Age` = min(age), 
            `Max Age` = max(age), 
            `Avg Trials` = mean(n_trials[!is.na(long_window_accuracy)]),
            `Avg RT Trials` = mean(n_trials_rt[!is.na(rt)]),
            `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
            `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
  arrange(desc(`N subjects`)) 
```

```{r eval=FALSE}
xtable(dataset_characteristics, 
       caption = "Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.")
```

\begin{table*}[t!]
\centering
\begin{tabular}{rlrrrrrrrll}
  \hline
 & dataset\_name & N subjects & N admins & Mean Age & Min Age & Max Age & Avg Trials & Avg RT Trials & CDIs & longitudinal \\ 
  \hline
1 & reflook\_v4 & 310 & 310 & 36.82 & 12.24 & 60.00 & 6.21 & 2.87 &  &  \\ 
  2 & yurovsky\_2017 & 282 & 282 & 25.64 & 12.59 & 58.65 & 5.95 & 2.79 &  &  \\ 
  3 & weaver\_zettersten\_2024 & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 & x & x \\ 
  4 & fernald\_marchman\_2012 & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.78 & 7.17 & x & x \\ 
  5 & sander-montant\_2022 & 122 & 122 & 21.94 & 12.02 & 31.11 & 10.07 & 4.20 & x &  \\ 
  6 & frank\_tablet\_2016 & 104 & 104 & 33.89 & 12.13 & 59.84 & 5.87 & 2.74 &  &  \\ 
  7 & baumgartner\_2014 & 100 & 100 & 12.01 & 12.00 & 13.00 & 4.00 & 2.31 & x &  \\ 
  8 & fmw\_2013 &  80 & 179 & 20.04 & 17.00 & 26.00 & 21.26 & 8.44 & x & x \\ 
  9 & adams\_marchman\_2018 &  69 & 711 & 23.57 & 12.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
  10 & potter\_canine &  67 &  67 & 23.76 & 21.00 & 27.00 & 10.38 & 3.80 & x &  \\ 
  11 & fernald\_totlot &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.22 & 6.05 & x & x \\ 
  12 & pomper\_prime &  63 &  63 & 39.73 & 37.90 & 42.00 & 12.98 & 5.34 &  &  \\ 
  13 & pomper\_saffran\_2016 &  60 &  60 & 44.27 & 41.00 & 47.00 & 7.55 & 3.30 &  &  \\ 
  14 & swingley\_aslin\_2002 &  50 &  50 & 15.09 & 14.13 & 16.00 & 11.74 & 3.79 & x &  \\ 
  15 & pomper\_salientme &  44 &  44 & 40.11 & 38.00 & 43.00 & 5.30 & 2.34 &  &  \\ 
  16 & perry\_cowpig &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 &  &  \\ 
  17 & ronfard\_2021 &  40 &  40 & 19.95 & 18.00 & 24.00 & 18.54 & 7.62 & x &  \\ 
  18 & bacon\_gendercues &  38 &  38 & 22.87 & 22.00 & 24.00 & 18.16 & 8.00 & x &  \\ 
  19 & garrison\_bergelson\_2020 &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.94 & 9.59 & x &  \\ 
  20 & pomper\_yumme &  32 &  32 & 26.38 & 25.00 & 28.00 & 7.62 & 2.31 &  &  \\ 
  21 & mahr\_coartic &  29 &  29 & 20.83 & 18.10 & 23.80 & 24.38 & 8.86 & x &  \\ 
  22 & ferguson\_eyetrackingr &  28 &  28 & 19.71 & 18.02 & 21.86 & 5.54 & 2.33 & x &  \\ 
  23 & potter\_remix &  23 &  44 & 22.59 & 18.00 & 29.00 & 6.07 & 2.58 & x & x \\ 
  24 & newman\_genderdistractor &  19 &  19 & 30.12 & 29.23 & 30.84 & 13.11 & 4.94 &  &  \\ 
   \hline
\end{tabular}
\caption{Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.\label{tab:datasets}} 
\end{table*}

# Results {.unnumbered}

## Speed and accuracy of word recognition increase {.unnumbered}

```{r acc_mods}
d_trial$age_s <- scale(d_trial$age)[,1]
d_trial$log_age_s <- scale(d_trial$log_age)[,1]

acc_mods_lmer <- list(lwa_lin = lmer(long_window_accuracy ~ age_s  +
                                     + (age_s | dataset_name) + (1 | subject_id), data = d_trial),
                      lwa_log = lmer(long_window_accuracy ~ log_age_s +
                                     + (log_age_s | dataset_name) + (1 | subject_id), data = d_trial),
                      swa_lin = lmer(short_window_accuracy ~ age_s  +
                                     + (age_s | dataset_name) + (1 | subject_id), data = d_trial),
                      swa_log = lmer(short_window_accuracy ~ log_age_s +
                                     + (log_age_s | dataset_name) + (1 | subject_id), data = d_trial))

acc_mods_lmer_summary <- map_df(acc_mods_lmer, ~broom.mixed::glance(.x)) |>
  mutate(model = names(acc_mods_lmer), 
         r2 = map_dbl(acc_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
```

```{r rt_mods}
rt_mods_lmer <- list(log_lin = lmer(log_rt ~ age_s + 
                                      (age_s | dataset_name) + (1 | subject_id), data = d_trial),
                     log_log = lmer(log_rt ~ log_age_s + 
                                      (log_age_s | dataset_name) + (1 | subject_id), data = d_trial),
                     lin_lin = lmer(rt ~ age_s + 
                                      (age_s | dataset_name) + (1 | subject_id), data = d_trial),
                     lin_log = lmer(rt ~ log_age_s +  
                                      (log_age_s | dataset_name) + (1 | subject_id), data = d_trial))

rt_mods_lmer_summary <- map_df(rt_mods_lmer, ~broom.mixed::glance(.x)) |>
  mutate(model = names(rt_mods_lmer), 
         r2 = map_dbl(rt_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
```

We were first interested in developmental changes in speed and accuracy. We computed both reaction times and accuracies following standard practices in the literature [@fernald2008]. Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms).

Our first question was about the functional form of the relationships between age, speed, and accuracy. To investigate this question, we fit linear mixed effects models predicting accuracy and reaction time on each trial across the full dataset with random slopes of age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of reaction time. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text. All findings reported here hold for both window sizes, however.] and the best fitting model of speed predicted log reaction time as a function of log age as well (See Supplemental Tables 1 and 2 for model fit statistics).

Figure \ref{fig:devchange} shows these relationships. Log reaction time decreased significantly with age (`r papaja::apa_print(rt_mods_lmer[[2]])$full_result$log_age_s`) and accuracy increased significantly with age  (`r papaja::apa_print(acc_mods_lmer[[2]])$full_result$log_age_s`). In sum, we see continuing improvements in word recognition across the full age range in our dataset that appear roughly linear in the logarithm of age.  These logarithmic relationships follow theoretical expectations that both speed and accuracy should gradually asymptote to mature levels of performance as in skill learning more generally [@anderson1982,@snoddy1926]. With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.

```{r devchange, fig.cap="Participant-level accuracy and reaction time (log), plotted by age (log). The solid green line shows a linear fit and associated confidence interval. Individual dotted lines show linear fits for those datasets spanning six or more months of age. The dashed line for accuracy shows chance-level looking (.5)", fig.env="figure*", fig.height = 2.5}

datasets_with_age_variation <- d_trial |>
  group_by(dataset_name) |>
  summarise(age_range = max(age) - min(age)) |>
  filter(age_range >= AGEVAR_CUTOFF)

d_sub_agevar <- d_sub |>
  filter(dataset_name %in% datasets_with_age_variation$dataset_name)

a <- ggplot(d_sub,
       aes(x = age, y = long_window_accuracy)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm") +
  geom_smooth(data = d_sub_agevar,
              aes(group = dataset_name, col = dataset_name), se=FALSE,
              alpha = .1, method = "lm", size = .5, lty = 2) + 
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() +
  scale_color_discrete(guide = FALSE) + 
  ylab("Accuracy") + 
  xlab("Age (months)")

b <- ggplot(d_sub,
       aes(x = age, y = rt)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm") +
  geom_smooth(data = d_sub_agevar,
              aes(group = dataset_name, col = dataset_name), se=FALSE,
              alpha = .1, method = "lm", size = .5, lty = 2) + 
  scale_x_log10() +
  scale_y_log10() +
  scale_color_discrete(guide = FALSE) + 
  theme(legend.position = "bottom") + 
  ylab("Reaction Time (log)") + 
  xlab("Age (months)")

cowplot::plot_grid(a,b)
```

## Variability of word recognition decreases {.unnumbered}

```{r var_mods}
acc_var_mod <- lmer(long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) + 
                 (1 | subject_id), 
               data = d_sub)

rt_var_mod <- lmer(log_rt_var ~ log_age_s + (log_age_s | dataset_name) + 
                 (1 | subject_id), 
               data = d_sub)
```

One further hallmark of increasing skill is a decrease in task-relevant variability [@todorov2002optimal]. Both within and across datasets, variation in accuracy and reaction time decreased across datasets, again relatively continuously across the developmental range we examined (Figure \ref{fig:variance}). We fit mixed effects models to the standard deviation of both accuracy and reaction time for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant. For both speed and accuracy, variability decreased with age (RT: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`). Thus, as well as being faster and more accurate, older children appear to be more consistent in their online word recognition.

```{r variance, fig.cap="Participant-level variability in accuracy and reaction time (log RT), plotted by age (log). Plotting conventions are as in Figure 1.", fig.env="figure*", fig.height = 2.5}
c <- ggplot(d_sub,
       aes(x = age, y = long_window_acc_var)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm") +
  geom_smooth(data = d_sub_agevar,
              aes(group = dataset_name, col = dataset_name), se=FALSE,
              alpha = .1, method = "lm", size = .5, lty = 2) + 
  scale_x_log10() +
  scale_color_discrete(guide = FALSE) + 
  ylab("SD of Accuracy)") + 
  xlab("Age (months)")

d <- ggplot(d_sub,
       aes(x = age, y = log_rt_var)) + 
  geom_point(alpha = .05) + 
  geom_smooth(method = "lm") +
  geom_smooth(data = d_sub_agevar,
              aes(group = dataset_name, col = dataset_name), se=FALSE,
              alpha = .1, method = "lm", size = .5, lty = 2) + 
  scale_x_log10() +
  scale_color_discrete(guide = FALSE) + 
  theme(legend.position = "bottom") + 
  ylab("SD of log RT)") + 
  xlab("Age (months)")

cowplot::plot_grid(c,d)
```

## Speed and accuracy covary with vocabulary {.unnumbered}

We were next interested in whether the various different aspects of word recognition -- including accuracy, reaction time, and the variability of each of these -- were related to other aspects of early language ability. Of the studies in our database, `r sum(str_length(dataset_characteristics$CDIs)>0)` gathered parent reports about children's early vocabulary using the MacArthur-Bates Communicative Development Inventory (CDI), a popular survey instrument that provides a reliable and valid holistic picture of children's early language [@marchman2023;@frank2021]. Different forms of the CDI can be used to measure either receptive and expressive vocabulary (for children up to 18 months) or expressive vocabulary only (for children 16 -- 30 months).

```{r efa}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age, log_age)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                       ~ age_scale(.x, age)))

parallel_ana <- capture.output(fa.parallel(select(d_sub_mat, -dataset_name, -age, -log_age), fa = "fa", 
            use = "pairwise.complete.obs", plot=FALSE))
```

```{r cfa}
fa3_model <- "vocab =~ prod + comp
              accuracy =~ long_window_accuracy + long_window_acc_var
              speed =~ rt + rt_var"

fit3 <- cfa(fa3_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

fit_stats <- suppressMessages(summary(fit3, fit.measures=TRUE, standardize=TRUE))
```

We fit a series of factor analytic models to explore the dimensionality of the data. Initial exploratory factor analysis using paralell analysis suggested that three factors explained substantial variance in the data. Due to missingness of data, we used confirmatory factory analysis with full information maximum likelihood to find the best set of loadings. The best fitting model was a three-factor model with factors for speed (reaction time and reaction time variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI). Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit_stats$fit["rmsea.pvalue"], 3)`).

```{r age_regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age. Squares show  Solid lines show regression coefficients and factor loadings, and dotted lines show covariances (reflecting age-residualized correlations). Abbreviations: acc = accuracy, prod = production vocabulary, comp = comprehension vocabulary.", fig.height = 3}

d_sub_mat_s_renamed <- d_sub_mat_s |>
  rename(acc = long_window_accuracy, 
         acc_sd = long_window_acc_var, 
         rt_sd = rt_var)

fa3_age_model <- "
# measurement model
vocab =~ prod + comp
accuracy =~ acc + acc_sd
speed =~ rt + rt_sd

# regressions
vocab ~ log_age
accuracy ~ log_age
speed ~ log_age
"

fit3_age <- cfa(fa3_age_model, d_sub_mat_s_renamed, std.lv=TRUE, missing='fiml')

fit3_age_summary <- suppressMessages(summary(fit3_age, fit.measures=TRUE, standardize=TRUE))
layout_age = matrix(nrow=5, ncol = 6,
                data = c(NA,NA,"log_age",NA, NA, NA,
                         NA, NA, NA, NA, NA, NA, 
                         "speed",NA,"accuracy",NA,"vocab",NA,
                         NA, NA, NA, NA, NA, NA, 
                         "rt_sd","rt", "acc",
                         "acc_sd", "prod","comp"), 
                byrow = TRUE)

graph_sem(model = fit3_age, text_size = 2, layout = t(layout_age))
```

Figure \ref{fig:age_regression} shows a regression model fit to this confirmatory factor analysis, with log age predicting each latent variable. This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age). All three latent factors were significantly related to one another, with speed and accuracy showing strong negative covariance ($\beta=$ `r fit3_age_summary$pe[21,"est"]`, SE = `r fit3_age_summary$pe[21,"se"]`, $p < .0001$) and weaker but significant covariation between speed and vocabulary ($\beta=$ `r fit3_age_summary$pe[20,"est"]`, SE = `r fit3_age_summary$pe[20,"se"]`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r fit3_age_summary$pe[19,"est"]`, SE = `r fit3_age_summary$pe[19,"se"]`, $p < .0001$). This model supports the idea that individual variation in speed and accuracy of word recognition is concurrently related to vocabulary beyond effects of age.

## Speed of processing relates to vocabulary growth {.unnumbered}

```{r reliability}
MONTH_CUTOFF <- 2

longitudinal <- d_sub |>
  group_by(dataset_name, subject_id) |>
  count() |>
  filter(n > 1)

d_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)))

d_reliability <- d_long |>
  filter(delta_t <= MONTH_CUTOFF, 
         admin_num <= 2) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "admin_num",
              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
                              "long_window_acc_var", "prod", "comp"))

dataset_reliabilities <- d_reliability |>
  group_by(dataset_name) |>
  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
            acc = cor(long_window_accuracy_1, 
                          long_window_accuracy_2, use = "pairwise.complete.obs")) |>
  filter(dataset_name != "fernald_totlot")

# note this is not correct for CDIs because some CDIs can get grouped with two admins, inflating correlations
global_reliabilities <- d_reliability |>
  ungroup() |>
  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
            acc = cor(long_window_accuracy_1, 
                          long_window_accuracy_2, use = "pairwise.complete.obs"))
```

While we fit the latent variable models above to the full cross-sectional dataset, we are most interested in within-child developmental changes, as measured within individual studies. We first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than two months apart. Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$), suggesting that, across studies, LWL measures from individual sessions provide relatively noisy measurements of individual children even if individual studies are able to achieve higher reliability by averaging across measurement occasions [e.g., @marchman2008].

```{r prep_long_data}
d_sub_long <- d_sub |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
                         "long_window_acc_var", "prod", "comp")),
                ~ age_scale(.x,age))) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0, diff(age)))

d_sub_long_t <- d_sub_long |> 
  mutate(t = ifelse(delta_t > MONTH_CUTOFF, "t2", "t1")) |>
  group_by(subject_id, dataset_name, t) |>
  summarise(across(all_of(c("rt", "rt_var", "long_window_accuracy",
                            "long_window_acc_var", "prod", "comp")),
                   ~ mean(.x, na.rm=TRUE))) |> 
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
                         "long_window_acc_var", "prod", "comp")),
                ~ ifelse(is.nan(.x), NA, .x)))

d_sub_wide <- d_sub_long_t |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "t",
              values_from = c("rt", "rt_var", "long_window_accuracy", 
                              "long_window_acc_var", "prod", "comp"))
```

```{r}
fa3_model_long <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + s1*comp_t1
accuracy_t1 =~ 1*long_window_accuracy_t1 + s2*long_window_acc_var_t1
speed_t1 =~ 1*rt_t1 + s3*rt_var_t1

vocab_t2 =~ 1*prod_t2 + s1*comp_t2
accuracy_t2 =~ 1*long_window_accuracy_t2 + s2*long_window_acc_var_t2
speed_t2 =~ 1*rt_t2 + s3*rt_var_t2

# longitudinal relationships
vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2

# means for the latents
vocab_t1 ~ 1
vocab_t2 ~ 1
accuracy_t1 ~ 1
accuracy_t2 ~ 1
speed_t1 ~ 1
speed_t2 ~ 1

# residual variance for the latents
vocab_t1 ~~ NA*vocab_t1
vocab_t2 ~~ NA*vocab_t2
accuracy_t1 ~~ NA*accuracy_t1
accuracy_t2 ~~ NA*accuracy_t2
speed_t1 ~~ NA*speed_t1
speed_t2 ~~ NA*speed_t2
"

fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')

fit3_long_summary <- suppressMessages(summary(fit3_long, fit.measures=TRUE, standardize=TRUE))
```

Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure longitudinal relationships between LWL and vocabulary via a longitudinal latent factor model. We fit the same three-factor model, but this time averaging longitudinal administrations after splitting them into two categories: those datapoints within a two-month time period (which we designate as $t_1$) and those datapoints between three and six months from the initial measurement ($t_2$). To ensure measurement invariance, we coupled the factor loadings across the two time-points. Next, to investigate coupling between the different measures, we fit regressions linking the two time points.

The fitted longitudinal model is shown in \ref{fig:longitudinal}. Overall fit statistics were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit3_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit3_long_summary$fit["rmsea.pvalue"], 3)`). Each latent factor was significantly coupled across time, again indicating moderate longitudinal stability of these abilities.

```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Coupled change model."}
layout_long = matrix(nrow=7, ncol = 7,
                     data = c("rt_var_t1","rt_t1", "long_window_accuracy_t1", "long_window_acc_var_t1", "prod_t1","comp_t1", NA,
                               NA, NA, NA, NA, NA, NA,  NA,
                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
                              NA, NA, NA, NA, NA, NA,  NA,
                               NA,"speed_t2",NA,"accuracy_t2",NA,NA,"vocab_t2",
                               NA, NA, NA, NA, NA, NA,  NA,
                               NA,"rt_var_t2","rt_t2", "long_window_accuracy_t2", "long_window_acc_var_t2", "prod_t2","comp_t2"), 
                     byrow = TRUE)
graph_sem(model = fit3_long, text_size = 3, layout = t(layout_long))
```

```{r}
fa3_model_long_control <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + s1*comp_t1
accuracy_t1 =~ 1*long_window_accuracy_t1 + s2*long_window_acc_var_t1
speed_t1 =~ 1*rt_t1 + s3*rt_var_t1

vocab_t2 =~ 1*prod_t2 + s1*comp_t2
accuracy_t2 =~ 1*long_window_accuracy_t2 + s2*long_window_acc_var_t2
speed_t2 =~ 1*rt_t2 + s3*rt_var_t2

# longitudinal relationships
vocab_t2 ~ vocab_t1 + 0*speed_t1 + accuracy_t1
speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2

# means for the latents
vocab_t1 ~ 1
vocab_t2 ~ 1
accuracy_t1 ~ 1
accuracy_t2 ~ 1
speed_t1 ~ 1
speed_t2 ~ 1

# residual variance for the latents
vocab_t1 ~~ NA*vocab_t1
vocab_t2 ~~ NA*vocab_t2
accuracy_t1 ~~ NA*accuracy_t1
accuracy_t2 ~~ NA*accuracy_t2
speed_t1 ~~ NA*speed_t1
speed_t2 ~~ NA*speed_t2
"

fit3_long_control <- sem(fa3_model_long_control, d_sub_wide, std.lv=TRUE, missing='fiml')
speed_vocab_anova <- anova(fit3_long, fit3_long_control)
speed_vocab_anova$p = speed_vocab_anova$`Pr(>Chisq)`
speed_vocab_anova$c2d = speed_vocab_anova$`Chisq diff`
```

Critically, LWL processing speed at time 1 predicted vocabulary at time 2 ($\beta=$ `r fit3_long_summary$pe[14,"est"]`, SE = `r fit3_long_summary$pe[21,"se"]`, $p =$ `r fit3_long_summary$pe[14,"pvalue"]`). This model significantly improved fit compared with a model that set this coupling parameter to zero but was otherwise unchanged ($\chi^2 =$ `r round(speed_vocab_anova$c2d[2], 2)`, $p = $ `r round(speed_vocab_anova$p[2], 4)`). These findings are consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2012,@Weisleder2013].

# Discussion {.unnumbered}

Spoken word recognition has been argued to be a fundamental part of language development, but there is still a lack of clarity about how this ability changes across early childhood and how it relates to other aspects of language development. Here we used a large open dataset of experimental measurements to address these questions. Across all data, the development of both speed and accuracy of word recognition was approximately linear with the logarithm of age, that is, gradually converging on mature levels of processing efficiency. Further, older children showed less variability in accuracy and reaction time across trials, indicating more robust execution of this skilled behavior. Word recognition is not a unifactorial ability, however. Our best fitting model showed two distinct components: speed and accuracy. Notably, the reaction time variables grouped together, while accuracy and parent reports of vocabulary loaded on the same latent construct. These two latent constructs -- speed and accuracy -- both showed approximately linear change in the log of age both within and across children and datasets.

These data suggest graded, continual changes in word recognition starting in early childhood and continuing at least through elementary school. Notably, the experiments included in our datasets used simple words that children of this age almost surely "know" (e.g., "ball," which is learned very early, or occasionally "scissors" or "zebra," which are learned slightly later). Even for these "easy" words, careful measurement reveals increasing efficiency.

<!-- % address causal indeterminacy issue  -->
<!-- % https://www.degruyter.com/document/doi/10.1515/jci-2022-0074/html -->

<!-- DISCUSS G EXPLANATION - fernald fundamentally speedier processor -->

<!-- INPUT VS MATURATION? PROBABLY BOTH -->

Consistent with this idea,  trial-to-trial variability in children's cognitive task performance decreases with age and has some predictive relationships with skill in other tasks, including language [@judd2021working;@galeano2018variability;@kautto2023introducing].

What explains variability in language performance? Integrating information from parent-report measures of vocabulary development with infant behavioral data, we show that infants' word recognition.

<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
\showmatmethods
\showacknow
\pnasbreak

\newpage

<!-- # Supplemental Figures {.unnumbered} -->

<!-- ```{r} -->
<!-- kbl(acc_mods_lmer_summary, digits = 3, longtable=FALSE) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- kbl(rt_mods_lmer_summary, digits = 3) -->
<!-- ``` -->

```{r reliability_table, eval=FALSE}
xtable(reliabilities)
```

<!-- % latex table generated in R 4.3.1 by xtable 1.8-4 package -->
<!-- % Fri Nov 15 12:22:42 2024 -->
<!-- \begin{table}[ht] -->
<!-- \centering -->
<!-- \begin{tabular}{rlrrrrrr} -->
<!--   \hline -->
<!--  & dataset name & rt & rt\_var & acc & acc\_var & prod & comp \\  -->
<!--   \hline -->
<!-- 1 & adams\_marchman\_2018 & 0.37 & 0.25 & 0.56 & 0.60 &  & 0.98 \\  -->
<!--   2 & fernald\_marchman\_2012 & 0.45 & 0.18 & 0.54 & 0.33 & 0.99 & 0.99 \\  -->
<!--   3 & fmw\_2013 & 0.40 & -0.07 & 0.25 & 0.47 & 0.96 & 0.98 \\  -->
<!--   4 & potter\_remix & 0.17 & 0.64 & 0.32 & 0.32 & 1.00 &  \\  -->
<!--   5 & weaver\_zettersten\_2024 & 0.30 & 0.18 & 0.41 & 0.45 &  &  \\  -->
<!--    \hline -->
<!-- \end{tabular} -->
<!-- \caption{\label{reliabilities} Overall test-retest reliabilities (Pearson correlation) for each of the measures for longitudinal administrations within two months of one another.   -->
<!-- \end{table} -->

### 2024-11-15 helper/common.R

@@ -19,11 +19,17 @@ knitr::opts_chunk$set(cache.extra = knitr::rand_seed, cache = TRUE,
 options(dplyr.summarise.inform = FALSE)
 
 
-.font <- "Source Sans Pro"
-theme_set(theme_bw(base_size = 14, base_family = .font))
+# .font <- "Source Sans Pro"
+# theme_set(theme_bw(base_size = 14, base_family = .font))
+theme_set(theme_bw(base_size = 10))
 theme_update(panel.grid = element_blank(),
              strip.background = element_blank(),
              legend.key = element_blank(),
              panel.border = element_blank(),
              axis.line = element_line(),
              strip.text = element_text(face = "bold"))
+
+# helper function for scaling to the variance of a particular age group
+age_scale <- function (x, age, min = 16, max = 20) {
+  (x - mean(x, na.rm=TRUE)) / sd(x[age >= min & age <= max], na.rm=TRUE)
+}

### 2024-11-15 paper/paper.Rmd
paper

@@
 Combining 24 datasets with data from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development, both within and across studies.
-Factor analysis revealed a cross-sectional coupling of word recognition speed and accuracy to parent-reported vocabulary and a longitudinal relationship such that children with faster language processing showed larger vocabulary growth.
-Together, these findings are consistent with the view that language processing is –– among other things –– a skill, and language learning is a process of skill learning.
+Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy to parent-reported vocabulary as well as a longitudinal relationship such that children with faster language processing tend to show larger vocabulary growth.
+Together, these findings support the view that language processing is –– among other things –– a skill, and that this skill plays a role in early language learning.
 
@@
 The efficiency with which children recognize words provides a measurement of their real-time language processing, which has been argued to be a key part of the learning process.
-Here we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age.
-This trajectory is consistent with the idea that understanding language is a skill that improves gradually with practice.
+Here we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
+Understanding language is a skill that improves gradually with practice.
 
@@
 \dropcap{C}hildren acquiring a language are learning a body of knowledge -- a set of words and the ways they are combined -- but they are also learning to deploy this knowledge in the myriad complex, noisy, and fast-moving environments in which language is used.
-As children enter their second year, language explodes onto the scene; children's vocabulary and grammatical abilities grow rapidly in tandem [@bates1994developmental;@frank2021].
-This growth in knowledge is also accompanied by changes in their language processing: they become quicker and more accurate in recognizing words and matching them with their referents [@fernald1998;@peter2019;@bergelson2020comprehension].
+As children enter their second year, language explodes onto the scene; both vocabulary and grammatical abilities grow rapidly and in tandem [@bates1994developmental;@frank2021].
+This growth in knowledge is also accompanied by changes in  language processing: children become quicker and more accurate in recognizing words and matching them with their referents [@fernald1998;@peter2019;@bergelson2020comprehension].
 
@@
 
-The version of this method that is used with children goes by many names, including the "intermodal preferential looking paradigm" and the "looking while listening" paradigm (or LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
-In these experiments, children are typically shown a series of trials in which two images are displayed side by side and they are asked to find one of them.
+\begin{table*}[h!]
+\centering
+\begin{tabular}{rlrrrrrrrll}
+  \hline
+ & dataset\_name & N subjects & N admins & Mean Age & Min Age & Max Age & Avg Trials & Avg RT Trials & CDIs & longitudinal \\ 
+  \hline
+1 & reflook\_v4 & 310 & 310 & 36.82 & 12.24 & 60.00 & 6.21 & 2.87 &  &  \\ 
+  2 & yurovsky\_2017 & 282 & 282 & 25.64 & 12.59 & 58.65 & 5.95 & 2.79 &  &  \\ 
+  3 & weaver\_zettersten\_2024 & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 & x & x \\ 
+  4 & fernald\_marchman\_2012 & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.78 & 7.17 & x & x \\ 
+  5 & sander-montant\_2022 & 122 & 122 & 21.94 & 12.02 & 31.11 & 10.07 & 4.20 & x &  \\ 
+  6 & frank\_tablet\_2016 & 104 & 104 & 33.89 & 12.13 & 59.84 & 5.87 & 2.74 &  &  \\ 
+  7 & baumgartner\_2014 & 100 & 100 & 12.01 & 12.00 & 13.00 & 4.00 & 2.31 & x &  \\ 
+  8 & fmw\_2013 &  80 & 179 & 20.04 & 17.00 & 26.00 & 21.26 & 8.44 & x & x \\ 
+  9 & adams\_marchman\_2018 &  69 & 711 & 23.57 & 12.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
+  10 & potter\_canine &  67 &  67 & 23.76 & 21.00 & 27.00 & 10.38 & 3.80 & x &  \\ 
+  11 & fernald\_totlot &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.22 & 6.05 & x & x \\ 
+  12 & pomper\_prime &  63 &  63 & 39.73 & 37.90 & 42.00 & 12.98 & 5.34 &  &  \\ 
+  13 & pomper\_saffran\_2016 &  60 &  60 & 44.27 & 41.00 & 47.00 & 7.55 & 3.30 &  &  \\ 
+  14 & swingley\_aslin\_2002 &  50 &  50 & 15.09 & 14.13 & 16.00 & 11.74 & 3.79 & x &  \\ 
+  15 & pomper\_salientme &  44 &  44 & 40.11 & 38.00 & 43.00 & 5.30 & 2.34 &  &  \\ 
+  16 & perry\_cowpig &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 &  &  \\ 
+  17 & ronfard\_2021 &  40 &  40 & 19.95 & 18.00 & 24.00 & 18.54 & 7.62 & x &  \\ 
+  18 & bacon\_gendercues &  38 &  38 & 22.87 & 22.00 & 24.00 & 18.16 & 8.00 & x &  \\ 
+  19 & garrison\_bergelson\_2020 &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.94 & 9.59 & x &  \\ 
+  20 & pomper\_yumme &  32 &  32 & 26.38 & 25.00 & 28.00 & 7.62 & 2.31 &  &  \\ 
+  21 & mahr\_coartic &  29 &  29 & 20.83 & 18.10 & 23.80 & 24.38 & 8.86 & x &  \\ 
+  22 & ferguson\_eyetrackingr &  28 &  28 & 19.71 & 18.02 & 21.86 & 5.54 & 2.33 & x &  \\ 
+  23 & potter\_remix &  23 &  44 & 22.59 & 18.00 & 29.00 & 6.07 & 2.58 & x & x \\ 
+  24 & newman\_genderdistractor &  19 &  19 & 30.12 & 29.23 & 30.84 & 13.11 & 4.94 &  &  \\ 
+   \hline
+\end{tabular}
+\caption{Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.\label{tab:datasets}} 
+\end{table*}
+
+The version of this method that is used with children goes by many names, including the "intermodal preferential looking paradigm" and the "looking while listening" paradigm (LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
+In LWL experiments, children are typically shown a series of trials in which two images are displayed side by side and they are asked to find one of them.
 For example, a ball and a book might be shown, and the child might hear "Look at the ball!
-Can you find it?" Accuracy is then computed as the proportion of time they fixate the correct image within a fixed window after the onset of the noun ("ball" in this case).
-Reaction time is typically computed only on trials in which the child is fixating the distractor image (the book); in these cases, the average time it takes to shift fixation to the target is used as an index of processing speed.
+Can you find it?".
+Accuracy is then computed as the proportion of time they fixate the correct image within a fixed window after the onset of the noun ("ball" in this case).
+Reaction time is computed only on trials in which the child is fixating the distractor image (the book); in these cases, the average time it takes to shift fixation to the target is used as an index of processing speed.
 The seminal work using this method showed that children's speed and accuracy both increase rapidly across the second year [@fernald1998;@reznick1990visual].
@@
 These samples provide neither the range of ages nor the number of participants to estimate the broader dynamics of how word recognition changes developmentally and how it connects with broader changes across the lifespan [@mcmurray2022slow;@colby2023efficiency].
-Further, previous findings do not shed light on whether processing efficiency indexes a separate construct and might vary independently from vocabulary size, or whether it is tightly coupled to vocabulary such that children with larger vocabularies are also better language processors.
-
-% suggest that speed of processing is related to concurrent and later language ability [@peter2019;@marchman2008].
-This work is consistent with several explanations, however.
-First, it is possible that processing speed indexes a different construct than accuracy.
-On this kind of account, speedier processors may be differentiated from those children with a larger vocabulary; identifying these children thus helps to increase predictive accuracy.
-An alternative account, however, is that processing speed and accuracy reflect two aspects of how a single construct is translated into behavior (in this case, eye movements).
-On this second account, the increase in predictive accuracy by using reaction times is accounted for because each has independent measurement error -- the inclusion of both into a model thus increases precision.
-Thus, the factor structure of early language processing -- and in particular, whether there is evidence for orthogonal variation in early language related to processing speed -- is an open question.
+One influential recent hypothesis refers to language learning as a process of learning the skill of real-time processing  [@christiansen2016;@chater2018].
+On this hypothesis, we should expect to see many of the signatures of expertise and skill learning in language processing: speed of processing should decrease asymptotically with age [@kail1991;@anderson1982] and trial-to-trial variability should decrease [@todorov2002optimal].
+Further, previous findings have provided limited and sometimes conflicting evidence on the relation between word recognition and language learning [@Weisleder2013;@peter2019;@newbury2016interrelationships].
+To the extent that there are consistent relations between vocabulary and word recognition, these should be visible in a larger dataset.
 
@@
 ```
-
-\begin{table*}[t!]
-\centering
-\begin{tabular}{rlrrrrrrrll}
-  \hline
- & dataset\_name & N subjects & N admins & Mean Age & Min Age & Max Age & Avg Trials & Avg RT Trials & CDIs & longitudinal \\ 
-  \hline
-1 & reflook\_v4 & 310 & 310 & 36.82 & 12.24 & 60.00 & 6.21 & 2.87 &  &  \\ 
-  2 & yurovsky\_2017 & 282 & 282 & 25.64 & 12.59 & 58.65 & 5.95 & 2.79 &  &  \\ 
-  3 & weaver\_zettersten\_2024 & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 & x & x \\ 
-  4 & fernald\_marchman\_2012 & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.78 & 7.17 & x & x \\ 
-  5 & sander-montant\_2022 & 122 & 122 & 21.94 & 12.02 & 31.11 & 10.07 & 4.20 & x &  \\ 
-  6 & frank\_tablet\_2016 & 104 & 104 & 33.89 & 12.13 & 59.84 & 5.87 & 2.74 &  &  \\ 
-  7 & baumgartner\_2014 & 100 & 100 & 12.01 & 12.00 & 13.00 & 4.00 & 2.31 & x &  \\ 
-  8 & fmw\_2013 &  80 & 179 & 20.04 & 17.00 & 26.00 & 21.26 & 8.44 & x & x \\ 
-  9 & adams\_marchman\_2018 &  69 & 711 & 23.57 & 12.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
-  10 & potter\_canine &  67 &  67 & 23.76 & 21.00 & 27.00 & 10.38 & 3.80 & x &  \\ 
-  11 & fernald\_totlot &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.22 & 6.05 & x & x \\ 
-  12 & pomper\_prime &  63 &  63 & 39.73 & 37.90 & 42.00 & 12.98 & 5.34 &  &  \\ 
-  13 & pomper\_saffran\_2016 &  60 &  60 & 44.27 & 41.00 & 47.00 & 7.55 & 3.30 &  &  \\ 
-  14 & swingley\_aslin\_2002 &  50 &  50 & 15.09 & 14.13 & 16.00 & 11.74 & 3.79 & x &  \\ 
-  15 & pomper\_salientme &  44 &  44 & 40.11 & 38.00 & 43.00 & 5.30 & 2.34 &  &  \\ 
-  16 & perry\_cowpig &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 &  &  \\ 
-  17 & ronfard\_2021 &  40 &  40 & 19.95 & 18.00 & 24.00 & 18.54 & 7.62 & x &  \\ 
-  18 & bacon\_gendercues &  38 &  38 & 22.87 & 22.00 & 24.00 & 18.16 & 8.00 & x &  \\ 
-  19 & garrison\_bergelson\_2020 &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.94 & 9.59 & x &  \\ 
-  20 & pomper\_yumme &  32 &  32 & 26.38 & 25.00 & 28.00 & 7.62 & 2.31 &  &  \\ 
-  21 & mahr\_coartic &  29 &  29 & 20.83 & 18.10 & 23.80 & 24.38 & 8.86 & x &  \\ 
-  22 & ferguson\_eyetrackingr &  28 &  28 & 19.71 & 18.02 & 21.86 & 5.54 & 2.33 & x &  \\ 
-  23 & potter\_remix &  23 &  44 & 22.59 & 18.00 & 29.00 & 6.07 & 2.58 & x & x \\ 
-  24 & newman\_genderdistractor &  19 &  19 & 30.12 & 29.23 & 30.84 & 13.11 & 4.94 &  &  \\ 
-   \hline
-\end{tabular}
-\caption{Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.\label{tab:datasets}} 
-\end{table*}
 
@@
   ungroup() |>
-  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age, log_age)
+  select(dataset_name, log_rt, log_rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age, log_age)
 
@@
   ungroup() |>
-  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
+  mutate(across(all_of(c("log_rt", "log_rt_var", "long_window_accuracy", 
                          "long_window_acc_var", "prod", "comp")), 
@@
               accuracy =~ long_window_accuracy + long_window_acc_var
-              speed =~ rt + rt_var"
+              speed =~ log_rt + log_rt_var"
 
@@
          acc_sd = long_window_acc_var, 
-         rt_sd = rt_var)
+         log_rt_sd = log_rt_var)
 
@@
 accuracy =~ acc + acc_sd
-speed =~ rt + rt_sd
+speed =~ log_rt + log_rt_sd
 
@@
                          NA, NA, NA, NA, NA, NA, 
-                         "rt_sd","rt", "acc",
+                         "log_rt","log_rt_sd", "acc",
                          "acc_sd", "prod","comp"), 
@@
 This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age).
-All three latent factors were significantly related to one another, with speed and accuracy showing strong negative covariance ($\beta=$ `r fit3_age_summary$pe[21,"est"]`, SE = `r fit3_age_summary$pe[21,"se"]`, $p < .0001$) and weaker but significant covariation between speed and vocabulary ($\beta=$ `r fit3_age_summary$pe[20,"est"]`, SE = `r fit3_age_summary$pe[20,"se"]`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r fit3_age_summary$pe[19,"est"]`, SE = `r fit3_age_summary$pe[19,"se"]`, $p < .0001$).
+All three latent factors were significantly related to one another, with speed and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between speed and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
 This model supports the idea that individual variation in speed and accuracy of word recognition is concurrently related to vocabulary beyond effects of age.
@@
 d_sub_long <- d_sub |>
-  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
+  mutate(across(all_of(c("log_rt", "log_rt_var", "long_window_accuracy",
                          "long_window_acc_var", "prod", "comp")),
@@
   mutate(t = ifelse(delta_t > MONTH_CUTOFF, "t2", "t1")) |>
+  rename(acc = long_window_accuracy,
+         acc_sd = long_window_acc_var,
+         log_rt_sd = log_rt_var) |>
   group_by(subject_id, dataset_name, t) |>
-  summarise(across(all_of(c("rt", "rt_var", "long_window_accuracy",
-                            "long_window_acc_var", "prod", "comp")),
+  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                            "acc_sd", "prod", "comp")),
                    ~ mean(.x, na.rm=TRUE))) |> 
-  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
-                         "long_window_acc_var", "prod", "comp")),
+  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                         "acc_sd", "prod", "comp")),
                 ~ ifelse(is.nan(.x), NA, .x)))
@@
               names_from = "t",
-              values_from = c("rt", "rt_var", "long_window_accuracy", 
-                              "long_window_acc_var", "prod", "comp"))
+              values_from = c("log_rt", "log_rt_sd", "acc", 
+                              "acc_sd", "prod", "comp"))
 ```
@@
 vocab_t1 =~ 1*prod_t1 + s1*comp_t1
-accuracy_t1 =~ 1*long_window_accuracy_t1 + s2*long_window_acc_var_t1
-speed_t1 =~ 1*rt_t1 + s3*rt_var_t1
+accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
+speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1
 
 vocab_t2 =~ 1*prod_t2 + s1*comp_t2
-accuracy_t2 =~ 1*long_window_accuracy_t2 + s2*long_window_acc_var_t2
-speed_t2 =~ 1*rt_t2 + s3*rt_var_t2
+accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
+speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2
 
@@
 layout_long = matrix(nrow=7, ncol = 7,
-                     data = c("rt_var_t1","rt_t1", "long_window_accuracy_t1", "long_window_acc_var_t1", "prod_t1","comp_t1", NA,
+                     data = c("log_rt_sd_t1","log_rt_t1", "acc_t1", "acc_sd_t1", "prod_t1","comp_t1", NA,
                                NA, NA, NA, NA, NA, NA,  NA,
@@
                                NA, NA, NA, NA, NA, NA,  NA,
-                               NA,"rt_var_t2","rt_t2", "long_window_accuracy_t2", "long_window_acc_var_t2", "prod_t2","comp_t2"), 
+                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", "prod_t2","comp_t2"), 
                      byrow = TRUE)
@@
 
-Critically, LWL processing speed at time 1 predicted vocabulary at time 2 ($\beta=$ `r fit3_long_summary$pe[14,"est"]`, SE = `r fit3_long_summary$pe[21,"se"]`, $p =$ `r fit3_long_summary$pe[14,"pvalue"]`).
-This model significantly improved fit compared with a model that set this coupling parameter to zero but was otherwise unchanged ($\chi^2 =$ `r round(speed_vocab_anova$c2d[2], 2)`, $p = $ `r round(speed_vocab_anova$p[2], 4)`).
+Critically, LWL processing speed at time 1 predicted vocabulary at time 2 ($\beta=$ `r round(fit3_long_summary$pe[14,"est"],3)`, SE = `r round(fit3_long_summary$pe[21,"se"],3)`, $p =$ `r round(fit3_long_summary$pe[14,"pvalue"],3)`).
+This model significantly improved fit compared with a model that set this coupling parameter to zero but was otherwise unchanged ($\chi^2 = `r round(speed_vocab_anova$c2d[2], 2)`$, $p =  `r round(speed_vocab_anova$p[2], 4)`$).
 These findings are consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2012,@Weisleder2013].

### 2024-11-16 paper/paper.Rmd
updated intro

@@
 ---
-title: Continuous developmental changes in word recognition across early childhood
+title: Continuous developmental changes in word recognition across early childhood support early language learning
 
@@
 abstract: |
-  Being a fluent language user involves not just knowing how to combine words into meaningful sentences, but also being able to understand language as it unfolds in time.
+  Being a fluent language user involves understanding language as it unfolds in time.
 How does this skill develop over the course of early childhood?
@@
 Very young children with incomplete knowledge may not be able to point to the correct referent of a word, but they may still have some representation of word meaning [@bergelson2012].
-Eye tracking has thus emerged as a key method that allows the measurement of language comprehension with high temporal resolution: both adults and children reliably fixate the referent of a spoken word soon after it is spoken [@tanenhaus1995integration;@altmann1999incremental;@fernald1998;@fernald2008].
+Eye tracking has thus emerged as a key method that allows the measurement of language comprehension with high temporal resolution: both adults and children reliably fixate the referent of a word soon after it is used [@tanenhaus1995integration;@altmann1999incremental;@fernald1998;@fernald2008;@macdonald2018].
 The relative timecourse of fixation then can provide an index of both the comprehender's ability and the properties of the stimulus.
@@
 
-The version of this method that is used with children goes by many names, including the "intermodal preferential looking paradigm" and the "looking while listening" paradigm (LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
+The version of this method that is used with children goes by many names, including the intermodal preferential looking paradigm and the "looking while listening" paradigm (LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
 In LWL experiments, children are typically shown a series of trials in which two images are displayed side by side and they are asked to find one of them.
@@
 Reaction time is computed only on trials in which the child is fixating the distractor image (the book); in these cases, the average time it takes to shift fixation to the target is used as an index of processing speed.
-The seminal work using this method showed that children's speed and accuracy both increase rapidly across the second year [@fernald1998;@reznick1990visual].
+Early work using this method showed that children's speed and accuracy both increase rapidly across the second year [@fernald1998;@reznick1990visual], and related methods have provided a window into how children process phonological [@mani2011], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] competition as well as how their lexical representations develop [@Swingley2002].
 
-<!-- % MCF: we go back and forth now between word recognition and language processing. is this oK? -->
-<!-- % MCF: I haven't adopted the fernald "efficiency" term, which groups speed and accuracy as one process... in part this seems to presuppose the answer to some bits of our factor analysis... -->
+Word recognition ability is hypothesized to play a key role in language learning.
+Each word that a child experiences is an opportunity to learn; measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking].
+These learning opportunities are likely helpful to different degrees, however [@kachergis2022toward;@mcmurray2007defusing].
+<!-- Language input is more helpful for learning if it focused on the child's own interest and grounded in the particulars of their current environment -- for example, in a game, routine, or object -- and if it uses vocabulary and syntax that they know [@tomasello2001perceiving;@roy2015predicting;@cartmill2013quality].
+-->
+Recognizing incoming words and linking them with their referents is a prerequisite for learning.
+Consider a child hearing the utterance "can you put the ball in the box?" The faster and more accurately the child can recognize that the ball is a referent, the better they can use this evidence to help infer the speaker's intended meaning, perhaps making inferences about the meaning of "put" or "box" [@frank2009using].
+Consistent with this idea, one important study found that children's word recognition speed mediated the relationship between home language input and vocabulary growth [@Weisleder2013].
 
-Language processing ability is hypothesized to play a key role in language learning.
-Each word that a child hears is an opportunity to learn; measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking].
-These opportunities are likely helpful to different degrees, however [@kachergis2022toward;@mcmurray2007defusing].
-Language is likely to be a better learning signal if it is focused on the child's own interest, grounded in the particulars of their current environment -- for example, in a game, routine, or object -- and uses vocabulary and syntax that they know [@tomasello2001perceiving;@roy2015predicting;@cartmill2013quality].
-But processing is also prerequisite for learning.
-Consider a child hearing the utterance "can you put the ball in the box?" The faster and more accurately the child can recognize that the ball is a referent, the better they can use this evidence to help infer the speaker's intended meaning, perhaps making inferences about the meaning of "put" or "box" [@frank2009using].
-Consistent with this idea, in one study, children's language processing skill mediated the relationship between home language input and vocabulary growth [@Weisleder2013].
-
-<!-- % This method has provided an important window into how children process phonological [@???}, syntactic [@trueswell1999kindergarten}, and semantic [@mani} competition; how their lexical representations develop [@Swingley2002}; etc.
--->
-
-Speed of word recognition has been used as an index of individual differences in early childhood [@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow].
+Speed of word recognition has been used as an index of individual differences in early childhood [@fernald2006;@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow].
 Word recognition speed at 18 months can predict children's later language ability over and above measures of vocabulary size [@marchman2008], though these predictions may be limited to particular ages or processing assessments [@peter2019].
@@
 Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
-These samples provide neither the range of ages nor the number of participants to estimate the broader dynamics of how word recognition changes developmentally and how it connects with broader changes across the lifespan [@mcmurray2022slow;@colby2023efficiency].
-One influential recent hypothesis refers to language learning as a process of learning the skill of real-time processing  [@christiansen2016;@chater2018].
-On this hypothesis, we should expect to see many of the signatures of expertise and skill learning in language processing: speed of processing should decrease asymptotically with age [@kail1991;@anderson1982] and trial-to-trial variability should decrease [@todorov2002optimal].
-Further, previous findings have provided limited and sometimes conflicting evidence on the relation between word recognition and language learning [@Weisleder2013;@peter2019;@newbury2016interrelationships].
+These samples provide neither the breadth of ages nor the number of participants to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of language [as has been done with school-aged children: @mcmurray2022slow;@colby2023efficiency].
+Here we investigate two hypotheses.
+First, one influential recent hypothesis characterizes language learning more broadly as a process of learning the skill of real-time processing  [@christiansen2016;@chater2018].
+On this hypothesis, we should expect to see the signatures of expertise and skill learning in word recognition.
+We expect that accuracy will change linearly with the logarithm of age, reflecting gradual asymptotic convergence to mature levels of accuracy.
+In addition, the logarithm of processing speed should also decrease with the logarithm of age, potentially reflecting the  "power law of practice" [@snoddy1926;@kail1991;@anderson1982; cf @heathcote2000].
+Further, trial-to-trial variability in both speed and accuracy should decrease with increasing expertise [@todorov2002optimal].
+
+Second, previous findings have provided limited and sometimes conflicting evidence on the concurrent and predictive relations between word recognition and language learning.
+Initial reports showed strong predictive relationships between both speed and accuracy and later vocabulary growth [@fernald2006].
+Two other studies extended this result to infants born preterm [@fernald2012] and to late talkers [@marchman2016].
+Subsequent studies have primarily focused on speed of processing and found more mixed results, with reaction time measures only inconsistently predictive of later vocabulary outcomes [@peter2019;@newbury2016interrelationships;@lany2017].
 To the extent that there are consistent relations between vocabulary and word recognition, these should be visible in a larger dataset.
+Further, by examining the relationship between speed, accuracy, and vocabulary, it should be possible to assess the extent to which processing speed specifically plays a role in vocabulary growth.
 
@@
 suppressPackageStartupMessages(library(xtable))
-suppressPackageStartupMessages(library(kableExtra))
+# suppressPackageStartupMessages(library(kableExtra))
+suppressPackageStartupMessages(library(papaja))
 
@@
 
-To address these questions, we made use of Peekbank, an open database of  looking while listening data from young children, stored in a harmonized format [@zettersten2023peekbank].
-Aiming to map the development of language processing across ages 1 -- 6 years, we analyzed data from simple word recognition trials in which children were shown two pictures of concrete objects and heard an English word.
-(We focused on English purely for practical reasons -- we had very limited data from other languages).
+To address these questions, we made use of a new release of data in Peekbank, an open database of  looking while listening data from young children, stored in a harmonized format [@zettersten2023peekbank].
+We retrieved data from children ages 1--6 years.
+Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a word; these trials often constituted control conditions for experiments.
+We focus here on English purely for practical reasons -- the Peekbank dataset contains limited data from other languages.
+
 These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions).
@@
 These logarithmic relationships follow theoretical expectations that both speed and accuracy should gradually asymptote to mature levels of performance as in skill learning more generally [@anderson1982,@snoddy1926].
-With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.
+<!-- With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.
+-->
 
@@
                      byrow = TRUE)
-graph_sem(model = fit3_long, text_size = 3, layout = t(layout_long))
+graph_sem(model = fit3_long, text_size = 2, layout = t(layout_long))
 ```
@@
 
-Spoken word recognition has been argued to be a fundamental part of language development, but there is still a lack of clarity about how this ability changes across early childhood and how it relates to other aspects of language development.
-Here we used a large open dataset of experimental measurements to address these questions.
-Across all data, the development of both speed and accuracy of word recognition was approximately linear with the logarithm of age, that is, gradually converging on mature levels of processing efficiency.
+We used a new database aggregating results across many individual developmental eye-tracking experiments to investigate the developmental course of word recognition and its relationship with early language learning.
+We found continuous developmental changes across all measures, consistent with a characterization of word recognition as a skill.
+Speed and accuracy both improved asymptoticaly, with evidence that recognition speed showed the log-log relationship associated with the power law of practice, that is, gradually converging on mature levels of processing efficiency.
 Further, older children showed less variability in accuracy and reaction time across trials, indicating more robust execution of this skilled behavior.
-Word recognition is not a unifactorial ability, however.
-Our best fitting model showed two distinct components: speed and accuracy.
-Notably, the reaction time variables grouped together, while accuracy and parent reports of vocabulary loaded on the same latent construct.
-These two latent constructs -- speed and accuracy -- both showed approximately linear change in the log of age both within and across children and datasets.
+
+FACTOR PARAGRAPH
 
@@
 \newpage
-
-<!-- # Supplemental Figures {.unnumbered} -->
-
-<!-- ```{r} -->
-<!-- kbl(acc_mods_lmer_summary, digits = 3, longtable=FALSE) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- kbl(rt_mods_lmer_summary, digits = 3) -->
-<!-- ``` -->
-
-```{r reliability_table, eval=FALSE}
-xtable(reliabilities)
-```
-
-<!-- % latex table generated in R 4.3.1 by xtable 1.8-4 package -->
-<!-- % Fri Nov 15 12:22:42 2024 -->
-<!-- \begin{table}[ht] -->
-<!-- \centering -->
-<!-- \begin{tabular}{rlrrrrrr} -->
-<!--   \hline -->
-<!--  & dataset name & rt & rt\_var & acc & acc\_var & prod & comp \\  -->
-<!--   \hline -->
-<!-- 1 & adams\_marchman\_2018 & 0.37 & 0.25 & 0.56 & 0.60 &  & 0.98 \\  -->
-<!--   2 & fernald\_marchman\_2012 & 0.45 & 0.18 & 0.54 & 0.33 & 0.99 & 0.99 \\  -->
-<!--   3 & fmw\_2013 & 0.40 & -0.07 & 0.25 & 0.47 & 0.96 & 0.98 \\  -->
-<!--   4 & potter\_remix & 0.17 & 0.64 & 0.32 & 0.32 & 1.00 &  \\  -->
-<!--   5 & weaver\_zettersten\_2024 & 0.30 & 0.18 & 0.41 & 0.45 &  &  \\  -->
-<!--    \hline -->
-<!-- \end{tabular} -->
-<!-- \caption{\label{reliabilities} Overall test-retest reliabilities (Pearson correlation) for each of the measures for longitudinal administrations within two months of one another.   -->
-<!-- \end{table} -->

### 2024-11-16 paper/supplemental_information.Rmd
updated intro

---
title: "Supplemental Information"
output: pdf_document
header-includes:
  - \usepackage{threeparttablex}
  - \usepackage{booktabs}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
```

```{r}
knitr::purl("paper.Rmd", documentation = 0)
source("paper.R")
```

# Supplemental Figure 1: Age histograms

```{r, fig.env="figure*"}
ggplot(d_sub, aes(x = age))+ 
  geom_histogram(binwidth = 1) + 
  facet_wrap(~dataset_name, scale = "free_y")
```

# Supplemental Table 1: Model comparison for accuracy models

```{r results="asis"}
papaja::apa_table(acc_mods_lmer_summary)
```

# Supplemental Table 2: Model comparison for reaction time models

```{r results="asis"}
papaja::apa_table(rt_mods_lmer_summary)
```

# Supplemental Table 3: Reliabilities by dataset

```{r reliability_table, results="asis"}
xtable(dataset_reliabilities)
```

### 2024-11-17 5_longitudinal.qmd
bug fix and further development

--- 2024-10-28 5_longitudinal.qmd
@@
 title: "Longitudinal models"
-format: html
+format: 
+  html:
+    toc: true
+execute: 
+  cache: true
 ---
 
-Questions: 
-(from Nilam)
+Notes from Nilam:
+Look at this part of the latent growth tutorial #observed residual variances (constrained to equality)- math2 ~~ sigma2_u*math2
 
-1.
-Is the mean of the latent “language ability” factor scores different across 1st assessment, 2nd assessment, 3rd assessment, etc?
-This is basically an ANOVA on the latent factor scores – using “assessment number” as the categorical predictor.
-– an examination of “practice effects”.
+the two-wave model:
+change in inter-individual differences
+vs. differences in intra-individual change
 
-Is the mean of the latent “language ability” factor scores increasing with age?
-- This is basically a regression on the latent factor scores – and a growth model.
-- In which case we also implicitly test research questions about the variances
-- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?
+This markdown documents attempts to do longitudinal models for the Peekbank data.
 
-- A more nuanced version of the “practice effects" examination might use delta_t. 
-- Is the mean of the latent “language ability” factor different across retest interval?
-- This is basically a regression on the latent factor scores.
+```{r}
+library(here)
+source(here("helper","common.R"))
+library(lavaan)
+library(tidySEM)
 
-- In which case we also implicitly test research questions about the variances
-- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds")) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(admin_num = 1:n(), 
+         time_since_t0 = age - age[1],
+         delta_t = c(0, diff(age)))
 
-3. If we are interested in the covariances –
-- Is the structure of the latent “language ability” factor changing with age?
-- This again gets us into a multi-group SEM –
-- Testing whether the factor loading structure differs across groups.
-- This is most easily invoked with using categorical “time variable” – e.g, assessment number, age-group.
-- However, it is also possible to do using a continuous time variable (e.g., age).
-In this case we basically use a moving window to obtain the covariances and then basically use age as a moderator of the factor loadings to examine the age x factor loading interaction.
-
-# Descriptives
-
-## Longitudinal data
+```
 
@@
 d_sub_long <- d_sub |>
-  filter(subject_id %in% longitudinal$subject_id) |>
-  group_by(subject_id) |>
-  arrange(administration_id) |>
-  mutate(admin_num = 1:n(), 
-         delta_t = c(0,diff(age)))
+  filter(subject_id %in% longitudinal$subject_id)
 
 ggplot(d_sub_long, 
-       aes(x = delta_t, fill = as.factor(admin_num))) +
-  geom_histogram() + 
-  facet_wrap(~dataset_name)
+       aes(x = time_since_t0, fill = dataset_name)) +
+  geom_histogram(binwidth = 1) + 
+  # facet_wrap(~dataset_name) + 
+  xlab("Time since first test")
 ```
-## Datasets with CDIs
 
-Longitudinal data with CDI is just FMW2013 right now.
+# Reliability
+
+How to do this right? WRONG RIGHT NOW
 
 ```{r}
-fmw <- filter(d_sub_long, dataset_name == "fmw_2013") |>
-  mutate(t = ifelse(age < 23, "t1","t2")) |>
-  group_by(subject_id, t) |>
-  summarise(age = mean(age), 
-            log_rt = mean(log_rt), 
-            prod = mean(prod), 
-            long_window_accuracy = mean(long_window_accuracy),
-            delta_t = mean(delta_t)) |>
-  pivot_wider(names_from = "t", 
-              values_from = c("age","log_rt","prod","long_window_accuracy", "delta_t"))
+MONTH_CUTOFF <- 3
 
-summary(lm(prod_t2 ~ long_window_accuracy_t1 * log_rt_t1 + age_t2, 
-           data = fmw))
+d_reliability <- d_sub_long |>
+  filter(admin_num <= 2) |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "admin_num",
+              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
+                              "long_window_acc_var", "prod", "comp"))
+
+dataset_reliabilities <- d_reliability |>
+  group_by(dataset_name) |>
+  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
+            acc = cor(long_window_accuracy_1, 
+                          long_window_accuracy_2, use = "pairwise.complete.obs")) |>
+  filter(dataset_name != "fernald_totlot") 
 ```
 
-## Without CDIs
+# Longitudinal factor analysis
 
 ```{r}
-ggplot(d_sub_long, 
-       aes(x = age, y = long_window_accuracy, group = subject_id, col = dataset_name)) +
-  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
-  geom_smooth(aes(group = 1)) + 
-  scale_y_log10() + 
-  geom_hline(lty = 2, yintercept = .5)
+d_sub_s <- d_sub |>
+  ungroup() |>
+  select(dataset_name, subject_id, administration_id, age, 
+         time_since_t0, delta_t,
+         log_rt, log_rt_var, 
+         long_window_accuracy, long_window_acc_var, prod, comp, ) |>
+  rename(acc = long_window_accuracy, 
+         acc_sd = long_window_acc_var, 
+         log_rt_sd = log_rt_var) |>
+  ungroup() |>
+  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc", "acc_sd", 
+                         "prod", "comp")), 
+                       ~ age_scale(.x, age))) 
+```
+
+Check on amount of data.
+
+# Two-time point
+
+```{r}
+d_sub_wide <- d_sub_s |> 
+  mutate(t = cut(time_since_t0, breaks = c(0, 3, 6, 30), 
+                 include.lowest = TRUE)) |>
+  filter(t != "(6,30]") |>
+  mutate(t = as.numeric(t)) |>
+  group_by(subject_id, dataset_name, t) |>
+  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                            "acc_sd", "prod", "comp")),
+                   ~ mean(.x, na.rm=TRUE))) |> 
+  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                         "acc_sd", "prod", "comp")),
+                ~ ifelse(is.nan(.x), NA, .x))) |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "t",
+              values_from = c("log_rt", "log_rt_sd", "acc", 
+                              "acc_sd", "prod", "comp"), 
+              names_prefix = "t")
+
+d_sub_wide |>
+  ungroup() |>
+  summarise(across(log_rt_t1:comp_t2, 
+                   ~sum(!is.na(.x))))
+```
+
+```{r}
+fa3_model_long <- "
+# measurement model
+vocab_t1 =~ 1*prod_t1 + s1*comp_t1
+accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
+speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1
+
+vocab_t2 =~ 1*prod_t2 + s1*comp_t1
+accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
+speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2
+
+# longitudinal relationships
+vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
+
+# factor covariances
+vocab_t1 ~~ accuracy_t1
+vocab_t1 ~~ speed_t1
+accuracy_t1 ~~ speed_t1
+
+vocab_t2 ~~ accuracy_t2
+vocab_t2 ~~ speed_t2
+accuracy_t2 ~~ speed_t2
+
+# means for the latents
+vocab_t1 ~ 1
+vocab_t2 ~ 1
+accuracy_t1 ~ 1
+accuracy_t2 ~ 1
+speed_t1 ~ 1
+speed_t2 ~ 1
+
+# residual variance for the latents
+vocab_t1 ~~ NA*vocab_t1
+vocab_t2 ~~ NA*vocab_t2
+accuracy_t1 ~~ NA*accuracy_t1
+accuracy_t2 ~~ NA*accuracy_t2
+speed_t1 ~~ NA*speed_t1
+speed_t2 ~~ NA*speed_t2
+"
+
+fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(fit3_long, fit.measures=TRUE, standardize=TRUE)
+
+```
+```{r}
+layout_long = matrix(nrow=7, ncol = 7,
+                     data = c("log_rt_sd_t1","log_rt_t1", "acc_t1", "acc_sd_t1", "prod_t1","comp_t1", NA,
+                               NA, NA, NA, NA, NA, NA,  NA,
+                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
+                              NA, NA, NA, NA, NA, NA,  NA,
+                               NA,"speed_t2",NA,"accuracy_t2",NA,NA,"vocab_t2",
+                               NA, NA, NA, NA, NA, NA,  NA,
+                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", "prod_t2", "comp_t2"), 
+                     byrow = TRUE)
+graph_sem(model = fit3_long, text_size = 2, layout = t(layout_long))
+```
+
+# Full longitudinal model
+
+```{r}
+d_sub_wide <- d_sub_s |> 
+  mutate(t = cut(time_since_t0, breaks = c(0, 3, 6, 9, 12, 30), 
+                 include.lowest = TRUE)) |>
+  filter(t != "(12,30]") |>
+  mutate(t = as.numeric(t)) |>
+  group_by(subject_id, dataset_name, t) |>
+  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                            "acc_sd", "prod", "comp")),
+                   ~ mean(.x, na.rm=TRUE))) |> 
+  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                         "acc_sd", "prod", "comp")),
+                ~ ifelse(is.nan(.x), NA, .x))) |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "t",
+              values_from = c("log_rt", "log_rt_sd", "acc", 
+                              "acc_sd", "prod", "comp"), 
+              names_prefix = "t")
+
+d_sub_wide |>
+  ungroup() |>
+  summarise(across(log_rt_t1:comp_t4, 
+                   ~sum(!is.na(.x))))
+```
+
+```{r}
+fa3_model_long <- "
+# measurement model
+vocab_t1 =~ 1*prod_t1 + s1*comp_t1
+accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
+speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1
+
+vocab_t2 =~ 1*prod_t2 
+accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
+speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2
+
+vocab_t3 =~ 1*prod_t3 
+accuracy_t3 =~ 1*acc_t3 + s2*acc_sd_t3
+speed_t3 =~ 1*log_rt_t3 + s3*log_rt_sd_t3
+
+vocab_t4 =~ 1*prod_t4 
+accuracy_t4 =~ 1*acc_t4 + s2*acc_sd_t4
+speed_t4 =~ 1*log_rt_t4 + s3*log_rt_sd_t4
+
+# factor covariances
+vocab_t1 ~~ accuracy_t1
+vocab_t1 ~~ speed_t1
+accuracy_t1 ~~ speed_t1
+
+vocab_t2 ~~ accuracy_t2
+vocab_t2 ~~ speed_t2
+accuracy_t2 ~~ speed_t2
+
+vocab_t3 ~~ accuracy_t3
+vocab_t3 ~~ speed_t3
+accuracy_t3 ~~ speed_t3
+
+vocab_t4 ~~ accuracy_t4
+vocab_t4 ~~ speed_t4
+accuracy_t4 ~~ speed_t4
+
+# means for the latents
+vocab_t1 ~ 1
+vocab_t2 ~ 1
+vocab_t3 ~ 1
+vocab_t4 ~ 1
+accuracy_t1 ~ 1
+accuracy_t2 ~ 1
+accuracy_t3 ~ 1
+accuracy_t4 ~ 1
+speed_t1 ~ 1
+speed_t2 ~ 1
+speed_t3 ~ 1
+speed_t4 ~ 1
+
+# regressions
+accuracy_intercept =~ 1*accuracy_t1 + 1*accuracy_t2 + 1*accuracy_t3 + 1*accuracy_t4 
+accuracy_slope =~ 1*accuracy_t1 + 2*accuracy_t2 + 3*accuracy_t3 + 4*accuracy_t4
+speed_intercept =~ 1*speed_t1 + 1*speed_t2 + 1*speed_t3 + 1*speed_t4 
+speed_slope =~ 1*speed_t1 + 2*speed_t2 + 3*speed_t3 + 4*speed_t4
+vocab_intercept =~ 1*vocab_t1 + 1*vocab_t2 + 1*vocab_t3 + 1*vocab_t4 
+vocab_slope =~ 1*vocab_t1 + 2*vocab_t2 + 3*vocab_t3 + 4*vocab_t4
+
+# parameters for the regressions
+vocab_intercept ~ 1
+vocab_slope ~ 1
+accuracy_intercept ~ 1
+accuracy_slope ~ 1
+speed_intercept ~ 1
+speed_slope ~ 1
+
+# # residual variance for the latents
+# vocab_t1 ~~ NA*vocab_t1
+# vocab_t2 ~~ NA*vocab_t2
+# accuracy_t1 ~~ NA*accuracy_t1
+# accuracy_t2 ~~ NA*accuracy_t2
+# speed_t1 ~~ NA*speed_t1
+# speed_t2 ~~ NA*speed_t2
+"
+
+fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(fit3_long, fit.measures=TRUE, standardize=TRUE)
 
@@
 ```{r}
-ggplot(filter(d_sub_long, n_trials_rt > 3),
-       aes(x = age, y = log_rt, group = subject_id, col = dataset_name)) +
-  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
-  geom_smooth(aes(group = 1)) + 
-  scale_y_log10() 
-```
-# Q1: age-related change in the latent score
+layout_dt = matrix(nrow=11, ncol = 7,
+                   data = c("prod_t1", "prod_t2", "prod_t3", "prod_t4", "vocab_intercept", NA, NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "comp_t1", NA, NA, NA, "vocab_slope", NA,NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "accuracy_t1", "accuracy_t2", "accuracy_t3", "accuracy_t4", "accuracy_intercept", NA, NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "acc_sd_t1", "acc_sd_t2", "acc_sd_t3", "accuracy_sd_t4", "accuracy_slope", NA, NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "log_rt_t1", "log_rt_t2", "log_rt_t3", "log_rt_t4", "speed_int", NA, NA,
+                            NA, NA, NA, NA, NA, NA, NA,
+                            "log_rt_sd_t1", "log_rt_sd_t2", "log_rt_sd_t3", "log_rt_sd_t4", "speed_slope", NA, NA),
+                   byrow = TRUE)
 
-Growth in latent language factor. This is the same as the previous factor analysis, but we only look within child and within dataset.
-
-## 1 factor approach
-
-This has the benefits of simplicity, even though we know that the 1F model is "wrong."
-
-```{r}
-fa_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod + comp"
-fit <- cfa(fa_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
-
-d_sub$f1 <- lavPredict(fit, newdata = d_sub_mat_s)
+graph_sem(model = fit3_long, text_size = 3, layout = layout_dt)
 
 ```
-
-```{r}
-ggplot(d_sub |>
-         filter(subject_id %in% longitudinal$subject_id), 
-       aes(x = age, y = f1, group = subject_id, col = dataset_name)) + 
-  geom_point(alpha = .4) + 
-  geom_line(alpha = .4) + 
-  geom_smooth(aes(group = dataset_name), method = "lm")
-```
-Note that nesting age slopes doesn't converge. Could go Bayesian if we cared...
-
-```{r}
-long_mod <- lmer(f1 ~ age + (1 | subject_id) + (1 | dataset_name), 
-                 data = d_sub)
-
-summary(long_mod)
-```
-
-## 2 factor approach
-
-Let's try looking at the speed/accuracy model across time.
-
-```{r}
-fa2_model <-  "accuracy =~ long_window_accuracy + long_window_acc_var + comp + prod
-               speed =~ rt + rt_var"
-
-fit2 <- cfa(fa2_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
-
-preds <- predict(fit2)
-
-d_sub$accuracy <- preds[,1]
-d_sub$speed <- preds[,2]
-```
-
-Merge with longitudinal data.
-
-```{r}
-d_long <- d_sub |>
-  filter(subject_id %in% longitudinal$subject_id) |>
-  group_by(subject_id) |>
-  arrange(administration_id) |>
-  mutate(admin_num = 1:n(), 
-         delta_t = c(0,diff(age)), 
-         speed_0 = speed[1]) |>
-  group_by(dataset_name) |>
-  mutate(speed_0_high = speed_0 > median(speed_0, na.rm=TRUE))
-```
-
-Plot.
-
-```{r}
-ggplot(d_long |>
-         pivot_longer(speed:accuracy, names_to = "variable", values_to = "value"), 
-       aes(x = age, y = value, group = subject_id, col = dataset_name)) + 
-  geom_point(alpha = .4) + 
-  geom_line(alpha = .4) + 
-  geom_smooth(aes(group = dataset_name), method = "lm") + 
-  scale_x_log10() +
-  facet_wrap(~variable) +
-  theme(legend.position = "bottom") + 
-  xlab("Age (months)") + 
-  ylab("Latent score") 
-```
-Change in the latents over age.
-
-```{r}
-long_mod_acc <- lmer(accuracy ~ age + (1 | subject_id) + (1 | dataset_name), 
-                 data = d_sub)
-long_mod_speed <- lmer(speed ~ age + (1 | subject_id) + (1 | dataset_name), 
-                 data = d_sub)
-
-summary(long_mod_acc)
-summary(long_mod_speed)
-```
-
-# Q2: relations between latents and other factors
-
-Longitudinal model.
-
-```{r}
-lmod <- lmer(accuracy ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
-     data = d_long)
-
-summary(lmod)
-```
-
-```{r}
-ggplot(d_long, 
-       aes(x = age, y = accuracy, group = subject_id, col = dataset_name, lty = speed_0_high)) + 
-  geom_line(alpha = .4) + 
-  geom_smooth(aes(group = speed_0_high), method = "lm") + 
-  facet_wrap(~dataset_name, scales="free_x")
-```

### 2024-11-17 paper/paper.Rmd
bug fix and further development

@@
 suppressPackageStartupMessages(library(xtable))
-# suppressPackageStartupMessages(library(kableExtra))
 suppressPackageStartupMessages(library(papaja))
@@
 source(here("helper","common.R"))
+
 d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
   mutate(log_age = log(age))
-d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds")) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(admin_num = 1:n(), 
+         time_since_t0 = age - age[1],
+         delta_t = c(0, diff(age)))
 
@@
 These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions).
-Table \ref{tab:datasets} shows the characteristics of individual datasets.
-The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allow us to take a more detailed look at early language processing than has previously been possible.
+Table \ref{tab:datasets} shows the characteristics of individual datasets (see also Figures S1 and S2).
+The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed exploration of the development of word recognition than has previously been possible.
 
@@
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms).
-
-Our first question was about the functional form of the relationships between age, speed, and accuracy.
+Our first question was about the functional form of the relationships between age, speed, and accuracy (see Table S-XYZ for raw correlations between variables).
 To investigate this question, we fit linear mixed effects models predicting accuracy and reaction time on each trial across the full dataset with random slopes of age by study and random intercepts by participant.
@@
 The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text.
-All findings reported here hold for both window sizes, however.] and the best fitting model of speed predicted log reaction time as a function of log age as well (See Supplemental Tables 1 and 2 for model fit statistics).
+All findings reported here hold for both window sizes, however (see SI: Short Window Analyses).] and the best fitting model of speed predicted log reaction time as a function of log age as well (See Tables S1 and S2 for model fit statistics).
 
@@
 
-## Speed and accuracy covary with vocabulary {.unnumbered}
+## Speed and accuracy relate to vocabulary size {.unnumbered}
 
@@
 We fit a series of factor analytic models to explore the dimensionality of the data.
-Initial exploratory factor analysis using paralell analysis suggested that three factors explained substantial variance in the data.
+Initial exploratory factor analysis using paralell analysis suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
 Due to missingness of data, we used confirmatory factory analysis with full information maximum likelihood to find the best set of loadings.
 The best fitting model was a three-factor model with factors for speed (reaction time and reaction time variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI).
-Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit_stats$fit["rmsea.pvalue"], 3)`).
+Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`)`; see SI: Alternative Factor Structures).
 
 ```{r age_regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age.
-Squares show  Solid lines show regression coefficients and factor loadings, and dotted lines show covariances (reflecting age-residualized correlations).
+Observed variables are notated as squares and latent variables are notated as circles Factor loadings and regression coefficients are shown with straight, solid lines; covariances are shown with dashed lines; residual variances are shown as solid circular connections.
+Stars show conventional levels of statistical significance, e.g.
+* indicates p < .05, ** indicates p < .01, and *** indicates p < .001.
+Covariances reflect age-residualized correlations between variables.
 Abbreviations: acc = accuracy, prod = production vocabulary, comp = comprehension vocabulary.", fig.height = 3}
@@
   group_by(subject_id) |>
-  arrange(administration_id) |>
+  arrange(age) |>
   mutate(admin_num = 1:n(), 
-         delta_t = c(0,diff(age)))
+         delta_t = age - age[1])
 
@@
 ```{r prep_long_data}
-d_sub_long <- d_sub |>
-  mutate(across(all_of(c("log_rt", "log_rt_var", "long_window_accuracy",
-                         "long_window_acc_var", "prod", "comp")),
-                ~ age_scale(.x,age))) |>
-  group_by(subject_id) |>
-  arrange(administration_id) |>
-  mutate(admin_num = 1:n(), 
-         delta_t = c(0, diff(age)))
-
-d_sub_long_t <- d_sub_long |> 
-  mutate(t = ifelse(delta_t > MONTH_CUTOFF, "t2", "t1")) |>
+# rename and scale variables
+d_sub_s <- d_sub |>
+  ungroup() |>
+  select(dataset_name, subject_id, administration_id, age, 
+         time_since_t0, delta_t,
+         log_rt, log_rt_var, 
+         long_window_accuracy, long_window_acc_var, prod, comp, ) |>
   rename(acc = long_window_accuracy, 
@@
          log_rt_sd = log_rt_var) |>
+  ungroup() |>
+  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc", "acc_sd", 
+                         "prod", "comp")), 
+                       ~ age_scale(.x, age)))
+
+d_sub_wide <- d_sub_s |> 
+  mutate(t = cut(time_since_t0, breaks = c(0, 6, 12, 30), 
+                 include.lowest = TRUE)) |>
+  filter(t != "(12,30]") |>
+  mutate(t = as.numeric(t)) |>
   group_by(subject_id, dataset_name, t) |>
@@
                          "acc_sd", "prod", "comp")),
-                ~ ifelse(is.nan(.x), NA, .x)))
-
-d_sub_wide <- d_sub_long_t |>
+                ~ ifelse(is.nan(.x), NA, .x))) |>
   pivot_wider(id_cols = c("subject_id","dataset_name"), 
@@
               values_from = c("log_rt", "log_rt_sd", "acc", 
-                              "acc_sd", "prod", "comp"))
+                              "acc_sd", "prod", "comp"), 
+              names_prefix = "t")
+
 ```
@@
 
-vocab_t2 =~ 1*prod_t2 + s1*comp_t2
+vocab_t2 =~ 1*prod_t2 
 accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
@@
 Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure longitudinal relationships between LWL and vocabulary via a longitudinal latent factor model.
-We fit the same three-factor model, but this time averaging longitudinal administrations after splitting them into two categories: those datapoints within a two-month time period (which we designate as $t_1$) and those datapoints between three and six months from the initial measurement ($t_2$).
+We fit the same three-factor model, but this time averaging longitudinal administrations after splitting them into two categories: those datapoints within a two-month time period (which we designate as $t_1$) and those datapoints between three and six months from the initial measurement ($t_2$) (see SI: Sensitivity Analysis for Longitudinal Models).
 To ensure measurement invariance, we coupled the factor loadings across the two time-points.
@@
 
-The fitted longitudinal model is shown in \ref{fig:longitudinal}.
+The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}.
 Overall fit statistics were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit3_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit3_long_summary$fit["rmsea.pvalue"], 3)`).
@@
 
-```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Coupled change model."}
+```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between latent factors.
+Variables marked t1 are measures from two or fewer months from study initiation; variables marked t2 are measures from 3--6 months after study initiation.
+Factor loadings are constrained to be equivalent between t1 and t2.
+Plotting conventions are as in Figure 3."}
 layout_long = matrix(nrow=7, ncol = 7,
@@
                                NA, NA, NA, NA, NA, NA,  NA,
-                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", "prod_t2","comp_t2"), 
+                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", NA, "prod_t2"), 
                      byrow = TRUE)
@@
 This model significantly improved fit compared with a model that set this coupling parameter to zero but was otherwise unchanged ($\chi^2 = `r round(speed_vocab_anova$c2d[2], 2)`$, $p =  `r round(speed_vocab_anova$p[2], 4)`$).
-These findings are consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2012,@Weisleder2013].
+These findings are consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
 
@@
 
-We used a new database aggregating results across many individual developmental eye-tracking experiments to investigate the developmental course of word recognition and its relationship with early language learning.
-We found continuous developmental changes across all measures, consistent with a characterization of word recognition as a skill.
-Speed and accuracy both improved asymptoticaly, with evidence that recognition speed showed the log-log relationship associated with the power law of practice, that is, gradually converging on mature levels of processing efficiency.
+How does word recognition change across early childhood, and how does it relate to language learning?
+We investigated these questions using a new dataset of developmental eye-tracking measurements.
+We found continuous developmental changes from ages 1 -- 6 years.
+Speed and accuracy both improved asymptotically, with evidence that recognition speed showed the log-log relationship associated with the power law of practice, that is, gradually converging on mature levels of processing efficiency.
 Further, older children showed less variability in accuracy and reaction time across trials, indicating more robust execution of this skilled behavior.
+Speed and accuracy were both related to vocabulary size concurrently, and processing speed was also related to later vocabulary growth.
+
+By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had samples sizes at least an order of magnitude smaller than ours.
+The presence of multiple studies with longitudinal data,
+
+These findings are consistent with theories that posit that language learning is a process of skill acquisition, in which children become skillful at quickly converting ephemeral signals into meaning [@christiansen2016].
 

### 2024-11-17 paper/supplemental_information.Rmd
bug fix and further development

@@
 
-```{r, echo=FALSE}
-knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+library(here)
+source(here("helper","common.R"))
 ```
 
-```{r}
+```{r, echo=FALSE}
 knitr::purl("paper.Rmd", documentation = 0)
-source("paper.R")
+suppressMessages(source("paper.R"))
 ```
 
-# Supplemental Figure 1: Age histograms
+# Dataset description
 
@@
 
-# Supplemental Table 1: Model comparison for accuracy models
+```{r, fig.cap="Number of administrations for datasets with repeated measurements."}
+longitudinal <- d_sub |>
+  group_by(dataset_name, subject_id) |>
+  count() |>
+  filter(n > 1)
+
+d_sub_long <- d_sub |>
+  filter(subject_id %in% longitudinal$subject_id) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(admin_num = 1:n(), 
+         delta_t = age - age[1])
+
+ggplot(d_sub_long, 
+       aes(x = delta_t, fill = dataset_name)) +
+  geom_histogram(binwidth = 1) + 
+  # facet_wrap(~dataset_name) + 
+  xlab("Age since first test")
+```
+
+# Speed and accuracy
+
+Correlations between variables
+
+```{r, results = "asis"}
+
+cor(select(d_sub, age, log_age, rt, log_rt, 
+           long_window_accuracy, short_window_accuracy, prod, comp), 
+    use = "pairwise.complete.obs") |>
+  round(2) |>
+  papaja::apa_table()
+
+```
 
@@
 
-# Supplemental Table 2: Model comparison for reaction time models
-
 ```{r results="asis"}
@@
 
-# Supplemental Table 3: Reliabilities by dataset
+# Exploratory Factor Analysis
+
+- Full EFA results
+
+# Alternative Factor Structures
+
+- Alternative factor models: variances together, two factor, etc. etc.
+
+# Sensitivity Analysis for Longitudinal Models
+
+THIS IS WRONG
 
@@
 ```
+
+Sensitivity analysis for age split point (e.g., split at 3 months)
+
+Sensitivity analysis for measurement invariance (e.g., not constraining loadings)
+
+# Analyses using short window accuracy
+
+## Longitudinal factor analysis
+
+## Longitudinal factor analysis

### 2024-12-13 5_longitudinal.qmd
paper and longitudinal updates

@@
 
-Notes from Nilam:
-Look at this part of the latent growth tutorial #observed residual variances (constrained to equality)- math2 ~~ sigma2_u*math2
-
-the two-wave model:
-change in inter-individual differences
-vs. differences in intra-individual change
-
 This markdown documents attempts to do longitudinal models for the Peekbank data.
@@
 library(tidySEM)
+library(nlme)
+library(emmeans)
+library(lme4)
+library(lmerTest)
 
@@
 
-# Reliability
+# Draft 1: Replicate growth models from Fernald & Marchman 2012
 
-How to do this right? WRONG RIGHT NOW
+First, we try replicating the original growth models.
+
+## Just FM2012
 
 ```{r}
-MONTH_CUTOFF <- 3
-
-d_reliability <- d_sub_long |>
-  filter(admin_num <= 2) |>
-  pivot_wider(id_cols = c("subject_id","dataset_name"), 
-              names_from = "admin_num",
-              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
-                              "long_window_acc_var", "prod", "comp"))
-
-dataset_reliabilities <- d_reliability |>
-  group_by(dataset_name) |>
-  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
-            acc = cor(long_window_accuracy_1, 
-                          long_window_accuracy_2, use = "pairwise.complete.obs")) |>
-  filter(dataset_name != "fernald_totlot") 
+fm2012 <- filter(d_sub, dataset_name == "fernald_marchman_2012") |>
+  group_by(subject_id) |>
+  mutate(age_group = case_when(age < 21 ~ 18, 
+                               age > 21 & age < 27 ~ 24, 
+                               .default = 30)) |>
+  arrange(age) |>
+  mutate(rt_t0 = rt[1]) |>
+  ungroup() |>
+  mutate(rt_t0_group = ifelse(rt_t0 < median(rt_t0), "low", "high"))
 ```
 
-# Longitudinal factor analysis
+```{r}
+ggplot(fm2012, 
+       aes(x = age_group, y= prod)) + 
+  geom_line(alpha = .2, aes(group = subject_id)) + 
+  stat_summary(aes(col = rt_t0_group)) + 
+  geom_smooth(aes(col = rt_t0_group), method = "lm", formula = y ~ poly(x, 2))
+```
+
+This looks similar (but maybe a bit worse with log rt.
+
+```{r}
+fm_mod <- lmer(prod ~ age * rt_t0 + (age | subject_id), 
+     data = fm2012, 
+     control = lmerControl(optimizer = "bobyqa"))
+
+summary(fm_mod)
+```
+
+Note that this model looks a lot better with quadratic growth. (They used quadratics in the original).
+
+```{r}
+fm_mod <- lmer(prod ~ age * rt_t0 + I(age^2) * rt_t0 + (age | subject_id), 
+     data = fm2012, 
+     control = lmerControl(optimizer = "bobyqa"))
+
+summary(fm_mod)
+```
+
+## All data
+
+Let's try the same model with all the data.
+
+```{r}
+d_growth <- d_sub |>
+  filter(!is.na(prod), subject_id %in% longitudinal$subject_id) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(rt_t0 = log_rt[1]) |>
+  ungroup() |>
+  mutate(rt_t0_group = ifelse(rt_t0 < median(rt_t0, na.rm=TRUE), "low", "high"))
+
+ggplot(filter(d_growth, !is.na(rt_t0)), 
+              aes(x = age, y= prod)) +
+  geom_line(alpha = .2, aes(group = subject_id)) +
+  geom_smooth(aes(col = rt_t0_group), method = "lm", formula = y ~ x + I(x^2))
+```
+
+```{r}
+ggplot(filter(d_growth, !is.na(rt_t0)), 
+              aes(x = age, y= prod)) +
+  geom_line(alpha = .2, aes(group = subject_id)) +
+  geom_smooth(aes(col = rt_t0_group), method = "lm", formula = y ~ x + I(x^2))
+```
+
+Linear.
+
+```{r}
+d_growth$age_15 <- d_growth$age - 15
+all_mod <- lmer(prod ~ age_15 *rt_t0 +  (age  | subject_id) + (age | dataset_name), 
+     data = d_growth, 
+     control = lmerControl(optimizer = "bobyqa"))
+
+summary(all_mod)
+```
+
+Again better with quadratics. (Won't converge with nested quadratics in random effects).
+
+```{r}
+d_growth$age_15 <- d_growth$age - 15
+all_mod <- lmer(prod ~ age_15 *rt_t0  + I(age_15^2) * rt_t0 + (age  | subject_id) + (age | dataset_name), 
+     data = d_growth, 
+     control = lmerControl(optimizer = "bobyqa"))
+
+summary(all_mod)
+```
+
+So what we see here is that RT at t0 is very predictive of something about growth.
+
+On the other hand, these models don't take into account RT at t1 or t2. So they don't tell us if RT is a stable trait that's associated with growth...
+
+# Draft 2: Two-time points
+
+So we try fitting two factor analyses and looking at coupling over time.
 
@@
   select(dataset_name, subject_id, administration_id, age, 
-         time_since_t0, delta_t,
+         time_since_t0, delta_t, admin_num,
          log_rt, log_rt_var, 
@@
 Check on amount of data.
-
-# Two-time point
 
@@
 summary(fit3_long, fit.measures=TRUE, standardize=TRUE)
-
 ```
@@
 
-# Full longitudinal model
+# Draft 3: Full longitudinal model
 
@@
 ```
+## with just the observed variables
+
+```{r}
+slopes_model_long <- "
+
+# regressions
+accuracy_intercept =~ 1*acc_t1 + 1*acc_t2 + 1*acc_t3 + 1*acc_t4 
+accuracy_slope =~ 1*acc_t1 + 2*acc_t2 + 3*acc_t3 + 4*acc_t4
+speed_intercept =~ 1*log_rt_t1 + 1*log_rt_t2 + 1*log_rt_t3 + 1*log_rt_t4 
+speed_slope =~ 1*log_rt_t1 + 2*log_rt_t2 + 3*log_rt_t3 + 4*log_rt_t4
+vocab_intercept =~ 1*prod_t1 + 1*prod_t2 + 1*prod_t3 + 1*prod_t4 
+vocab_slope =~ 1*prod_t1 + 2*prod_t2 + 3*prod_t3 + 4*prod_t4
+"
+
+slopes_long <- growth(slopes_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(slopes_long, fit.measures=TRUE, standardize=TRUE)
+
+```
+
+```{r}
+layout <- read.csv(here("misc/layout_slopes.csv"), header = FALSE)
+
+graph_sem(model = slopes_long, text_size = 3, layout=t(layout))
+
+```
+
+This is still linear. We can add quadratic?
+
+```{r}
+d_sub_wide |>
+  select(subject_id, prod_t1, prod_t2, prod_t3, prod_t4) |>
+  pivot_longer(starts_with("prod"), names_to = "variable", values_to = "prod") |>
+  mutate(t = as.numeric(str_sub(variable, -1))) |>
+  ggplot(aes(x = t, y = prod)) + 
+  geom_line(aes(group = subject_id), alpha = .1) + 
+  geom_smooth(aes(group = 1), method = "lm") + 
+  geom_smooth(aes(group = 1), method = "lm", formula = y ~ poly(x, 2), col = "green")
+```
+
+But it seems unnecessary for this view of the data. Arguably we're losing a lot by doing it this way.
+
+## try finer-grained rebinning
+
+Let's try putting prod back to its natural units?
+
+```{r}
+d_sub_wide_fine <- d_sub_s |>
+  mutate(t = cut(time_since_t0, breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 30), 
+                 include.lowest = TRUE)) |>
+  filter(t != "(12,30]") |>
+  mutate(t = as.numeric(t)) |>
+  group_by(subject_id, dataset_name, t) |>
+  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                            "acc_sd", "prod", "comp")),
+                   ~ mean(.x, na.rm=TRUE))) |> 
+  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
+                         "acc_sd", "prod", "comp")),
+                ~ ifelse(is.nan(.x), NA, .x))) |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "t",
+              values_from = c("log_rt", "log_rt_sd", "acc", 
+                              "acc_sd", "prod", "comp"), 
+              names_prefix = "t")
+
+d_sub_wide_fine |>
+  ungroup() |>
+  summarise(across(log_rt_t1:comp_t4, 
+                   ~sum(!is.na(.x))))
+```
+
+```{r}
+d_sub_wide_fine |>
+  select(subject_id, starts_with("prod")) |>
+  pivot_longer(starts_with("prod"), names_to = "variable", values_to = "prod") |>
+  mutate(t = as.numeric(str_sub(variable, -1))) |>
+  ggplot(aes(x = t, y = prod)) + 
+  geom_line(aes(group = subject_id), alpha = .1) 
+```
+
+```{r}
+slopes_model_long <- "
+
+# regressions
+accuracy_intercept =~ 1*acc_t1 + 1*acc_t2 + 1*acc_t3 + 1*acc_t4 + 1*acc_t5 
+accuracy_slope =~ 1*acc_t1 + 2*acc_t2 + 3*acc_t3 + 4*acc_t4 + 5*acc_t5 
+speed_intercept =~ 1*log_rt_t1 + 1*log_rt_t2 + 1*log_rt_t3 + 1*log_rt_t4 + 1*log_rt_t5 
+speed_slope =~ 1*log_rt_t1 + 2*log_rt_t2 + 3*log_rt_t3 + 4*log_rt_t4 + 5*log_rt_t5
+vocab_intercept =~ 1*prod_t1 + 1*prod_t2 + 1*prod_t3 + 1*prod_t4 + 1*prod_t5 
+vocab_slope =~ 1*prod_t1 + 2*prod_t2 + 3*prod_t3 + 4*prod_t4 + 5*prod_t5
+
+# let variances of each latent vary
+accuracy_intercept ~~ NA*accuracy_intercept
+accuracy_slope ~~ NA*accuracy_slope
+speed_intercept ~~ NA*speed_intercept
+speed_slope ~~ NA*speed_slope
+vocab_intercept ~~ NA*vocab_intercept
+vocab_slope ~~ NA*vocab_slope
+"
+
+slopes_long <- growth(slopes_model_long, d_sub_wide_fine, std.lv=TRUE, missing='fiml')
+
+summary(slopes_long, fit.measures=TRUE, standardize=TRUE)
+
+```
+
+```{r}
+layout <- read.csv(here("misc/layout_slopes_5.csv"), header = FALSE)
+
+graph_sem(model = slopes_long, text_size = 3, layout=t(layout))
+
+```
+
+## with lots of latent factor structure
+
+This one still seems not quite right.
 
@@
 
-# factor covariances
-vocab_t1 ~~ accuracy_t1
-vocab_t1 ~~ speed_t1
-accuracy_t1 ~~ speed_t1
+# means for the latents
+# vocab_t1 ~ 1
+# vocab_t2 ~ 1
+# vocab_t3 ~ 1
+# vocab_t4 ~ 1
+# accuracy_t1 ~ 1
+# accuracy_t2 ~ 1
+# accuracy_t3 ~ 1
+# accuracy_t4 ~ 1
+# speed_t1 ~ 1
+# speed_t2 ~ 1
+# speed_t3 ~ 1
+# speed_t4 ~ 1
 
-vocab_t2 ~~ accuracy_t2
-vocab_t2 ~~ speed_t2
-accuracy_t2 ~~ speed_t2
-
-vocab_t3 ~~ accuracy_t3
-vocab_t3 ~~ speed_t3
-accuracy_t3 ~~ speed_t3
-
-vocab_t4 ~~ accuracy_t4
-vocab_t4 ~~ speed_t4
-accuracy_t4 ~~ speed_t4
-
-# means for the latents
-vocab_t1 ~ 1
-vocab_t2 ~ 1
-vocab_t3 ~ 1
-vocab_t4 ~ 1
-accuracy_t1 ~ 1
-accuracy_t2 ~ 1
-accuracy_t3 ~ 1
-accuracy_t4 ~ 1
-speed_t1 ~ 1
-speed_t2 ~ 1
-speed_t3 ~ 1
-speed_t4 ~ 1
+# autoregression for the latents
+# vocab_t2 ~ 1 * vocab_t1
+# vocab_t3 ~ 1 * vocab_t2
+# vocab_t4 ~ 1 * vocab_t3
+# accuracy_t2 ~ 1 * accuracy_t1
+# accuracy_t3 ~ 1 * accuracy_t2
+# accuracy_t4 ~ 1 * accuracy_t3
+# speed_t2 ~ 1 * speed_t1
+# speed_t3 ~ 1 * speed_t2
+# speed_t4 ~ 1 * speed_t3
 
@@
 
-# # residual variance for the latents
-# vocab_t1 ~~ NA*vocab_t1
-# vocab_t2 ~~ NA*vocab_t2
-# accuracy_t1 ~~ NA*accuracy_t1
-# accuracy_t2 ~~ NA*accuracy_t2
-# speed_t1 ~~ NA*speed_t1
-# speed_t2 ~~ NA*speed_t2
+# residual variance for the latents
+vocab_t1 ~~ NA*vocab_t1
+vocab_t2 ~~ NA*vocab_t2
+vocab_t3 ~~ NA*vocab_t3
+vocab_t4 ~~ NA*vocab_t4
+accuracy_t1 ~~ NA*accuracy_t1
+accuracy_t2 ~~ NA*accuracy_t2
+accuracy_t3 ~~ NA*accuracy_t3
+accuracy_t4 ~~ NA*accuracy_t4
+speed_t1 ~~ NA*speed_t1
+speed_t2 ~~ NA*speed_t2
+speed_t3 ~~ NA*speed_t3
+speed_t4 ~~ NA*speed_t4
+
+# residual variance for the other latents
+vocab_intercept ~~ NA*vocab_intercept
+vocab_slope ~~ NA*vocab_slope
+accuracy_intercept ~~ NA*accuracy_intercept
+accuracy_slope ~~ NA*accuracy_slope
+speed_intercept ~~ NA*speed_intercept
+speed_slope ~~ NA*speed_slope
+
+# set observed intercepts to zero - these should go into the latents
+prod_t1 ~ 0
+prod_t2 ~ 0
+prod_t3 ~ 0
+prod_t4 ~ 0
+comp_t1 ~ 0 
+log_rt_t1 ~ 0
+log_rt_t2 ~ 0
+log_rt_t3 ~ 0
+log_rt_t4 ~ 0
+log_rt_sd_t1 ~ 0 
+log_rt_sd_t2 ~ 0
+log_rt_sd_t3 ~ 0 
+log_rt_sd_t4 ~ 0 
+acc_t1 ~ 0 
+acc_t2 ~ 0 
+acc_t3 ~ 0 
+acc_t4 ~ 0 
+acc_sd_t1 ~ 0
+acc_sd_t2 ~ 0
+acc_sd_t3 ~ 0
+acc_sd_t4 ~ 0
 "
@@
 ```{r}
-layout_dt = matrix(nrow=11, ncol = 7,
-                   data = c("prod_t1", "prod_t2", "prod_t3", "prod_t4", "vocab_intercept", NA, NA,
-                            NA, NA, NA, NA, NA, NA, NA,
-                            "comp_t1", NA, NA, NA, "vocab_slope", NA,NA,
-                            NA, NA, NA, NA, NA, NA, NA,
-                            "accuracy_t1", "accuracy_t2", "accuracy_t3", "accuracy_t4", "accuracy_intercept", NA, NA,
-                            NA, NA, NA, NA, NA, NA, NA,
-                            "acc_sd_t1", "acc_sd_t2", "acc_sd_t3", "accuracy_sd_t4", "accuracy_slope", NA, NA,
-                            NA, NA, NA, NA, NA, NA, NA,
-                            "log_rt_t1", "log_rt_t2", "log_rt_t3", "log_rt_t4", "speed_int", NA, NA,
-                            NA, NA, NA, NA, NA, NA, NA,
-                            "log_rt_sd_t1", "log_rt_sd_t2", "log_rt_sd_t3", "log_rt_sd_t4", "speed_slope", NA, NA),
-                   byrow = TRUE)
+layout <- read.csv(here("misc/layout_dt3.csv"), header = FALSE)
 
-graph_sem(model = fit3_long, text_size = 3, layout = layout_dt)
+graph_sem(model = fit3_long, text_size = 3, layout=t(layout))
 
 ```
+
+# Draft 4: Multi-level modeling of a single variable
+
+Let's try NLME.
+
+```{r}
+d_sub_prod <- filter(d_sub, !is.na(prod)) |>
+  group_by(subject_id) |>
+  mutate(log_rt_0 = log_rt[1],
+         acc_0 = long_window_accuracy[1])
+
+library(nlme)
+
+mod1 <- lme(prod ~ age, 
+            random = ~ 1 | dataset_name / subject_id,
+            na.action = na.omit, 
+            data = d_sub_prod)
+
+summary(mod1)
+```
+
+Plotting model.
+
+```{r}
+ages <- seq(min(d_sub_prod$age), max(d_sub_prod$age), .1)
+datasets <- unique(d_sub_prod$dataset_name)
+
+# fixed effects with CIs
+newdata_global <- tibble(age = ages) 
+global_preds <- emmeans(mod1, ~ age, at = newdata_global, CIs = TRUE) |> 
+  summary() |>
+  tibble()
+
+# random effects
+dataset_preds <- d_sub |>
+  filter(!is.na(prod)) |> # filter on predictor
+  group_by(dataset_name) |>
+  summarise(min_age = min(age), 
+            max_age = max(age)) |>
+  group_by(dataset_name) |>
+  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1)))) |>
+  select(-min_age, -max_age) |>
+  unnest(col = "df") |>
+  ungroup()
+
+dataset_preds$pred <- predict(mod1, 
+                              newdata = dataset_preds, 
+                              level = c(0,1))$predict.dataset_name
+
+ggplot(d_sub_prod, aes(x = age, y = prod, col = dataset_name)) + 
+  geom_point(alpha = .1) + 
+  geom_line(data = global_preds, aes(y = emmean), col = "blue", size = 1.5) +
+  geom_ribbon(data = global_preds, 
+              aes(ymin = lower.CL, ymax = upper.CL, y = emmean), col = NA,
+              fill = "blue",alpha = .1) + 
+  geom_line(data = dataset_preds, aes(y = pred))
+```
+
+## Logistics
+
+So I guess we need good parameters.
+
+```{r}
+d_sub$logistic = SSlogis(d_sub$age, Asym = 1, xmid = 0, scal = 3)
+
+ggplot(d_sub, aes(x = age, y = prod)) + 
+  geom_point(alpha = 1.) + 
+  geom_line(aes(x = age, y = logistic))
+
+```
+
+Use a logistic growth model.
+
+```{r}
+mod2 <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
+             data = d_sub,
+             fixed = xmid + scale ~ 1,
+             random = xmid + scale ~ 1,
+             groups = ~ dataset_name,
+             start = c(xmid = 20, scale = 3),
+             na.action = na.exclude, 
+             control = list(maxIter = 1000, tolerance = 10))
+
+summary(mod2)
+
+# fixed effects (no CIs)
+global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
+                                      max(d_sub_prod$age), .1))
+global_preds$pred <- predict(mod2, newdata = global_preds, level = 0)
+
+# random effects
+dataset_preds <- d_sub |>
+  filter(!is.na(prod)) |> # filter on predictor
+  group_by(dataset_name) |>
+  summarise(min_age = min(age), 
+            max_age = max(age)) |>
+  group_by(dataset_name) |>
+  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1)))) |>
+  select(-min_age, -max_age) |>
+  unnest(col = "df") |>
+  ungroup()
+
+dataset_preds$pred <- predict(mod2, 
+                              newdata = dataset_preds, 
+                              level = c(0,1))$predict.dataset_name
+
+ggplot(d_sub_prod, aes(x = age, y = prod, col = dataset_name)) + 
+  geom_point(alpha = .1) + 
+  geom_line(data = global_preds, aes(y = pred), col = "blue", size = 1.5) +
+  geom_line(data = dataset_preds, aes(y = pred))
+```
+## Interaction with t0 RT
+
+Now let's use t0 processing attributes in a regression.
+
+```{r}
+d_sub_prod <- d_sub_prod |>
+  group_by(subject_id) |>
+  mutate(log_rt_0 = log_rt[1],
+         acc_0 = long_window_accuracy[1])
+
+mod3 <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
+             data = d_sub_prod,
+             fixed = xmid + scale ~ log_rt_0,
+             random = xmid + scale ~ 1,
+             groups = ~ dataset_name,
+             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 0, `scale:log_rt_0` = 3),
+             na.action = na.exclude, 
+             control = list(maxIter = 1000, tolerance = 10))
+
+summary(mod3)
+```
+
+```{r}
+# fixed effects (no CIs)
+global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
+                                      max(d_sub_prod$age), .1), 
+                            log_rt_0 = c(quantile(d_sub_prod$log_rt, 
+                                                  c(.1,.25,.5,.75,.9), 
+                                                  na.rm=TRUE)))
+global_preds$pred <- predict(mod3, newdata = global_preds, level = 0)
+
+# random effects
+# dataset_preds <- d_sub |>
+#   filter(!is.na(prod)) |> # filter on predictor
+#   group_by(dataset_name) |>
+#   summarise(min_age = min(age), 
+#             max_age = max(age)) |>
+#   group_by(dataset_name) |>
+#   mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1)))) |>
+#   select(-min_age, -max_age) |>
+#   unnest(col = "df") |>
+#   ungroup()
+#   
+# dataset_preds$pred <- predict(mod3, 
+#                               newdata = dataset_preds, 
+#                               level = c(0,1))$predict.dataset_name
+
+ggplot(d_sub_prod, aes(x = age, y = prod)) + 
+  geom_point(alpha = .1) + 
+  geom_line(data = global_preds, aes(y = pred, group = log_rt_0, col = as.factor(log_rt_0)))
+# +
+#   geom_line(data = dataset_preds, aes(y = pred))
+```
+
+## Interaction with accuracy
+
+```{r}
+mod4 <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
+             data = d_sub_prod,
+             fixed = xmid + scale ~ log_rt_0 + acc_0,
+             random = xmid + scale ~ 1,
+             groups = ~ dataset_name,
+             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 0, `scale:log_rt_0` = 3, 
+                       `xmid:acc_0` = 0, `scale:acc_0` = 3),
+             na.action = na.exclude, 
+             control = list(maxIter = 1000, tolerance = 10))
+
+summary(mod4)
+```
+## power function with offset
+
+So I guess we need good parameters.
+
+```{r}
+power_growth <- function(X, a, b, c)
+{
+  return((a * (X-c)^b))
+}
+
+d_sub_prod$power = power_growth(d_sub_prod$age, a = 1, b = 2.3, c = 12)
+
+ggplot(d_sub_prod, aes(x = age, y = prod*680)) + 
+  geom_point(alpha = 1.) + 
+  geom_line(aes(x = age, y = power))
+
+```
+
+Now model.
+
+```{r}
+d_sub_prod$prod_680 <- d_sub_prod$prod * 680
+mod5 <- nlme(model = prod_680 ~ power_growth(age, a, b, c),
+             data = d_sub_prod,
+             fixed = a + b + c ~ 1,
+             random = a + b + c ~ 1,
+             groups = ~ dataset_name,
+             start = c(a = 1, b = 2.3, c = 12),
+             na.action = na.exclude)
+
+summary(mod5)
+
+# fixed effects (no CIs)
+global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
+                                      max(d_sub_prod$age), .1))
+global_preds$pred <- predict(mod5, newdata = global_preds, level = 0)
+
+# random effects
+dataset_preds <- d_sub |>
+  filter(!is.na(prod)) |> # filter on predictor
+  group_by(dataset_name) |>
+  summarise(min_age = min(age), 
+            max_age = max(age)) |>
+  group_by(dataset_name) |>
+  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1)))) |>
+  select(-min_age, -max_age) |>
+  unnest(col = "df") |>
+  ungroup()
+
+dataset_preds$pred <- predict(mod5, 
+                              newdata = dataset_preds, 
+                              level = c(0,1))$predict.dataset_name
+
+ggplot(d_sub_prod, aes(x = age, y = prod_680, col = dataset_name)) + 
+  geom_point(alpha = .1) + 
+  geom_line(data = global_preds, aes(y = pred), col = "blue", size = 1.5) +
+  geom_line(data = dataset_preds, aes(y = pred))
+```
+
+Having trouble fitting this model - throws an error sadly.
+
+```{r eval=FALSE}
+d_sub_prod$prod_681 <- d_sub_prod$prod_680 + 1
+mod5 <- nlme(model = prod_681 ~ power_growth(age, a, b, c),
+             data = d_sub_prod,
+             fixed = a + b + c ~ log_rt_0,
+             random = a + b + c ~ 1,
+             groups = ~ dataset_name,
+             start = c(a = 1, b = 2.3, c = 12, 
+                       `a:log_rt_0` = 0.1, `b:log_rt_0` = 0.1, `c:log_rt_0` = 0.1),
+             na.action = na.exclude)
+
+summary(mod5)
+```

### 2024-12-13 paper/paper.Rmd
paper and longitudinal updates

@@
 Combining 24 datasets with data from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development, both within and across studies.
-Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy to parent-reported vocabulary as well as a longitudinal relationship such that children with faster language processing tend to show larger vocabulary growth.
-Together, these findings support the view that language processing is –– among other things –– a skill, and that this skill plays a role in early language learning.
+Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy to parent-reported vocabulary.
+Across a range of models, speed, accuracy, and vocabulary also show coupled growth such that children with faster initial language processing tend to show larger vocabulary growth.
+Together, these findings support the view that language processing is a skill, and that this skill plays a role in early language learning.
 
@@
 
-Word recognition ability is hypothesized to play a key role in language learning.
+Word recognition ability, as measured by LWL, is hypothesized to play a key role in language learning.
 Each word that a child experiences is an opportunity to learn; measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking].
-These learning opportunities are likely helpful to different degrees, however [@kachergis2022toward;@mcmurray2007defusing].
+<!-- These learning opportunities are likely helpful to different degrees, however [@kachergis2022toward;@mcmurray2007defusing].  -->
 <!-- Language input is more helpful for learning if it focused on the child's own interest and grounded in the particulars of their current environment -- for example, in a game, routine, or object -- and if it uses vocabulary and syntax that they know [@tomasello2001perceiving;@roy2015predicting;@cartmill2013quality].
@@
 Here we investigate two hypotheses.
+
 First, one influential recent hypothesis characterizes language learning more broadly as a process of learning the skill of real-time processing  [@christiansen2016;@chater2018].
@@
 Table \ref{tab:datasets} shows the characteristics of individual datasets (see also Figures S1 and S2).
-The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed exploration of the development of word recognition than has previously been possible.
+The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed  study of the development of word recognition than has previously been possible.
+While our analysis is exploratory in nature, it is guided by the two hypotheses outlined above: the presence of signatures of skill learning in word recognition, and 2) linkages between word recognition and vocabulary.
 
@@
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms).
-Our first question was about the functional form of the relationships between age, speed, and accuracy (see Table S-XYZ for raw correlations between variables).
+Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI for raw correlations between variables).
 To investigate this question, we fit linear mixed effects models predicting accuracy and reaction time on each trial across the full dataset with random slopes of age by study and random intercepts by participant.
@@
 The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text.
-All findings reported here hold for both window sizes, however (see SI: Short Window Analyses).] and the best fitting model of speed predicted log reaction time as a function of log age as well (See Tables S1 and S2 for model fit statistics).
+All findings reported here hold for both window sizes, however (see SI).] and the best fitting model of speed predicted log reaction time as a function of log age as well (see SI).
 
@@
 ```{r reliability}
-MONTH_CUTOFF <- 2
+MONTH_CUTOFF <- 3
 
@@
   mutate(admin_num = 1:n(), 
-         delta_t = age - age[1])
+         time_since_t0 = age - age[1],
+         delta_t = c(0, diff(age)))
 
 d_reliability <- d_long |>
-  filter(delta_t <= MONTH_CUTOFF, 
+  filter(time_since_t0 <= MONTH_CUTOFF, 
          admin_num <= 2) |>
@@
             acc = cor(long_window_accuracy_1, 
-                          long_window_accuracy_2, use = "pairwise.complete.obs")) |>
-  filter(dataset_name != "fernald_totlot")
+                          long_window_accuracy_2, use = "pairwise.complete.obs"))
 
@@
 While we fit the latent variable models above to the full cross-sectional dataset, we are most interested in within-child developmental changes, as measured within individual studies.
-We first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than two months apart.
+We first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
 Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$), suggesting that, across studies, LWL measures from individual sessions provide relatively noisy measurements of individual children even if individual studies are able to achieve higher reliability by averaging across measurement occasions [e.g., @marchman2008].
+
+```{r}
+d_growth <- d_sub |>
+  filter(!is.na(prod), subject_id %in% longitudinal$subject_id) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(rt_t0 = log_rt[1]) |>
+  ungroup() |>
+  mutate(rt_t0_group = ifelse(rt_t0 < median(rt_t0, na.rm=TRUE), "low", "high"))
+
+d_growth$age_15 <- d_growth$age - 15
+growth_mod <- lmer(prod ~ poly(age_15,2) *rt_t0 + (age  | subject_id) + (age | dataset_name), 
+     data = d_growth, 
+     control = lmerControl(optimizer = "bobyqa"))
+
+# summary(growth_mod)
+```
+
+Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure relationships between LWL and vocabulary using longitudinal growth models.
+We began by reproducing the analysis reported in @fernald2012, in which longitudinal growth in productive vocabulary was predicted based on reaction time at study initiation.
+We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, reaction time at initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset).
+This model revealed a significant effect of $t_0$ reaction time (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ reaction time and the quadratic age predictor (`r papaja::apa_print(acc_var_mod)$full_result$polyage_15_22_rt_t0`).
+This analysis suggests that children with faster reaction times showed both larger vocabularies and faster vocabulary growth over time, confirming that findings from one of the studies in the sample were robust across the full dataset.
 
@@
 
-d_sub_wide <- d_sub_s |> 
-  mutate(t = cut(time_since_t0, breaks = c(0, 6, 12, 30), 
+d_sub_wide_fine <- d_sub_s |>
+  mutate(t = cut(time_since_t0, breaks = c(0, 2, 4, 6, 8, 10, 30), 
                  include.lowest = TRUE)) |>
   filter(t != "(12,30]") |>
-  mutate(t = as.numeric(t)) |>
+  mutate(t = as.numeric(t) -1) |>
   group_by(subject_id, dataset_name, t) |>
@@
 
+slopes_model_long <- "
+
+# regressions
+accuracy_intercept =~ 1*acc_t0 + 1*acc_t1 + 1*acc_t2 + 1*acc_t3 + 1*acc_t4
+accuracy_slope =~ 1*acc_t0 + 2*acc_t1 + 3*acc_t2 + 4*acc_t3 + 5*acc_t4
+speed_intercept =~ 1*log_rt_t0 + 1*log_rt_t1 + 1*log_rt_t2 + 1*log_rt_t3 + 1*log_rt_t4
+speed_slope =~ 1*log_rt_t0 + 2*log_rt_t1 + 3*log_rt_t2 + 4*log_rt_t3 + 5*log_rt_t4
+vocab_intercept =~ 1*prod_t0 + 1*prod_t1 + 1*prod_t2 + 1*prod_t3 + 1*prod_t4
+vocab_slope =~ 1*prod_t0 + 2*prod_t1 + 3*prod_t2 + 4*prod_t3 + 5*prod_t4
+# let variances of each latent vary
+accuracy_intercept ~~ NA*accuracy_intercept
+accuracy_slope ~~ NA*accuracy_slope
+speed_intercept ~~ NA*speed_intercept
+speed_slope ~~ NA*speed_slope
+vocab_intercept ~~ NA*vocab_intercept
+vocab_slope ~~ NA*vocab_slope
+"
+
+slopes_long <- growth(slopes_model_long, d_sub_wide_fine, std.lv=TRUE, missing='fiml')
+
+slopes_long_summary <- suppressMessages(summary(slopes_long, fit.measures=TRUE, standardize=TRUE))
 ```
 
-```{r}
-fa3_model_long <- "
-# measurement model
-vocab_t1 =~ 1*prod_t1 + s1*comp_t1
-accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
-speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1
+On the other hand, it is possible that differences in predicted growth trajectories is due to overall coupling between vocabulary size and language processing, rather than a specifically predictive relationship between $t_0$ reaction time and vocabulary growth.
+To test this relationship, we used longitudinal structural equation models.
+We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months ($t_0 ...
+t_4$) and fit individual growth across each of these variables.
+We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset.
+(We present a model of coupled growth in the observed variables here for simplicity; see SI for a comparable model using growth in the latent factors).
+The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}.
+Overall fit statistics were generally acceptable (Confirmatory fit index: `r round(slopes_long_summary$fit["cfi"], 3)`, RMSE: `r round(slopes_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(slopes_long_summary$fit["rmsea.pvalue"], 3)`).
 
-vocab_t2 =~ 1*prod_t2 
-accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
-speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2
+Our key question of interest concerned coupling between the parameters of these growth models.
+Consistent with the idea that overall faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
+Notably, there was no coupling between growth in reaction time and growth in vocabulary, suggesting that this relationship was stable across development ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
+These findings are further evidence consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
 
-# longitudinal relationships
-vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
-speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
-accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
-
-# factor covariances
-vocab_t1 ~~ accuracy_t1
-vocab_t1 ~~ speed_t1
-accuracy_t1 ~~ speed_t1
-
-vocab_t2 ~~ accuracy_t2
-vocab_t2 ~~ speed_t2
-accuracy_t2 ~~ speed_t2
-
-# means for the latents
-vocab_t1 ~ 1
-vocab_t2 ~ 1
-accuracy_t1 ~ 1
-accuracy_t2 ~ 1
-speed_t1 ~ 1
-speed_t2 ~ 1
-
-# residual variance for the latents
-vocab_t1 ~~ NA*vocab_t1
-vocab_t2 ~~ NA*vocab_t2
-accuracy_t1 ~~ NA*accuracy_t1
-accuracy_t2 ~~ NA*accuracy_t2
-speed_t1 ~~ NA*speed_t1
-speed_t2 ~~ NA*speed_t2
-"
-
-fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
-
-fit3_long_summary <- suppressMessages(summary(fit3_long, fit.measures=TRUE, standardize=TRUE))
+```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between growth parameters. "}
+layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
+graph_sem(model = slopes_long, text_size = 2, layout=t(layout))
 ```
-
-Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure longitudinal relationships between LWL and vocabulary via a longitudinal latent factor model.
-We fit the same three-factor model, but this time averaging longitudinal administrations after splitting them into two categories: those datapoints within a two-month time period (which we designate as $t_1$) and those datapoints between three and six months from the initial measurement ($t_2$) (see SI: Sensitivity Analysis for Longitudinal Models).
-To ensure measurement invariance, we coupled the factor loadings across the two time-points.
-Next, to investigate coupling between the different measures, we fit regressions linking the two time points.
-
-The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}.
-Overall fit statistics were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit3_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit3_long_summary$fit["rmsea.pvalue"], 3)`).
-Each latent factor was significantly coupled across time, again indicating moderate longitudinal stability of these abilities.
-
-```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between latent factors.
-Variables marked t1 are measures from two or fewer months from study initiation; variables marked t2 are measures from 3--6 months after study initiation.
-Factor loadings are constrained to be equivalent between t1 and t2.
-Plotting conventions are as in Figure 3."}
-layout_long = matrix(nrow=7, ncol = 7,
-                     data = c("log_rt_sd_t1","log_rt_t1", "acc_t1", "acc_sd_t1", "prod_t1","comp_t1", NA,
-                               NA, NA, NA, NA, NA, NA,  NA,
-                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
-                              NA, NA, NA, NA, NA, NA,  NA,
-                               NA,"speed_t2",NA,"accuracy_t2",NA,NA,"vocab_t2",
-                               NA, NA, NA, NA, NA, NA,  NA,
-                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", NA, "prod_t2"), 
-                     byrow = TRUE)
-graph_sem(model = fit3_long, text_size = 2, layout = t(layout_long))
-```
-
-```{r}
-fa3_model_long_control <- "
-# measurement model
-vocab_t1 =~ 1*prod_t1 + s1*comp_t1
-accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
-speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1
-
-vocab_t2 =~ 1*prod_t2 + s1*comp_t2
-accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
-speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2
-
-# longitudinal relationships
-vocab_t2 ~ vocab_t1 + 0*speed_t1 + accuracy_t1
-speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
-accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
-
-# factor covariances
-vocab_t1 ~~ accuracy_t1
-vocab_t1 ~~ speed_t1
-accuracy_t1 ~~ speed_t1
-
-vocab_t2 ~~ accuracy_t2
-vocab_t2 ~~ speed_t2
-accuracy_t2 ~~ speed_t2
-
-# means for the latents
-vocab_t1 ~ 1
-vocab_t2 ~ 1
-accuracy_t1 ~ 1
-accuracy_t2 ~ 1
-speed_t1 ~ 1
-speed_t2 ~ 1
-
-# residual variance for the latents
-vocab_t1 ~~ NA*vocab_t1
-vocab_t2 ~~ NA*vocab_t2
-accuracy_t1 ~~ NA*accuracy_t1
-accuracy_t2 ~~ NA*accuracy_t2
-speed_t1 ~~ NA*speed_t1
-speed_t2 ~~ NA*speed_t2
-"
-
-fit3_long_control <- sem(fa3_model_long_control, d_sub_wide, std.lv=TRUE, missing='fiml')
-speed_vocab_anova <- anova(fit3_long, fit3_long_control)
-speed_vocab_anova$p = speed_vocab_anova$`Pr(>Chisq)`
-speed_vocab_anova$c2d = speed_vocab_anova$`Chisq diff`
-```
-
-Critically, LWL processing speed at time 1 predicted vocabulary at time 2 ($\beta=$ `r round(fit3_long_summary$pe[14,"est"],3)`, SE = `r round(fit3_long_summary$pe[21,"se"],3)`, $p =$ `r round(fit3_long_summary$pe[14,"pvalue"],3)`).
-This model significantly improved fit compared with a model that set this coupling parameter to zero but was otherwise unchanged ($\chi^2 = `r round(speed_vocab_anova$c2d[2], 2)`$, $p =  `r round(speed_vocab_anova$p[2], 4)`$).
-These findings are consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
 
@@
 
+NOTES FOR FUTURE DISCUSSION
+
 By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had samples sizes at least an order of magnitude smaller than ours.
@@
 
-FACTOR PARAGRAPH
+<!-- FACTOR PARAGRAPH -->
 

### 2024-12-13 helper/common.R

@@ -1,7 +1,7 @@
 # source me for all analyses 
 
 library(tidyverse)
-library(lme4)
+# library(lme4)
 library(knitr) # for kable
 # library(ggrepel)
 library(ggthemes)

### 2025-01-08 paper/paper.Rmd
finished draft

@@
 author:
-  - name: Alice Anonymous
-    affiliation: a,1,2
-  - name: Bob Security
-    affiliation: a,b
+  - name: Michael C. Frank
+    affiliation: a
+  - name: Other peekbankers to be added
+    affiliation: b
+  - name: Martin Zettersten
+    affiliation: c
 address:
   - code: a
-    address: Some Institute of Technology, Department, Street, City, State, Zip
+    address: Department of Psychology, Stanford University, Stanford, CA, USA
   - code: b
-    address: Another University Department, Street, City, State, Zip
+    address: Many other peekbankers affiliations
+  - code: c
+    address: Department of Cognitive Science, University of California, San Diego, CA, USA
 
@@
 Here we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
-Understanding language is a skill that improves gradually with practice.
+These data suggest that understanding language is a skill that improves gradually with practice.
 
@@
 Reaction time is computed only on trials in which the child is fixating the distractor image (the book); in these cases, the average time it takes to shift fixation to the target is used as an index of processing speed.
-Early work using this method showed that children's speed and accuracy both increase rapidly across the second year [@fernald1998;@reznick1990visual], and related methods have provided a window into how children process phonological [@mani2011], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] competition as well as how their lexical representations develop [@Swingley2002].
+Early work using this method showed that children's speed and accuracy both increase rapidly across the second year [@fernald1998;@reznick1990visual].
+Related methods have provided a window into how children process phonological [@mani2011], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] competition as well as how their lexical representations develop [@Swingley2002].
 
@@
 
-<!-- % MCF: is this question statement sharp enough?  -->
-
 Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
-These samples provide neither the breadth of ages nor the number of participants to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of language [as has been done with school-aged children: @mcmurray2022slow;@colby2023efficiency].
+These samples provide neither the breadth of ages nor the number of participants to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of language [as has been done with school-aged children; @mcmurray2022slow;@colby2023efficiency].
 Here we investigate two hypotheses.
 
-First, one influential recent hypothesis characterizes language learning more broadly as a process of learning the skill of real-time processing  [@christiansen2016;@chater2018].
-On this hypothesis, we should expect to see the signatures of expertise and skill learning in word recognition.
-We expect that accuracy will change linearly with the logarithm of age, reflecting gradual asymptotic convergence to mature levels of accuracy.
-In addition, the logarithm of processing speed should also decrease with the logarithm of age, potentially reflecting the  "power law of practice" [@snoddy1926;@kail1991;@anderson1982; cf @heathcote2000].
-Further, trial-to-trial variability in both speed and accuracy should decrease with increasing expertise [@todorov2002optimal].
+First, one influential recent theory characterizes language learning more broadly as a process of learning the skill of real-time processing  [@christiansen2016;@chater2018].
+Based on this theory, we should expect to see the signatures of expertise and skill learning in word recognition.
+Accuracy should change linearly with the logarithm of age, reflecting gradual asymptotic convergence to mature levels of accuracy.
+In addition, the logarithm of processing speed should decrease with the logarithm of age, potentially reflecting the  "power law of practice" [@snoddy1926;@kail1991;@anderson1982; cf @heathcote2000].
+Finally, trial-to-trial variability in both speed and accuracy should decrease with increasing expertise [@todorov2002optimal].
 
@@
 
-To address these questions, we made use of a new release of data in Peekbank, an open database of looking while listening data from young children, stored in a harmonized format [@zettersten2023peekbank].
+# Results {.unnumbered}
+
+To address these questions, we made use of a new release of data in Peekbank, an open database of LWL from young children, stored in a harmonized format [@zettersten2023peekbank].
 We retrieved data from children ages 1--6 years.
-Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a word; these trials often constituted control conditions for experiments.
+Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a word (typically embedded in a simple carrier phrase such as "look at the ..."); these trials often constituted control conditions for experiments.
 We focus here on English purely for practical reasons -- the Peekbank dataset contains limited data from other languages.
@@
 ```
-
-# Results {.unnumbered}
 
@@
 
-We were first interested in developmental changes in speed and accuracy.
+We began by characterizing developmental changes in speed and accuracy.
 We computed both reaction times and accuracies following standard practices in the literature [@fernald2008].
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms).
+
 Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI for raw correlations between variables).
@@
 We fit a series of factor analytic models to explore the dimensionality of the data.
-Initial exploratory factor analysis using paralell analysis suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
+Initial exploratory factor analysis using parallel analysis suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
 Due to missingness of data, we used confirmatory factory analysis with full information maximum likelihood to find the best set of loadings.
@@
 
-While we fit the latent variable models above to the full cross-sectional dataset, we are most interested in within-child developmental changes, as measured within individual studies.
-We first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
+While we fit the latent variable models above to the full cross-sectional dataset, we were most interested in within-child developmental changes, as measured within individual studies.
+To probe these changes, we first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
 Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$), suggesting that, across studies, LWL measures from individual sessions provide relatively noisy measurements of individual children even if individual studies are able to achieve higher reliability by averaging across measurement occasions [e.g., @marchman2008].
@@
 This model revealed a significant effect of $t_0$ reaction time (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ reaction time and the quadratic age predictor (`r papaja::apa_print(acc_var_mod)$full_result$polyage_15_22_rt_t0`).
-This analysis suggests that children with faster reaction times showed both larger vocabularies and faster vocabulary growth over time, confirming that findings from one of the studies in the sample were robust across the full dataset.
+This analysis suggests that children with faster reaction times show both larger vocabularies and faster vocabulary growth over time, confirming that findings from @fernald2012 were robust across the full dataset.
 
@@
 Consistent with the idea that overall faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
-Notably, there was no coupling between growth in reaction time and growth in vocabulary, suggesting that this relationship was stable across development ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
-These findings are further evidence consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
+There was not significant coupling between growth in reaction time and growth in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
+These abilities might grow independently, but we cannot rule out other possibilities.
+First, the longitudinal data we had might not allow sufficiently precise estimates of growth slopes, or second, since vocabulary growth is non-linear, the linear model we used here might not be as sensitive to non-linear changes.
+
+In sum, these findings provide further evidence consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
+Children with greater skill in word recognition learn words faster.
 
@@
 We found continuous developmental changes from ages 1 -- 6 years.
-Speed and accuracy both improved asymptotically, with evidence that recognition speed showed the log-log relationship associated with the power law of practice, that is, gradually converging on mature levels of processing efficiency.
-Further, older children showed less variability in accuracy and reaction time across trials, indicating more robust execution of this skilled behavior.
+Speed and accuracy both improved asymptotically, with evidence that recognition speed showed the log-log relationship associated with the "power law of practice", that is, gradually converging on mature levels of processing efficiency.
+Further, trial-to-trial variability decreased, consistent with both the literature on skill learning [@todorov2002optimal] and other work on developmental changes in variability [@judd2021working;@galeano2018variability;@kautto2023introducing].
 Speed and accuracy were both related to vocabulary size concurrently, and processing speed was also related to later vocabulary growth.
 
-NOTES FOR FUTURE DISCUSSION
+Together, our findings are consistent with theories that posit that language learning is a process of skill acquisition, in which children become adept at quickly converting ephemeral signals into meaning [@christiansen2016].
+This skill develops gradually over the course of early childhood and supports word learning.
+Further, our results point to consistency between skill development in early childhood and the continued refinement of language processing and language knowledge during middle childhood [@mcmurray2022slow;@colby2023efficiency].
 
 By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had samples sizes at least an order of magnitude smaller than ours.
-The presence of multiple studies with longitudinal data,
+In contrast to individual studies, which typically only have statistical power to test specific contrasts, our "big data" approach provided the sample sizes necessary to fit complex structural equation models and to compare different functional forms for developmental change.
+Because early language is so variable [@frank2021], these kinds of samples -- thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning.
 
-These findings are consistent with theories that posit that language learning is a process of skill acquisition, in which children become skillful at quickly converting ephemeral signals into meaning [@christiansen2016].
+At the same time, our approach is both observational and exploratory.
+Thus, we cannot untangle a variety of different causal models for the variation we observed.
+Early word recognition skill could lead to faster word learning, due to skilled children gleaning more information from the same signal [@fernald2006].
+But faster children could also be faster due to their larger vocabulary and stronger lexical representations.
+These two causal directions could also interact reciprocally, leading to a "rich get richer process" in which children with larger vocabularies process faster, and their faster processing helps them increase their vocabulary size more rapidly.
+Or a third shared factor -- perhaps general cognitive ability -- could underpin both processes.
+Our cross-sectional data cannot distinguish these hypotheses even in principle [@vanderweele2023dimensional], and our longitudinal data are likely too sparse to distinguish such complex causal models.
+Thus, these questions remain an open target for dense longitudinal data collection.
 
-<!-- FACTOR PARAGRAPH -->
+Our findings here are limited in their generalizability by the convenience samples that were used in the individual studies in the Peekbank database.
+These studies typically (but not always) represent children from well-educated parents living in university-adjacent communities.
+We would not expect that specific numerical parameters estimated in this aggregate convenience sample would generalize to other samples.
+Nevertheless, the consistency of the trends we observed across datasets suggest that our qualitative conclusions are robust to some significant sociodemographic variation.
 
-These data suggest graded, continual changes in word recognition starting in early childhood and continuing at least through elementary school.
-Notably, the experiments included in our datasets used simple words that children of this age almost surely "know" (e.g., "ball," which is learned very early, or occasionally "scissors" or "zebra," which are learned slightly later).
-Even for these "easy" words, careful measurement reveals increasing efficiency.
-
-<!-- % address causal indeterminacy issue  -->
-<!-- % https://www.degruyter.com/document/doi/10.1515/jci-2022-0074/html -->
-
-<!-- DISCUSS G EXPLANATION - fernald fundamentally speedier processor -->
-
-<!-- INPUT VS MATURATION? PROBABLY BOTH -->
-
-Consistent with this idea,  trial-to-trial variability in children's cognitive task performance decreases with age and has some predictive relationships with skill in other tasks, including language [@judd2021working;@galeano2018variability;@kautto2023introducing].
-
-What explains variability in language performance? Integrating information from parent-report measures of vocabulary development with infant behavioral data, we show that infants' word recognition.
+More broadly, our results here suggest the continued importance of the looking-while-listening paradigm as an index of children's language processing abilities.
+If language learning is, at least in part, a process of skill learning, then measurement of this skill is a critical window into understanding the remarkable process of language learning.
 

### 2025-02-04 paper/paper.Rmd
Veronica messing with graphs

@@
 
+#library(RColorBrewer)
+
+  mycolors = c("#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E","#1B9E77", "#D95F02")
+  # 4 colors from brewer scale "Dark2" that I thought had decent contrast against the grey/black background dots
+
 a <- ggplot(d_sub,
        aes(x = age, y = long_window_accuracy)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth(method = "lm") +
+  geom_point(alpha = .02) + 
   geom_smooth(data = d_sub_agevar,
               aes(group = dataset_name, col = dataset_name), se=FALSE,
-              alpha = .1, method = "lm", size = .5, lty = 2) + 
+              alpha = 1, method = "lm", size = .5, lty = 2) + 
+    geom_smooth(method = "lm") +
   geom_hline(yintercept = .5, lty = 2) + 
   scale_x_log10() +
-  scale_color_discrete(guide = FALSE) + 
+  scale_color_manual(values=mycolors, guide = FALSE) + 
   ylab("Accuracy") + 
@@
        aes(x = age, y = rt)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth(method = "lm") +
+  geom_point(alpha = .02) + 
   geom_smooth(data = d_sub_agevar,
@@
               alpha = .1, method = "lm", size = .5, lty = 2) + 
+    geom_smooth(method = "lm") +
   scale_x_log10() +
   scale_y_log10() +
-  scale_color_discrete(guide = FALSE) + 
+  scale_color_manual(values=mycolors, guide = FALSE) + 
   theme(legend.position = "bottom") + 
@@
 ```{r variance, fig.cap="Participant-level variability in accuracy and reaction time (log RT), plotted by age (log). Plotting conventions are as in Figure 1.", fig.env="figure*", fig.height = 2.5}
+  mycolors = c("#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E","#1B9E77", "#D95F02")
+
 c <- ggplot(d_sub,
        aes(x = age, y = long_window_acc_var)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth(method = "lm") +
+  geom_point(alpha = .02) + 
   geom_smooth(data = d_sub_agevar,
@@
               alpha = .1, method = "lm", size = .5, lty = 2) + 
+    geom_smooth(method = "lm") +
   scale_x_log10() +
-  scale_color_discrete(guide = FALSE) + 
+  scale_color_manual(values=mycolors, guide = FALSE) + 
   ylab("SD of Accuracy)") + 
@@
        aes(x = age, y = log_rt_var)) + 
-  geom_point(alpha = .05) + 
-  geom_smooth(method = "lm") +
+  geom_point(alpha = .02) + 
   geom_smooth(data = d_sub_agevar,
@@
               alpha = .1, method = "lm", size = .5, lty = 2) + 
+    geom_smooth(method = "lm") +
   scale_x_log10() +
-  scale_color_discrete(guide = FALSE) + 
+  scale_color_manual(values=mycolors, guide = FALSE) + 
   theme(legend.position = "bottom") + 
@@
 graph_sem(model = fit3_age, text_size = 2, layout = t(layout_age))
+
+nodes <- get_nodes(fit3_age)
+
+edges <- get_edges(fit3_age) |> filter(label!="1.00") |>
+
+graph_sem(edges=edges, nodes=nodes, text_size = 2, layout = t(layout_age),
+          rect_width=2,
+          rect_height=1.2,
+          ellipses_width=1.5,
+          ellipses_height=1.5,
+          variance_diameter=.9,
+          spacing_x=2
+)
 ```
@@
 ```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between growth parameters. "}
-layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
-graph_sem(model = slopes_long, text_size = 2, layout=t(layout))
+#layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
+#graph_sem(model = slopes_long, text_size = 2, layout=t(layout))
+graph_sem(model = slopes_long, text_size = 2)
+
 ```

### 2025-02-04 paper/paper.Rmd
messing around with plotting

@@
 source(here("helper","common.R"))
+
+source(here("helper/modified_sem_plot.R"))
 
@@
 
-graph_sem(model = fit3_age, text_size = 2, layout = t(layout_age))
+node_text <- tibble(name=c("acc", "acc_sd", "accuracy", "comp", "log_age", "log_rt", "log_rt_sd", "prod", "speed", "vocab"), pretty=c("Accuracy", "Accuracy (sd)", "Accuracy", "Comprehension", "Age (log months)", "log RT", "log RT (SD)", "Production", "Speed", "Vocabulary"))
 
-nodes <- get_nodes(fit3_age)
+nodes <- get_nodes(fit3_age) |> tibble() |> left_join(node_text) |> mutate(label=ifelse(shape=="oval",pretty, str_c(pretty, "\n", est_sig)))
 
-edges <- get_edges(fit3_age) |> filter(label!="1.00") |>
+edges <- get_edges(fit3_age) |> filter(label!="1.00") |> mutate(label_location=.5)
 
-graph_sem(edges=edges, nodes=nodes, text_size = 2, layout = t(layout_age),
-          rect_width=2,
-          rect_height=1.2,
-          ellipses_width=1.5,
-          ellipses_height=1.5,
-          variance_diameter=.9,
-          spacing_x=2
-)
+edges[17,"curvature"]=50
+edges[17,"label_location"]=.6
+
+source(here("helper/modified_sem_plot.R"))
+
+#View(edges)
+prepare_graph(edges=edges, nodes=nodes, text_size = 2.5, layout = t(layout_age),
+          rect_width=4.2,
+          rect_height=3,
+          ellipses_width=3,
+          ellipses_height=3,
+          variance_diameter=2,
+          spacing_y=4,
+          spacing_x=4,
+          fix_coord=FALSE) |> plot_happy_arrows()
 ```
@@
 ```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between growth parameters. "}
-#layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
-#graph_sem(model = slopes_long, text_size = 2, layout=t(layout))
-graph_sem(model = slopes_long, text_size = 2)
+layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
 
+node_text <- tribble(~name, ~ pretty,
+                     "acc_t0", "t0 Accuracy",
+                     "acc_t1", "t1 Accuracy",
+                     "acc_t2", "t2 Accuracy",
+                     "acc_t3", "t3 Accuracy",
+                     "acc_t4", "t4 Accuracy",
+                     "accuracy_intercept", "Accuracy intercept",
+                     "accuracy_slope", "Accuracy slope",
+                     "log_rt_t0", "t0 log RT",
+                     "log_rt_t1", "t1 log RT",
+                     "log_rt_t2", "t2 log RT",
+                     "log_rt_t3", "t3 log RT",
+                     "log_rt_t4", "t4 log RT",
+                     "prod_t0" , "t0 Production",
+                     "prod_t1" , "t1 Production",
+                     "prod_t2" , "t2 Production" ,
+                     "prod_t3", "t3 Production",         
+                     "prod_t4", "t4 Production" ,
+                     "speed_intercept" , "Speed intercept" ,  
+                     "speed_slope" , "Speed slope",
+                     "vocab_intercept", "Vocabulary intercept",
+                     "vocab_slope", "Vocabulary slope" )
+
+nodes <- get_nodes(slopes_long) |> tibble() |> left_join(node_text) |> mutate(label=pretty)
+
+edges <- get_edges(slopes_long) |> mutate(label_location=.5) |> mutate(label=ifelse(arrow=="last", NA, label)) |>  mutate(connect_from=ifelse(to==from, "right",NA),
+                    connect_to=connect_from) |> 
+  mutate(color="black")
+
+edges[52,"curvature"]=85
+edges[52,"color"]="blue"
+edges[61,"curvature"]=85
+edges[61,"color"]="blue"
+edges[66,"curvature"]=85
+edges[66,"color"]="blue"
+
+edges[54,"curvature"]=85
+edges[63,"curvature"]=85
+
+edges[53,"curvature"]=80
+edges[53,"color"]="red"
+edges[55,"curvature"]=82
+edges[55,"color"]="red"
+edges[62,"curvature"]=80
+edges[62,"color"]="red"
+
+edges[58,"curvature"]=80
+edges[58,"color"]="green"
+edges[60,"curvature"]=82
+edges[60,"color"]="green"
+edges[65,"curvature"]=80
+edges[65,"color"]="green"
+
+edges[59,"curvature"]=77
+
+edges[64,"curvature"]=75
+edges[56,"curvature"]=75
+edges[57,"curvature"]=75
+
+source(here("helper/modified_sem_plot.R"))
+#View(edges)
+prepare_graph(edges=edges, nodes=nodes, text_size = 2, layout = t(layout),angle=0,
+          rect_width=1.5,
+          rect_height=2,
+          ellipses_width=1.5,
+          ellipses_height=4,
+          variance_diameter=1,
+          spacing_y=3,
+          spacing_x=2,
+          fix_coord=FALSE) |> 
+  edit_edges({connect_from=case_when(
+    (to==from & from %in% c("accuracy_intercept","accuracy_slope","speed_slope",
+                          "speed_intercept", "vocab_slope", "vocab_intercept"))~"bottom",
+    to==from ~ "right",
+    !is.na(curvature)~"left",
+    T ~ "right")}) |> 
+  plot_happy_arrows()
 ```

### 2025-02-04 helper/modified_sem_plot.R

@@ -1,0 +1,1140 @@
+
+plot_happy_arrows <- function(x, y, ...){
+  df_nodes <- x$nodes
+  df_edges <- x$edges
+  if("show" %in% names(df_nodes)){
+    df_nodes <- df_nodes[df_nodes$show, , drop = FALSE]
+    if(!all(df_edges$from %in% df_nodes$name & df_edges$to %in% df_nodes$name)){
+      message("Some edges involve nodes with argument 'show = FALSE'. These were dropped.")
+      df_edges <- df_edges[df_edges$from %in% df_nodes$name & df_edges$to %in% df_nodes$name, , drop = FALSE]
+    }
+  }
+
+  if("show" %in% names(df_edges)){
+    df_edges <- df_edges[df_edges$show, , drop = FALSE]
+  }
+  if(nrow(df_nodes) == 0){
+    stop("No nodes left to plot.")
+  }
+  rect_width <- x$rect_width
+  rect_height <- x$rect_height
+  ellipses_width <- x$ellipses_width
+  ellipses_height <- x$ellipses_height
+  variance_diameter <- x$variance_diameter
+  spacing_x <- x$spacing_x
+  spacing_y <- x$spacing_y
+  text_size <- x$text_size
+  fix_coord <- x$fix_coord
+
+  # Check dfs ---------------------------------------------------------------
+  numeric_cols <- c("curvature")
+  df_edges[numeric_cols[which(numeric_cols %in% names(df_edges))]] <- lapply(df_edges[numeric_cols[which(numeric_cols %in% names(df_edges))]], as.numeric)
+
+  if(nrow(df_edges)){
+    connect_points <- .connect_points(df_nodes, df_edges)
+
+    df_edges <- cbind(df_edges, connect_points)
+
+    df_edges <- cbind(df_edges, setNames(data.frame(t(apply(connect_points, 1, function(x){(x[1:2]+x[3:4])/2}))), c("text_x", "text_y")))
+  }
+
+
+  # Make plot ---------------------------------------------------------------
+
+  p <- ggplot(NULL)
+
+  if(any(df_edges$from == df_edges$to)){
+    p <- .plot_variances(p, df = df_edges[df_edges$from == df_edges$to, ], text_size = text_size, diameter = variance_diameter)
+  }
+  if(any(!df_edges$from == df_edges$to)){
+    p <- .plot_edges(p, df_edges[!df_edges$from == df_edges$to, ], text_size)
+  }
+
[…]

### 2025-02-05 paper/paper.Rmd
more plotting tweaks -- growth curves and arrows

--- 2025-02-04 paper/paper.Rmd
@@
 
+source(here("helper/modified_sem_plot.R"))
+
 d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
@@
 
+```{r growth, fig.cap="TODO veronica messing around.
+Dark blue is fastest RT, dark red is slowest RT -- not sure what the ideal legend is.", fig.env="figure*", fig.height = 2.5, fig.width=4, out.width=".5\\textwidth"}
+
+library(nlme)
+
+datasets_with_age_variation <- d_trial |>
+  group_by(dataset_name) |>
+  summarise(age_range = max(age) - min(age)) |>
+  filter(age_range >= AGEVAR_CUTOFF)
+
+  d_sub_prod <- filter(d_sub, !is.na(prod)) |>
+  group_by(subject_id) |>
+  mutate(log_rt_0 = log_rt[1],
+         acc_0 = long_window_accuracy[1]) |> 
+  filter(!is.na(log_rt_0))
+
+mod3 <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
+             data = d_sub_prod,
+             fixed = xmid + scale ~ log_rt_0,
+             random = xmid + scale ~ 1,
+             groups = ~ dataset_name,
+             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 1, `scale:log_rt_0` = 3),
+             na.action = na.exclude, 
+             control = list(maxIter = 1000, tolerance = 10))
+
+# fixed effects (no CIs)
+global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
+                                      max(d_sub_prod$age), .1), 
+                            log_rt_0 = c(quantile(d_sub_prod$log_rt, 
+                                                  c(.1,.25,.5,.75,.9), 
+                                                  na.rm=TRUE)))
+global_preds$pred <- predict(mod3, newdata = global_preds, level = 0)
+
+#random effects
+dataset_preds <- d_sub |>
+  filter(!is.na(prod)) |> # filter on predictor
+  group_by(dataset_name) |>
+  summarise(min_age = min(age),
+            max_age = max(age)) |>
+  group_by(dataset_name) |>
+  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0 = c(quantile(d_sub_prod$log_rt, 
+                                                  c(.1,.25,.5,.75,.9), 
+                                                  na.rm=TRUE))))) |>
+  select(-min_age, -max_age) |>
+  unnest(col = "df") |>
+  ungroup()
+
+library(viridis)
+dataset_preds$pred <- predict(mod3,
+                              newdata = dataset_preds,
+                              level = c(0,1))$predict.dataset_name
+
+ggplot(d_sub_prod, aes(x = age, y = prod, col = as.factor(log_rt_0))) + 
+  geom_point(alpha = .02, color="black") + 
+  geom_line(data = dataset_preds |>   filter(dataset_name %in% datasets_with_age_variation$dataset_name), aes(y = pred, group = interaction(log_rt_0, dataset_name)), lty="dashed", alpha=.5)+
+    geom_line(data = global_preds, aes(y = pred, group = log_rt_0, col = as.factor(log_rt_0)), size=1)+
+  scale_color_viridis(discrete=T, option="H")+
+  theme(legend.position = "none")+
+  labs(y="Production", x="Age in months")
+# +
+#   geom_line(data = dataset_preds, aes(y = pred))
+```
+
 ## Variability of word recognition decreases {.unnumbered}
@@
 
-graph_sem(model = fit3_age, text_size = 2, layout = t(layout_age))
+node_text <- tibble(name=c("acc", "acc_sd", "accuracy", "comp", "log_age", "log_rt", "log_rt_sd", "prod", "speed", "vocab"), pretty=c("Accuracy", "Accuracy (sd)", "Accuracy", "Comprehension", "Age (log months)", "log RT", "log RT (SD)", "Production", "Speed", "Vocabulary"))
 
-nodes <- get_nodes(fit3_age)
+nodes <- get_nodes(fit3_age) |> tibble() |> left_join(node_text) |> mutate(label=ifelse(shape=="oval",pretty, str_c(pretty, "\n", est_sig)))
 
-edges <- get_edges(fit3_age) |> filter(label!="1.00") |>
+edges <- get_edges(fit3_age) |> filter(label!="1.00") |> mutate(label_location=.5) |> mutate(label_hjust=.5, label_vjust=.5, label_lineheight=1.5)
 
-graph_sem(edges=edges, nodes=nodes, text_size = 2, layout = t(layout_age),
-          rect_width=2,
-          rect_height=1.2,
-          ellipses_width=1.5,
-          ellipses_height=1.5,
-          variance_diameter=.9,
-          spacing_x=2
-)
+edges[9, "label_hjust"]=1
+edges[7, "label_hjust"]=1
+
+edges[18, "label_hjust"]=0
+edges[16, "label_hjust"]=0
+
+edges[17,"curvature"]=50
+edges[17,"label_location"]=.6
+
+source(here("helper/modified_sem_plot.R"))
+
+#View(edges)
+prepare_graph(edges=edges, nodes=nodes, text_size = 2.5, layout = t(layout_age),
+          rect_width=4.2,
+          rect_height=3,
+          ellipses_width=3,
+          ellipses_height=3,
+          variance_height=2,
+          variance_width=1.5,
+          arrow_angle=15,
+          arrow_length=.1,
+          var_arrow_angle=15,
+          var_arrow_length=.1,
+          spacing_y=4,
+          spacing_x=4,
+          fix_coord=FALSE) |> plot_happy_arrows()
 ```
@@
 ```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between growth parameters. "}
-#layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
-#graph_sem(model = slopes_long, text_size = 2, layout=t(layout))
-graph_sem(model = slopes_long, text_size = 2)
+layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
 
+node_text <- tribble(~name, ~ pretty, ~fill, 
+                     "acc_t0", "t0 Accuracy", "lightblue",
+                     "acc_t1", "t1 Accuracy", "lightblue",
+                     "acc_t2", "t2 Accuracy","lightblue",
+                     "acc_t3", "t3 Accuracy","lightblue",
+                     "acc_t4", "t4 Accuracy","lightblue",
+                     "accuracy_intercept", "Accuracy intercept","lightblue",
+                     "accuracy_slope", "Accuracy slope","lightblue",
+                     "log_rt_t0", "t0 log RT","pink",
+                     "log_rt_t1", "t1 log RT","pink",
+                     "log_rt_t2", "t2 log RT","pink",
+                     "log_rt_t3", "t3 log RT","pink",
+                     "log_rt_t4", "t4 log RT","pink",
+                     "prod_t0" , "t0 Production", "lightgreen",
+                     "prod_t1" , "t1 Production",  "lightgreen",
+                     "prod_t2" , "t2 Production" ,  "lightgreen",
+                     "prod_t3", "t3 Production",          "lightgreen",
+                     "prod_t4", "t4 Production" , "lightgreen",
+                     "speed_intercept" , "Speed intercept" ,  "pink",
+                     "speed_slope" , "Speed slope","pink",
+                     "vocab_intercept", "Vocabulary intercept", "lightgreen",
+                     "vocab_slope", "Vocabulary slope",  "lightgreen" ) |> 
+  mutate(alpha=.5)
+
+nodes <- get_nodes(slopes_long) |> tibble() |> left_join(node_text) |> mutate(label=pretty)
+
+edges <- get_edges(slopes_long) |> mutate(label_location=.5) |> mutate(label=ifelse(arrow=="last", NA, label)) |>  mutate(connect_from=ifelse(to==from, "right",NA),
+                    connect_to=connect_from) |> 
+  mutate(color="black", label_hjust=.5, label_location=.5)
+
+edges[52,"curvature"]=85
+edges[52,"color"]="blue"
+edges[61,"curvature"]=85
+edges[61,"color"]="blue"
+edges[66,"curvature"]=85
+edges[66,"color"]="blue"
+edges[52, "label_hjust"]=0
+edges[61, "label_hjust"]=0
+edges[66, "label_hjust"]=0
+
+edges[54,"curvature"]=85
+edges[54, "label_hjust"]=0
+
+edges[63,"curvature"]=85
+edges[63, "label_hjust"]=0
+
+edges[53,"curvature"]=80
+edges[53,"color"]="red"
+edges[55,"curvature"]=82
+edges[55,"label_location"]=.55
+
+edges[55,"color"]="red"
+edges[62,"curvature"]=80
+edges[62,"color"]="red"
+edges[53, "label_location"]=.6
+edges[53, "label_hjust"]=0.4
+
+edges[58,"curvature"]=80
+edges[58,"color"]="darkgreen"
+edges[60,"curvature"]=82
+edges[60,"color"]="darkgreen"
+edges[60,"label_location"]=.45
+edges[65,"curvature"]=80
+edges[65,"color"]="darkgreen"
+edges[65, "label_location"]=.4
+edges[53, "label_hjust"]=0.4
+
+edges[59,"curvature"]=77
+
+edges[64,"curvature"]=75
+edges[56,"curvature"]=75
+edges[57,"curvature"]=75
+
+source(here("helper/modified_sem_plot.R"))
+#View(edges)
+prepare_graph(edges=edges, nodes=nodes, text_size = 2, layout = t(layout),angle=0,
+          rect_width=1.5,
+          rect_height=2,
+          ellipses_width=1.5,
+          ellipses_height=3,
+          variance_width= .5,
+          variance_height=1.3,
+          var_arrow_angle=10,
+          var_arrow_length=.05,
+          arrow_angle=15,
+          arrow_length=.07,
+          variance_diameter=1,
+          spacing_y=2.5,
+          spacing_x=1.5,
+          fix_coord=FALSE) |> 
+  edit_edges({connect_from=case_when(
+    (to==from & from %in% c("accuracy_intercept","accuracy_slope","speed_slope",
+                          "speed_intercept", "vocab_slope", "vocab_intercept"))~"bottom",
+    to==from ~ "right",
+    !is.na(curvature)~"left",
+    T ~ "right")}) |> 
+  plot_happy_arrows()
 ```

### 2025-02-05 helper/modified_sem_plot.R

@@ -20,7 +20,12 @@ plot_happy_arrows <- function(x, y, ...){
   rect_height <- x$rect_height
   ellipses_width <- x$ellipses_width
   ellipses_height <- x$ellipses_height
-  variance_diameter <- x$variance_diameter
+  variance_width <- x$variance_width
+  variance_height <- x$variance_height
+  arrow_angle <- x$arrow_angle
+  arrow_length <- x$arrow_length
+  var_arrow_angle <- x$var_arrow_angle
+  var_arrow_length <- x$var_arrow_length
   spacing_x <- x$spacing_x
   spacing_y <- x$spacing_y
   text_size <- x$text_size
@@ -44,10 +49,11 @@ plot_happy_arrows <- function(x, y, ...){
   p <- ggplot(NULL)
 
   if(any(df_edges$from == df_edges$to)){
-    p <- .plot_variances(p, df = df_edges[df_edges$from == df_edges$to, ], text_size = text_size, diameter = variance_diameter)
+    p <- .plot_variances(p, df = df_edges[df_edges$from == df_edges$to, ], text_size = text_size, height = variance_height, width=variance_width,
+                         arrow_angle=var_arrow_angle, arrow_length=var_arrow_length)
   }
   if(any(!df_edges$from == df_edges$to)){
-    p <- .plot_edges(p, df_edges[!df_edges$from == df_edges$to, ], text_size)
+    p <- .plot_edges(p, df_edges[!df_edges$from == df_edges$to, ], text_size = text_size, arrow_angle= arrow_angle, arrow_length=arrow_length)
   }
 
   p <- .plot_nodes(p, df = df_nodes, text_size = text_size, ellipses_width = ellipses_width, ellipses_height = ellipses_height)
@@ -583,30 +589,31 @@ match.call.defaults <- function(...) {
   match.call(sys.function(sys.parent()), call)
 }
 
-.plot_variances <- function(p, df, text_size, diameter, ...) {
+.plot_variances <- function(p, df, text_size, width, height, arrow_angle, arrow_length, ...) {
   npoints <- 20
-  radius <- diameter / 2
   xlabel <- xcenter <- df$edge_xmin
   ylabel <- ycenter <- df$edge_ymin
+  
   offset <- rep(1.5, length(xcenter))
+  
   ycenter[df$connect_from == "top"] <-
-    ycenter[df$connect_from == "top"] + radius
+    ycenter[df$connect_from == "top"] + height/2
   ylabel[df$connect_from == "top"] <-
[…]

### 2025-02-05 paper/paper.Rmd
sem aesthetic tweaks

@@
 
+acc_color <- "#B2DF8A"
+rt_color <- "#FB9A99"
+prod_color <- "#FFFF99"
+
+intercept_color <- "#1F78B4"
+intercept_background <- "#A6CEE3"
+slope_color <- "#FF7F00"
+slope_background <- "#FDBF6F"
+
+#"#A6CEE3" "#1F78B4" "#B2DF8A" "#33A02C"
+# [5] "#FB9A99" "#E31A1C" "#FDBF6F" "#FF7F00"
+# [9] "#CAB2D6" "#6A3D9A" "#FFFF99" "#B15928"
 node_text <- tribble(~name, ~ pretty, ~fill, 
-                     "acc_t0", "t0 Accuracy", "lightblue",
-                     "acc_t1", "t1 Accuracy", "lightblue",
-                     "acc_t2", "t2 Accuracy","lightblue",
-                     "acc_t3", "t3 Accuracy","lightblue",
-                     "acc_t4", "t4 Accuracy","lightblue",
-                     "accuracy_intercept", "Accuracy intercept","lightblue",
-                     "accuracy_slope", "Accuracy slope","lightblue",
-                     "log_rt_t0", "t0 log RT","pink",
-                     "log_rt_t1", "t1 log RT","pink",
-                     "log_rt_t2", "t2 log RT","pink",
-                     "log_rt_t3", "t3 log RT","pink",
-                     "log_rt_t4", "t4 log RT","pink",
-                     "prod_t0" , "t0 Production", "lightgreen",
-                     "prod_t1" , "t1 Production",  "lightgreen",
-                     "prod_t2" , "t2 Production" ,  "lightgreen",
-                     "prod_t3", "t3 Production",          "lightgreen",
-                     "prod_t4", "t4 Production" , "lightgreen",
-                     "speed_intercept" , "Speed intercept" ,  "pink",
-                     "speed_slope" , "Speed slope","pink",
-                     "vocab_intercept", "Vocabulary intercept", "lightgreen",
-                     "vocab_slope", "Vocabulary slope",  "lightgreen" ) |> 
-  mutate(alpha=.5)
+                     "acc_t0", "t0 Accuracy", acc_color,
+                     "acc_t1", "t1 Accuracy", acc_color,
+                     "acc_t2", "t2 Accuracy",acc_color,
+                     "acc_t3", "t3 Accuracy",acc_color,
+                     "acc_t4", "t4 Accuracy",acc_color,
+                     "accuracy_intercept", "Accuracy intercept",acc_color,
+                     "accuracy_slope", "Accuracy slope",acc_color,
+                     "log_rt_t0", "t0 log RT",rt_color,
+                     "log_rt_t1", "t1 log RT",rt_color,
+                     "log_rt_t2", "t2 log RT",rt_color,
+                     "log_rt_t3", "t3 log RT",rt_color,
+                     "log_rt_t4", "t4 log RT",rt_color,
+                     "prod_t0" , "t0 Production", prod_color,
+                     "prod_t1" , "t1 Production",  prod_color,
+                     "prod_t2" , "t2 Production" ,  prod_color,
+                     "prod_t3", "t3 Production",         prod_color,
+                     "prod_t4", "t4 Production" , prod_color,
+                     "speed_intercept" , "Speed intercept" , rt_color,
+                     "speed_slope" , "Speed slope",rt_color,
+                     "vocab_intercept", "Vocabulary intercept", prod_color,
+                     "vocab_slope", "Vocabulary slope",  prod_color ) |> 
+  mutate(alpha=.1)
 
@@
                     connect_to=connect_from) |> 
-  mutate(color="black", label_hjust=.5, label_location=.5)
+  mutate(color="black", label_hjust=.5, label_location=.5, label_fill="white")
 
 edges[52,"curvature"]=85
-edges[52,"color"]="blue"
+#edges[52,"color"]="blue"
 edges[61,"curvature"]=85
-edges[61,"color"]="blue"
+#edges[61,"color"]="blue"
 edges[66,"curvature"]=85
-edges[66,"color"]="blue"
+#edges[66,"color"]="blue"
 edges[52, "label_hjust"]=0
@@
 edges[53,"curvature"]=80
-edges[53,"color"]="red"
+edges[53,"color"]=intercept_color
+edges[53, "label_fill"]=intercept_background
 edges[55,"curvature"]=82
@@
 
-edges[55,"color"]="red"
+edges[55,"color"]=intercept_color
+edges[55, "label_fill"]=intercept_background
+
 edges[62,"curvature"]=80
-edges[62,"color"]="red"
+edges[62,"color"]=intercept_color
+edges[62, "label_fill"]=intercept_background
+
 edges[53, "label_location"]=.6
@@
 edges[58,"curvature"]=80
-edges[58,"color"]="darkgreen"
+edges[58,"color"]=slope_color
+edges[58, "label_fill"]=slope_background
+
 edges[60,"curvature"]=82
-edges[60,"color"]="darkgreen"
+edges[60,"color"]=slope_color
+edges[60, "label_fill"]=slope_background
 edges[60,"label_location"]=.45
 edges[65,"curvature"]=80
-edges[65,"color"]="darkgreen"
+edges[65,"color"]=slope_color
+edges[65, "label_fill"]=slope_background
 edges[65, "label_location"]=.4

### 2025-02-05 helper/modified_sem_plot.R

@@ -990,7 +990,7 @@ match.call.defaults <- function(...) {
     names(df) <- gsub("^label_", "", names(df))
     # Prepare aesthetics ------------------------------------------------------
     if(!"fill" %in% names(df)){
-      df$fill = "white"
+      df$fill = NA
     }
     if(!"size" %in% names(df)){
       df$size <- text_size

### 2025-02-06 paper/paper.Rmd
following recs from doc

--- 2025-02-04 paper/paper.Rmd
@@
 ---
-title: Continuous developmental changes in word recognition across early childhood support early language learning
+title: Continuous developmental changes in word recognition support language learning across early childhood
 
@@
 How does this skill develop over the course of early childhood?
-And how does facility in word recognition relate to the growth of vocabulary?
+And how does facility in word recognition relate to the growth of vocabulary knowledge?
 We address these questions using data from Peekbank, an open database of eye-tracking experiments on children's early word recognition.
-Combining 24 datasets with data from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development, both within and across studies.
+Combining 24 datasets from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development, both within and across studies.
 Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy to parent-reported vocabulary.
 Across a range of models, speed, accuracy, and vocabulary also show coupled growth such that children with faster initial language processing tend to show larger vocabulary growth.
-Together, these findings support the view that language processing is a skill, and that this skill plays a role in early language learning.
+Together, these findings support the view that language processing is a skill and that this skill plays a role in supporting early language learning.
 
@@
 The efficiency with which children recognize words provides a measurement of their real-time language processing, which has been argued to be a key part of the learning process.
-Here we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
+Her, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
 These data suggest that understanding language is a skill that improves gradually with practice.
@@
 
-source(here("helper/modified_sem_plot.R"))
-
 d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
@@
 
-#library(RColorBrewer)
-
-  mycolors = c("#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E","#1B9E77", "#D95F02")
-  # 4 colors from brewer scale "Dark2" that I thought had decent contrast against the grey/black background dots
-
 a <- ggplot(d_sub,
        aes(x = age, y = long_window_accuracy)) + 
-  geom_point(alpha = .02) + 
+  geom_point(alpha = .05) + 
+  geom_smooth(method = "lm") +
   geom_smooth(data = d_sub_agevar,
               aes(group = dataset_name, col = dataset_name), se=FALSE,
-              alpha = 1, method = "lm", size = .5, lty = 2) + 
-    geom_smooth(method = "lm") +
+              alpha = .1, method = "lm", size = .5, lty = 2) + 
   geom_hline(yintercept = .5, lty = 2) + 
   scale_x_log10() +
-  scale_color_manual(values=mycolors, guide = FALSE) + 
+  scale_color_discrete(guide = FALSE) + 
   ylab("Accuracy") + 
@@
        aes(x = age, y = rt)) + 
-  geom_point(alpha = .02) + 
+  geom_point(alpha = .05) + 
+  geom_smooth(method = "lm") +
   geom_smooth(data = d_sub_agevar,
@@
               alpha = .1, method = "lm", size = .5, lty = 2) + 
-    geom_smooth(method = "lm") +
   scale_x_log10() +
   scale_y_log10() +
-  scale_color_manual(values=mycolors, guide = FALSE) + 
+  scale_color_discrete(guide = FALSE) + 
   theme(legend.position = "bottom") + 
@@
 ```{r variance, fig.cap="Participant-level variability in accuracy and reaction time (log RT), plotted by age (log). Plotting conventions are as in Figure 1.", fig.env="figure*", fig.height = 2.5}
-  mycolors = c("#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E","#1B9E77", "#D95F02")
-
 c <- ggplot(d_sub,
        aes(x = age, y = long_window_acc_var)) + 
-  geom_point(alpha = .02) + 
+  geom_point(alpha = .05) + 
+  geom_smooth(method = "lm") +
   geom_smooth(data = d_sub_agevar,
@@
               alpha = .1, method = "lm", size = .5, lty = 2) + 
-    geom_smooth(method = "lm") +
   scale_x_log10() +
-  scale_color_manual(values=mycolors, guide = FALSE) + 
+  scale_color_discrete(guide = FALSE) + 
   ylab("SD of Accuracy)") + 
@@
        aes(x = age, y = log_rt_var)) + 
-  geom_point(alpha = .02) + 
+  geom_point(alpha = .05) + 
+  geom_smooth(method = "lm") +
   geom_smooth(data = d_sub_agevar,
@@
               alpha = .1, method = "lm", size = .5, lty = 2) + 
-    geom_smooth(method = "lm") +
   scale_x_log10() +
-  scale_color_manual(values=mycolors, guide = FALSE) + 
+  scale_color_discrete(guide = FALSE) + 
   theme(legend.position = "bottom") + 
@@
 
-node_text <- tibble(name=c("acc", "acc_sd", "accuracy", "comp", "log_age", "log_rt", "log_rt_sd", "prod", "speed", "vocab"), pretty=c("Accuracy", "Accuracy (sd)", "Accuracy", "Comprehension", "Age (log months)", "log RT", "log RT (SD)", "Production", "Speed", "Vocabulary"))
-
-nodes <- get_nodes(fit3_age) |> tibble() |> left_join(node_text) |> mutate(label=ifelse(shape=="oval",pretty, str_c(pretty, "\n", est_sig)))
-
-edges <- get_edges(fit3_age) |> filter(label!="1.00") |> mutate(label_location=.5)
-
-edges[17,"curvature"]=50
-edges[17,"label_location"]=.6
-
-source(here("helper/modified_sem_plot.R"))
-
-#View(edges)
-prepare_graph(edges=edges, nodes=nodes, text_size = 2.5, layout = t(layout_age),
-          rect_width=4.2,
-          rect_height=3,
-          ellipses_width=3,
-          ellipses_height=3,
-          variance_diameter=2,
-          spacing_y=4,
-          spacing_x=4,
-          fix_coord=FALSE) |> plot_happy_arrows()
+graph_sem(model = fit3_age, text_size = 2, layout = t(layout_age))
 ```
@@
 layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
-
-node_text <- tribble(~name, ~ pretty,
-                     "acc_t0", "t0 Accuracy",
-                     "acc_t1", "t1 Accuracy",
-                     "acc_t2", "t2 Accuracy",
-                     "acc_t3", "t3 Accuracy",
-                     "acc_t4", "t4 Accuracy",
-                     "accuracy_intercept", "Accuracy intercept",
-                     "accuracy_slope", "Accuracy slope",
-                     "log_rt_t0", "t0 log RT",
-                     "log_rt_t1", "t1 log RT",
-                     "log_rt_t2", "t2 log RT",
-                     "log_rt_t3", "t3 log RT",
-                     "log_rt_t4", "t4 log RT",
-                     "prod_t0" , "t0 Production",
-                     "prod_t1" , "t1 Production",
-                     "prod_t2" , "t2 Production" ,
-                     "prod_t3", "t3 Production",         
-                     "prod_t4", "t4 Production" ,
-                     "speed_intercept" , "Speed intercept" ,  
-                     "speed_slope" , "Speed slope",
-                     "vocab_intercept", "Vocabulary intercept",
-                     "vocab_slope", "Vocabulary slope" )
-
-nodes <- get_nodes(slopes_long) |> tibble() |> left_join(node_text) |> mutate(label=pretty)
-
-edges <- get_edges(slopes_long) |> mutate(label_location=.5) |> mutate(label=ifelse(arrow=="last", NA, label)) |>  mutate(connect_from=ifelse(to==from, "right",NA),
-                    connect_to=connect_from) |> 
-  mutate(color="black")
-
-edges[52,"curvature"]=85
-edges[52,"color"]="blue"
-edges[61,"curvature"]=85
-edges[61,"color"]="blue"
-edges[66,"curvature"]=85
-edges[66,"color"]="blue"
-
-edges[54,"curvature"]=85
-edges[63,"curvature"]=85
-
-edges[53,"curvature"]=80
-edges[53,"color"]="red"
-edges[55,"curvature"]=82
-edges[55,"color"]="red"
-edges[62,"curvature"]=80
-edges[62,"color"]="red"
-
-edges[58,"curvature"]=80
-edges[58,"color"]="green"
-edges[60,"curvature"]=82
-edges[60,"color"]="green"
-edges[65,"curvature"]=80
-edges[65,"color"]="green"
-
-edges[59,"curvature"]=77
-
-edges[64,"curvature"]=75
-edges[56,"curvature"]=75
-edges[57,"curvature"]=75
-
-source(here("helper/modified_sem_plot.R"))
-#View(edges)
-prepare_graph(edges=edges, nodes=nodes, text_size = 2, layout = t(layout),angle=0,
-          rect_width=1.5,
-          rect_height=2,
-          ellipses_width=1.5,
-          ellipses_height=4,
-          variance_diameter=1,
-          spacing_y=3,
-          spacing_x=2,
-          fix_coord=FALSE) |> 
-  edit_edges({connect_from=case_when(
-    (to==from & from %in% c("accuracy_intercept","accuracy_slope","speed_slope",
-                          "speed_intercept", "vocab_slope", "vocab_intercept"))~"bottom",
-    to==from ~ "right",
-    !is.na(curvature)~"left",
-    T ~ "right")}) |> 
-  plot_happy_arrows()
+graph_sem(model = slopes_long, text_size = 2, layout=t(layout))
 ```

### 2025-02-06 5_longitudinal.qmd

@@
   mutate(log_rt_0 = log_rt[1],
-         acc_0 = long_window_accuracy[1])
+         acc_0 = long_window_accuracy[1]) |> 
+  filter(!is.na(log_rt_0))
 
@@
              groups = ~ dataset_name,
-             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 0, `scale:log_rt_0` = 3),
+             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 1, `scale:log_rt_0` = 3),
              na.action = na.exclude, 
@@
 
-summary(mod3)
 ```

### 2025-02-06 paper/paper.Rmd
recache and revisions

@@
   - name: Michael C. Frank
+    affiliation: a, 1
+  - name: Virginia A. Marchman
     affiliation: a
-  - name: Other peekbankers to be added
+  - name: Claire Augusta Bergey
     affiliation: b
+  - name: Veronica Boyce
+    affiliation: a
+  - name: Mika Braginsky
+    affiliation: a
+  - name: Jess Mankewitz
+    affiliation: c
+  - name: Stephan Meylan
+    affiliation: d
+  - name: Ben Prystawski
+    affiliation: a
+  - name: Nilam Ram
+    affiliation: a, e
+  - name: Robert Z. Sparks
+    affiliation: a
+  - name: Adrian Steffan
+    affiliation: f
+  - name: Alvin Wei Ming Tan
+    affiliation: a
   - name: Martin Zettersten
-    affiliation: c
+    affiliation: f
 address:
@@
   - code: b
-    address: Many other peekbankers affiliations
+    address: Department of Linguistics, Stanford University, Stanford, CA, USA
   - code: c
-    address: Department of Cognitive Science, University of California, San Diego, CA, USA
+    address: Department of Psychology, University of Wisconsin, Madison, WI, USA
+  - code: d
+    address: Department of Linguistics, University of California, Berkeley, CA, USA
+  - code: e
+    address: Department of Communication, Stanford University, Stanford, CA, USA
+  - code: f
+    address: Department of Communication, Stanford University, Stanford, CA, USA
+  - code: g
+    address: Department of Psychology, Ludwig Maximilian University, Munich, Germany
 
 corresponding_author:
-  code: 2
-  text: "To whom correspondence should be addressed. E-mail: bob@email.com"
+  code: 1
+  text: "To whom correspondence should be addressed. E-mail: mcfrank@stanford.edu"
 
@@
 abstract: |
-  Being a fluent language user involves understanding language as it unfolds in time.
+  Being a fluent language user involves recognizing words as they unfold in time.
 How does this skill develop over the course of early childhood?
 And how does facility in word recognition relate to the growth of vocabulary knowledge?
-We address these questions using data from Peekbank, an open database of eye-tracking experiments on children's early word recognition.
-Combining 24 datasets from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development, both within and across studies.
-Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy to parent-reported vocabulary.
-Across a range of models, speed, accuracy, and vocabulary also show coupled growth such that children with faster initial language processing tend to show larger vocabulary growth.
-Together, these findings support the view that language processing is a skill and that this skill plays a role in supporting early language learning.
+We address these questions using data from Peekbank, an open database of experiments measuring children's eye-movements during early word recognition.
+Combining 24 datasets from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development.
+Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy with parent-reported vocabulary.
+Across a range of models, speed, accuracy, and vocabulary also show coupled growth such that children with faster initial word recognition tend to show faster vocabulary growth.
+Together, these findings support the view that word recognition is a skill that develops gradually across early childhood and that this skill plays a role in supporting early language learning.
 
 significance: |
-  Early language development is a key part of children's development that also sets the stage for later academic success.
-The efficiency with which children recognize words provides a measurement of their real-time language processing, which has been argued to be a key part of the learning process.
+  The efficiency with which children recognize words provides a measurement of their real-time language processing, which has been argued to be a key part of the language learning process.
 Her, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
@@
 As children enter their second year, language explodes onto the scene; both vocabulary and grammatical abilities grow rapidly and in tandem [@bates1994developmental;@frank2021].
-This growth in knowledge is also accompanied by changes in  language processing: children become quicker and more accurate in recognizing words and matching them with their referents [@fernald1998;@peter2019;@bergelson2020comprehension].
+This growth in knowledge is also accompanied by changes in language processing efficiency: children become quicker and more accurate in recognizing words and matching them with their referents [@fernald1998;@peter2019;@bergelson2020comprehension].
 
@@
 Eye tracking has thus emerged as a key method that allows the measurement of language comprehension with high temporal resolution: both adults and children reliably fixate the referent of a word soon after it is used [@tanenhaus1995integration;@altmann1999incremental;@fernald1998;@fernald2008;@macdonald2018].
-The relative timecourse of fixation then can provide an index of both the comprehender's ability and the properties of the stimulus.
+The relative timecourse of fixation then can provide an index of the comprehender's ability or the difference between two stimulus conditions.
 
-\begin{table*}[h!]
+The version of this method that is used with children goes by many names, including the intermodal preferential looking paradigm and the "looking while listening" paradigm (LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
+In LWL experiments, children are typically shown a series of trials in which two images are displayed side by side and they are asked to find one of them.
+For example, a ball and a book might be shown, and the child might hear "Look at the ball!
+Can you find it?".
+Accuracy is then computed as the proportion of time they fixate the correct image within a fixed window after the onset of the noun ("ball" in this case).
+Reaction time is computed only on trials in which the child is fixating on the distractor image (the book) before word onset; in these cases, the average time it takes until the child shifts fixation from the distractor to the target is used as an index of processing speed.
+Early work using this method showed that both children's speed and accuracy increase rapidly across the second year [@fernald1998;@reznick1990visual].
+Related methods have provided a window into how children process phonological [@mani2011], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] information as well as how their lexical representations develop [@Swingley2002].
+
+Word recognition ability, as measured by LWL, is hypothesized to play a key role in language learning.
+Each word that a child experiences is an opportunity to learn; measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking].
+Recognizing incoming words and linking them with their referents is a prerequisite for learning.
+Consider a child hearing the utterance "Can you put the ball in the box?" The faster and more accurately the child can recognize that the ball is a referent, the better they can use this evidence to help infer the speaker's intended meaning, perhaps making inferences about the meaning of "put" or "box" [@frank2009using].
+Consistent with this idea, one important study found that children's word recognition speed mediated the relationship between home language input and vocabulary growth [@Weisleder2013].
+
+\begin{table*}[ht]
 \centering
@@
   2 & yurovsky\_2017 & 282 & 282 & 25.64 & 12.59 & 58.65 & 5.95 & 2.79 &  &  \\ 
-  3 & weaver\_zettersten\_2024 & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 & x & x \\ 
-  4 & fernald\_marchman\_2012 & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.78 & 7.17 & x & x \\ 
+  3 & weaver\_zettersten\_2024 & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 &  & x \\ 
+  4 & fernald\_marchman\_2012 & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.89 & 7.21 & x & x \\ 
   5 & sander-montant\_2022 & 122 & 122 & 21.94 & 12.02 & 31.11 & 10.07 & 4.20 & x &  \\ 
@@
   8 & fmw\_2013 &  80 & 179 & 20.04 & 17.00 & 26.00 & 21.26 & 8.44 & x & x \\ 
-  9 & adams\_marchman\_2018 &  69 & 711 & 23.57 & 12.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
+  9 & adams\_marchman\_2018 &  69 & 711 & 23.58 & 13.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
   10 & potter\_canine &  67 &  67 & 23.76 & 21.00 & 27.00 & 10.38 & 3.80 & x &  \\ 
-  11 & fernald\_totlot &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.22 & 6.05 & x & x \\ 
+  11 & fernald\_totlot &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.25 & 6.06 & x & x \\ 
   12 & pomper\_prime &  63 &  63 & 39.73 & 37.90 & 42.00 & 12.98 & 5.34 &  &  \\ 
@@
   15 & pomper\_salientme &  44 &  44 & 40.11 & 38.00 & 43.00 & 5.30 & 2.34 &  &  \\ 
-  16 & perry\_cowpig &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 &  &  \\ 
+  16 & perry\_cowpig &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 & x &  \\ 
   17 & ronfard\_2021 &  40 &  40 & 19.95 & 18.00 & 24.00 & 18.54 & 7.62 & x &  \\ 
   18 & bacon\_gendercues &  38 &  38 & 22.87 & 22.00 & 24.00 & 18.16 & 8.00 & x &  \\ 
-  19 & garrison\_bergelson\_2020 &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.94 & 9.59 & x &  \\ 
+  19 & garrison\_bergelson\_2020 &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.47 & 9.41 & x &  \\ 
   20 & pomper\_yumme &  32 &  32 & 26.38 & 25.00 & 28.00 & 7.62 & 2.31 &  &  \\ 
@@
   \hline
+    & Total & 1963 & 3554 & 24.73 & 12.00 & 60.00 & 12.74 & 5.05 & 15 & 6 \\ 
+   \hline
 \end{tabular}
-\caption{Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions.\label{tab:datasets}} 
+\caption{Characteristics of included datasets from Peekbank.
+`Admins` denotes separate experimental sessions.
+`CDIs` refers to whether the dataset contains parent report vocabulary data from the MacArthur-Bates Communicative Development Inventory.}
 \end{table*}
 
-The version of this method that is used with children goes by many names, including the intermodal preferential looking paradigm and the "looking while listening" paradigm (LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
-In LWL experiments, children are typically shown a series of trials in which two images are displayed side by side and they are asked to find one of them.
-For example, a ball and a book might be shown, and the child might hear "Look at the ball!
-Can you find it?".
-Accuracy is then computed as the proportion of time they fixate the correct image within a fixed window after the onset of the noun ("ball" in this case).
-Reaction time is computed only on trials in which the child is fixating the distractor image (the book); in these cases, the average time it takes to shift fixation to the target is used as an index of processing speed.
-Early work using this method showed that children's speed and accuracy both increase rapidly across the second year [@fernald1998;@reznick1990visual].
-Related methods have provided a window into how children process phonological [@mani2011], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] competition as well as how their lexical representations develop [@Swingley2002].
-
-Word recognition ability, as measured by LWL, is hypothesized to play a key role in language learning.
-Each word that a child experiences is an opportunity to learn; measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking].
-<!-- These learning opportunities are likely helpful to different degrees, however [@kachergis2022toward;@mcmurray2007defusing].  -->
-<!-- Language input is more helpful for learning if it focused on the child's own interest and grounded in the particulars of their current environment -- for example, in a game, routine, or object -- and if it uses vocabulary and syntax that they know [@tomasello2001perceiving;@roy2015predicting;@cartmill2013quality].
--->
-Recognizing incoming words and linking them with their referents is a prerequisite for learning.
-Consider a child hearing the utterance "can you put the ball in the box?" The faster and more accurately the child can recognize that the ball is a referent, the better they can use this evidence to help infer the speaker's intended meaning, perhaps making inferences about the meaning of "put" or "box" [@frank2009using].
-Consistent with this idea, one important study found that children's word recognition speed mediated the relationship between home language input and vocabulary growth [@Weisleder2013].
-
-Speed of word recognition has been used as an index of individual differences in early childhood [@fernald2006;@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow].
-Word recognition speed at 18 months can predict children's later language ability over and above measures of vocabulary size [@marchman2008], though these predictions may be limited to particular ages or processing assessments [@peter2019].
-Further, faster processing at age two is predictive of whether "late talkers" catch up to their peers or require further intervention [@fernald2012].
+Word recognition speed has been used as an index of individual differences in early childhood [@fernald2006;@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow].
+Over and above measures of vocabulary size, word recognition speed at 18 months predicts children's standardized test scores years later [@marchman2008], though these predictions may be limited to particular ages or processing assessments [@peter2019].
+Further, faster processing at 18 months is predictive of whether "late talkers" catch up to their peers or require further intervention [@fernald2012].
 Critically, all of these assessments use words that children are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.
@@
 Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
-These samples provide neither the breadth of ages nor the number of participants to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of language [as has been done with school-aged children; @mcmurray2022slow;@colby2023efficiency].
+These samples provide neither the breadth of ages nor the number of participants needed to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of language [as has been done with school-aged children; @colby2023efficiency; @mcmurray2022slow].
 Here we investigate two hypotheses.
 
-First, one influential recent theory characterizes language learning more broadly as a process of learning the skill of real-time processing  [@christiansen2016;@chater2018].
+First, one influential theory characterizes language learning as a process of learning the skill of real-time processing [@christiansen2016;@chater2018].
 Based on this theory, we should expect to see the signatures of expertise and skill learning in word recognition.
@@
 Second, previous findings have provided limited and sometimes conflicting evidence on the concurrent and predictive relations between word recognition and language learning.
-Initial reports showed strong predictive relationships between both speed and accuracy and later vocabulary growth [@fernald2006].
-Two other studies extended this result to infants born preterm [@fernald2012] and to late talkers [@marchman2016].
-Subsequent studies have primarily focused on speed of processing and found more mixed results, with reaction time measures only inconsistently predictive of later vocabulary outcomes [@peter2019;@newbury2016interrelationships;@lany2017].
+Initial reports showed strong predictive relationships between both speed and accuracy and later vocabulary growth [@fernald2006], with replications in infants born preterm [@marchman2016] and late talkers [@fernald2012].
+Subsequent studies have primarily focused on speed of processing and found more mixed results, with reaction time measures found to be only inconsistently predictive of later vocabulary outcomes [@peter2019;@newbury2016interrelationships;@lany2018].
 To the extent that there are consistent relations between vocabulary and word recognition, these should be visible in a larger dataset.
@@
 We retrieved data from children ages 1--6 years.
-Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a word (typically embedded in a simple carrier phrase such as "look at the ..."); these trials often constituted control conditions for experiments.
+Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ..."); these trials often constituted control conditions for experiments.
 We focus here on English purely for practical reasons -- the Peekbank dataset contains limited data from other languages.
@@
 The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed study of the development of word recognition than has previously been possible.
-While our analysis is exploratory in nature, it is guided by the two hypotheses outlined above: the presence of signatures of skill learning in word recognition, and 2) linkages between word recognition and vocabulary.
+While our analysis is exploratory in nature, it is guided by the two hypotheses outlined above: the presence of 1) signatures of skill learning in word recognition, and 2) linkages between word recognition and vocabulary.
 
@@
   arrange(desc(`N subjects`))
+
+dataset_characteristics <- dataset_characteristics |>
+  bind_rows(tibble(dataset_name = "Total", 
+                       `N subjects` = sum(dataset_characteristics$`N subjects`),
+                       `N admins` = sum(dataset_characteristics$`N admins`), 
+                       `Mean Age` = mean(dataset_characteristics$`Mean Age`),
+                       `Min Age` = min(dataset_characteristics$`Min Age`), 
+                       `Max Age` = max(dataset_characteristics$`Max Age`), 
+                       `Avg Trials` = mean(dataset_characteristics$`Avg Trials`),
+                       `Avg RT Trials` = mean(dataset_characteristics$`Avg RT Trials`),
+                       `CDIs` = as.character(sum(dataset_characteristics$CDIs != "")),
+                       `longitudinal` = as.character(sum(dataset_characteristics$longitudinal != ""))))
 ```
@@
 We began by characterizing developmental changes in speed and accuracy.
-We computed both reaction times and accuracies following standard practices in the literature [@fernald2008].
+We computed both RTs (reaction times) and accuracies following standard practices in the literature [@fernald2008].
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms).
+For each window, we averaged all fixation within the window to compute a continuous proportion between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
 
 Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI for raw correlations between variables).
-To investigate this question, we fit linear mixed effects models predicting accuracy and reaction time on each trial across the full dataset with random slopes of age by study and random intercepts by participant.
-We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of reaction time.
+To investigate this question, we fit linear mixed effects models predicting accuracy and RT on each trial across the full dataset with random slopes of child age by study and random intercepts by participant.
+We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
 The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text.
-All findings reported here hold for both window sizes, however (see SI).] and the best fitting model of speed predicted log reaction time as a function of log age as well (see SI).
+All findings reported here hold for both window sizes, however (see SI).] and the best fitting model of speed predicted log RT as a function of log age as well (see SI).
 
 Figure \ref{fig:devchange} shows these relationships.
-Log reaction time decreased significantly with age (`r papaja::apa_print(rt_mods_lmer[[2]])$full_result$log_age_s`) and accuracy increased significantly with age  (`r papaja::apa_print(acc_mods_lmer[[2]])$full_result$log_age_s`).
+Log RT decreased significantly with age (`r papaja::apa_print(rt_mods_lmer[[2]])$full_result$log_age_s`) and accuracy increased significantly with age  (`r papaja::apa_print(acc_mods_lmer[[2]])$full_result$log_age_s`).
 In sum, we see continuing improvements in word recognition across the full age range in our dataset that appear roughly linear in the logarithm of age.
@@
 ```{r devchange, fig.cap="Participant-level accuracy and reaction time (log), plotted by age (log).
-The solid green line shows a linear fit and associated confidence interval.
+The solid blue line shows a linear fit and associated confidence interval.
 Individual dotted lines show linear fits for those datasets spanning six or more months of age.
@@
   geom_hline(yintercept = .5, lty = 2) + 
-  scale_x_log10() +
+  scale_x_log10(breaks = c(12,24,36,48,60)) +
   scale_color_manual(values=mycolors, guide = FALSE) + 
@@
     geom_smooth(method = "lm") +
-  scale_x_log10() +
+  scale_x_log10(breaks = c(12,24,36,48,60)) +
   scale_y_log10() +
@@
 ```{r growth, fig.cap="TODO veronica messing around.
-Dark blue is fastest RT, dark red is slowest RT -- not sure what the ideal legend is.", fig.env="figure*", fig.height = 2.5, fig.width=4, out.width=".5\\textwidth"}
+Dark blue is fastest RT, dark red is slowest RT -- not sure what the ideal legend is.", fig.env="figure*", fig.height = 2.5, fig.width=4, out.width=".5\\textwidth", include=FALSE}
 
@@
 One further hallmark of increasing skill is a decrease in task-relevant variability [@todorov2002optimal].
-Both within and across datasets, variation in accuracy and reaction time decreased across datasets, again relatively continuously across the developmental range we examined (Figure \ref{fig:variance}).
-We fit mixed effects models to the standard deviation of both accuracy and reaction time for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
-For both speed and accuracy, variability decreased with age (RT: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`).
-Thus, as well as being faster and more accurate, older children appear to be more consistent in their online word recognition.
+Both within and across datasets, within-individual variation in accuracy and RT decreased smoothly  across the developmental range we examined (Figure \ref{fig:variance}).
+We fit mixed effects models to the standard deviation of both accuracy and RT for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
+For both speed and accuracy, within-individual variability decreased with age (RT: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`).
+Thus, as well as being faster and more accurate, older children were more consistent in their online word recognition.
 
@@
     geom_smooth(method = "lm") +
-  scale_x_log10() +
+  scale_x_log10(breaks = c(12,24,36,48,60)) +
   scale_color_manual(values=mycolors, guide = FALSE) + 
-  ylab("SD of Accuracy)") + 
+  ylab("SD of Accuracy") + 
   xlab("Age (months)")
@@
     geom_smooth(method = "lm") +
-  scale_x_log10() +
+  scale_x_log10(breaks = c(12,24,36,48,60)) +
   scale_color_manual(values=mycolors, guide = FALSE) + 
   theme(legend.position = "bottom") + 
-  ylab("SD of log RT)") + 
+  ylab("SD of log RT") + 
   xlab("Age (months)")
@@
 
-We were next interested in whether the various different aspects of word recognition -- including accuracy, reaction time, and the variability of each of these -- were related to other aspects of early language ability.
-Of the studies in our database, `r sum(str_length(dataset_characteristics$CDIs)>0)` gathered parent reports about children's early vocabulary using the MacArthur-Bates Communicative Development Inventory (CDI), a popular survey instrument that provides a reliable and valid holistic picture of children's early language [@marchman2023;@frank2021].
+We were next interested in whether the various aspects of word recognition -- including accuracy, RT, and the variability of each of these -- were related to other aspects of early language ability.
+Of the studies in our database, `r sum(str_length(dataset_characteristics$CDIs)>0)` gathered parent reports about children's early vocabulary using the MacArthur-Bates Communicative Development Inventory (CDI), a popular survey instrument that provides a reliable and valid estimate of children's early vocabulary [@marchman2023;@frank2021].
 Different forms of the CDI can be used to measure either receptive and expressive vocabulary (for children up to 18 months) or expressive vocabulary only (for children 16 -- 30 months).
@@
 Initial exploratory factor analysis using parallel analysis suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
-Due to missingness of data, we used confirmatory factory analysis with full information maximum likelihood to find the best set of loadings.
-The best fitting model was a three-factor model with factors for speed (reaction time and reaction time variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI).
-Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`)`; see SI: Alternative Factor Structures).
+Due to missingness of data, we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
+The best fitting model was a three-factor model with factors for speed (RT and RT variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI).
+Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`); see SI: Alternative Factor Structures).
 
@@
 
-node_text <- tibble(name=c("acc", "acc_sd", "accuracy", "comp", "log_age", "log_rt", "log_rt_sd", "prod", "speed", "vocab"), pretty=c("Accuracy", "Accuracy (sd)", "Accuracy", "Comprehension", "Age (log months)", "log RT", "log RT (SD)", "Production", "Speed", "Vocabulary"))
+node_text <- tibble(name=c("acc", "acc_sd", "accuracy", "comp", "log_age", "log_rt", "log_rt_sd", "prod", "speed", "vocab"), pretty=c("Accuracy", "Accuracy (SD)", "Accuracy", "Comprehension", "Age (log months)", "log RT", "log RT (SD)", "Production", "Speed", "Vocabulary"))
 
@@
 edges[17,"label_location"]=.6
+edges[, "label_fill"]="white"
 
@@
 This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age).
-All three latent factors were significantly related to one another, with speed and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between speed and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
-This model supports the idea that individual variation in speed and accuracy of word recognition is concurrently related to vocabulary beyond effects of age.
+All three latent factors were significantly related to one another, with RT and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
+This model supports the idea that individual variation in speed and accuracy of word recognition is concurrently related to vocabulary beyond the effects of age.
 
@@
 d_growth$age_15 <- d_growth$age - 15
+
 growth_mod <- lmer(prod ~ poly(age_15,2) *rt_t0 + (age  | subject_id) + (age | dataset_name), 
@@
 Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure relationships between LWL and vocabulary using longitudinal growth models.
-We began by reproducing the analysis reported in @fernald2012, in which longitudinal growth in productive vocabulary was predicted based on reaction time at study initiation.
-We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, reaction time at initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset).
-This model revealed a significant effect of $t_0$ reaction time (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ reaction time and the quadratic age predictor (`r papaja::apa_print(acc_var_mod)$full_result$polyage_15_22_rt_t0`).
-This analysis suggests that children with faster reaction times show both larger vocabularies and faster vocabulary growth over time, confirming that findings from @fernald2012 were robust across the full dataset.
+We began by reproducing the analysis reported in @fernald2012, in which longitudinal growth in productive vocabulary was predicted based on RT during the initial session of the study.
+We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset).
+This model revealed a significant effect of $t_0$ RT (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ RT and the quadratic age predictor (`r papaja::apa_print(growth_mod)$full_result$polyage_15_22_rt_t0`).
+This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time, confirming that findings from @fernald2012 were robust across the full dataset.
 
@@
 
-On the other hand, it is possible that differences in predicted growth trajectories is due to overall coupling between vocabulary size and language processing, rather than a specifically predictive relationship between $t_0$ reaction time and vocabulary growth.
+On the other hand, it is possible that differences in predicted growth trajectories are due to overall coupling between vocabulary size and language processing, rather than a specifically predictive relationship between $t_0$ RT and vocabulary growth.
 To test this relationship, we used longitudinal structural equation models.
@@
 Consistent with the idea that overall faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
-There was not significant coupling between growth in reaction time and growth in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
+There was not significant coupling between growth in RT and growth in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
 These abilities might grow independently, but we cannot rule out other possibilities.
@@
 
-How does word recognition change across early childhood, and how does it relate to language learning?
-We investigated these questions using a new dataset of developmental eye-tracking measurements.
+How does word recognition change across early childhood and how does it relate to language learning?
+We investigated these questions using a new, large-scale dataset of developmental eye-tracking measurements compiled across many prior studies.
 We found continuous developmental changes from ages 1 -- 6 years.
@@
 Further, trial-to-trial variability decreased, consistent with both the literature on skill learning [@todorov2002optimal] and other work on developmental changes in variability [@judd2021working;@galeano2018variability;@kautto2023introducing].
-Speed and accuracy were both related to vocabulary size concurrently, and processing speed was also related to later vocabulary growth.
+Speed and accuracy were both related to vocabulary size concurrently and processing speed was also related to later vocabulary growth.
 
@@
 
-By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had samples sizes at least an order of magnitude smaller than ours.
-In contrast to individual studies, which typically only have statistical power to test specific contrasts, our "big data" approach provided the sample sizes necessary to fit complex structural equation models and to compare different functional forms for developmental change.
-Because early language is so variable [@frank2021], these kinds of samples -- thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning.
+By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had sample sizes at least an order of magnitude smaller than ours.
+In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to fit complex structural equation models and to compare different functional forms for developmental change.
+Because early language is so variable [@frank2021], these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [bergelson2023;@frank2017;@frank2021;@manybabies2020].
 
 At the same time, our approach is both observational and exploratory.
-Thus, we cannot untangle a variety of different causal models for the variation we observed.
-Early word recognition skill could lead to faster word learning, due to skilled children gleaning more information from the same signal [@fernald2006].
-But faster children could also be faster due to their larger vocabulary and stronger lexical representations.
-These two causal directions could also interact reciprocally, leading to a "rich get richer process" in which children with larger vocabularies process faster, and their faster processing helps them increase their vocabulary size more rapidly.
-Or a third shared factor -- perhaps general cognitive ability -- could underpin both processes.
+Thus, we cannot untangle the range of different causal models that explain the variation we observed.
+First, early word recognition skill could lead to faster word learning, due to skilled children gleaning more information from the same signal [@fernald2006].
+But second, faster children could also be faster due to their larger vocabulary and stronger lexical representations.
+These two causal directions could also interact reciprocally, leading to a "rich get richer" process in which children with larger vocabularies process faster, and their faster processing helps them increase their vocabulary size more rapidly.
+Finally, a third shared factor -- perhaps general cognitive ability -- could underpin both processes.
 Our cross-sectional data cannot distinguish these hypotheses even in principle [@vanderweele2023dimensional], and our longitudinal data are likely too sparse to distinguish such complex causal models.
@@
 We would not expect that specific numerical parameters estimated in this aggregate convenience sample would generalize to other samples.
-Nevertheless, the consistency of the trends we observed across datasets suggest that our qualitative conclusions are robust to some significant sociodemographic variation.
+Nevertheless, the consistency of the trends we observed across datasets suggests that our qualitative conclusions are robust to some significant sociodemographic variation.
 

### 2025-03-13 paper/paper.Rmd
keywords+typo

@@
   The efficiency with which children recognize words provides a measurement of their real-time language processing, which has been argued to be a key part of the language learning process.
-Her, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
+Here, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
 These data suggest that understanding language is a skill that improves gradually with practice.
@@
 keywords:
-  - one
-  - two
-  - optional
-  - optional
-  - optional
+  - language development
+  - word recognition
+  - developmental change
+  - skill acquisition
+  - eye-tracking
 

### 2025-03-13 paper/paper.tex
keywords+typo

\documentclass[9pt,twocolumn,twoside,]{pnas-new}

% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
%From Pandoc 3.1.8
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\templatetype{pnasresearcharticle}  % Choose template

\title{Continuous developmental changes in word recognition support
language learning across early childhood}

\author[a, 1]{Michael C. Frank}
\author[a]{Virginia A. Marchman}
\author[b]{Claire Augusta Bergey}
\author[a]{Veronica Boyce}
\author[a]{Mika Braginsky}
\author[c]{Jess Mankewitz}
\author[d]{Stephan Meylan}
\author[a]{Ben Prystawski}
\author[a, e]{Nilam Ram}
\author[a]{Robert Z. Sparks}
\author[f]{Adrian Steffan}
\author[a]{Alvin Wei Ming Tan}
\author[f]{Martin Zettersten}

  \affil[a]{Department of Psychology, Stanford University, Stanford, CA,
USA}
  \affil[b]{Department of Linguistics, Stanford University, Stanford,
CA, USA}
  \affil[c]{Department of Psychology, University of Wisconsin, Madison,
WI, USA}
  \affil[d]{Department of Linguistics, University of California,
Berkeley, CA, USA}
  \affil[e]{Department of Communication, Stanford University, Stanford,
CA, USA}
  \affil[f]{Department of Communication, Stanford University, Stanford,
CA, USA}
  \affil[g]{Department of Psychology, Ludwig Maximilian University,
Munich, Germany}

% Please give the surname of the lead author for the running footer
\leadauthor{Anonymous}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{The efficiency with which children recognize
words provides a measurement of their real-time language processing,
which has been argued to be a key part of the language learning process.
Her, we use a large dataset of eye-tracking data from many different
experiments to map out the development of word recognition, finding that
this process becomes faster, more accurate, and less variable in the
period between one and six years of age, and that children who are
better at word recognition tend to have larger and faster growing
vocabularies. These data suggest that understanding language is a skill
that improves gradually with practice.}

\authorcontributions{Please provide details of author contributions
here.}

\correspondingauthor{\textsuperscript{1} To whom correspondence should
be addressed. E-mail:
\href{mailto:mcfrank@stanford.edu}{\nolinkurl{mcfrank@stanford.edu}}}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  one |  two |  optional |  optional |  optional  }

\begin{abstract}
Being a fluent language user involves recognizing words as they unfold
in time. How does this skill develop over the course of early childhood?
And how does facility in word recognition relate to the growth of
vocabulary knowledge? We address these questions using data from
Peekbank, an open database of experiments measuring children's
eye-movements during early word recognition. Combining 24 datasets from
almost 2,000 children ages 1--6 years, we show that word recognition
becomes faster, more accurate, and less variable across development.
Factor analysis reveals a cross-sectional coupling of word recognition
speed and accuracy with parent-reported vocabulary. Across a range of
models, speed, accuracy, and vocabulary also show coupled growth such
that children with faster initial word recognition tend to show faster
vocabulary growth. Together, these findings support the view that word
recognition is a skill that develops gradually across early childhood
and that this skill plays a role in supporting early language learning.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.

\acknow{This work was supported by a grant from the Jacobs Foundation.}

\dropcap{C}hildren acquiring a language are learning a body of knowledge
-- a set of words and the ways they are combined -- but they are also
learning to deploy this knowledge in the myriad complex, noisy, and
fast-moving environments in which language is used. As children enter
their second year, language explodes onto the scene; both vocabulary and
grammatical abilities grow rapidly and in tandem (1, 2). This growth in
knowledge is also accompanied by changes in language processing
efficiency: children become quicker and more accurate in recognizing
words and matching them with their referents (3--5).

Yet unlike language production, which is manifest via overt behavior,
evidence for word recognition is often more subtle. Very young children
with incomplete knowledge may not be able to point to the correct
referent of a word, but they may still have some representation of word
meaning (6). Eye tracking has thus emerged as a key method that allows
the measurement of language comprehension with high temporal resolution:
both adults and children reliably fixate the referent of a word soon
after it is used (3, 7--10). The relative timecourse of fixation then
can provide an index of the comprehender's ability or the difference
between two stimulus conditions.

The version of this method that is used with children goes by many
names, including the intermodal preferential looking paradigm and the
``looking while listening'' paradigm (LWL, the name we adopt here) (9,
11, 12). In LWL experiments, children are typically shown a series of
trials in which two images are displayed side by side and they are asked
to find one of them. For example, a ball and a book might be shown, and
the child might hear ``Look at the ball! Can you find it?''. Accuracy is
then computed as the proportion of time they fixate the correct image
within a fixed window after the onset of the noun (``ball'' in this
case). Reaction time is computed only on trials in which the child is
fixating on the distractor image (the book) before word onset; in these
cases, the average time it takes until the child shifts fixation from
the distractor to the target is used as an index of processing speed.
Early work using this method showed that both children's speed and
accuracy increase rapidly across the second year (3, 12). Related
methods have provided a window into how children process phonological
(13), syntactic (14), and semantic (15, 16) information as well as how
their lexical representations develop (17).

Word recognition ability, as measured by LWL, is hypothesized to play a
key role in language learning. Each word that a child experiences is an
opportunity to learn; measurements of children's language input at home
are consistently associated with their vocabulary size (18, 19).
Recognizing incoming words and linking them with their referents is a
prerequisite for learning. Consider a child hearing the utterance ``Can
you put the ball in the box?'' The faster and more accurately the child
can recognize that the ball is a referent, the better they can use this
evidence to help infer the speaker's intended meaning, perhaps making
inferences about the meaning of ``put'' or ``box'' (20). Consistent with
this idea, one important study found that children's word recognition
speed mediated the relationship between home language input and
vocabulary growth (21).

\begin{table*}[ht]
\centering
\begin{tabular}{rlrrrrrrrll}
  \hline
 & dataset\_name & N subjects & N admins & Mean Age & Min Age & Max Age & Avg Trials & Avg RT Trials & CDIs & longitudinal \\ 
  \hline
1 & reflook\_v4 & 310 & 310 & 36.82 & 12.24 & 60.00 & 6.21 & 2.87 &  &  \\ 
  2 & yurovsky\_2017 & 282 & 282 & 25.64 & 12.59 & 58.65 & 5.95 & 2.79 &  &  \\ 
  3 & weaver\_zettersten\_2024 & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 &  & x \\ 
  4 & fernald\_marchman\_2012 & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.89 & 7.21 & x & x \\ 
  5 & sander-montant\_2022 & 122 & 122 & 21.94 & 12.02 & 31.11 & 10.07 & 4.20 & x &  \\ 
  6 & frank\_tablet\_2016 & 104 & 104 & 33.89 & 12.13 & 59.84 & 5.87 & 2.74 &  &  \\ 
  7 & baumgartner\_2014 & 100 & 100 & 12.01 & 12.00 & 13.00 & 4.00 & 2.31 & x &  \\ 
  8 & fmw\_2013 &  80 & 179 & 20.04 & 17.00 & 26.00 & 21.26 & 8.44 & x & x \\ 
  9 & adams\_marchman\_2018 &  69 & 711 & 23.58 & 13.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
  10 & potter\_canine &  67 &  67 & 23.76 & 21.00 & 27.00 & 10.38 & 3.80 & x &  \\ 
  11 & fernald\_totlot &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.25 & 6.06 & x & x \\ 
  12 & pomper\_prime &  63 &  63 & 39.73 & 37.90 & 42.00 & 12.98 & 5.34 &  &  \\ 
  13 & pomper\_saffran\_2016 &  60 &  60 & 44.27 & 41.00 & 47.00 & 7.55 & 3.30 &  &  \\ 
  14 & swingley\_aslin\_2002 &  50 &  50 & 15.09 & 14.13 & 16.00 & 11.74 & 3.79 & x &  \\ 
  15 & pomper\_salientme &  44 &  44 & 40.11 & 38.00 & 43.00 & 5.30 & 2.34 &  &  \\ 
  16 & perry\_cowpig &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 & x &  \\ 
  17 & ronfard\_2021 &  40 &  40 & 19.95 & 18.00 & 24.00 & 18.54 & 7.62 & x &  \\ 
  18 & bacon\_gendercues &  38 &  38 & 22.87 & 22.00 & 24.00 & 18.16 & 8.00 & x &  \\ 
  19 & garrison\_bergelson\_2020 &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.47 & 9.41 & x &  \\ 
  20 & pomper\_yumme &  32 &  32 & 26.38 & 25.00 & 28.00 & 7.62 & 2.31 &  &  \\ 
  21 & mahr\_coartic &  29 &  29 & 20.83 & 18.10 & 23.80 & 24.38 & 8.86 & x &  \\ 
  22 & ferguson\_eyetrackingr &  28 &  28 & 19.71 & 18.02 & 21.86 & 5.54 & 2.33 & x &  \\ 
  23 & potter\_remix &  23 &  44 & 22.59 & 18.00 & 29.00 & 6.07 & 2.58 & x & x \\ 
  24 & newman\_genderdistractor &  19 &  19 & 30.12 & 29.23 & 30.84 & 13.11 & 4.94 &  &  \\ 
  \hline
    & Total & 1963 & 3554 & 24.73 & 12.00 & 60.00 & 12.74 & 5.05 & 15 & 6 \\ 
   \hline
\end{tabular}
\caption{Characteristics of included datasets from Peekbank. `Admins` denotes separate experimental sessions. `CDIs` refers to whether the dataset contains parent report vocabulary data from the MacArthur-Bates Communicative Development Inventory.} 
\end{table*}

Word recognition speed has been used as an index of individual
differences in early childhood (4, 22--25) and beyond (26--28). Over and
above measures of vocabulary size, word recognition speed at 18 months
predicts children's standardized test scores years later (23), though
these predictions may be limited to particular ages or processing
assessments (4). Further, faster processing at 18 months is predictive
of whether ``late talkers'' catch up to their peers or require further
intervention (24). Critically, all of these assessments use words that
children are reported to understand and produce -- they are not indices
of vocabulary size but rather of how quickly and accurately they can
recognize a spoken word and use it to guide their visual attention to a
referent.

Yet individual experiments measuring processing with young children
typically recruit relatively small samples in a restricted range of
ages. These samples provide neither the breadth of ages nor the number
of participants needed to estimate the broader dynamics of how word
recognition changes developmentally and how it connects with other
aspects of language {[}as has been done with school-aged children; (26);
(28){]}. Here we investigate two hypotheses.

First, one influential theory characterizes language learning as a
process of learning the skill of real-time processing (29, 30). Based on
this theory, we should expect to see the signatures of expertise and
skill learning in word recognition. Accuracy should change linearly with
the logarithm of age, reflecting gradual asymptotic convergence to
mature levels of accuracy. In addition, the logarithm of processing
speed should decrease with the logarithm of age, potentially reflecting
the ``power law of practice'' (31--33, cf 34). Finally, trial-to-trial
variability in both speed and accuracy should decrease with increasing
expertise (35).

Second, previous findings have provided limited and sometimes
conflicting evidence on the concurrent and predictive relations between
word recognition and language learning. Initial reports showed strong
predictive relationships between both speed and accuracy and later
vocabulary growth (22), with replications in infants born preterm (36)
and late talkers (24). Subsequent studies have primarily focused on
speed of processing and found more mixed results, with reaction time
measures found to be only inconsistently predictive of later vocabulary
outcomes (4, 25, 37). To the extent that there are consistent relations
between vocabulary and word recognition, these should be visible in a
larger dataset. Further, by examining the relationship between speed,
accuracy, and vocabulary, it should be possible to assess the extent to
which processing speed specifically plays a role in vocabulary growth.

\section*{Results}\label{results}
\addcontentsline{toc}{section}{Results}

To address these questions, we made use of a new release of data in
Peekbank, an open database of LWL from young children, stored in a
harmonized format (38). We retrieved data from children ages 1--6 years.
Although experiments in Peekbank include a variety of different
experimental manipulations, we analyzed data only from simple word
recognition trials in which children were shown two pictures of concrete
objects and heard a label for an object (typically embedded in a simple
carrier phrase such as ``Look at the \ldots{}''); these trials often
constituted control conditions for experiments. We focus here on English
purely for practical reasons -- the Peekbank dataset contains limited
data from other languages.

These criteria yielded 24 datasets, including 1963 children and 3554
administrations of the LWL procedure (some datasets were longitudinal or
involved multiple closely-spaced testing sessions). Table
\ref{tab:datasets} shows the characteristics of individual datasets (see
also Figures S1 and S2). The size of the combined dataset, the unified
data processing pipeline, and the fact that individual studies used very
similar implementations of the LWL experimental paradigm, all allowed us
to make a more detailed study of the development of word recognition
than has previously been possible. While our analysis is exploratory in
nature, it is guided by the two hypotheses outlined above: the presence
of 1) signatures of skill learning in word recognition, and 2) linkages
between word recognition and vocabulary.

\subsection*{Speed and accuracy of word recognition
increase}\label{speed-and-accuracy-of-word-recognition-increase}
\addcontentsline{toc}{subsection}{Speed and accuracy of word recognition
increase}

We began by characterizing developmental changes in speed and accuracy.
We computed both RTs (reaction times) and accuracies following standard
practices in the literature (9). Because there is no consensus about the
length of time windows for the computation of accuracy, we considered
both a shorter window (from 200 -- 2200 ms after noun onset) and a
longer window (from 200 -- 4000 ms). For each window, we averaged all
fixation within the window to compute a continuous proportion between 0
(no fixation on the target during the window) and 1 (total fixation on
the target during the window) on every trial.

Our first question was about the functional form of the relationships
between age, speed, and accuracy (see SI for raw correlations between
variables). To investigate this question, we fit linear mixed effects
models predicting accuracy and RT on each trial across the full dataset
with random slopes of child age by study and random intercepts by
participant. We compared models that included both long and short
accuracy windows, as well as logarithmic and linear effects of age, and
logarithmic and linear transformations of RT. The best fitting model of
accuracy predicted long window accuracy as a function of the logarithm
of age,\footnote{This longer window yielded overall higher cross-trial
  reliability as well, and so we report long window results in the main
  text. All findings reported here hold for both window sizes, however
  (see SI).} and the best fitting model of speed predicted log RT as a
function of log age as well (see SI).

Figure \ref{fig:devchange} shows these relationships. Log RT decreased
significantly with age (\(\hat{\beta} = -0.11\), 95\% CI
\([-0.14, -0.08]\), \(t(12.86) = -8.39\), \(p < .001\)) and accuracy
increased significantly with age (\(\hat{\beta} = 0.07\), 95\% CI
\([0.06, 0.08]\), \(t(17.34) = 12.77\), \(p < .001\)). In sum, we see
continuing improvements in word recognition across the full age range in
our dataset that appear roughly linear in the logarithm of age. These
logarithmic relationships follow theoretical expectations that both
speed and accuracy should gradually asymptote to mature levels of
performance as in skill learning more generally (31).

\begin{figure*}
\includegraphics{paper_files/figure-latex/devchange-1} \caption{Participant-level accuracy and reaction time (log), plotted by age (log). The solid blue line shows a linear fit and associated confidence interval. Individual dotted lines show linear fits for those datasets spanning six or more months of age. The dashed line for accuracy shows chance-level looking (.5)}\label{fig:devchange}
\end{figure*}

\subsection*{Variability of word recognition
decreases}\label{variability-of-word-recognition-decreases}
\addcontentsline{toc}{subsection}{Variability of word recognition
decreases}

One further hallmark of increasing skill is a decrease in task-relevant
variability (35). Both within and across datasets, within-individual
variation in accuracy and RT decreased smoothly across the developmental
range we examined (Figure \ref{fig:variance}). We fit mixed effects
models to the standard deviation of both accuracy and RT for each
testing session for each participant, including random slopes of log age
by dataset and random intercepts for each participant. For both speed
and accuracy, within-individual variability decreased with age (RT:
\(\hat{\beta} = -0.03\), 95\% CI \([-0.05, -0.02]\),
\(t(12.48) = -5.30\), \(p < .001\); accuracy: \(\hat{\beta} = -0.03\),
95\% CI \([-0.04, -0.03]\), \(t(13.52) = -11.90\), \(p < .001\)). Thus,
as well as being faster and more accurate, older children were more
consistent in their online word recognition.

\begin{figure*}
\includegraphics{paper_files/figure-latex/variance-1} \caption{Participant-level variability in accuracy and reaction time (log RT), plotted by age (log). Plotting conventions are as in Figure 1.}\label{fig:variance}
\end{figure*}

\subsection*{Speed and accuracy relate to vocabulary
size}\label{speed-and-accuracy-relate-to-vocabulary-size}
\addcontentsline{toc}{subsection}{Speed and accuracy relate to
vocabulary size}

We were next interested in whether the various aspects of word
recognition -- including accuracy, RT, and the variability of each of
these -- were related to other aspects of early language ability. Of the
studies in our database, 16 gathered parent reports about children's
early vocabulary using the MacArthur-Bates Communicative Development
Inventory (CDI), a popular survey instrument that provides a reliable
and valid estimate of children's early vocabulary (2, 39). Different
forms of the CDI can be used to measure either receptive and expressive
vocabulary (for children up to 18 months) or expressive vocabulary only
(for children 16 -- 30 months).

We fit a series of factor analytic models to explore the dimensionality
of the data. Initial exploratory factor analysis using parallel analysis
suggested that three factors explained substantial variance in the data
(see SI: Factor Analysis). Due to missingness of data, we used
confirmatory factor analysis with full information maximum likelihood to
find the best set of loadings. The best fitting model was a three-factor
model with factors for speed (RT and RT variability), accuracy (accuracy
and accuracy variability), and vocabulary (comprehension and production
from the CDI). Fit statistics for this model were generally good
(Confirmatory fit index: 0.975, RMSE: 0.06); see SI: Alternative Factor
Structures).

\begin{figure*}
\includegraphics{paper_files/figure-latex/age_regression-1} \caption{Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age. Observed variables are notated as squares and latent variables are notated as circles Factor loadings and regression coefficients are shown with straight, solid lines; covariances are shown with dashed lines; residual variances are shown as solid circular connections. Stars show conventional levels of statistical significance, e.g. * indicates p < .05, ** indicates p < .01, and *** indicates p < .001. Covariances reflect age-residualized correlations between variables. Abbreviations: acc = accuracy, prod = production vocabulary, comp = comprehension vocabulary.}\label{fig:age_regression}
\end{figure*}

Figure \ref{fig:age_regression} shows a regression model fit to this
confirmatory factor analysis, with log age predicting each latent
variable. This regression model allows interpretation of the covariances
between latent factors as partial correlations (controlling for age).
All three latent factors were significantly related to one another, with
RT and accuracy showing strong negative covariance (\(\beta=\) -0.734,
SE = 0.034, \(p < .0001\)) and weaker but significant covariation
between RT and vocabulary (\(\beta=\) -0.294, SE = 0.036, \(p < .0001\))
and accuracy and vocabulary (\(\beta=\) 0.369, SE = 0.033,
\(p < .0001\)). This model supports the idea that individual variation
in speed and accuracy of word recognition is concurrently related to
vocabulary beyond the effects of age.

\subsection*{Speed of processing relates to vocabulary
growth}\label{speed-of-processing-relates-to-vocabulary-growth}
\addcontentsline{toc}{subsection}{Speed of processing relates to
vocabulary growth}

While we fit the latent variable models above to the full
cross-sectional dataset, we were most interested in within-child
developmental changes, as measured within individual studies. To probe
these changes, we first examined test-retest reliability for our primary
variables of interest by calculating Pearson correlations between pairs
of administrations given no more than three months apart. Test-retest
correlations were significant but relatively modest
(\(\rho_{acc} = 0.455\), \(\rho_{rt} =  0.407\)), suggesting that,
across studies, LWL measures from individual sessions provide relatively
noisy measurements of individual children even if individual studies are
able to achieve higher reliability by averaging across measurement
occasions (e.g., 23).

Given this relatively low reliability in our longitudinal datasets
(which also tended to be from younger children), we chose to measure
relationships between LWL and vocabulary using longitudinal growth
models. We began by reproducing the analysis reported in (24), in which
longitudinal growth in productive vocabulary was predicted based on RT
during the initial session of the study. We fit a mixed effects model
predicting growth in vocabulary as a quadratic function of age, RT at
study initiation (\(t_0\)), and their interaction (as well as random
effects of age by participant and by dataset). This model revealed a
significant effect of \(t_0\) RT (\(\hat{\beta} = -0.14\), 95\% CI
\([-0.21, -0.06]\), \(t(294.12) = -3.61\), \(p < .001\)) and an
interaction between \(t_0\) RT and the quadratic age predictor
(\(\hat{\beta} = 1.44\), 95\% CI \([0.76, 2.11]\),
\(t(1036.95) = 4.15\), \(p < .001\)). This analysis suggests that
children with faster initial RTs show both larger vocabularies and
faster vocabulary growth over time, confirming that findings from (24)
were robust across the full dataset.

On the other hand, it is possible that differences in predicted growth
trajectories are due to overall coupling between vocabulary size and
language processing, rather than a specifically predictive relationship
between \(t_0\) RT and vocabulary growth. To test this relationship, we
used longitudinal structural equation models. We separated the
longitudinal speed, accuracy, and vocabulary data into two-month bins
spanning up to 10 months (\(t_0 ... t_4\)) and fit individual growth
across each of these variables. We used full-information maximum
likelihood to handle the substantial missing data caused by the
different longitudinal sampling schemes of studies in our dataset. (We
present a model of coupled growth in the observed variables here for
simplicity; see SI for a comparable model using growth in the latent
factors). The fitted longitudinal model is shown in Figure
\ref{fig:longitudinal}. Overall fit statistics were generally acceptable
(Confirmatory fit index: 0.883, RMSE: 0.028, RMSE \(p\)-value: 1).

Our key question of interest concerned coupling between the parameters
of these growth models. Consistent with the idea that overall faster
processing is related to vocabulary growth, we saw significant coupling
between processing speed intercepts and vocabulary growth slopes
(\(\beta=\) -0.136, SE = 0.052, \(p =\) 0.008) as well as a variety of
other couplings. There was not significant coupling between growth in RT
and growth in vocabulary (\(\beta=\) -0.006, SE = 0.014, \(p =\) 0.674).
These abilities might grow independently, but we cannot rule out other
possibilities. First, the longitudinal data we had might not allow
sufficiently precise estimates of growth slopes, or second, since
vocabulary growth is non-linear, the linear model we used here might not
be as sensitive to non-linear changes.

In sum, these findings provide further evidence consistent with the
claim that faster processing is related to longitudinal growth in
vocabulary (21, 22). Children with greater skill in word recognition
learn words faster.

\begin{figure*}
\includegraphics{paper_files/figure-latex/longitudinal-1} \caption{Structural equation model showing longitudinal couplings between growth parameters. }\label{fig:longitudinal}
\end{figure*}

\section*{Discussion}\label{discussion}
\addcontentsline{toc}{section}{Discussion}

How does word recognition change across early childhood and how does it
relate to language learning? We investigated these questions using a
new, large-scale dataset of developmental eye-tracking measurements
compiled across many prior studies. We found continuous developmental
changes from ages 1 -- 6 years. Speed and accuracy both improved
asymptotically, with evidence that recognition speed showed the log-log
relationship associated with the ``power law of practice'', that is,
gradually converging on mature levels of processing efficiency. Further,
trial-to-trial variability decreased, consistent with both the
literature on skill learning (35) and other work on developmental
changes in variability (40--42). Speed and accuracy were both related to
vocabulary size concurrently and processing speed was also related to
later vocabulary growth.

Together, our findings are consistent with theories that posit that
language learning is a process of skill acquisition, in which children
become adept at quickly converting ephemeral signals into meaning (29).
This skill develops gradually over the course of early childhood and
supports word learning. Further, our results point to consistency
between skill development in early childhood and the continued
refinement of language processing and language knowledge during middle
childhood (26, 28).

By aggregating data from many pre-existing studies, we were able to
overcome the limitations of prior investigations, which typically had
sample sizes at least an order of magnitude smaller than ours. In
contrast to individual studies, which typically have at best the
statistical power to test one or two specific contrasts, our ``big
data'' approach provided the sample sizes necessary to fit complex
structural equation models and to compare different functional forms for
developmental change. Because early language is so variable (2), these
kinds of samples -- with thousands, rather than dozens of children --
are likely to be required to gain further insight into the psychometrics
of early language learning
{[}\href{mailto:bergelson2023;@frank2017}{\nolinkurl{bergelson2023;@frank2017}};(2);(43){]}.

At the same time, our approach is both observational and exploratory.
Thus, we cannot untangle the range of different causal models that
explain the variation we observed. First, early word recognition skill
could lead to faster word learning, due to skilled children gleaning
more information from the same signal (22). But second, faster children
could also be faster due to their larger vocabulary and stronger lexical
representations. These two causal directions could also interact
reciprocally, leading to a ``rich get richer'' process in which children
with larger vocabularies process faster, and their faster processing
helps them increase their vocabulary size more rapidly. Finally, a third
shared factor -- perhaps general cognitive ability -- could underpin
both processes. Our cross-sectional data cannot distinguish these
hypotheses even in principle (44), and our longitudinal data are likely
too sparse to distinguish such complex causal models. Thus, these
questions remain an open target for dense longitudinal data collection.

Our findings here are limited in their generalizability by the
convenience samples that were used in the individual studies in the
Peekbank database. These studies typically (but not always) represent
children from well-educated parents living in university-adjacent
communities. We would not expect that specific numerical parameters
estimated in this aggregate convenience sample would generalize to other
samples. Nevertheless, the consistency of the trends we observed across
datasets suggests that our qualitative conclusions are robust to some
significant sociodemographic variation.

More broadly, our results here suggest the continued importance of the
looking-while-listening paradigm as an index of children's language
processing abilities. If language learning is, at least in part, a
process of skill learning, then measurement of this skill is a critical
window into understanding the remarkable process of language learning.

\showmatmethods
\showacknow
\pnasbreak

\newpage

\phantomsection\label{refs}
\begin{CSLReferences}{0}{1}
\bibitem[\citeproctext]{ref-bates1994developmental}
\CSLLeftMargin{1. }%
\CSLRightInline{Bates E, et al. (1994) Developmental and stylistic
variation in the composition of early vocabulary. \emph{Journal of child
language} 21(1):85--123.}

\bibitem[\citeproctext]{ref-frank2021}
\CSLLeftMargin{2. }%
\CSLRightInline{Frank MC, Braginsky M, Yurovsky D, Marchman VA (2021)
\emph{{Variability and Consistency in Early Language Learning: The
Wordbank Project}} (MIT Press, Cambridge, MA).}

\bibitem[\citeproctext]{ref-fernald1998}
\CSLLeftMargin{3. }%
\CSLRightInline{Fernald A, Pinto JP, Swingley D, Weinberg A, McRoberts
GW (1998) Rapid gains in speed of verbal processing by infants in the
2nd year. \emph{Psychological Science} 9(3):228--231.}

\bibitem[\citeproctext]{ref-peter2019}
\CSLLeftMargin{4. }%
\CSLRightInline{Peter MS, et al. (2019) Does speed of processing or
vocabulary size predict later language growth in toddlers?
\emph{Cognitive Psychology} 115:101238.}

\bibitem[\citeproctext]{ref-bergelson2020comprehension}
\CSLLeftMargin{5. }%
\CSLRightInline{Bergelson E (2020) The comprehension boost in early word
learning: Older infants are better learners. \emph{Child development
perspectives} 14(3):142--149.}

\bibitem[\citeproctext]{ref-bergelson2012}
\CSLLeftMargin{6. }%
\CSLRightInline{Bergelson E, Swingley D (2012)
\href{https://www.ncbi.nlm.nih.gov/pubmed/22331874}{{At 6-9 months,
human infants know the meanings of many common nouns.}}
\emph{Proceedings of the National Academy of Sciences}
109(9):3253--3258.}

\bibitem[\citeproctext]{ref-tanenhaus1995integration}
\CSLLeftMargin{7. }%
\CSLRightInline{Tanenhaus MK, Spivey-Knowlton MJ, Eberhard KM, Sedivy JC
(1995) Integration of visual and linguistic information in spoken
language comprehension. \emph{Science} 268(5217):1632--1634.}

\bibitem[\citeproctext]{ref-altmann1999incremental}
\CSLLeftMargin{8. }%
\CSLRightInline{Altmann GT, Kamide Y (1999) Incremental interpretation
at verbs: Restricting the domain of subsequent reference.
\emph{Cognition} 73(3):247--264.}

\bibitem[\citeproctext]{ref-fernald2008}
\CSLLeftMargin{9. }%
\CSLRightInline{Fernald A, Zangl R, Portillo AL, Marchman VA (2008)
{Looking while listening: Using eye movements to monitor spoken language
comprehension by infants and young children}. \emph{Developmental
Psycholinguistics: On-Line Methods in Children's Language Processing},
eds Sekerina IA, Fernandez EM, Clahsen H (John Benjamins, Amsterdam), pp
97--135.}

\bibitem[\citeproctext]{ref-macdonald2018}
\CSLLeftMargin{10. }%
\CSLRightInline{MacDonald K, LaMarr T, Corina D, Marchman VA, Fernald A
(2018) Real-time lexical comprehension in young children learning
american sign language. \emph{Developmental science} 21(6):e12672.}

\bibitem[\citeproctext]{ref-hirsh1996intermodal}
\CSLLeftMargin{11. }%
\CSLRightInline{Hirsh-Pasek K, Golinkoff RM (1996) The intermodal
preferential looking paradigm: A window onto emerging language
comprehension.}

\bibitem[\citeproctext]{ref-reznick1990visual}
\CSLLeftMargin{12. }%
\CSLRightInline{Reznick JS (1990) Visual preference as a test of infant
word comprehension. \emph{Applied Psycholinguistics} 11(2):145--166.}

\bibitem[\citeproctext]{ref-mani2011}
\CSLLeftMargin{13. }%
\CSLRightInline{Mani N, Plunkett K (2011) Phonological priming and
cohort effects in toddlers. \emph{Cognition} 121(2):196--206.}

\bibitem[\citeproctext]{ref-trueswell1999kindergarten}
\CSLLeftMargin{14. }%
\CSLRightInline{Trueswell JC, Sekerina I, Hill NM, Logrip ML (1999) The
kindergarten-path effect: Studying on-line sentence processing in young
children. \emph{Cognition} 73(2):89--134.}

\bibitem[\citeproctext]{ref-mani2010}
\CSLLeftMargin{15. }%
\CSLRightInline{Mani N, Plunkett K (2010) In the infant's mind's ear:
Evidence for implicit naming in 18-month-olds. \emph{Psychological
science} 21(7):908--913.}

\bibitem[\citeproctext]{ref-bergelson2017}
\CSLLeftMargin{16. }%
\CSLRightInline{Bergelson E, Aslin RN (2017) Nature and origins of the
lexicon in 6-mo-olds. \emph{Proceedings of the National Academy of
Sciences} 114(49):12916--12921.}

\bibitem[\citeproctext]{ref-Swingley2002}
\CSLLeftMargin{17. }%
\CSLRightInline{Swingley D, Aslin RN (2002)
\href{https://doi.org/10.1111/1467-9280.00485}{{Lexical neighborhoods
and the word-form representations of 14-month-olds}}.
\emph{Psychological Science} 13(5):480--484.}

\bibitem[\citeproctext]{ref-hart1995meaningful}
\CSLLeftMargin{18. }%
\CSLRightInline{Hart B, Risley TR (1995) Meaningful differences in the
everyday experience of young american children.}

\bibitem[\citeproctext]{ref-anderson2021linking}
\CSLLeftMargin{19. }%
\CSLRightInline{Anderson NJ, Graham SA, Prime H, Jenkins JM, Madigan S
(2021) Linking quality and quantity of parental linguistic input to
child language skills: A meta-analysis. \emph{Child Development}
92(2):484--501.}

\bibitem[\citeproctext]{ref-frank2009using}
\CSLLeftMargin{20. }%
\CSLRightInline{Frank MC, Goodman ND, Tenenbaum JB (2009) Using
speakers' referential intentions to model early cross-situational word
learning. \emph{Psychological science} 20(5):578--585.}

\bibitem[\citeproctext]{ref-Weisleder2013}
\CSLLeftMargin{21. }%
\CSLRightInline{Weisleder A, Fernald A (2013)
\href{https://doi.org/10.1177/0956797613488145}{{Talking to children
matters: Early language experience strengthens processing and builds
vocabulary}}. \emph{Psychological Science} 24(11):2143--2152.}

\bibitem[\citeproctext]{ref-fernald2006}
\CSLLeftMargin{22. }%
\CSLRightInline{Fernald A, Perfors A, Marchman VA (2006) Picking up
speed in understanding: Speech processing efficiency and vocabulary
growth across the 2nd year. \emph{Developmental psychology} 42(1):98.}

\bibitem[\citeproctext]{ref-marchman2008}
\CSLLeftMargin{23. }%
\CSLRightInline{Marchman VA, Fernald A (2008)
\href{https://www.ncbi.nlm.nih.gov/pubmed/18466367}{{Speed of word
recognition and vocabulary knowledge in infancy predict cognitive and
language outcomes in later childhood.}} \emph{Developmental science}
11(3):F9--16.}

\bibitem[\citeproctext]{ref-fernald2012}
\CSLLeftMargin{24. }%
\CSLRightInline{Fernald A, Marchman VA (2012)
\href{https://www.ncbi.nlm.nih.gov/pubmed/22172209}{{Individual
differences in lexical processing at 18 months predict vocabulary growth
in typically developing and late-talking toddlers.}} \emph{Child
development} 83(1):203--22.}

\bibitem[\citeproctext]{ref-newbury2016interrelationships}
\CSLLeftMargin{25. }%
\CSLRightInline{Newbury J, Klee T, Stokes SF, Moran C (2016)
Interrelationships between working memory, processing speed, and
language development in the age range 2--4 years. \emph{Journal of
Speech, Language, and Hearing Research} 59(5):1146--1158.}

\bibitem[\citeproctext]{ref-colby2023efficiency}
\CSLLeftMargin{26. }%
\CSLRightInline{Colby SE, McMurray B (2023) Efficiency of spoken word
recognition slows across the adult lifespan. \emph{Cognition}
240:105588.}

\bibitem[\citeproctext]{ref-jeppsen2022development}
\CSLLeftMargin{27. }%
\CSLRightInline{Jeppsen C, Baxelbaum K, Tomblin B, Klein K, McMurray B
(2022) The development of lexical processing: Real-time phonological
competition and semantic activation in school age children.
\emph{Quarterly Journal of Experimental Psychology}:17470218241244799.}

\bibitem[\citeproctext]{ref-mcmurray2022slow}
\CSLLeftMargin{28. }%
\CSLRightInline{McMurray B, Apfelbaum KS, Tomblin JB (2022) The slow
development of real-time processing: Spoken-word recognition as a
crucible for new thinking about language acquisition and language
disorders. \emph{Current Directions in Psychological Science}
31(4):305--315.}

\bibitem[\citeproctext]{ref-christiansen2016}
\CSLLeftMargin{29. }%
\CSLRightInline{Christiansen MH, Chater N (2016) The now-or-never
bottleneck: A fundamental constraint on language. \emph{Behavioral and
brain sciences} 39:e62.}

\bibitem[\citeproctext]{ref-chater2018}
\CSLLeftMargin{30. }%
\CSLRightInline{Chater N, Christiansen MH (2018) Language acquisition as
skill learning. \emph{Current opinion in behavioral sciences}
21:205--208.}

\bibitem[\citeproctext]{ref-snoddy1926}
\CSLLeftMargin{31. }%
\CSLRightInline{Snoddy GS (1926) Learning and stability: A
psychophysiological analysis of a case of motor learning with clinical
applications. \emph{Journal of Applied Psychology} 10(1):1.}

\bibitem[\citeproctext]{ref-kail1991}
\CSLLeftMargin{32. }%
\CSLRightInline{Kail R (1991) Processing time declines exponentially
during childhood and adolescence. \emph{Developmental psychology}
27(2):259.}

\bibitem[\citeproctext]{ref-anderson1982}
\CSLLeftMargin{33. }%
\CSLRightInline{Anderson JR (1982) Acquisition of cognitive skill.
\emph{Psychological review} 89(4):369.}

\bibitem[\citeproctext]{ref-heathcote2000}
\CSLLeftMargin{34. }%
\CSLRightInline{Heathcote A, Brown S, Mewhort DJ (2000) The power law
repealed: The case for an exponential law of practice. \emph{Psychonomic
bulletin \& review} 7(2):185--207.}

\bibitem[\citeproctext]{ref-todorov2002optimal}
\CSLLeftMargin{35. }%
\CSLRightInline{Todorov E, Jordan MI (2002) Optimal feedback control as
a theory of motor coordination. \emph{Nature neuroscience}
5(11):1226--1235.}

\bibitem[\citeproctext]{ref-marchman2016}
\CSLLeftMargin{36. }%
\CSLRightInline{Marchman VA, Adams KA, Loi EC, Fernald A, Feldman HM
(2016) Early language processing efficiency predicts later receptive
vocabulary outcomes in children born preterm. \emph{Child
Neuropsychology} 22(6):649--665.}

\bibitem[\citeproctext]{ref-lany2018}
\CSLLeftMargin{37. }%
\CSLRightInline{Lany J (2018) Lexical-processing efficiency leverages
novel word learning in infants and toddlers. \emph{Developmental
science} 21(3):e12569.}

\bibitem[\citeproctext]{ref-zettersten2023peekbank}
\CSLLeftMargin{38. }%
\CSLRightInline{Zettersten M, et al. (2023) Peekbank: An open,
large-scale repository for developmental eye-tracking data of children's
word recognition. \emph{Behavior Research Methods} 55(5):2485--2500.}

\bibitem[\citeproctext]{ref-marchman2023}
\CSLLeftMargin{39. }%
\CSLRightInline{Marchman VA, Dale PS, Fenson L (2023)
\emph{MacArthur-bates communicative development inventories userś guide
and technical manual, third edition} (Brookes).}

\bibitem[\citeproctext]{ref-judd2021working}
\CSLLeftMargin{40. }%
\CSLRightInline{Judd N, Klingberg T, Sjöwall D (2021) Working memory
capacity, variability, and response to intervention at age 6 and its
association to inattention and mathematics age 9. \emph{Cognitive
Development} 58:101013.}

\bibitem[\citeproctext]{ref-galeano2018variability}
\CSLLeftMargin{41. }%
\CSLRightInline{Galeano Weber EM, Dirk J, Schmiedek F (2018) Variability
in the precision of children's spatial working memory. \emph{Journal of
Intelligence} 6(1):8.}

\bibitem[\citeproctext]{ref-kautto2023introducing}
\CSLLeftMargin{42. }%
\CSLRightInline{Kautto A, Railo H, Mainela-Arnold E (2023) Introducing
the intra-individual variability hypothesis in explaining individual
differences in language development.}

\bibitem[\citeproctext]{ref-manybabies2020}
\CSLLeftMargin{43. }%
\CSLRightInline{The ManyBabies Consortium (2020) {Quantifying sources of
variability in infancy research using the infant-directed speech
preference}. \emph{Advances in Methods and Practices in Psychological
Science} 3(1):24--52.}

\bibitem[\citeproctext]{ref-vanderweele2023dimensional}
\CSLLeftMargin{44. }%
\CSLRightInline{VanderWeele TJ, Batty CJ (2023) On the dimensional
indeterminacy of one-wave factor analysis under causal effects.
\emph{Journal of Causal Inference} 11(1):20220074.}

\end{CSLReferences}

% Bibliography
% \bibliography{pnas-sample}

\end{document}

### 2025-03-13 paper.R
fix paths for rendering supplement

@@ -1,0 +1,558 @@
+suppressPackageStartupMessages(library(here))
+suppressPackageStartupMessages(library(psych))
+suppressPackageStartupMessages(library(lmerTest))
+suppressPackageStartupMessages(library(lavaan))
+suppressPackageStartupMessages(library(tidySEM))
+suppressPackageStartupMessages(library(xtable))
+suppressPackageStartupMessages(library(papaja))
+
+source(here("helper","common.R"))
+
+source(here("helper/modified_sem_plot.R"))
+
+d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
+  mutate(log_age = log(age))
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds")) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(admin_num = 1:n(), 
+         time_since_t0 = age - age[1],
+         delta_t = c(0, diff(age)))
+
+d_trial$age_s <- scale(d_trial$age)[,1]
+d_trial$log_age_s <- scale(d_trial$log_age)[,1]   
+d_sub$age_s <- scale(d_sub$age)[,1]
+d_sub$log_age_s <- scale(d_sub$log_age)[,1]   
+
+
+AGEVAR_CUTOFF <- 6
+
+knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
+
+dataset_characteristics <- d_sub |>
+  group_by(dataset_name) |>
+  summarise(`N subjects` = length(unique(subject_id)),
+            `N admins` = n(), 
+            `Mean Age` = mean(age), 
+            `Min Age` = min(age), 
+            `Max Age` = max(age), 
+            `Avg Trials` = mean(n_trials[!is.na(long_window_accuracy)]),
+            `Avg RT Trials` = mean(n_trials_rt[!is.na(rt)]),
+            `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
+            `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
+  arrange(desc(`N subjects`)) 
+
+dataset_characteristics <- dataset_characteristics |>
+  bind_rows(tibble(dataset_name = "Total", 
+                       `N subjects` = sum(dataset_characteristics$`N subjects`),
+                       `N admins` = sum(dataset_characteristics$`N admins`), 
+                       `Mean Age` = mean(dataset_characteristics$`Mean Age`),
[…]

### 2025-03-13 paper/paper.Rmd
fix a citation rendering issue

@@
 In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to fit complex structural equation models and to compare different functional forms for developmental change.
-Because early language is so variable [@frank2021], these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [bergelson2023;@frank2017;@frank2021;@manybabies2020].
+Because early language is so variable [@frank2021], these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [@bergelson2023;@frank2017;@frank2021;@manybabies2020].
 

### 2025-03-13 paper/paper.tex
fix a citation rendering issue

@@
 which has been argued to be a key part of the language learning process.
-Her, we use a large dataset of eye-tracking data from many different
+Here, we use a large dataset of eye-tracking data from many different
 experiments to map out the development of word recognition, finding that
@@
 % Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
- \keywords{  one |  two |  optional |  optional |  optional  }
+ \keywords{  language development |  word recognition |  developmental
+change |  skill acquisition |  eye-tracking  }
 
@@
 are likely to be required to gain further insight into the psychometrics
-of early language learning
-{[}\href{mailto:bergelson2023;@frank2017}{\nolinkurl{bergelson2023;@frank2017}};(2);(43){]}.
+of early language learning (2, 43--45).
 
@@
 both processes. Our cross-sectional data cannot distinguish these
-hypotheses even in principle (44), and our longitudinal data are likely
+hypotheses even in principle (46), and our longitudinal data are likely
 too sparse to distinguish such complex causal models. Thus, these
@@
 
+\bibitem[\citeproctext]{ref-bergelson2023}
+\CSLLeftMargin{43. }%
+\CSLRightInline{Bergelson E, et al. (2023) Everyday language input and
+production in 1,001 children from six continents. \emph{Proceedings of
+the National Academy of Sciences of the United States of America}
+120(52):e2300671120.}
+
+\bibitem[\citeproctext]{ref-frank2017}
+\CSLLeftMargin{44. }%
+\CSLRightInline{Frank MC, Braginsky M, Yurovsky D, Marchman VA (2017)
+{Wordbank: An open repository for developmental vocabulary data}.
+\emph{Journal of Child Language} 44(3):677--694.}
+
 \bibitem[\citeproctext]{ref-manybabies2020}
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{The ManyBabies Consortium (2020) {Quantifying sources of
@@
 \bibitem[\citeproctext]{ref-vanderweele2023dimensional}
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{VanderWeele TJ, Batty CJ (2023) On the dimensional

### 2025-03-13 paper/paper.Rmd
nlme

@@
 Critically, all of these assessments use words that children are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.
+Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
+These samples provide neither the breadth of ages nor the number of participants needed to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of early language development (see [@colby2023efficiency; @mcmurray2022slow] for examples of these analyses in school-aged children).
+To overcome this limitation, we turned to Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank].
+This resource allows us to gain an unprecedented view of the development of word recognition across a large sample of children.
 
-Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
-These samples provide neither the breadth of ages nor the number of participants needed to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of language [as has been done with school-aged children; @colby2023efficiency; @mcmurray2022slow].
-Here we investigate two hypotheses.
-
+Here we investigate two specific hypotheses.
 First, one influential theory characterizes language learning as a process of learning the skill of real-time processing [@christiansen2016;@chater2018].
@@
 suppressPackageStartupMessages(library(papaja))
+suppressPackageStartupMessages(library(viridis))
+suppressPackageStartupMessages(library(nlme))
 
@@
 
-To address these questions, we made use of a new release of data in Peekbank, an open database of LWL from young children, stored in a harmonized format [@zettersten2023peekbank].
-We retrieved data from children ages 1--6 years.
+We retrieved data from Peekbank, focusing on data from children ages 1--6 years.
 Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ..."); these trials often constituted control conditions for experiments.
@@
               aes(group = dataset_name, col = dataset_name), se=FALSE,
-              alpha = 1, method = "lm", size = .5, lty = 2) + 
+              alpha = 1, method = "lm", size = .5, lty = 3) + 
     geom_smooth(method = "lm") +
-  geom_hline(yintercept = .5, lty = 2) + 
+  geom_hline(yintercept = .5, lty = 3) + 
   scale_x_log10(breaks = c(12,24,36,48,60)) +
@@
               aes(group = dataset_name, col = dataset_name), se=FALSE,
-              alpha = .1, method = "lm", size = .5, lty = 2) + 
+              alpha = .1, method = "lm", size = .5, lty = 3) + 
     geom_smooth(method = "lm") +
@@
 cowplot::plot_grid(a,b)
-```
-
-```{r growth, fig.cap="TODO veronica messing around.
-Dark blue is fastest RT, dark red is slowest RT -- not sure what the ideal legend is.", fig.env="figure*", fig.height = 2.5, fig.width=4, out.width=".5\\textwidth", include=FALSE}
-
-library(nlme)
-
-datasets_with_age_variation <- d_trial |>
-  group_by(dataset_name) |>
-  summarise(age_range = max(age) - min(age)) |>
-  filter(age_range >= AGEVAR_CUTOFF)
-
-  d_sub_prod <- filter(d_sub, !is.na(prod)) |>
-  group_by(subject_id) |>
-  mutate(log_rt_0 = log_rt[1],
-         acc_0 = long_window_accuracy[1]) |> 
-  filter(!is.na(log_rt_0))
-
-mod3 <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
-             data = d_sub_prod,
-             fixed = xmid + scale ~ log_rt_0,
-             random = xmid + scale ~ 1,
-             groups = ~ dataset_name,
-             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 1, `scale:log_rt_0` = 3),
-             na.action = na.exclude, 
-             control = list(maxIter = 1000, tolerance = 10))
-
-# fixed effects (no CIs)
-global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
-                                      max(d_sub_prod$age), .1), 
-                            log_rt_0 = c(quantile(d_sub_prod$log_rt, 
-                                                  c(.1,.25,.5,.75,.9), 
-                                                  na.rm=TRUE)))
-global_preds$pred <- predict(mod3, newdata = global_preds, level = 0)
-
-#random effects
-dataset_preds <- d_sub |>
-  filter(!is.na(prod)) |> # filter on predictor
-  group_by(dataset_name) |>
-  summarise(min_age = min(age),
-            max_age = max(age)) |>
-  group_by(dataset_name) |>
-  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0 = c(quantile(d_sub_prod$log_rt, 
-                                                  c(.1,.25,.5,.75,.9), 
-                                                  na.rm=TRUE))))) |>
-  select(-min_age, -max_age) |>
-  unnest(col = "df") |>
-  ungroup()
-
-library(viridis)
-dataset_preds$pred <- predict(mod3,
-                              newdata = dataset_preds,
-                              level = c(0,1))$predict.dataset_name
-
-ggplot(d_sub_prod, aes(x = age, y = prod, col = as.factor(log_rt_0))) + 
-  geom_point(alpha = .02, color="black") + 
-  geom_line(data = dataset_preds |>   filter(dataset_name %in% datasets_with_age_variation$dataset_name), aes(y = pred, group = interaction(log_rt_0, dataset_name)), lty="dashed", alpha=.5)+
-    geom_line(data = global_preds, aes(y = pred, group = log_rt_0, col = as.factor(log_rt_0)), size=1)+
-  scale_color_viridis(discrete=T, option="H")+
-  theme(legend.position = "none")+
-  labs(y="Production", x="Age in months")
-# +
-#   geom_line(data = dataset_preds, aes(y = pred))
 ```
@@
 
-growth_mod <- lmer(prod ~ poly(age_15,2) *rt_t0 + (age  | subject_id) + (age | dataset_name), 
+# note that it's singular to add an age by dataset random effect
+growth_mod <- lmer(prod ~ poly(age_15,2) *rt_t0 + (age  | subject_id) + (1 | dataset_name), 
      data = d_growth, 
@@
 This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time, confirming that findings from @fernald2012 were robust across the full dataset.
+We confirmed this analysis using a non-linear growth model with a logistic shape, which shows a better fit to vocabulary size than the quadratic model (see SI: Non-Linear Growth Model).
+Figure \ref{fig:growth} shows predictions from this model, confirming the differentiation of growth curves for children with higher and lower initial reaction time.
+
+```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for the 25th and 75th percentile on initial reaction times.
+Solid lines show global model estimates and dashed lines show estimates for individual datasets with sufficient age variation as above.", fig.env="figure", fig.height = 2.5, fig.width=4, out.width=".5\\textwidth", include=TRUE}
+
+d_sub_prod <- filter(d_sub, !is.na(prod)) |>
+  group_by(subject_id) |>
+  mutate(log_rt_0 = log_rt[1],
+         acc_0 = long_window_accuracy[1]) |> 
+  filter(!is.na(log_rt_0))
+
+mod_nlme <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
+             data = d_sub_prod,
+             fixed = xmid + scale ~ log_rt_0,
+             random = xmid + scale ~ 1,
+             groups = ~ dataset_name,
+             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 1, `scale:log_rt_0` = 3),
+             na.action = na.exclude, 
+             control = list(maxIter = 1000, tolerance = 10))
+
+# fixed effects (no CIs)
+global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
+                                      max(d_sub_prod$age), .1), 
+                            log_rt_0 = c(quantile(d_sub_prod$log_rt, 
+                                                  c(.25,.75), 
+                                                  na.rm=TRUE)))
+global_preds$pred <- predict(mod_nlme, newdata = global_preds, level = 0)
+
+#random effects
+dataset_preds <- d_sub |>
+  filter(!is.na(prod)) |> # filter on predictor
+  group_by(dataset_name) |>
+  summarise(min_age = min(age),
+            max_age = max(age)) |>
+  group_by(dataset_name) |>
+  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0 = c(quantile(d_sub_prod$log_rt, 
+                                                  c(.25,.75), 
+                                                  na.rm=TRUE))))) |>
+  select(-min_age, -max_age) |>
+  unnest(col = "df") |>
+  ungroup()
+
+dataset_preds$pred <- predict(mod_nlme,
+                              newdata = dataset_preds,
+                              level = c(0,1))$predict.dataset_name
+
+ggplot(d_sub_prod, aes(x = age, y = prod, col = as.factor(log_rt_0))) + 
+  geom_point(alpha = .02, color="black") + 
+  geom_line(data = dataset_preds |>   
+              filter(dataset_name %in% datasets_with_age_variation$dataset_name), 
+            aes(y = pred, group = interaction(log_rt_0, dataset_name)), lty="dashed", alpha=.5)+
+  geom_line(data = global_preds, aes(y = pred, group = log_rt_0, col = as.factor(log_rt_0)), size=1)+
+  scale_color_viridis(discrete=T, option = "H")+
+  theme(legend.position = "none")+
+  labs(y="Production", x="Age in months")
+# +
+#   geom_line(data = dataset_preds, aes(y = pred))
+```
 
@@
 To test this relationship, we used longitudinal structural equation models.
-We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months ($t_0 ...
-t_4$) and fit individual growth across each of these variables.
+We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months ($t_0, ..., t_4$) and fit individual growth across each of these variables.
 We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset.
@@
 Consistent with the idea that overall faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
-There was not significant coupling between growth in RT and growth in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
-These abilities might grow independently, but we cannot rule out other possibilities.
+On the other hand, there was not significant coupling between *growth* in RT and *growth* in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
+This null effect could be interpreted as being consistent with these abilities growing independently, but we cannot rule out other possibilities.
 First, the longitudinal data we had might not allow sufficiently precise estimates of growth slopes, or second, since vocabulary growth is non-linear, the linear model we used here might not be as sensitive to non-linear changes.
 
-In sum, these findings provide further evidence consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
+In sum, these findings provide evidence consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
 Children with greater skill in word recognition learn words faster.

### 2025-03-13 paper/supplemental_information.Rmd
fix paths for rendering supplement
nlme

@@
 ```{r, echo=FALSE}
-knitr::purl("paper.Rmd", documentation = 0)
-suppressMessages(source("paper.R"))
+knitr::purl(here("paper","paper.Rmd"), documentation = 0)
+suppressMessages(source(here("paper.R")))
 ```
@@
 
+# Non-linear Growth Models
+
+To test for the differentiation of vocabulary growth based on initial reaction time, we used the package `nlme` to fit a logistic growth model to the production data.
+This model has two parameters for the logistic curve, a scale and an intercept.
+Both were allowed to interact with initial reaction time.
+We also included random effects of logistic intercept and scale by participant and a grouping term across datasets.
+This model showed a significant effect of initial reaction time on the intercept of the logistic growth curve, but not on its scale (see Table \ref{tab:nlme}).
+
+% latex table generated in R 4.4.2 by xtable 1.8-4 package
+% Thu Mar 13 13:10:30 2025
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrr}
+  \hline
+ & Value & Std Error & DF & $t$-value & $p$-value \\ 
+  \hline
+Growth Intercept & 2.26 & 2.70 & 1766.00 & 0.84 & 0.40 \\ 
+  Growth Intercept $\times$ $t_0$ RT & 3.30 & 0.38 & 1766.00 & 8.59 & 0.00 \\ 
+  Growth Scale & 2.05 & 2.60 & 1766.00 & 0.79 & 0.43 \\ 
+  Growth Scale $\times$ $t_0$ RT & 0.28 & 0.38 & 1766.00 & 0.75 & 0.46 \\ 
+   \hline
+\end{tabular}
+\caption{Fixed effects estimates from logistic growth model. \label{tab:nlme}}
+\end{table}
+
+Interpretation of growth in both this model and the linear growth model in the main text is complicated by the fact that the CDI form puts a ceiling on the total number of words that can be recorded; both the quadratic growth functions and the logistic functions come together at the form ceiling.
+Thus, a shift in quadratic growth in the linear model and a shift in intercept in the logistic model both point to the same overall effect, which is faster growth at the point of maximal sensitivity of the CDI.
+Neither model can estimate whether the overall growth trajectory is different beyond the range of the CDI.
+Thus, although these models might initially seem to be in conflict, we believe that they actually point to the same phenomenon, which is perhaps better described by the longitudinal SEM model reported in the main text.
+Children with greater skill in word recognition show an overall positive shift in the growth trajectory of vocabulary development.
+
+```{r}
+# xtable(summary(mod_nlme)$tTable)
+```
+
 # Sensitivity Analysis for Longitudinal Models

### 2025-03-13 1_get_data.qmd

@@
 1. classic short: 200 - 2000
-2. ICC maximizing long: 500 - 4000 - our analysis suggests this maximizes stability of individual differences (albeit not necessarily differences in language)
+2. ICC maximizing long: 200 - 4000 - our analysis suggests this maximizes stability of individual differences (albeit not necessarily differences in language)
 
@@
   mutate(short_window = t_norm > 200 & t_norm <= 2000,
-         long_window = t_norm > 500 & t_norm <= 4000) |>
+         long_window = t_norm > 200 & t_norm <= 4000) |>
   group_by(dataset_name, dataset_id, administration_id, subject_id,
@@
             prop_rts = mean(!is.na(rt)), 
-            log_age = log(age)) |>
+            log_age = log(mean(age,na.rm=TRUE))) |>
   arrange(desc(n)) |>

### 2025-03-13 paper/paper.Rmd
nilam

@@
 Consider a child hearing the utterance "Can you put the ball in the box?" The faster and more accurately the child can recognize that the ball is a referent, the better they can use this evidence to help infer the speaker's intended meaning, perhaps making inferences about the meaning of "put" or "box" [@frank2009using].
-Consistent with this idea, one important study found that children's word recognition speed mediated the relationship between home language input and vocabulary growth [@Weisleder2013].
+Consistent with this idea, one important study found that children's word recognition speed mediated the longitudinal relationship between home language input and vocabulary growth [@Weisleder2013].
 
@@
 Subsequent studies have primarily focused on speed of processing and found more mixed results, with reaction time measures found to be only inconsistently predictive of later vocabulary outcomes [@peter2019;@newbury2016interrelationships;@lany2018].
-To the extent that there are consistent relations between vocabulary and word recognition, these should be visible in a larger dataset.
+A larger dataset should allow us to make a more definitive test of the presence of these relationships.
 Further, by examining the relationship between speed, accuracy, and vocabulary, it should be possible to assess the extent to which processing speed specifically plays a role in vocabulary growth.
@@
 
-Figure \ref{fig:devchange} shows these relationships.
+Figure \ref{fig:devchange} shows these age gradients.
 Log RT decreased significantly with age (`r papaja::apa_print(rt_mods_lmer[[2]])$full_result$log_age_s`) and accuracy increased significantly with age  (`r papaja::apa_print(acc_mods_lmer[[2]])$full_result$log_age_s`).
@@
 One further hallmark of increasing skill is a decrease in task-relevant variability [@todorov2002optimal].
-Both within and across datasets, within-individual variation in accuracy and RT decreased smoothly  across the developmental range we examined (Figure \ref{fig:variance}).
+Both within and across datasets, within-individual variation in accuracy and RT decreased across the developmental range we examined (Figure \ref{fig:variance}).
 We fit mixed effects models to the standard deviation of both accuracy and RT for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
 For both speed and accuracy, within-individual variability decreased with age (RT: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`).
-Thus, as well as being faster and more accurate, older children were more consistent in their online word recognition.
+Thus, as well as being faster and more accurate, older children were more consistent in their online word recognition than younger children.
 
@@
 
-We fit a series of factor analytic models to explore the dimensionality of the data.
-Initial exploratory factor analysis using parallel analysis suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
-Due to missingness of data, we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
-The best fitting model was a three-factor model with factors for speed (RT and RT variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI).
+We fit a series of factor analytic models to explore the dimensionality of the parent report and child LWL data.
+Initial exploratory factor analysis using parallel analysis to select the number of factors suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
+To better accommodate missing data under the assumption of data missing at random (e.g., due to dataset), we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
+The best fitting model of those we examined was a three-factor model with factors for speed (RT and RT variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI).
 Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`); see SI: Alternative Factor Structures).
@@
 This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age).
-All three latent factors were significantly related to one another, with RT and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
-This model supports the idea that individual variation in speed and accuracy of word recognition is concurrently related to vocabulary beyond the effects of age.
+The non-age related variance of all three latent factors was significantly related to that of the other factors, with RT and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
+This model supports the idea that variation in speed and accuracy of word recognition is related to differences in parent-reported vocabulary beyond the effects of age.
+These analyses all treat between-person differences, however, rather than modeling change in these factors within individuals.
 
@@
 
-While we fit the latent variable models above to the full cross-sectional dataset, we were most interested in within-child developmental changes, as measured within individual studies.
-To probe these changes, we first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
+To probe within-person changes, we first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
 Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$), suggesting that, across studies, LWL measures from individual sessions provide relatively noisy measurements of individual children even if individual studies are able to achieve higher reliability by averaging across measurement occasions [e.g., @marchman2008].
@@
 Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure relationships between LWL and vocabulary using longitudinal growth models.
-We began by reproducing the analysis reported in @fernald2012, in which longitudinal growth in productive vocabulary was predicted based on RT during the initial session of the study.
+We began by reproducing the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in RT during the initial session of the study.
 We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset).
@@
 
-Our key question of interest concerned coupling between the parameters of these growth models.
+Our key question of interest concerned coupling among the (latent) intercepts and slopes of these growth models.
 Consistent with the idea that overall faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
@@
 This null effect could be interpreted as being consistent with these abilities growing independently, but we cannot rule out other possibilities.
-First, the longitudinal data we had might not allow sufficiently precise estimates of growth slopes, or second, since vocabulary growth is non-linear, the linear model we used here might not be as sensitive to non-linear changes.
+First, the longitudinal data we had might not allow sufficiently precise estimates of growth slopes, or second, since vocabulary growth is non-linear, the linear model we used here might not have captured coupling among nonlinear aspects of developmental change.
 
-In sum, these findings provide evidence consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].
+In sum, these findings provide evidence consistent with the claim that differences in processing speed are related to differences in the rate of age-related change in vocabulary [@fernald2006;@Weisleder2013].
 Children with greater skill in word recognition learn words faster.
@@
 We investigated these questions using a new, large-scale dataset of developmental eye-tracking measurements compiled across many prior studies.
-We found continuous developmental changes from ages 1 -- 6 years.
-Speed and accuracy both improved asymptotically, with evidence that recognition speed showed the log-log relationship associated with the "power law of practice", that is, gradually converging on mature levels of processing efficiency.
-Further, trial-to-trial variability decreased, consistent with both the literature on skill learning [@todorov2002optimal] and other work on developmental changes in variability [@judd2021working;@galeano2018variability;@kautto2023introducing].
+The age-gradients for speed and accuracy indicate that both improved asymptotically.
+Gradients for recognition speed were consistent with the log-log relationship associated with the "power law of practice", that is, with a gradual convergence on mature levels of processing efficiency.
+Further, the age gradient suggested that trial-to-trial variability decreases with age, consistent with both the literature on skill learning [@todorov2002optimal] and other work on developmental changes in variability [@judd2021working;@galeano2018variability;@kautto2023introducing].
 Speed and accuracy were both related to vocabulary size concurrently and processing speed was also related to later vocabulary growth.
@@
 By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had sample sizes at least an order of magnitude smaller than ours.
-In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to fit complex structural equation models and to compare different functional forms for developmental change.
+In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to explore the relationships between different variables.
 Because early language is so variable [@frank2021], these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [@bergelson2023;@frank2017;@frank2021;@manybabies2020].
+Future work must also explore how the functional forms we observed here between individuals reflect processes of within-person change.
+Although the Peekbank dataset includes a variety of longitudinal data, most reflect a small number of measurements; denser longitudinal data collection is required to better estimate  within-person growth models.
 

### 2025-03-13 paper/paper.Rmd
minor other tweaks

@@
 `Admins` denotes separate experimental sessions.
-`CDIs` refers to whether the dataset contains parent report vocabulary data from the MacArthur-Bates Communicative Development Inventory.}
+`CDIs` refers to whether the dataset contains parent report vocabulary data from the MacArthur-Bates Communicative Development Inventory.
+\label{tab:datasets}}
 \end{table*}
@@
 Here we investigate two specific hypotheses.
-First, one influential theory characterizes language learning as a process of learning the skill of real-time processing [@christiansen2016;@chater2018].
+First, one influential theory posits that language learning is a process of skill learning, in which the primary challenge of the child is learning to converse with others in real time [@christiansen2016;@chater2018].
 Based on this theory, we should expect to see the signatures of expertise and skill learning in word recognition.
@@
 
-```{r reliability}
-MONTH_CUTOFF <- 3
-
+```{r}
 longitudinal <- d_sub |>
@@
 
-d_long <- d_sub |>
-  filter(subject_id %in% longitudinal$subject_id) |>
-  group_by(subject_id) |>
-  arrange(age) |>
-  mutate(admin_num = 1:n(), 
-         time_since_t0 = age - age[1],
-         delta_t = c(0, diff(age)))
-
-d_reliability <- d_long |>
-  filter(time_since_t0 <= MONTH_CUTOFF, 
-         admin_num <= 2) |>
-  pivot_wider(id_cols = c("subject_id","dataset_name"), 
-              names_from = "admin_num",
-              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
-                              "long_window_acc_var", "prod", "comp"))
-
-dataset_reliabilities <- d_reliability |>
-  group_by(dataset_name) |>
-  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
-            acc = cor(long_window_accuracy_1, 
-                          long_window_accuracy_2, use = "pairwise.complete.obs"))
-
-# note this is not correct for CDIs because some CDIs can get grouped with two admins, inflating correlations
-global_reliabilities <- d_reliability |>
-  ungroup() |>
-  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
-            acc = cor(long_window_accuracy_1, 
-                          long_window_accuracy_2, use = "pairwise.complete.obs"))
-```
-
-To probe within-person changes, we first examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
-Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$), suggesting that, across studies, LWL measures from individual sessions provide relatively noisy measurements of individual children even if individual studies are able to achieve higher reliability by averaging across measurement occasions [e.g., @marchman2008].
-
-```{r}
 d_growth <- d_sub |>
@@
 
-Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure relationships between LWL and vocabulary using longitudinal growth models.
-We began by reproducing the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in RT during the initial session of the study.
+We began investigating relationships between LWL and vocabulary by using longitudinal growth models.
+We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in RT during the initial session of the study.
 We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset).

### 2025-03-13 paper/supplemental_information.Rmd
minor other tweaks

@@
 ```{r, fig.cap="Number of administrations for datasets with repeated measurements."}
-longitudinal <- d_sub |>
-  group_by(dataset_name, subject_id) |>
-  count() |>
-  filter(n > 1)
 
@@
 ```
+# Test retest reliability
+
+```{r reliability}
+MONTH_CUTOFF <- 3
+
+longitudinal <- d_sub |>
+  group_by(dataset_name, subject_id) |>
+  count() |>
+  filter(n > 1)
+
+d_long <- d_sub |>
+  filter(subject_id %in% longitudinal$subject_id) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(admin_num = 1:n(), 
+         time_since_t0 = age - age[1],
+         delta_t = c(0, diff(age)))
+
+d_reliability <- d_long |>
+  filter(time_since_t0 <= MONTH_CUTOFF, 
+         admin_num <= 2) |>
+  pivot_wider(id_cols = c("subject_id","dataset_name"), 
+              names_from = "admin_num",
+              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
+                              "long_window_acc_var", "prod", "comp"))
+
+dataset_reliabilities <- d_reliability |>
+  group_by(dataset_name) |>
+  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
+            acc = cor(long_window_accuracy_1, 
+                          long_window_accuracy_2, use = "pairwise.complete.obs"))
+
+# note this is not correct for CDIs because some CDIs can get grouped with two admins, inflating correlations
+global_reliabilities <- d_reliability |>
+  ungroup() |>
+  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
+            acc = cor(long_window_accuracy_1, 
+                          long_window_accuracy_2, use = "pairwise.complete.obs"))
+```
+
+We examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
+Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$).
+These reliabilities were biased downwards by three factors, however.
+First, longitudinal assessments sometimes use variable items between testing sessions, leading to item-related variance in measurement.
+Second, even three months can lead to substantial change in some children's language abilities, thus correlations are attenuated by true change as well as measurement error.
+Third, longitudinal data in the dataset come primarily from the youngest children and hence are likely to show overall higher measurement error due to variability in children's behavior and an overall lower number of trials.
 

### 2025-03-13 paper/paper.Rmd
updates to Peekbank language

@@
 To overcome this limitation, we turned to Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank].
+This dataset unifies and carefully curates a large amount of eye-tracking data from studies with infants and toddlers, representing cumulatively over 12 million individual samples of children's eye movements while processing language.
 This resource allows us to gain an unprecedented view of the development of word recognition across a large sample of children.
@@
 We retrieved data from Peekbank, focusing on data from children ages 1--6 years.
-Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ..."); these trials often constituted control conditions for experiments.
+Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ..."); these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations.
 We focus here on English purely for practical reasons -- the Peekbank dataset contains limited data from other languages.
@@
 By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had sample sizes at least an order of magnitude smaller than ours.
+Our approach was to build on the time-consuming and meticulous data collection from previous infant and toddler eye-tracking studies --- representing cumulatively many thousands of hours of in-lab data collection and hand-annotation of the resulting videos of child looking behavior --- by harmonizing these data into a single, large-scale database.
+This approach illustrates how building harmonized databases can be especially powerful when composed of high-effort and high-quality datasets that are smaller in scope, maximizing the impact of previous data collection efforts and allowing us to ask broader questions about developmental change.
 In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to explore the relationships between different variables.

### 2025-03-13 paper/paper.Rmd
lots of renaming

--- 2025-03-13 paper/paper.Rmd
@@
 Further, faster processing at 18 months is predictive of whether "late talkers" catch up to their peers or require further intervention [@fernald2012].
-Critically, all of these assessments use words that children are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.
+Typically, these assessments use words that children are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.
 Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
@@
 Here we investigate two specific hypotheses.
-First, one influential theory posits that language learning is a process of skill learning, in which the primary challenge of the child is learning to converse with others in real time [@christiansen2016;@chater2018].
-Based on this theory, we should expect to see the signatures of expertise and skill learning in word recognition.
+First, one influential theory posits that language learning is a process of skill learning, in which the child is learning the skill of fluent conversation with other language users [@christiansen2016;@chater2018].
+In this theory, the major information processing challenge of language learning is that incoming language is ephemeral and must be processed quickly before it is lost (the "now-or-never bottleneck").
+On this kind of account, we should expect to see the signatures of expertise and skill learning in word recognition, which is one of the primary skills involved in processing incoming language in real time.
 Accuracy should change linearly with the logarithm of age, reflecting gradual asymptotic convergence to mature levels of accuracy.
-In addition, the logarithm of processing speed should decrease with the logarithm of age, potentially reflecting the "power law of practice" [@snoddy1926;@kail1991;@anderson1982; cf @heathcote2000].
+In addition, we might observe what is known as the "power law of practice," the ubiquitous regularity that, for many skills, the logarithm of reaction time decreases with the logarithm of experience [@snoddy1926;@kail1991;@anderson1982; cf.
+@heathcote2000].
+In our case, we expect that chronological age is a proxy for experience and so the logarithm of processing speed should decrease linearly with the logarithm of age.
 Finally, trial-to-trial variability in both speed and accuracy should decrease with increasing expertise [@todorov2002optimal].
@@
 We began by characterizing developmental changes in speed and accuracy.
-We computed both RTs (reaction times) and accuracies following standard practices in the literature [@fernald2008].
-Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms).
-For each window, we averaged all fixation within the window to compute a continuous proportion between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
+We computed both RTs (reaction times) and accuracies (proportion target looking) following standard practices in the literature [@fernald2008].
+Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms).
+For each window, we averaged all fixation within the window to compute a continuous proportion target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
 
 Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI for raw correlations between variables).
-To investigate this question, we fit linear mixed effects models predicting accuracy and RT on each trial across the full dataset with random slopes of child age by study and random intercepts by participant.
+(In this initial investigation, we treat the observed variables RT and target looking as direct measures of the constructs speed and accuracy, but we later estimate them as latent variables.) We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant.
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
@@
 Figure \ref{fig:devchange} shows these age gradients.
-Log RT decreased significantly with age (`r papaja::apa_print(rt_mods_lmer[[2]])$full_result$log_age_s`) and accuracy increased significantly with age  (`r papaja::apa_print(acc_mods_lmer[[2]])$full_result$log_age_s`).
+Log RT decreased significantly with age, reflecting increasing speed (`r papaja::apa_print(rt_mods_lmer[[2]])$full_result$log_age_s`) and accuracy also increased significantly with age  (`r papaja::apa_print(acc_mods_lmer[[2]])$full_result$log_age_s`).
 In sum, we see continuing improvements in word recognition across the full age range in our dataset that appear roughly linear in the logarithm of age.
 These logarithmic relationships follow theoretical expectations that both speed and accuracy should gradually asymptote to mature levels of performance as in skill learning more generally [@anderson1982,@snoddy1926].
-<!-- With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.
--->
 
-```{r devchange, fig.cap="Participant-level accuracy and reaction time (log), plotted by age (log).
+```{r devchange, fig.cap="Participant-level target looking and reaction time (log), plotted by age (log).
+Longitudinal datapoints are connected by lines.
 The solid blue line shows a linear fit and associated confidence interval.
@@
   geom_point(alpha = .02) + 
+  geom_line(aes(group = subject_id), alpha = .05) + 
   geom_smooth(data = d_sub_agevar,
@@
   scale_color_manual(values=mycolors, guide = FALSE) + 
-  ylab("Accuracy") + 
+  ylab("Proportion Target Looking") + 
   xlab("Age (months)")
@@
 One further hallmark of increasing skill is a decrease in task-relevant variability [@todorov2002optimal].
-Both within and across datasets, within-individual variation in accuracy and RT decreased across the developmental range we examined (Figure \ref{fig:variance}).
-We fit mixed effects models to the standard deviation of both accuracy and RT for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
-For both speed and accuracy, within-individual variability decreased with age (RT: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`).
+Both within and across datasets, within-individual variation in speed and accuracy decreased across the developmental range we examined (Figure \ref{fig:variance}).
+We fit mixed effects models to the standard deviation of both speed and accuracy for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
+For both speed and accuracy, within-individual variability decreased with age (speed: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`).
 Thus, as well as being faster and more accurate, older children were more consistent in their online word recognition than younger children.
 
-```{r variance, fig.cap="Participant-level variability in accuracy and reaction time (log RT), plotted by age (log). Plotting conventions are as in Figure 1.", fig.env="figure*", fig.height = 2.5}
+```{r variance, fig.cap="Participant-level variability in target looking and reaction time (log RT), plotted by age (log).
+Plotting conventions are as in Figure 1.", fig.env="figure*", fig.height = 2.5}
   mycolors = c("#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E","#1B9E77", "#D95F02")
@@
   scale_color_manual(values=mycolors, guide = FALSE) + 
-  ylab("SD of Accuracy") + 
+  ylab("SD of Prop Target Looking") + 
   xlab("Age (months)")
@@
 
-We were next interested in whether the various aspects of word recognition -- including accuracy, RT, and the variability of each of these -- were related to other aspects of early language ability.
+We were next interested in whether the various aspects of word recognition -- including speed, accuracy, and the variability of each of these -- were related to other aspects of early language ability.
 Of the studies in our database, `r sum(str_length(dataset_characteristics$CDIs)>0)` gathered parent reports about children's early vocabulary using the MacArthur-Bates Communicative Development Inventory (CDI), a popular survey instrument that provides a reliable and valid estimate of children's early vocabulary [@marchman2023;@frank2021].
@@
 To better accommodate missing data under the assumption of data missing at random (e.g., due to dataset), we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
-The best fitting model of those we examined was a three-factor model with factors for speed (RT and RT variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI).
+The best fitting model of those we examined was a three-factor model with factors for speed (RT and RT variability), accuracy (proportion looking to target on each trial and associated variability of this measure), and vocabulary (comprehension and production from the CDI).
 Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`); see SI: Alternative Factor Structures).
@@
 
-node_text <- tibble(name=c("acc", "acc_sd", "accuracy", "comp", "log_age", "log_rt", "log_rt_sd", "prod", "speed", "vocab"), pretty=c("Accuracy", "Accuracy (SD)", "Accuracy", "Comprehension", "Age (log months)", "log RT", "log RT (SD)", "Production", "Speed", "Vocabulary"))
+node_text <- tibble(name=c("acc", "acc_sd", "accuracy", "comp", "log_age", "log_rt", "log_rt_sd", "prod", "speed", "vocab"), pretty=c("Target Looking", "Target Looking (SD)", "Accuracy", "Comprehension", "Age (log months)", "log RT", "log RT (SD)", "Production", "Speed", "Vocabulary"))
 
@@
 This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age).
-The non-age related variance of all three latent factors was significantly related to that of the other factors, with RT and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
+The non-age related variance of all three latent factors was significantly related to that of the other factors, with speed and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
 This model supports the idea that variation in speed and accuracy of word recognition is related to differences in parent-reported vocabulary beyond the effects of age.
@@
 We began investigating relationships between LWL and vocabulary by using longitudinal growth models.
-We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in RT during the initial session of the study.
+We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in speed during the initial session of the study.
 We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset).
@@
 ```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for the 25th and 75th percentile on initial reaction times.
+Individual longitudinal trajectories are shown in light gray.
 Solid lines show global model estimates and dashed lines show estimates for individual datasets with sufficient age variation as above.", fig.env="figure", fig.height = 2.5, fig.width=4, out.width=".5\\textwidth", include=TRUE}
@@
 ggplot(d_sub_prod, aes(x = age, y = prod, col = as.factor(log_rt_0))) + 
-  geom_point(alpha = .02, color="black") + 
+  geom_point(alpha = .03, pch = ".", col = "black") + 
+  geom_line(aes(group = subject_id), alpha = .02, col = "black") + 
   geom_line(data = dataset_preds |>   
@@
 Consistent with the idea that overall faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
-On the other hand, there was not significant coupling between *growth* in RT and *growth* in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
+On the other hand, there was not significant coupling between *growth* in speed and *growth* in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
 This null effect could be interpreted as being consistent with these abilities growing independently, but we cannot rule out other possibilities.
@@
 node_text <- tribble(~name, ~ pretty, ~fill, 
-                     "acc_t0", "t0 Accuracy", acc_color,
-                     "acc_t1", "t1 Accuracy", acc_color,
-                     "acc_t2", "t2 Accuracy",acc_color,
-                     "acc_t3", "t3 Accuracy",acc_color,
-                     "acc_t4", "t4 Accuracy",acc_color,
+                     "acc_t0", "t0 Target Looking", acc_color,
+                     "acc_t1", "t1 Target Looking", acc_color,
+                     "acc_t2", "t2 Target Looking",acc_color,
+                     "acc_t3", "t3 Target Looking",acc_color,
+                     "acc_t4", "t4 Target Looking",acc_color,
                      "accuracy_intercept", "Accuracy intercept",acc_color,
@@
 Future work must also explore how the functional forms we observed here between individuals reflect processes of within-person change.
-Although the Peekbank dataset includes a variety of longitudinal data, most reflect a small number of measurements; denser longitudinal data collection is required to better estimate  within-person growth models.
 
-At the same time, our approach is both observational and exploratory.
+Our approach is both observational and exploratory.
 Thus, we cannot untangle the range of different causal models that explain the variation we observed.

### 2025-03-13 paper/paper.Rmd

@@
 Future work must also explore how the functional forms we observed here between individuals reflect processes of within-person change.
+Although the Peekbank dataset includes a variety of longitudinal data, most reflect a small number of measurements; denser longitudinal data collection is required to better estimate  within-person growth models.
 

### 2025-03-13 paper/paper.Rmd
materials and methods pass

@@
 Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy with parent-reported vocabulary.
-Across a range of models, speed, accuracy, and vocabulary also show coupled growth such that children with faster initial word recognition tend to show faster vocabulary growth.
+Across a range of longitudinal models, speed, accuracy, and vocabulary also show coupled growth such that children with faster word recognition tend to show faster vocabulary growth.
 Together, these findings support the view that word recognition is a skill that develops gradually across early childhood and that this skill plays a role in supporting early language learning.
@@
 Here, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
-These data suggest that understanding language is a skill that improves gradually with practice.
+These data suggest that understanding language in real-time is an important part of eearly language learning and that it improves gradually with practice across early childhood.
 
@@
 Early work using this method showed that both children's speed and accuracy increase rapidly across the second year [@fernald1998;@reznick1990visual].
-Related methods have provided a window into how children process phonological [@mani2011], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] information as well as how their lexical representations develop [@Swingley2002].
+Related methods have provided a window into how children process phonological [@mani2011], lexical [@Swingley2002], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] information.
 
 Word recognition ability, as measured by LWL, is hypothesized to play a key role in language learning.
-Each word that a child experiences is an opportunity to learn; measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking].
-Recognizing incoming words and linking them with their referents is a prerequisite for learning.
-Consider a child hearing the utterance "Can you put the ball in the box?" The faster and more accurately the child can recognize that the ball is a referent, the better they can use this evidence to help infer the speaker's intended meaning, perhaps making inferences about the meaning of "put" or "box" [@frank2009using].
+Measurements of children's language input at home are consistently associated with their vocabulary size [@hart1995meaningful;@anderson2021linking].
+The mechanism posited to drive this association is that each word that a child experiences is an opportunity to learn.
+But each word must first be recognized during the short window of time when it is present in the child's memory.
+Consider a child hearing the utterance "Can you put the ball in the crate?" The faster and more accurately the child can recognize the word "ball", the better they can use this evidence to help infer the speaker's intended meaning, perhaps in the process making inferences about the meaning of the less familiar word "crate" [@frank2009using].
 Consistent with this idea, one important study found that children's word recognition speed mediated the longitudinal relationship between home language input and vocabulary growth [@Weisleder2013].
@@
 Word recognition speed has been used as an index of individual differences in early childhood [@fernald2006;@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow].
-Over and above measures of vocabulary size, word recognition speed at 18 months predicts children's standardized test scores years later [@marchman2008], though these predictions may be limited to particular ages or processing assessments [@peter2019].
+Over and above measures of vocabulary size, word recognition speed at 18 months predicts children's standardized test scores years later [@marchman2008].
 Further, faster processing at 18 months is predictive of whether "late talkers" catch up to their peers or require further intervention [@fernald2012].
-Typically, these assessments use words that children are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.
+Critically, these assessments use words that children at the target age are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.
+
 Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
@@
 Accuracy should change linearly with the logarithm of age, reflecting gradual asymptotic convergence to mature levels of accuracy.
-In addition, we might observe what is known as the "power law of practice," the ubiquitous regularity that, for many skills, the logarithm of reaction time decreases with the logarithm of experience [@snoddy1926;@kail1991;@anderson1982; cf.
+In addition, we might observe what is known as the "power law of practice," the regularity found in many cases of skill learning that the logarithm of reaction time decreases with the logarithm of experience [@snoddy1926;@kail1991;@anderson1982; cf.
 @heathcote2000].
 In our case, we expect that chronological age is a proxy for experience and so the logarithm of processing speed should decrease linearly with the logarithm of age.
-Finally, trial-to-trial variability in both speed and accuracy should decrease with increasing expertise [@todorov2002optimal].
+Finally, trial-to-trial variability in both speed and accuracy should decrease with increasing expertise, as is found in studies of motor expertise [@todorov2002optimal].
 
@@
 
-We retrieved data from Peekbank, focusing on data from children ages 1--6 years.
-Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ..."); these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations.
-We focus here on English purely for practical reasons -- the Peekbank dataset contains limited data from other languages.
-
+We retrieved data from Peekbank, focusing on data from English-speaking children ages 1--6  years and on simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ...").
 These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions).
@@
 
-We began by characterizing developmental changes in speed and accuracy.
-We computed both RTs (reaction times) and accuracies (proportion target looking) following standard practices in the literature [@fernald2008].
+We began by characterizing age gradients in speed and accuracy across children.
+We computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms).
 For each window, we averaged all fixation within the window to compute a continuous proportion target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
+In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy; in subsequent analyses we estimate latent variables representing these constructs.
 
 Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI for raw correlations between variables).
-(In this initial investigation, we treat the observed variables RT and target looking as direct measures of the constructs speed and accuracy, but we later estimate them as latent variables.) We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant.
+We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant.
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
@@
 
-We began investigating relationships between LWL and vocabulary by using longitudinal growth models.
+To investigate within-person relationships between LWL and vocabulary, we began by using longitudinal growth models.
 We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in speed during the initial session of the study.
@@
 This model revealed a significant effect of $t_0$ RT (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ RT and the quadratic age predictor (`r papaja::apa_print(growth_mod)$full_result$polyage_15_22_rt_t0`).
-This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time, confirming that findings from @fernald2012 were robust across the full dataset.
-We confirmed this analysis using a non-linear growth model with a logistic shape, which shows a better fit to vocabulary size than the quadratic model (see SI: Non-Linear Growth Model).
+This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time, confirming that findings of @fernald2012 were robust across the full set of longitudinal data in our dataset.
+
+We confirmed this analysis using a non-linear growth model with a logistic shape, which provides a better fit to vocabulary size than the quadratic model (see SI: Non-Linear Growth Model) [@frank2021].
 Figure \ref{fig:growth} shows predictions from this model, confirming the differentiation of growth curves for children with higher and lower initial reaction time.
@@
 We investigated these questions using a new, large-scale dataset of developmental eye-tracking measurements compiled across many prior studies.
-The age-gradients for speed and accuracy indicate that both improved asymptotically.
+The age-gradients for speed and accuracy indicate that both improve asymptotically.
 Gradients for recognition speed were consistent with the log-log relationship associated with the "power law of practice", that is, with a gradual convergence on mature levels of processing efficiency.
 Further, the age gradient suggested that trial-to-trial variability decreases with age, consistent with both the literature on skill learning [@todorov2002optimal] and other work on developmental changes in variability [@judd2021working;@galeano2018variability;@kautto2023introducing].
-Speed and accuracy were both related to vocabulary size concurrently and processing speed was also related to later vocabulary growth.
+Speed and accuracy were both related to vocabulary size concurrently and processing speed was also related longitudinally to later vocabulary growth.
 
@@
 Our approach was to build on the time-consuming and meticulous data collection from previous infant and toddler eye-tracking studies --- representing cumulatively many thousands of hours of in-lab data collection and hand-annotation of the resulting videos of child looking behavior --- by harmonizing these data into a single, large-scale database.
-This approach illustrates how building harmonized databases can be especially powerful when composed of high-effort and high-quality datasets that are smaller in scope, maximizing the impact of previous data collection efforts and allowing us to ask broader questions about developmental change.
+This approach illustrates how building harmonized databases can be especially powerful when composed of high-effort and high-quality datasets that are smaller in scope, maximizing the impact of previous data collection efforts and allowing us to ask broader questions about developmental change [@frank2021].
 In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to explore the relationships between different variables.
-Because early language is so variable [@frank2021], these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [@bergelson2023;@frank2017;@frank2021;@manybabies2020].
+Because early language is so variable, these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [@bergelson2023;@frank2017;@frank2021;@manybabies2020].
+
+Our approach is both observational and exploratory.
+Thus, we cannot untangle the range of different causal models that explain the variation we observed.
+First, early word recognition skill could lead to faster word learning, but faster children could also be faster due to their larger vocabulary and stronger lexical representations.
+These two causal directions could also interact reciprocally, leading to a "rich get richer" process in which children with larger vocabularies process faster, and their faster processing helps them increase their vocabulary size more rapidly.
+Finally, a third shared factor -- perhaps general cognitive ability -- could underpin both processes.
+Our cross-sectional data cannot distinguish these hypotheses even in principle [@vanderweele2023dimensional], and our longitudinal data are likely too sparse to distinguish such complex causal models.
 Future work must also explore how the functional forms we observed here between individuals reflect processes of within-person change.
@@
 
-Our approach is both observational and exploratory.
-Thus, we cannot untangle the range of different causal models that explain the variation we observed.
-First, early word recognition skill could lead to faster word learning, due to skilled children gleaning more information from the same signal [@fernald2006].
-But second, faster children could also be faster due to their larger vocabulary and stronger lexical representations.
-These two causal directions could also interact reciprocally, leading to a "rich get richer" process in which children with larger vocabularies process faster, and their faster processing helps them increase their vocabulary size more rapidly.
-Finally, a third shared factor -- perhaps general cognitive ability -- could underpin both processes.
-Our cross-sectional data cannot distinguish these hypotheses even in principle [@vanderweele2023dimensional], and our longitudinal data are likely too sparse to distinguish such complex causal models.
-Thus, these questions remain an open target for dense longitudinal data collection.
-
-Our findings here are limited in their generalizability by the convenience samples that were used in the individual studies in the Peekbank database.
+Nevertheless, consistency of the trends we observed across datasets suggests that our qualitative conclusions are robust to some significant sociodemographic variation.
+Nevertheless, these findings are still limited in their generalizability by the convenience samples that were used in the studies aggregated in Peekbank.
 These studies typically (but not always) represent children from well-educated parents living in university-adjacent communities.
-We would not expect that specific numerical parameters estimated in this aggregate convenience sample would generalize to other samples.
-Nevertheless, the consistency of the trends we observed across datasets suggests that our qualitative conclusions are robust to some significant sociodemographic variation.
+We would not expect that specific numerical parameters estimated in our aggregate convenience sample would generalize to other samples.
 
@@
 If language learning is, at least in part, a process of skill learning, then measurement of this skill is a critical window into understanding the remarkable process of language learning.
+
+# Materials and Methods {.unnumbered}
+
+## Data {.unnumbered}
+
+We included information from `r length(unique(d_sub$subject_id))` unique participants across `r length(unique(d_sub$dataset_name))` datasets.
+Dataset information is given in Table \ref{tab:datasets}.Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials; these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations.
+We focus here on English purely for practical reasons -- the Peekbank dataset at present contains limited data from other languages.
+
+## Analytic methods {.unnumbered}
+
+We used `lme4` to fit linear mixed effects models, `nlme` to fit non-linear models, and `lavaan` to fit structural equation models. Full model specifications are available in the linked repository.
+
+## Code and Data Availability {.unnumbered}
+
+We retrieved all data from Peekbank release 2025.1 using the `peekbankr` package.
+All code and data necessary to reproduce this manuscript are available at [http://github.com/peekbank/peekbank-development]().
 

### 2025-03-13 paper/paper.Rmd
si cleanup

@@
 These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions).
-Table \ref{tab:datasets} shows the characteristics of individual datasets (see also Figures S1 and S2).
+Table \ref{tab:datasets} shows the characteristics of individual datasets (see also SI: Dataset Description).
 The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed study of the development of word recognition than has previously been possible.
@@
 For each window, we averaged all fixation within the window to compute a continuous proportion target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
-In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy; in subsequent analyses we estimate latent variables representing these constructs.
+In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy (see SI: Test-Retest Reliability); in subsequent analyses we estimate latent variables representing these constructs.
 
-Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI for raw correlations between variables).
+Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables).
 We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant.
@@
 The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text.
-All findings reported here hold for both window sizes, however (see SI).] and the best fitting model of speed predicted log RT as a function of log age as well (see SI).
+All findings reported here hold for both window sizes, however (see SI: Short Window Analysis).] and the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison).
 
@@
 We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset.
-(We present a model of coupled growth in the observed variables here for simplicity; see SI for a comparable model using growth in the latent factors).
+(We present a model of coupled growth in the observed variables here for simplicity; see SI: Latent Factor Growth for a comparable model using growth in the latent factors).
 The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}.

### 2025-03-13 paper/supplemental_information.Rmd
lots of renaming
si cleanup

@@
 
-# Dataset description
+# Dataset Description
 
@@
 ```
-# Test retest reliability
+# Test-Retest Reliability
 
@@
 
-# Speed and accuracy
-
-Correlations between variables
+# Correlations
 
@@
 
+# Functional Form Model Comparison
+
 ```{r results="asis"}
@@
 
-# Exploratory Factor Analysis
+With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.
 
-- Full EFA results
+# Factor Analysis
+
+```{r}
+d_sub_mat <- d_sub |>
+  ungroup() |>
+  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)
+
+d_sub_mat_s <- d_sub_mat |>
+  ungroup() |>
+  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
+                         "long_window_acc_var", "prod", "comp")), 
+                       ~ age_scale(.x, age))) 
+fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
+            use = "pairwise.complete.obs")
+```
+
+```{r}
+fs <- fa(select(d_sub_mat, -dataset_name), nfactor = 3,
+            use = "pairwise", rotate = "varimax")
+
+as.data.frame(unclass(loadings(fs))) |> 
+  kable(caption = "Factor loadings for the exploratory three factor solution using varimax rotation.", digits = 2)
+```
 
@@
 
-# Sensitivity Analysis for Longitudinal Models
+# Latent Factor Growth
 
-THIS IS WRONG
+```{r}
+fa3_model_long <- "
+# measurement model
+vocab_t1 =~ 1*prod_t1 + s1*comp_t1
+accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
+speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1
 
-```{r reliability_table, results="asis"}
-xtable(dataset_reliabilities)
+vocab_t2 =~ 1*prod_t2 
+accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
+speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2
+
+vocab_t3 =~ 1*prod_t3 
+accuracy_t3 =~ 1*acc_t3 + s2*acc_sd_t3
+speed_t3 =~ 1*log_rt_t3 + s3*log_rt_sd_t3
+
+vocab_t4 =~ 1*prod_t4 
+accuracy_t4 =~ 1*acc_t4 + s2*acc_sd_t4
+speed_t4 =~ 1*log_rt_t4 + s3*log_rt_sd_t4
+
+# means for the latents
+# vocab_t1 ~ 1
+# vocab_t2 ~ 1
+# vocab_t3 ~ 1
+# vocab_t4 ~ 1
+# accuracy_t1 ~ 1
+# accuracy_t2 ~ 1
+# accuracy_t3 ~ 1
+# accuracy_t4 ~ 1
+# speed_t1 ~ 1
+# speed_t2 ~ 1
+# speed_t3 ~ 1
+# speed_t4 ~ 1
+
+# autoregression for the latents
+# vocab_t2 ~ 1 * vocab_t1
+# vocab_t3 ~ 1 * vocab_t2
+# vocab_t4 ~ 1 * vocab_t3
+# accuracy_t2 ~ 1 * accuracy_t1
+# accuracy_t3 ~ 1 * accuracy_t2
+# accuracy_t4 ~ 1 * accuracy_t3
+# speed_t2 ~ 1 * speed_t1
+# speed_t3 ~ 1 * speed_t2
+# speed_t4 ~ 1 * speed_t3
+
+# regressions
+accuracy_intercept =~ 1*accuracy_t1 + 1*accuracy_t2 + 1*accuracy_t3 + 1*accuracy_t4 
+accuracy_slope =~ 1*accuracy_t1 + 2*accuracy_t2 + 3*accuracy_t3 + 4*accuracy_t4
+speed_intercept =~ 1*speed_t1 + 1*speed_t2 + 1*speed_t3 + 1*speed_t4 
+speed_slope =~ 1*speed_t1 + 2*speed_t2 + 3*speed_t3 + 4*speed_t4
+vocab_intercept =~ 1*vocab_t1 + 1*vocab_t2 + 1*vocab_t3 + 1*vocab_t4 
+vocab_slope =~ 1*vocab_t1 + 2*vocab_t2 + 3*vocab_t3 + 4*vocab_t4
+
+# parameters for the regressions
+vocab_intercept ~ 1
+vocab_slope ~ 1
+accuracy_intercept ~ 1
+accuracy_slope ~ 1
+speed_intercept ~ 1
+speed_slope ~ 1
+
+# residual variance for the latents
+vocab_t1 ~~ NA*vocab_t1
+vocab_t2 ~~ NA*vocab_t2
+vocab_t3 ~~ NA*vocab_t3
+vocab_t4 ~~ NA*vocab_t4
+accuracy_t1 ~~ NA*accuracy_t1
+accuracy_t2 ~~ NA*accuracy_t2
+accuracy_t3 ~~ NA*accuracy_t3
+accuracy_t4 ~~ NA*accuracy_t4
+speed_t1 ~~ NA*speed_t1
+speed_t2 ~~ NA*speed_t2
+speed_t3 ~~ NA*speed_t3
+speed_t4 ~~ NA*speed_t4
+
+# residual variance for the other latents
+vocab_intercept ~~ NA*vocab_intercept
+vocab_slope ~~ NA*vocab_slope
+accuracy_intercept ~~ NA*accuracy_intercept
+accuracy_slope ~~ NA*accuracy_slope
+speed_intercept ~~ NA*speed_intercept
+speed_slope ~~ NA*speed_slope
+
+# set observed intercepts to zero - these should go into the latents
+prod_t1 ~ 0
+prod_t2 ~ 0
+prod_t3 ~ 0
+prod_t4 ~ 0
+comp_t1 ~ 0 
+log_rt_t1 ~ 0
+log_rt_t2 ~ 0
+log_rt_t3 ~ 0
+log_rt_t4 ~ 0
+log_rt_sd_t1 ~ 0 
+log_rt_sd_t2 ~ 0
+log_rt_sd_t3 ~ 0 
+log_rt_sd_t4 ~ 0 
+acc_t1 ~ 0 
+acc_t2 ~ 0 
+acc_t3 ~ 0 
+acc_t4 ~ 0 
+acc_sd_t1 ~ 0
+acc_sd_t2 ~ 0
+acc_sd_t3 ~ 0
+acc_sd_t4 ~ 0
+"
+
+fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
+
+summary(fit3_long, fit.measures=TRUE, standardize=TRUE)
+
 ```
 
-Sensitivity analysis for age split point (e.g., split at 3 months)
+```{r}
+layout <- read.csv(here("misc/layout_dt3.csv"), header = FALSE)
 
-Sensitivity analysis for measurement invariance (e.g., not constraining loadings)
+graph_sem(model = fit3_long, text_size = 3, layout=t(layout))
 
-# Analyses using short window accuracy
+```
 
-## Longitudinal factor analysis
+# Short Window Analysis
 
-## Longitudinal factor analysis
+Need to redo all figures using short windows ugh.

### 2025-03-14 paper/supplemental_information.Rmd
supplement

@@
 ---
-title: "Supplemental Information"
+title: "Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood"
 output: pdf_document
@@
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
+knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
 library(here)
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
 knitr::purl(here("paper","paper.Rmd"), documentation = 0)
@@
 
-```{r, fig.env="figure*"}
+Figure \ref{fig:dataset} gives the age distribution for each paper within our full dataset.
+
+```{r dataset, fig.env="figure*"}
 ggplot(d_sub, aes(x = age))+ 
@@
 
-```{r, fig.cap="Number of administrations for datasets with repeated measurements."}
+Figure \ref{fig:longitudinal_admins} shows the distribution of measurement intervals for longitudinal studies within the dataset.
+
+```{r longitudinal_admins, fig.cap="Number of administrations for datasets with repeated measurements."}
 
@@
 
-```{r, results = "asis"}
+Table \ref{tab:corrs} shows pairwise correlations between the primary variables of interest in the dataset.
+
+```{r corrs, results = "asis"}
 
@@
 
-```{r results="asis"}
+Table \ref{tab:accmods} shows model comparison measures between for different models of the functional form for accuracy and Table \ref{tab:rtmods} shows the same for reaction time.
+Note that all of these comparisons With our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns [@heathcote2000].
+
+\begin{table}[tbp]
+\begin{center}
+\begin{threeparttable}
+\begin{tabular}{lllllllll}
+\toprule
+n.
+obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{r2}\\
+\midrule
+48354 & 0.28 & -8,338.83 & 16,691.65 & 16,753.16 & 16,677.65 & 48347 & lwa\_lin & 0.14\\
+48354 & 0.28 & -8,305.39 & 16,624.78 & 16,686.28 & 16,610.78 & 48347 & lwa\_log & 0.12\\
+50244 & 0.31 & -12,486.60 & 24,987.21 & 25,048.98 & 24,973.21 & 50237 & swa\_lin & 0.09\\
+50244 & 0.31 & -12,458.77 & 24,931.54 & 24,993.31 & 24,917.54 & 50237 & swa\_log & 0.08\\
+\bottomrule
+\end{tabular}
+\end{threeparttable}
+\end{center}
+\end{table}
+
+\begin{table}[tbp]
+\begin{center}
+\begin{threeparttable}
+\begin{tabular}{lllllllll}
+\toprule
+n.
+obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{r2}\\
+\midrule
+18940 & 0.45 & -12,414.69 & 24,843.38 & 24,898.32 & 24,829.38 & 18933 & log\_lin & 0.21\\
+18940 & 0.45 & -12,393.27 & 24,800.53 & 24,855.48 & 24,786.53 & 18933 & log\_log & 0.21\\
+18940 & 570.49 & -147,644.43 & 295,302.87 & 295,357.81 & 295,288.87 & 18933 & lin\_lin & 0.18\\
+18940 & 570.02 & -147,622.67 & 295,259.34 & 295,314.29 & 295,245.34 & 18933 & lin\_log & 0.18\\
+\bottomrule
+\end{tabular}
+\end{threeparttable}
+\end{center}
+\end{table}
+
+```{r accmods, results="asis", include = FALSE}
 papaja::apa_table(acc_mods_lmer_summary)
@@
 
-```{r results="asis"}
+```{r results="asis", include = FALSE}
 papaja::apa_table(rt_mods_lmer_summary)
 ```
-
-With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.
 
@@
 
-fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')
+fit3_long <- sem(fa3_model_long, d_sub_wide_fine, std.lv=TRUE, missing='fiml')
 
@@
 layout <- read.csv(here("misc/layout_dt3.csv"), header = FALSE)
-
 graph_sem(model = fit3_long, text_size = 3, layout=t(layout))

### 2025-03-14 paper/paper.Rmd
SI

@@
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
-The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text.
-All findings reported here hold for both window sizes, however (see SI: Short Window Analysis).] and the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison).
+The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison).
+Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analysis.
 
@@
 
+We excluded trials entirely if they were missing data on more than 50% of timepoints, and excluded RTs if they were based on fewer than 50% of timepoints in the short analytic window (200 -- 2000 ms).
+We also removed RTs shorter than 367 ms, as these were unlikely to be generated based on the specific linguistic stimulus.
+We then excluded participants from the analysis if they contributed fewer than four accuracy measurements or fewer than 2 reaction time measurements.
+At the participant level, these steps together led to `r round(mean(is.na(d_sub$rt)), 2)`% missingness for RTs and `r round(mean(is.na(d_sub$long_window_accuracy)), 2)`% missingness for long window accuracies.
+
 ## Analytic methods {.unnumbered}

### 2025-03-14 paper/supplemental_information.Rmd
SI

--- 2025-03-13 paper/supplemental_information.Rmd
@@
 ---
-title: "Supplemental Information"
-output: pdf_document
+title: "Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood"
+output: 
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
 header-includes:
@@
 
-```{r, echo=FALSE}
-knitr::purl(here("paper","paper.Rmd"), documentation = 0)
-suppressMessages(source(here("paper.R")))
+```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
+paper <- suppressMessages(knitr::purl(here("paper","paper.Rmd"), documentation = 0))
+suppressMessages(source(paper))
 ```
@@
 
-```{r, fig.env="figure*"}
+Figure \ref{fig:dataset} gives the age distribution for each separate dataset.
+
+```{r dataset, fig.cap="\\label{fig:dataset} Age distribution for each dataset, using three-month bins.", fig.pos="h!"}
 ggplot(d_sub, aes(x = age))+ 
-  geom_histogram(binwidth = 1) + 
-  facet_wrap(~dataset_name, scale = "free_y")
+  geom_histogram(binwidth = 3) + 
+  facet_wrap(~dataset_name, scale = "free_y") +
+  xlab("Number of participants") + 
+  ylab("Age (months)")
 ```
 
-```{r, fig.cap="Number of administrations for datasets with repeated measurements."}
+Figure \ref{fig:longitudinal} shows the distribution of measurement intervals for longitudinal studies within the dataset.
+
+```{r longitudinal, fig.cap="\\label{fig:longitudinal} Number of administrations for datasets with repeated measurements, plotted by dataset.", fig.pos="h!"}
 
@@
   geom_histogram(binwidth = 1) + 
+  scale_fill_solarized(name = "Dataset") +
+  theme(legend.position = "bottom") + 
   # facet_wrap(~dataset_name) + 
-  xlab("Age since first test")
+  xlab("Age since first test") + 
+  ylab("Number of participants")
 ```
+
 # Test-Retest Reliability
@@
               names_from = "admin_num",
-              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
-                              "long_window_acc_var", "prod", "comp"))
+              values_from = c("log_rt", "long_window_accuracy", 
+                              "short_window_accuracy"))
 
@@
   summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
-            acc = cor(long_window_accuracy_1, 
-                          long_window_accuracy_2, use = "pairwise.complete.obs"))
+            acc_long = cor(long_window_accuracy_1, 
+                           long_window_accuracy_2, use = "pairwise.complete.obs"),
+            acc_short = cor(short_window_accuracy_1, 
+                            short_window_accuracy_2, use = "pairwise.complete.obs"))
 
-# note this is not correct for CDIs because some CDIs can get grouped with two admins, inflating correlations
 global_reliabilities <- d_reliability |>
@@
   summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
-            acc = cor(long_window_accuracy_1, 
-                          long_window_accuracy_2, use = "pairwise.complete.obs"))
+            acc_long = cor(long_window_accuracy_1, 
+                           long_window_accuracy_2, use = "pairwise.complete.obs"),
+            acc_short = cor(short_window_accuracy_1, 
+                            short_window_accuracy_2, use = "pairwise.complete.obs"))
 ```
@@
 We examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart.
-Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$).
+Test-retest correlations were significant but relatively modest: $\rho_{long window acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{short window acc} = `r round(global_reliabilities[3], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$.
 These reliabilities were biased downwards by three factors, however.
@@
 
-```{r, results = "asis"}
+Table \ref{tab:corrs} shows pairwise correlations between the primary variables of interest in the dataset.
+
+\begin{table}[tbp]
+\begin{center}
+\begin{threeparttable}
+\begin{tabular}{llllllllll}
+\toprule
+& \multicolumn{1}{l}{age} & \multicolumn{1}{l}{log age} & \multicolumn{1}{l}{RT} & \multicolumn{1}{l}{log RT} & \multicolumn{1}{l}{long acc} & \multicolumn{1}{l}{short acc } & \multicolumn{1}{l}{prod} & \multicolumn{1}{l}{comp}\\
+\midrule
+age &  1.00 &  &  &  &  &  &  & \\
+log age  & 0.98 & 1.00 &  &  &  &  &  &\\
+rt  & -0.33 & -0.35 & 1.00 &  &  &  &  & \\
+log rt  & -0.34 & -0.36 & 0.96 & 1.00 &  &  &  & \\
+long window accuracy  & 0.44 & 0.48 & -0.48 & -0.46 & 1.00 &  &  & \\
+short window accuracy & 0.38 & 0.43 & -0.62 & -0.61 & 0.82 & 1.00 &  & \\
+production vocabulary & 0.72 & 0.70 & -0.31 & -0.33 & 0.51 & 0.45 & 1.00 & \\
+comprehension vocabulary & 0.42 & 0.42 & -0.25 & -0.24 & 0.24 & 0.24 & 0.59 & 1.00\\
+\bottomrule
+\end{tabular}
+\end{threeparttable}
+\end{center}
+\caption{Pairwise correlations between primary variables of interest. \label{tab:corrs}}
+\end{table}
+
+```{r corrs, results = "asis", include = FALSE}
 
@@
 
-```{r results="asis"}
+Table \ref{tab:accmods} shows model comparison measures between for different models of the functional form of the relationship between accuracy and age and Table \ref{tab:rtmods} shows the same for reaction time.
+Age gradients are estimated substantially better with long window accuracies.
+Note that there are a greater number of observations for short window accuracies due to less missing data.
+We speculate that, on average, more participants looked away from the screen towards the end of trials, leading to a greater number of exclusions of long window trials based on the 50% criterion.
+Note that the total percentage of trials excluded is still small for both measures: `r round(mean(is.na(d_trial$long_window_accuracy)), 2)`% for long window accuracy and `r round(mean(is.na(d_trial$short_window_accuracy)), 2)`% for short window accuracy.
+
+\begin{table}[tbp]
+\begin{center}
+\begin{threeparttable}
+\begin{tabular}{lllllllll}
+\toprule
+n.
+obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
+\midrule
+48354 & 0.28 & -8,338.83 & 16,691.65 & 16,753.16 & 16,677.65 & 48347 & Long window, linear age & 0.14\\
+48354 & 0.28 & -8,305.39 & 16,624.78 & 16,686.28 & 16,610.78 & 48347 & Long window, log age & 0.12\\
+50244 & 0.31 & -12,486.60 & 24,987.21 & 25,048.98 & 24,973.21 & 50237 & Short window, linear age & 0.09\\
+50244 & 0.31 & -12,458.77 & 24,931.54 & 24,993.31 & 24,917.54 & 50237 & Short window, log age & 0.08\\
+\bottomrule
+\end{tabular}
+\end{threeparttable}
+\end{center}
+\caption{Model comparison metrics for different functional forms of the relationship between accuracy and age. \label{tab:accmods}}
+\end{table}
+
+\begin{table}[tbp]
+\begin{center}
+\begin{threeparttable}
+\begin{tabular}{lllllllll}
+\toprule
+n.
+obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
+\midrule
+18940 & 0.45 & -12,414.69 & 24,843.38 & 24,898.32 & 24,829.38 & 18933 & Log RT, linear age & 0.21\\
+18940 & 0.45 & -12,393.27 & 24,800.53 & 24,855.48 & 24,786.53 & 18933 & Log RT, log age & 0.21\\
+18940 & 570.49 & -147,644.43 & 295,302.87 & 295,357.81 & 295,288.87 & 18933 & Linear RT, linear age & 0.18\\
+18940 & 570.02 & -147,622.67 & 295,259.34 & 295,314.29 & 295,245.34 & 18933 & Linear RT, log age & 0.18\\
+\bottomrule
+\end{tabular}
+\end{threeparttable}
+\end{center}
+\caption{Model comparison metrics for different functional forms of the relationship between RT and age. \label{tab:rtmods}}
+\end{table}
+
+Note that with our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns (Heathcote, Brown, \& Mewhort, 2000).
+Our current results apply to the form of the age gradient as opposed to the form of any individual's pattern of developmental change.
+
+```{r accmods, results="asis", include = FALSE}
 papaja::apa_table(acc_mods_lmer_summary)
@@
 
-# Short Window Analysis
+<!-- # Short Window Analysis -->
 
-Need to redo all figures using short windows ugh.
+<!-- Need to redo all figures using short windows ugh.  -->
+# References
+
+Heathcote, A., Brown, S., \& Mewhort, D. J. K. (2000). The power law repealed: The case for an exponential law of practice. Psychonomic Bulletin & Review, 7(2), 185–207.

### 2025-03-14 paper.R

@@ -1,558 +1,0 @@
-suppressPackageStartupMessages(library(here))
-suppressPackageStartupMessages(library(psych))
-suppressPackageStartupMessages(library(lmerTest))
-suppressPackageStartupMessages(library(lavaan))
-suppressPackageStartupMessages(library(tidySEM))
-suppressPackageStartupMessages(library(xtable))
-suppressPackageStartupMessages(library(papaja))
-
-source(here("helper","common.R"))
-
-source(here("helper/modified_sem_plot.R"))
-
-d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
-  mutate(log_age = log(age))
-d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds")) |>
-  group_by(subject_id) |>
-  arrange(age) |>
-  mutate(admin_num = 1:n(), 
-         time_since_t0 = age - age[1],
-         delta_t = c(0, diff(age)))
-
-d_trial$age_s <- scale(d_trial$age)[,1]
-d_trial$log_age_s <- scale(d_trial$log_age)[,1]   
-d_sub$age_s <- scale(d_sub$age)[,1]
-d_sub$log_age_s <- scale(d_sub$log_age)[,1]   
-
-
-AGEVAR_CUTOFF <- 6
-
-knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
-
-dataset_characteristics <- d_sub |>
-  group_by(dataset_name) |>
-  summarise(`N subjects` = length(unique(subject_id)),
-            `N admins` = n(), 
-            `Mean Age` = mean(age), 
-            `Min Age` = min(age), 
-            `Max Age` = max(age), 
-            `Avg Trials` = mean(n_trials[!is.na(long_window_accuracy)]),
-            `Avg RT Trials` = mean(n_trials_rt[!is.na(rt)]),
-            `CDIs` = ifelse(!is.na(any(prod)), "x", ""), 
-            `longitudinal` = ifelse(any(duplicated(subject_id)), "x", "")) |>
-  arrange(desc(`N subjects`)) 
-
-dataset_characteristics <- dataset_characteristics |>
-  bind_rows(tibble(dataset_name = "Total", 
-                       `N subjects` = sum(dataset_characteristics$`N subjects`),
-                       `N admins` = sum(dataset_characteristics$`N admins`), 
-                       `Mean Age` = mean(dataset_characteristics$`Mean Age`),
[…]

### 2025-03-20 paper/paper.Rmd
add timecourse figure

@@
 suppressPackageStartupMessages(library(nlme))
+suppressPackageStartupMessages(library(cowplot))
 
@@
 
+#load d_aoi for timecourse plot
+d_aoi <- readRDS(here("cached_intermediates","1_d_aoi.Rds"))
 d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
@@
 
+```{r timecourse, fig.cap="Timecourse of word recognition at different ages.
+The x-axis shows time (in ms) from the onset of the target label (vertical solid line).
+Colored lines show the average increase in proportion target looking post label onset at each age bin (in months).
+The dashed horizontal line represents chance looking.
+Error bands represent standard errors of the mean.
+Grey backgrounds highlight the short and long time windows used in subsequent analyses.", fig.env="figure", fig.height = 3, fig.width=4, out.width=".5\\textwidth", include=TRUE}
+
+#summarize by subject
+d_timecourse_by_dataset_age_subj <- d_aoi |>
+  filter(t_norm>-2000) |> 
+  #create age bins cutoffs
+  mutate(
+    age_bin = cut(age, breaks = c(0,12,15,18,21,24,27,30,36,60), 
+                  labels = c("<12", "12-15","15-18", "18-21", "21-24", "24-27","27-30","30-36", ">36"))
+  ) |>
+  #downsample t_norm to 50ms bins
+  #mutate(t_norm_downsampled = round(t_norm/50)*50) |>
+  group_by(dataset_name,dataset_id,subject_id,age_bin,t_norm) |> 
+  #calculate proportion time spent looking at target
+  summarize(
+    n=n(),
+    accuracy = mean(correct, na.rm=TRUE))
+
+#by age bin, not representing dataset structure
+d_timecourse_by_age_overall <- d_timecourse_by_dataset_age_subj |>
+  filter(n>=5) |> 
+  group_by(age_bin,t_norm) |>
+  summarize(
+    n_subj=n(),
+    proportion_target_looking = mean(accuracy,na.rm=TRUE),
+    sem = sd(accuracy,na.rm=TRUE)/sqrt(n_subj)
+  ) |> 
+  rename(accuracy=proportion_target_looking)
+
+#overall plot
+d_timecourse_by_age_overall |> 
+  filter(age_bin!="<12") |> 
+  filter(t_norm>-1000) |> 
+  ggplot(aes(x=t_norm, y=accuracy, group=age_bin, color=age_bin)) +
+  geom_rect(xmin=2000,xmax=4000,ymin=0,ymax=1.2,fill="#DFDFDF",color="#DFDFDF",alpha=0.05)+
+  geom_rect(xmin=200,xmax=2000,ymin=0,ymax=1.2,fill="#F2F2F2",color="#F2F2F2",alpha=0.05)+
+  geom_smooth(stat="identity", aes(y=accuracy, ymin=accuracy-sem, ymax=accuracy+sem,fill=age_bin))+
+  #geom_smooth(stat="smooth")+
+  geom_hline(yintercept=0.5, linetype="dashed")+
+  geom_vline(xintercept=0, linetype="solid")+
+  #add text
+  annotate("text", x=50, y=0.85, label="Label Onset", hjust=0, vjust=0,size=3)+
+  annotate("text", x=400, y=0.99, label="SHORT WINDOW",hjust=0, vjust=0,size=2.5)+
+  annotate("text", x=1500, y=0.93, label="LONG WINDOW",hjust=0, vjust=0,size=2.5)+
+  geom_segment(aes(x = 1000, y = 0.98, xend = 200, yend = 0.98),
+               arrow = arrow(type = "closed", angle = 20, length = unit(0.05, "inches")), color="black")+
+  geom_segment(aes(x = 1000, y = 0.98, xend = 2000, yend = 0.98),
+               arrow = arrow(type = "closed", angle = 20, length = unit(0.05, "inches")), color="black")+
+  geom_segment(aes(x = 2000, y = 0.92, xend = 200, yend = 0.92),
+               arrow = arrow(type = "closed", angle = 20, length = unit(0.05, "inches")), color="black")+
+  geom_segment(aes(x = 2000, y = 0.92, xend = 4000, yend = 0.92),
+               arrow = arrow(type = "closed", angle = 20, length = unit(0.05, "inches")), color="black")+
+  scale_color_viridis_d(name="Age Bin\n(mos)")+
+  scale_fill_viridis_d(name="Age Bin\n(mos)")+
+  #facet_wrap(~age_bin)+
+  scale_y_continuous(breaks=seq(0.4,1,0.1),limits=c(0.4,1))+
+  scale_x_continuous(breaks=seq(-1000,4000,1000),limits=c(-1050,4000))+
+  xlab("Time from trial onset (ms)")+
+  ylab("Proportion Target Looking")+
+  theme_cowplot()+
+  theme(legend.key.size = unit(0.3, 'cm'), #change legend key size
+        legend.key.height = unit(0.3, 'cm'), #change legend key height
+        legend.key.width = unit(0.3, 'cm'), #change legend key width
+        legend.text = element_text(size=9),
+        legend.title = element_text(size=9),
+        legend.position=c(0.03,0.6),
+        axis.title = element_text(size=12),
+        axis.text = element_text(size=10))
+```
+
 ```{r acc_mods}
@@
 
-We began by characterizing age gradients in speed and accuracy across children.
-We computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
+We began by examining developmental changes in children's word recognition.
+Figure \ref{fig:timecourse} depicts the average timecourse at different ages across all datasets.
+Intuitively, these timecourses show gradual increases in accuracy (higher overall proportion target looking) and speed (faster looking to the target after hearing a label) as age increases.
+To systematically characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms).

### 2025-03-20 1_get_data.qmd

@@
 load(file = here("..", "peekbank-method", "cached_intermediates", "1_d_aoi.Rds"))
+#cache this here as an interim hack until we bring in the full pipeline
+saveRDS(d_aoi, here("cached_intermediates","1_d_aoi.Rds"))
 ```

### 2025-03-20 paper/paper.Rmd
paper updates

@@
   - name: Mika Braginsky
+    affiliation: a
+  - name: George Kachergis
     affiliation: a

### 2025-03-21 paper/paper.Rmd
fix a minor aesthetic issue with the timecourse

@@
   ggplot(aes(x=t_norm, y=accuracy, group=age_bin, color=age_bin)) +
-  geom_rect(xmin=2000,xmax=4000,ymin=0,ymax=1.2,fill="#DFDFDF",color="#DFDFDF",alpha=0.05)+
-  geom_rect(xmin=200,xmax=2000,ymin=0,ymax=1.2,fill="#F2F2F2",color="#F2F2F2",alpha=0.05)+
+  geom_rect(xmin=2000,xmax=4000,ymin=0,ymax=1.2,fill="#DFDFDF",color=NA,alpha=0.05)+
+  geom_rect(xmin=200,xmax=2000,ymin=0,ymax=1.2,fill="#F2F2F2",color=NA,alpha=0.05)+
   geom_smooth(stat="identity", aes(y=accuracy, ymin=accuracy-sem, ymax=accuracy+sem,fill=age_bin))+

### 2025-04-03 paper/paper.Rmd
author affil tweak

--- 2025-03-20 paper/paper.Rmd
@@
   - name: Martin Zettersten
-    affiliation: f
+    affiliation: g
 address:
@@
   - code: f
-    address: Department of Communication, Stanford University, Stanford, CA, USA
+    address: Department of Psychology, Ludwig Maximilian University, Munich, Germany
   - code: g
-    address: Department of Psychology, Ludwig Maximilian University, Munich, Germany
+    address: Department of Cognitive Science, University of California, San Diego, CA, USA
 

### 2025-04-03 paper/paper.Rmd
changes to paper

@@
 # For footer text
-lead_author_surname: Anonymous
+lead_author_surname: Frank
 
@@
 And how does facility in word recognition relate to the growth of vocabulary knowledge?
-We address these questions using data from Peekbank, an open database of experiments measuring children's eye-movements during early word recognition.
+We address these questions using data from Peekbank, an open database of experiments measuring children's eye movements during early word recognition.
 Combining 24 datasets from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development.
@@
   The efficiency with which children recognize words provides a measurement of their real-time language processing, which has been argued to be a key part of the language learning process.
-Here, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster growing vocabularies.
-These data suggest that understanding language in real-time is an important part of eearly language learning and that it improves gradually with practice across early childhood.
+Here, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster-growing vocabularies.
+These data suggest that understanding language in real time is an important part of eearly language learning and that it improves gradually with practice across early childhood.
 
@@
 output: rticles::pnas_article
+
+nocite: |
+  @yurovsky2017,
+  @yurovksky_2017,
+  @weaver2024,
+  @fernald2012,
+  @sander-montant2023,
+  @frank2016,
+  @baumgartner2014,
+  @fernald2013,
+  @Adams2018,
+  @PotterLewWilliamsUnpub,
+  @fernald2006,
+  @pomper2016,
+  @Swingley2002,
+  @pomper2019,
+  @perry2017,
+  @ronfard2022,
+  @bacon2022,
+  @Garrison2020,
+  @Mahr2015,
+  @ferguson2014,
+  @potter2019,
+  @newman2017
 ---
@@
 Eye tracking has thus emerged as a key method that allows the measurement of language comprehension with high temporal resolution: both adults and children reliably fixate the referent of a word soon after it is used [@tanenhaus1995integration;@altmann1999incremental;@fernald1998;@fernald2008;@macdonald2018].
-The relative timecourse of fixation then can provide an index of the comprehender's ability or the difference between two stimulus conditions.
+The relative timecourse of fixation then can provide an index of an individual comprehender's ability or be used to measure the difference between two stimulus conditions.
 
-The version of this method that is used with children goes by many names, including the intermodal preferential looking paradigm and the "looking while listening" paradigm (LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
-In LWL experiments, children are typically shown a series of trials in which two images are displayed side by side and they are asked to find one of them.
+The version of this method that is used with children goes by many names, including the "intermodal preferential looking" paradigm and the "looking while listening" paradigm (LWL, the name we adopt here) [@hirsh1996intermodal;@reznick1990visual;@fernald2008].
+In LWL experiments, children are typically shown two images displayed side by side and asked to find one of them.
 For example, a ball and a book might be shown, and the child might hear "Look at the ball!
 Can you find it?".
-Accuracy is then computed as the proportion of time they fixate the correct image within a fixed window after the onset of the noun ("ball" in this case).
-Reaction time is computed only on trials in which the child is fixating on the distractor image (the book) before word onset; in these cases, the average time it takes until the child shifts fixation from the distractor to the target is used as an index of processing speed.
+Accuracy is then computed as the proportion of time their eyes fixate the correct image within a fixed window after the onset of the noun ("ball" in this case).
+Reaction time is computed only on trials in which the child is fixating the distractor image (the book) at word onset; in these cases, the average time it takes for the child to shift fixation from the distractor to the target image is used as an index of processing speed.
 Early work using this method showed that both children's speed and accuracy increase rapidly across the second year [@fernald1998;@reznick1990visual].
@@
 But each word must first be recognized during the short window of time when it is present in the child's memory.
-Consider a child hearing the utterance "Can you put the ball in the crate?" The faster and more accurately the child can recognize the word "ball", the better they can use this evidence to help infer the speaker's intended meaning, perhaps in the process making inferences about the meaning of the less familiar word "crate" [@frank2009using].
+Consider a child hearing the utterance "Can you put the ball in the crate?" The faster and more accurately the child can recognize the word "ball", the better they can use this evidence to help infer the speaker's intended meaning, allowing possible inferences about the meaning of the less familiar word, "crate" [@frank2009using].
 Consistent with this idea, one important study found that children's word recognition speed mediated the longitudinal relationship between home language input and vocabulary growth [@Weisleder2013].
@@
   \hline
-1 & reflook\_v4 & 310 & 310 & 36.82 & 12.24 & 60.00 & 6.21 & 2.87 &  &  \\ 
-  2 & yurovsky\_2017 & 282 & 282 & 25.64 & 12.59 & 58.65 & 5.95 & 2.79 &  &  \\ 
-  3 & weaver\_zettersten\_2024 & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 &  & x \\ 
-  4 & fernald\_marchman\_2012 & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.89 & 7.21 & x & x \\ 
-  5 & sander-montant\_2022 & 122 & 122 & 21.94 & 12.02 & 31.11 & 10.07 & 4.20 & x &  \\ 
-  6 & frank\_tablet\_2016 & 104 & 104 & 33.89 & 12.13 & 59.84 & 5.87 & 2.74 &  &  \\ 
-  7 & baumgartner\_2014 & 100 & 100 & 12.01 & 12.00 & 13.00 & 4.00 & 2.31 & x &  \\ 
-  8 & fmw\_2013 &  80 & 179 & 20.04 & 17.00 & 26.00 & 21.26 & 8.44 & x & x \\ 
-  9 & adams\_marchman\_2018 &  69 & 711 & 23.58 & 13.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
-  10 & potter\_canine &  67 &  67 & 23.76 & 21.00 & 27.00 & 10.38 & 3.80 & x &  \\ 
-  11 & fernald\_totlot &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.25 & 6.06 & x & x \\ 
-  12 & pomper\_prime &  63 &  63 & 39.73 & 37.90 & 42.00 & 12.98 & 5.34 &  &  \\ 
-  13 & pomper\_saffran\_2016 &  60 &  60 & 44.27 & 41.00 & 47.00 & 7.55 & 3.30 &  &  \\ 
-  14 & swingley\_aslin\_2002 &  50 &  50 & 15.09 & 14.13 & 16.00 & 11.74 & 3.79 & x &  \\ 
-  15 & pomper\_salientme &  44 &  44 & 40.11 & 38.00 & 43.00 & 5.30 & 2.34 &  &  \\ 
-  16 & perry\_cowpig &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 & x &  \\ 
-  17 & ronfard\_2021 &  40 &  40 & 19.95 & 18.00 & 24.00 & 18.54 & 7.62 & x &  \\ 
-  18 & bacon\_gendercues &  38 &  38 & 22.87 & 22.00 & 24.00 & 18.16 & 8.00 & x &  \\ 
-  19 & garrison\_bergelson\_2020 &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.47 & 9.41 & x &  \\ 
-  20 & pomper\_yumme &  32 &  32 & 26.38 & 25.00 & 28.00 & 7.62 & 2.31 &  &  \\ 
-  21 & mahr\_coartic &  29 &  29 & 20.83 & 18.10 & 23.80 & 24.38 & 8.86 & x &  \\ 
-  22 & ferguson\_eyetrackingr &  28 &  28 & 19.71 & 18.02 & 21.86 & 5.54 & 2.33 & x &  \\ 
-  23 & potter\_remix &  23 &  44 & 22.59 & 18.00 & 29.00 & 6.07 & 2.58 & x & x \\ 
-  24 & newman\_genderdistractor &  19 &  19 & 30.12 & 29.23 & 30.84 & 13.11 & 4.94 &  &  \\ 
+1 & Yurovsky et al. (2017) & 310 & 310 & 36.82 & 12.24 & 60.00 & 6.21 & 2.87 &  &  \\ 
+  2 & Yurovsky et al. (2017b) & 282 & 282 & 25.64 & 12.59 & 58.65 & 5.95 & 2.79 &  &  \\ 
+  3 & Weaver \& Zettersten (2024) & 141 & 248 & 15.73 & 13.50 & 23.60 & 18.10 & 6.72 &  & x \\ 
+  4 & Fernald \& Marchman (2012) & 122 & 678 & 23.92 & 17.00 & 32.00 & 16.89 & 7.21 & x & x \\ 
+  5 & Sander-Montant et al. (2023) & 122 & 122 & 21.94 & 12.02 & 31.11 & 10.07 & 4.20 & x &  \\
+  6 & Frank et al. (2016) & 104 & 104 & 33.89 & 12.13 & 59.84 & 5.87 & 2.74 &  &  \\ 
+  7 & Baumgartner (2014) & 100 & 100 & 12.01 & 12.00 & 13.00 & 4.00 & 2.31 & x &  \\ 
+  8 & Fernald et al. (2013) &  80 & 179 & 20.04 & 17.00 & 26.00 & 21.26 & 8.44 & x & x \\ 
+  9 & Adams et al. (2018) &  69 & 711 & 23.58 & 13.00 & 38.00 & 18.44 & 7.92 & x & x \\ 
+  10 & Potter \& Lew Williams (n.d.) &  67 &  67 & 23.76 & 21.00 & 27.00 & 10.38 & 3.80 & x &  \\ 
+  11 & Fernald et al. (2006) &  63 & 229 & 19.68 & 15.00 & 25.00 & 15.25 & 6.06 & x & x \\ 
+  12 & Pomper \& Saffran (n.d.) &  63 &  63 & 39.73 & 37.90 & 42.00 & 12.98 & 5.34 &  &  \\ 
+  13 & Pomper \& Saffran (2016) &  60 &  60 & 44.27 & 41.00 & 47.00 & 7.55 & 3.30 &  &  \\ 
+  14 & Swingley \& Aslin (2002) &  50 &  50 & 15.09 & 14.13 & 16.00 & 11.74 & 3.79 & x &  \\ 
+  15 & Pomper \& Saffran (2019) &  44 &  44 & 40.11 & 38.00 & 43.00 & 5.30 & 2.34 &  &  \\ 
+  16 & Perry \& Saffan (2017) &  42 &  42 & 20.45 & 19.00 & 22.00 & 14.88 & 5.43 & x &  \\ 
+  17 & Ronfard, Wei, \& Rowe (2022) &  40 &  40 & 19.95 & 18.00 & 24.00 & 18.54 & 7.62 & x &  \\ 
+  18 & Bacon \& Saffran (2022) &  38 &  38 & 22.87 & 22.00 & 24.00 & 18.16 & 8.00 & x &  \\ 
+  19 & Garrison \& Bergelson (2020) &  35 &  35 & 14.46 & 12.00 & 18.00 & 27.47 & 9.41 & x &  \\ 
+  20 & Pomper \& Saffran (n.d.) &  32 &  32 & 26.38 & 25.00 & 28.00 & 7.62 & 2.31 &  &  \\ 
+  21 & Mahr et al. (2015) & 29 &  29 & 20.83 & 18.10 & 23.80 & 24.38 & 8.86 & x &  \\ 
+  22 & Ferguson, Graf, \& Waxman (2014) & 28 &  28 & 19.71 & 18.02 & 21.86 & 5.54 & 2.33 & x &  \\ 
+  23 & Potter et al. (2019) &  23 &  44 & 22.59 & 18.00 & 29.00 & 6.07 & 2.58 & x & x \\ 
+  24 & Newman \& Morini (2017)  &  19 &  19 & 30.12 & 29.23 & 30.84 & 13.11 & 4.94 &  &  \\ 
   \hline
@@
 \caption{Characteristics of included datasets from Peekbank.
-`Admins` denotes separate experimental sessions.
-`CDIs` refers to whether the dataset contains parent report vocabulary data from the MacArthur-Bates Communicative Development Inventory.
+"Admins" denotes separate experimental sessions.
+"CDIs" refers to whether the dataset contains parent report vocabulary data from the MacArthur-Bates Communicative Development Inventory.
 \label{tab:datasets}}
@@
 
-Word recognition speed has been used as an index of individual differences in early childhood [@fernald2006;@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow].
+Word recognition speed has also been used as an index of individual differences in early childhood [@fernald2006;@marchman2008;@peter2019;@fernald2012;@newbury2016interrelationships] and beyond [@colby2023efficiency;@jeppsen2022development;@mcmurray2022slow].
 Over and above measures of vocabulary size, word recognition speed at 18 months predicts children's standardized test scores years later [@marchman2008].
-Further, faster processing at 18 months is predictive of whether "late talkers" catch up to their peers or require further intervention [@fernald2012].
-Critically, these assessments use words that children at the target age are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately they can recognize a spoken word and use it to guide their visual attention to a referent.
+Further, faster processing at 18 months is predictive of whether "late talkers" catch up to their peers or could benefit from further intervention [@fernald2012].
+Critically, these assessments use words that children at the target age are reported to understand and produce -- they are not indices of vocabulary size but rather of how quickly and accurately the child can recognize a familiar spoken word and use it to guide their visual attention to a referent.
 
-Yet individual experiments measuring processing with young children typically recruit relatively small samples in a restricted range of ages.
-These samples provide neither the breadth of ages nor the number of participants needed to estimate the broader dynamics of how word recognition changes developmentally and how it connects with other aspects of early language development (see [@colby2023efficiency; @mcmurray2022slow] for examples of these analyses in school-aged children).
-To overcome this limitation, we turned to Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank].
-This dataset unifies and carefully curates a large amount of eye-tracking data from studies with infants and toddlers, representing cumulatively over 12 million individual samples of children's eye movements while processing language.
+Yet individual experiments measuring processing speed with young children typically recruit relatively small samples in a restricted range of ages.
+These samples provide neither the breadth of ages nor the number of participants needed to estimate how word recognition changes developmentally and how it connects with other aspects of early language development (see [@colby2023efficiency; @mcmurray2022slow] for examples of these analyses in school-aged children).
+To overcome this limitation, we created Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank].
+This dataset unifies and carefully curates a large amount of eye-tracking data from studies with infants and toddlers, representing cumulatively over 12 million individual samples of children's eye movements during real-time language processing.
 This resource allows us to gain an unprecedented view of the development of word recognition across a large sample of children.
 
-Here we investigate two specific hypotheses.
+We investigate two specific hypotheses here.
 First, one influential theory posits that language learning is a process of skill learning, in which the child is learning the skill of fluent conversation with other language users [@christiansen2016;@chater2018].
@@
 @heathcote2000].
-In our case, we expect that chronological age is a proxy for experience and so the logarithm of processing speed should decrease linearly with the logarithm of age.
+In our case, we expect that chronological age is a proxy for experience and so the logarithm of reaction time should decrease linearly with the logarithm of age.
 Finally, trial-to-trial variability in both speed and accuracy should decrease with increasing expertise, as is found in studies of motor expertise [@todorov2002optimal].
@@
 We retrieved data from Peekbank, focusing on data from English-speaking children ages 1--6 years and on simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ...").
+While other experimental manipulations and languages are included in the database, we narrowed our sample to English-speaking children because they are well-represented across our age range and excluded manipulations which aimed to capture phenomena other than simple concrete noun reference (e.g., adjective comprehension or novel word learning).
 These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions).
+
 Table \ref{tab:datasets} shows the characteristics of individual datasets (see also SI: Dataset Description).
-The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed study of the development of word recognition than has previously been possible.
-While our analysis is exploratory in nature, it is guided by the two hypotheses outlined above: the presence of 1) signatures of skill learning in word recognition, and 2) linkages between word recognition and vocabulary.
+The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm all allowed us to make a more detailed study of the development of word recognition than has previously been possible.
+While our analyses are exploratory in nature, they are guided by the two hypotheses outlined above: the presence of 1) signatures of skill learning in word recognition, and 2) linkages between word recognition and vocabulary.
 
@@
 Colored lines show the average increase in proportion target looking post label onset at each age bin (in months).
+Age bins are larger for older children due to decreased data density.
 The dashed horizontal line represents chance looking.
@@
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms).
-For each window, we averaged all fixation within the window to compute a continuous proportion target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
+For each window, we averaged all fixations within the window to compute a continuous proportion of target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
 In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy (see SI: Test-Retest Reliability); in subsequent analyses we estimate latent variables representing these constructs.
@@
 The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison).
-Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analysis.
+Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analyses.
 
@@
 In sum, we see continuing improvements in word recognition across the full age range in our dataset that appear roughly linear in the logarithm of age.
-These logarithmic relationships follow theoretical expectations that both speed and accuracy should gradually asymptote to mature levels of performance as in skill learning more generally [@anderson1982,@snoddy1926].
+These logarithmic relationships follow theoretical expectations that both speed and accuracy should gradually asymptote to mature levels of performance, as seen in skill learning more generally [@anderson1982,@snoddy1926].
 
@@
 The solid blue line shows a linear fit and associated confidence interval.
-Individual dotted lines show linear fits for those datasets spanning six or more months of age.
+Thin colored lines show linear fits for those datasets spanning six or more months of age.
 The dashed line for accuracy shows chance-level looking (.5)", fig.env="figure*", fig.height = 2.5}
@@
               aes(group = dataset_name, col = dataset_name), se=FALSE,
-              alpha = 1, method = "lm", size = .5, lty = 3) + 
+              alpha = 1, method = "lm", size = .25, lty = 1) + 
   geom_smooth(method = "lm") +
@@
               aes(group = dataset_name, col = dataset_name), se=FALSE,
-              alpha = .1, method = "lm", size = .5, lty = 3) + 
+              alpha = .1, method = "lm", size = .25, lty = 1) + 
   geom_smooth(method = "lm") +
@@
 Both within and across datasets, within-individual variation in speed and accuracy decreased across the developmental range we examined (Figure \ref{fig:variance}).
-We fit mixed effects models to the standard deviation of both speed and accuracy for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
+We fit mixed-effects models predicting the standard deviation of both speed and accuracy for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
 For both speed and accuracy, within-individual variability decreased with age (speed: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`).
-Thus, as well as being faster and more accurate, older children were more consistent in their online word recognition than younger children.
+Thus, as well as being faster and more accurate, older children were more consistent in their real-time word recognition than younger children.
 
@@
               aes(group = dataset_name, col = dataset_name), se=FALSE,
-              alpha = .1, method = "lm", size = .5, lty = 3) + 
+              alpha = .1, method = "lm", size = .25, lty = 1) + 
     geom_smooth(method = "lm") +
@@
 We were next interested in whether the various aspects of word recognition -- including speed, accuracy, and the variability of each of these -- were related to other aspects of early language ability.
-Of the studies in our database, `r sum(str_length(dataset_characteristics$CDIs)>0)` gathered parent reports about children's early vocabulary using the MacArthur-Bates Communicative Development Inventory (CDI), a popular survey instrument that provides a reliable and valid estimate of children's early vocabulary [@marchman2023;@frank2021].
+Of the studies in our database, `r sum(dataset_characteristics$CDIs == "x")` gathered parent reports about children's early vocabulary using the MacArthur-Bates Communicative Development Inventory (CDI), a popular survey instrument that provides a reliable and valid estimate of children's early vocabulary [@marchman2023;@frank2021].
 Different forms of the CDI can be used to measure either receptive and expressive vocabulary (for children up to 18 months) or expressive vocabulary only (for children 16 -- 30 months).
@@
 To better accommodate missing data under the assumption of data missing at random (e.g., due to dataset), we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
-The best fitting model of those we examined was a three-factor model with factors for speed (RT and RT variability), accuracy (proportion looking to target on each trial and associated variability of this measure), and vocabulary (comprehension and production from the CDI).
+The best fitting model was a three-factor model with factors for speed (RT and RT variability), accuracy (proportion looking to target on each trial and associated variability of this measure), and vocabulary (comprehension and production from the CDI).
 Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`); see SI: Alternative Factor Structures).
@@
 * indicates p < .05, ** indicates p < .01, and *** indicates p < .001.
-Covariances reflect age-residualized correlations between variables.
-Abbreviations: acc = accuracy, prod = production vocabulary, comp = comprehension vocabulary.", fig.height = 3}
+Covariances reflect age-residualized correlations between variables.", fig.height = 3}
 
@@
 The non-age related variance of all three latent factors was significantly related to that of the other factors, with speed and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$).
-This model supports the idea that variation in speed and accuracy of word recognition is related to differences in parent-reported vocabulary beyond the effects of age.
-These analyses all treat between-person differences, however, rather than modeling change in these factors within individuals.
+This model supports the idea that variation in speed and accuracy of word recognition is related to individual differences in parent-reported vocabulary beyond the effects of age.
+These analyses treat all data as between person, however, rather than modeling change in these factors within individuals.
 
@@
 
-To investigate within-person relationships between LWL and vocabulary, we began by using longitudinal growth models.
+To investigate within-person relationships between LWL and vocabulary, we began by fitting longitudinal growth models to the portion of the data containing multiple LWL sessions for individual children.
 We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in speed during the initial session of the study.
@@
 
-```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for the 25th and 75th percentile on initial reaction times.
+```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for the 25th (red) and 75th (blue) percentile on initial reaction times.
 Individual longitudinal trajectories are shown in light gray.
@@
 
-On the other hand, it is possible that differences in predicted growth trajectories are due to overall coupling between vocabulary size and language processing, rather than a specifically predictive relationship between $t_0$ RT and vocabulary growth.
+On the other hand, it is possible that differences in predicted growth trajectories are due to coupling between vocabulary size and language processing across the entire developmental period, rather than a predictive relationship specifically between $t_0$ RT and vocabulary growth.
 To test this relationship, we used longitudinal structural equation models.
-We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months ($t_0, ..., t_4$) and fit individual growth across each of these variables.
+We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months from the initial measurement (i.e., $t_0, ..., t_4$) and fit individual growth across each of these variables.
 We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset.
@@
 Our key question of interest concerned coupling among the (latent) intercepts and slopes of these growth models.
-Consistent with the idea that overall faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
+Consistent with our earlier analysis showing that faster processing is related to vocabulary growth, we saw significant coupling between processing speed intercepts and vocabulary growth slopes ($\beta=$ `r round(slopes_long_summary$pe[63,"est"],3)`, SE = `r round(slopes_long_summary$pe[63,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[63,"pvalue"],3)`) as well as a variety of other couplings.
 On the other hand, there was not significant coupling between *growth* in speed and *growth* in vocabulary ($\beta=$ `r round(slopes_long_summary$pe[65,"est"],3)`, SE = `r round(slopes_long_summary$pe[65,"se"],3)`, $p =$ `r round(slopes_long_summary$pe[65,"pvalue"],3)`).
-This null effect could be interpreted as being consistent with these abilities growing independently, but we cannot rule out other possibilities.
-First, the longitudinal data we had might not allow sufficiently precise estimates of growth slopes, or second, since vocabulary growth is non-linear, the linear model we used here might not have captured coupling among nonlinear aspects of developmental change.
+This null effect could be interpreted as being consistent with these abilities growing independently, but there are other possibilities.
+First, the longitudinal data we had might not have allowed sufficiently precise estimates of growth slopes, or second, since vocabulary growth is non-linear, the linear model we used here might not have captured coupling among nonlinear aspects of developmental change.
 
@@
 We investigated these questions using a new, large-scale dataset of developmental eye-tracking measurements compiled across many prior studies.
-The age-gradients for speed and accuracy indicate that both improve asymptotically.
-Gradients for recognition speed were consistent with the log-log relationship associated with the "power law of practice", that is, with a gradual convergence on mature levels of processing efficiency.
+The age-gradients for speed and accuracy indicated that both improve asymptotically.
+Gradients for recognition speed were consistent with the log-log relationship associated with the "power law of practice," that is, with a gradual convergence to mature levels of processing efficiency.
 Further, the age gradient suggested that trial-to-trial variability decreases with age, consistent with both the literature on skill learning [@todorov2002optimal] and other work on developmental changes in variability [@judd2021working;@galeano2018variability;@kautto2023introducing].
@@
 By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had sample sizes at least an order of magnitude smaller than ours.
-Our approach was to build on the time-consuming and meticulous data collection from previous infant and toddler eye-tracking studies --- representing cumulatively many thousands of hours of in-lab data collection and hand-annotation of the resulting videos of child looking behavior --- by harmonizing these data into a single, large-scale database.
+Our approach was to build on the time-consuming and meticulous data collection from previous infant and toddler eye-tracking studies -- representing cumulatively many thousands of hours of in-lab data collection and hand-annotation of the resulting videos of child looking behavior -- by harmonizing these data into a single, large-scale database.
 This approach illustrates how building harmonized databases can be especially powerful when composed of high-effort and high-quality datasets that are smaller in scope, maximizing the impact of previous data collection efforts and allowing us to ask broader questions about developmental change [@frank2021].
 In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to explore the relationships between different variables.
-Because early language is so variable, these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [@bergelson2023;@frank2017;@frank2021;@manybabies2020].
+Because early language is so variable, these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [@bergelson2023;@frank2021;@manybabies2020].
 
@@
 
-Nevertheless, consistency of the trends we observed across datasets suggests that our qualitative conclusions are robust to some significant sociodemographic variation.
-Nevertheless, these findings are still limited in their generalizability by the convenience samples that were used in the studies aggregated in Peekbank.
+The consistency of the trends we observed across datasets suggests that our qualitative conclusions are robust to some significant cross-laboratory and cross-sociodemographic variation.
+Nevertheless, these findings are still limited in their generalizability by the convenience samples that were used in most of the studies aggregated in Peekbank.
 These studies typically (but not always) represent children from well-educated parents living in university-adjacent communities.
@@
 More broadly, our results here suggest the continued importance of the looking-while-listening paradigm as an index of children's language processing abilities.
-If language learning is, at least in part, a process of skill learning, then measurement of this skill is a critical window into understanding the remarkable process of language learning.
+If language learning is, at least in part, a process of skill learning, then measurement of this skill in larger samples provides a critical window into understanding the remarkable process of language learning.
 
@@
 We included information from `r length(unique(d_sub$subject_id))` unique participants across `r length(unique(d_sub$dataset_name))` datasets.
-Dataset information is given in Table \ref{tab:datasets}.Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed data only from simple word recognition trials; these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations.
+Dataset information is given in Table \ref{tab:datasets}.
+Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed only data from simple word recognition trials; these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations.
 We focus here on English purely for practical reasons -- the Peekbank dataset at present contains limited data from other languages.

### 2025-04-04 paper/paper.Rmd
stephan

@@
 We address these questions using data from Peekbank, an open database of experiments measuring children's eye movements during early word recognition.
-Combining 24 datasets from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development.
-Factor analysis reveals a cross-sectional coupling of word recognition speed and accuracy with parent-reported vocabulary.
-Across a range of longitudinal models, speed, accuracy, and vocabulary also show coupled growth such that children with faster word recognition tend to show faster vocabulary growth.
+Combining 24 datasets from almost 2,000 children ages 1--6 years, we show that word recognition becomes faster, more accurate, and less variable across development, consistent with a process of skill learning.
+Factor analysis reveals covariation of word recognition speed and accuracy with children's vocabulary size in cross-sectional analysis.
+Further, across a range of longitudinal models, speed, accuracy, and vocabulary show coupled growth such that children with faster word recognition tend to show faster vocabulary growth.
 Together, these findings support the view that word recognition is a skill that develops gradually across early childhood and that this skill plays a role in supporting early language learning.
@@
 Early work using this method showed that both children's speed and accuracy increase rapidly across the second year [@fernald1998;@reznick1990visual].
-Related methods have provided a window into how children process phonological [@mani2011], lexical [@Swingley2002], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] information.
+Related methods have provided a window into how children process phonological [@mani2011], morphological [@meylan2025], lexical [@Swingley2002], syntactic [@trueswell1999kindergarten], and semantic [@mani2010;@bergelson2017] information.
 
@@
 
-Yet individual experiments measuring processing speed with young children typically recruit relatively small samples in a restricted range of ages.
+Yet given the logistical hurdles involved in sampling from this population, individual experiments measuring processing speed with young children typically recruit relatively small samples in a restricted range of ages.
 These samples provide neither the breadth of ages nor the number of participants needed to estimate how word recognition changes developmentally and how it connects with other aspects of early language development (see [@colby2023efficiency; @mcmurray2022slow] for examples of these analyses in school-aged children).
@@
 Initial exploratory factor analysis using parallel analysis to select the number of factors suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
-To better accommodate missing data under the assumption of data missing at random (e.g., due to dataset), we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
+To better accommodate missing data under the assumption of data missing at random (e.g., missingness due to the age sampling schemes of the various datasets), we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
 The best fitting model was a three-factor model with factors for speed (RT and RT variability), accuracy (proportion looking to target on each trial and associated variability of this measure), and vocabulary (comprehension and production from the CDI).
@@
 We investigated these questions using a new, large-scale dataset of developmental eye-tracking measurements compiled across many prior studies.
-The age-gradients for speed and accuracy indicated that both improve asymptotically.
+The age gradients for speed and accuracy indicated that both improve asymptotically.
 Gradients for recognition speed were consistent with the log-log relationship associated with the "power law of practice," that is, with a gradual convergence to mature levels of processing efficiency.

### 2025-04-04 paper/supplemental_information.Rmd
fix flipped axis labels

@@
   facet_wrap(~dataset_name, scale = "free_y") +
-  xlab("Number of participants") + 
-  ylab("Age (months)")
+  xlab("Age (months)") + 
+  ylab("Number of participants")
 ```

### 2025-04-04 0_get_data.qmd
pipeline cleanup

---
title: "0_get_data.qmd"
format: html
---

Note that this code is reproduced from the `peekbank-method` repository to remove an external dependency that could compromise the reproducibility of this particular packages. But it may go out of date from that repository, which could continue to evolve.

```{r}
library(here)
library(peekbankr)
source(here("helper","common.R"))
source(here("helper","rt_helper.R"))

# min number of trials
ACC_MIN_TRIAL_CUTOFF <- 4
RT_MIN_TRIAL_CUTOFF <- 2
```

## Get data

```{r, eval=TRUE}
con <- connect_to_peekbank(db_version = "2025.1")
all_aoi_timepoints <- get_aoi_timepoints(connection = con, rle=FALSE)

# reload connection in case it is stale
con <- connect_to_peekbank(db_version = "2025.1")
all_stimuli <- collect(get_stimuli(connection = con))
all_administrations <- collect(get_administrations(connection = con))
all_subjects <- unpack_aux_data(collect(get_subjects(connection = con)))
all_trial_types <- collect(get_trial_types(connection = con))
all_trials <- unpack_aux_data(collect(get_trials(connection = con)))
all_datasets <- get_datasets(connection=con) %>% collect()
```

Now do the joins.

```{r}
aoi_data_joined <- all_aoi_timepoints |>
  right_join(all_administrations) |>
  right_join(all_subjects) |>
  right_join(all_trials) |>
  right_join(all_trial_types) |>
  mutate(stimulus_id = target_id) |>
  left_join(all_stimuli) |>
  select(dataset_name, subject_id, administration_id, trial_id, trial_order, dataset_id, 
         stimulus_id, distractor_id, t_norm, age, aoi, english_stimulus_label, 
         stimulus_novelty, target_side, vanilla_trial) %>%
  rename(target_label = english_stimulus_label, 
         target_id = stimulus_id) %>%
  left_join(all_stimuli %>%
              select(stimulus_id, dataset_id, 
                     stimulus_novelty, english_stimulus_label) %>%
              rename(distractor_id = stimulus_id, 
                     distractor_novelty = stimulus_novelty,
                     distractor_label = english_stimulus_label))

save(aoi_data_joined, file= here("cached_intermediates","0_aoi_data_joined.Rds"))
```

Check on number of datasets.

```{r}
load(here("cached_intermediates", "1_aoi_data_joined.Rds"))
length(unique(aoi_data_joined$dataset_name))
unique(aoi_data_joined$dataset_name)
```

Age distribution.

```{r}
ggplot(aoi_data_joined, aes(x = age, fill = dataset_name)) + 
  geom_histogram() 
```

We now have a `vanilla_trial` flag that we can use for filtering.

Take only English familiar word data and also remove datasets that aren't ready for primetime.

```{r}
d_joined <- aoi_data_joined |>
  filter(!(dataset_name %in% c("casillas_tseltal_2015", # tseltsal
                               "hurtado_2008", # spanish
                               "kartushina_2019", # norwegian
                               "weisleder_stl", # spanish
                               "xsectional_2007", # spanish
                               "gazetriggered_2020", # dutch
                               "byers-heinlein_2017", # french bilinguals
                               "moore_bergelson_2022_verb" #verbs
  ))) |>
  filter(age <= 60, 
         vanilla_trial == TRUE,
         # stimulus_novelty == "familiar", # these conditions should be redundant with "vanilla"
         # distractor_novelty == "familiar" # these conditions should be redundant with "vanilla"
         ((stimulus_novelty == "familiar" & distractor_novelty == "familiar") |
            dataset_name == "fernald_totlot") # fix for error in dataset
  ) |> 
  select(dataset_name, subject_id, administration_id, trial_id, trial_order,
         dataset_id, target_id, t_norm, age, aoi, target_label, distractor_label, target_side) |>
  mutate(correct = ifelse(aoi == "target", 1, 
                          ifelse(aoi == "distractor", 0, NA)), 
         target_label = str_to_lower(target_label),
         distractor_label = str_to_lower(distractor_label))
```

# Item cleanup

Do some further checks/cleanup of items.

Output targets and distractors.

```{r}
all_items <- tibble(item = sort(unique(c(d_joined$target_label, d_joined$distractor_label))))
write_csv(all_items, here("metadata", "all_items_2025.1.csv"))
```

Rename and exclude in a more general way. This is a manual step by which we have gone through and merged items by hand when we thought they should have the same general form.

```{r}
included_items <- read_csv(here("metadata","included_items_2025.1.csv"))

d_aoi <- d_joined |>
  left_join(included_items, by = join_by(target_label == item)) |>
  left_join(included_items, by = join_by(distractor_label == item), 
            suffix = c("_target","_distractor")) |>
  filter(include_target == 1, include_distractor == 1) |>
  mutate(target_label = ifelse(is.na(rename_to_target), 
                               target_label, rename_to_target), 
         distractor_label = ifelse(is.na(rename_to_distractor), 
                                   distractor_label, rename_to_distractor)) |>
  select(-include_target, -include_distractor, -rename_to_distractor, -rename_to_target)
```

# Timecourse cleanup

So datasets vary in missingness of data at the trial level because of:

1. internal exclusion decisions (e.g., fmw_2013, which excluded trials <50%)
2. amount of data given (adams_marchman_2018, which didn't provide times before -500ms)
3. padding of trials with NAs (e.g., pomper_saffran_2016, which has some padding later in the trials).

```{r}
d_missing <- d_aoi |>
  group_by(dataset_name, t_norm) |>
  summarise(prop_data = mean(!is.na(correct)), 
            n = n())

ggplot(d_missing,
       aes(x = t_norm, y = prop_data)) + 
  facet_wrap(~dataset_name) + 
  ylab("Proportion trials that are not missing") + 
  geom_line() + 
  geom_vline(xintercept = -4000, lty = 3) + 
  geom_vline(xintercept = 4000, lty = 3)
```

Note - there is a bit of a ramp down in terms of amount of data for many datasets towards the end. Here we decide to keep this for now, but maybe we want to clean up starts and ends at some point per dataset?

```{r}
d_missing_clipped <- d_missing |>
  filter(t_norm >= -4000, t_norm <= 4000)

ggplot(d_missing_clipped,
       aes(x = t_norm, y = prop_data)) + 
  facet_wrap(~dataset_name) + 
  ylab("Proportion trials that are not missing") + 
  geom_line() + 
  geom_hline(yintercept = .5, lty = 2)
```

Previously we filtered by hand, but for several of our analyses, we want all the data.

```{r}
d_aoi <- d_aoi |>
  filter(t_norm >= -4000, t_norm <= 4000)

save(d_aoi, file = here("cached_intermediates","1_d_aoi.Rds"))
```

## CDI data

Add instrument length and a CDI percent column.

```{r}
admin_info <- all_administrations %>% 
  distinct(subject_id, dataset_id, administration_id, age) |>
  # mutate(floor_age = floor(age)) |>
  # select(-age) |>
  left_join(all_datasets %>% select(dataset_id, dataset_name)) # get dataset names

eng_cdi_admins <- all_subjects %>%
  unnest(subject_aux_data) %>% 
  filter(!is.na(cdi_responses)) %>%
  unnest(cdi_responses) %>% 
  filter(language == "English (American)") |>
  # mutate(floor_age = floor(age)) %>%
  mutate(instrument_length = case_when(instrument_type=="ws" ~ 680,
                                       instrument_type=="wg" ~ 396,
                                       instrument_type=="wsshort" ~ 100,
                                       instrument_type=="wgshort" ~ 100,
                                       .default = NA),
         CDI_percent = rawscore / instrument_length) |>
  select(subject_id, age, measure, CDI_percent) |> 
  pivot_wider(names_from = "measure", values_from = "CDI_percent", 
              values_fill = NA, values_fn = mean)
```

Now we have one row per CDI administration - English only at the moment.

Let's try to join this with the administration data.

The issue is that we have no 1-1 relationship. Consider subject 289. They have many CDIs and many LWLs and they don't all match.

Our principles:
* we want to join CDIs to LWLs, so that means that the LWL admins are our base data frame (e.g., `left_join(admins, cdis)` and not the reverse)
* we want only one CDI per LWL admin, but we could in principle average for extra precision

Here's an example. Note that this example depends on the use of age for the join and only works because `adams_marchman_2018` provides rounded ages.

```{r}
left_join(filter(admin_info, subject_id == 289), 
          filter(eng_cdi_admins, subject_id == 289))

left_join(filter(admin_info, subject_id == 2709), 
          filter(eng_cdi_admins, subject_id == 2709))

```

Let's try a fuzzy join.

```{r}
fuzzyjoin::difference_left_join(filter(admin_info, subject_id == 289), 
                                filter(eng_cdi_admins, subject_id == 289), 
                                by = c("subject_id", "age"),
                                max_dist = 1)

fuzzyjoin::difference_left_join(filter(admin_info, subject_id == 2709), 
                                filter(eng_cdi_admins, subject_id == 2709), 
                                by = c("subject_id", "age"),
                                max_dist = 1) 
```

This results in too many rows because there could be multiple CDIs matching a single admin. We can fix this by averaging.

```{r}
fuzzyjoin::difference_left_join(filter(admin_info, subject_id == 289), 
                                filter(eng_cdi_admins, subject_id == 289), 
                                by = c("subject_id", "age"),
                                max_dist = 1) |>
  select(-age.y, -subject_id.y) |>
  rename(age = age.x, subject_id = subject_id.x) |>
  group_by(subject_id, administration_id, dataset_name, age) |>
  summarise(prod = mean(prod, na.rm = TRUE), 
            comp = mean(comp, na.rm = TRUE)) |>
  mutate(across(c("prod","comp"), ~ifelse(is.nan(.x), NA, .x))) 
```

Here's the actual join. Problem, we need exact match on the subject ID and distance of 1 on the age.

Subject 2709 has a CDI age and subject-id less than 1 off and improperly matches 2710. we fix this by multiplying by 10 (hack!).

```{r}
foo <- fuzzyjoin::difference_left_join(admin_info, eng_cdi_admins, 
                                            by = c("subject_id", "age"),
                                            max_dist = 1)
filter(foo, subject_id.x == 2709)

bar <- fuzzyjoin::difference_left_join(mutate(admin_info, subject_id = 10*subject_id),
                                       mutate(eng_cdi_admins, subject_id = 10*subject_id), 
                                            by = c("subject_id", "age"),
                                            max_dist = 1)

filter(bar, subject_id.x == 27090)

```

```{r}
# cdi_data <- left_join(admin_info, eng_cdi_admins) 
cdi_data <- fuzzyjoin::difference_left_join(mutate(admin_info, subject_id = 10*subject_id),
                                            mutate(eng_cdi_admins, subject_id = 10*subject_id),
                                            by = c("subject_id", "age"),
                                            max_dist = 1) |>
  select(-age.y, -subject_id.y) |>
  rename(age = age.x, subject_id = subject_id.x) |>
  mutate(subject_id = subject_id/10) |> # remove hack 
  group_by(subject_id, administration_id, dataset_name, age) |>
  summarise(prod = mean(prod, na.rm = TRUE), 
            comp = mean(comp, na.rm = TRUE)) |>
  mutate(across(c("prod","comp"), ~ifelse(is.nan(.x), NA, .x))) |>
  ungroup()

save(cdi_data, file=here("cached_intermediates", "0_cdi_subjects.Rds"))
```

```{r}
cdi_data |>
  filter(!is.na(prod)) |>
  ggplot(aes(x=age, y=prod)) +
  geom_point(alpha=.2) + 
  geom_smooth() + 
  facet_wrap(~dataset_name)
```

```{r}
cdi_data |>
  filter(!is.na(comp)) |>
  ggplot(aes(x=age, y=comp)) +
  geom_point(alpha=.2) + 
  geom_smooth() + 
  facet_wrap(~dataset_name)
```

### 2025-04-04 1_tidy_data.qmd
pipeline cleanup

--- 2025-03-13 1_get_data.qmd
@@
 ---
-title: "Data getter for peekbank analysis"
+title: "Data tidying for peekbank analysis"
 format: html
@@
 ```{r}
-load(file = here("..", "peekbank-method", "cached_intermediates", "1_d_aoi.Rds"))
+readRDS(file = here("cached_intermediates", "0_d_aoi.Rds"))
 ```
@@
 ```{r}
-load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))
+load(here("cached_intermediates", "0_cdi_subjects.Rds"))
 

### 2025-04-04 0_get_data.qmd
recached intermediates

@@
 
-save(aoi_data_joined, file= here("cached_intermediates","0_aoi_data_joined.Rds"))
+saveRDS(aoi_data_joined, file= here("cached_intermediates","0_aoi_data_joined.Rds"))
 ```
@@
 ```{r}
-load(here("cached_intermediates", "1_aoi_data_joined.Rds"))
+aoi_data_joined <- readRDS(here("cached_intermediates", "1_aoi_data_joined.Rds"))
 length(unique(aoi_data_joined$dataset_name))
@@
 
-save(d_aoi, file = here("cached_intermediates","1_d_aoi.Rds"))
+saveRDS(d_aoi, file = here("cached_intermediates","0_d_aoi.Rds"))
 ```
@@
 
-save(cdi_data, file=here("cached_intermediates", "0_cdi_subjects.Rds"))
+saveRDS(cdi_data, file=here("cached_intermediates", "0_cdi_subjects.Rds"))
 ```

### 2025-04-04 paper/paper.Rmd
recached intermediates

@@
 #load d_aoi for timecourse plot
-d_aoi <- readRDS(here("cached_intermediates","1_d_aoi.Rds"))
+d_aoi <- readRDS(here("cached_intermediates","0_d_aoi.Rds"))
 d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>

### 2025-04-04 paper/supplemental_information.Rmd
supplement factor structures

@@
 
-- Alternative factor models: variances together, two factor, etc. etc.
+In this section, we provide comparisons between the three-factor model we report in the main text and several alternative models, including:
+
+* a one-factor model;
+* a two-factor model with vocabulary separated from speed and accuracy;
+* a two-factor model with speed separated from accuracy and vocabulary; and 
+* a two-factor model with variability terms separated from speed, accuracy, and vocabulary.
+
+\noindent Table \ref{tab:modcomp} shows the result of these comparisons. The three-factor model shows the lowest AIC and BIC, as well as being significantly better fittign than the next-best model.
+
+```{r, eval=FALSE}
+fa1_model <-  "
+F1  =~ log_rt + log_rt_var + long_window_accuracy + long_window_acc_var + prod + comp"
+fa1_fit <- cfa(fa1_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
+
+fa2_model <-  "
+processing =~ log_rt + log_rt_var + long_window_accuracy + long_window_acc_var
+vocabulary =~ prod + comp"
+fa2_fit <- cfa(fa2_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
+
+fa2b_model <-  "
+accuracy =~ long_window_accuracy + long_window_acc_var + prod + comp
+speed =~ log_rt + log_rt_var"
+fa2b_fit <- cfa(fa2b_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
+
+fa2v_model <-  "
+language =~ log_rt + long_window_accuracy + prod + comp
+               variability =~  long_window_acc_var + log_rt_var"
+fa2v_fit <- cfa(fa2v_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
+
+fa3_model <-  "
+speed =~ log_rt  + log_rt_var
+accuracy =~ long_window_accuracy + long_window_acc_var 
+vocab =~ prod + comp"
+fa3_fit <- cfa(fa3_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')
+
+xtable(anova(fa1_fit, fa2_fit, fa2b_fit, fa2v_fit, fa3_fit), digits = 2)
+```
+<!-- xtable 1.8-4 package -->
+<!-- % Fri Apr  4 15:15:52 2025 -->
+\begin{table}[ht]
+\centering
+\caption{Model comparison for alternative factor structures.
+$p$-values show differences between adjacent models; no $p$-values are shown for comparisons between non-nested models.
+\label{tab:modcomp}}
+\begin{tabular}{lrrrrrrrr}
+  \hline
+ & Df & AIC & BIC & Chisq & Chisq diff & RMSEA & Df diff & Pr($>$Chisq) \\ 
+  \hline
+Three-factor  & 6 & 46346.69 & 46475.43 & 91.96 &  &  &  &  \\ 
+  Two-factor (vocab)  & 8 & 46397.74 & 46514.22 & 147.01 & 55.05 & 0.09 & 2 & $<$ 0.0001 \\ 
+  Two-factor (speed) & 8 & 46486.96 & 46603.44 & 236.23 & 89.22 & 0.00 & 0 &  \\ 
+  Two-factor (variability) & 8 & 46536.10 & 46652.58 & 285.37 & 49.14 & 0 & 0.00 &  \\ 
+  One-factor & 9 & 46535.49 & 46645.84 & 286.76 & 1.39 & 0.01 & 1 & 0.24 \\ 
+   \hline
+\end{tabular}
+\end{table}
 
@@
 
-% latex table generated in R 4.4.2 by xtable 1.8-4 package
-% Thu Mar 13 13:10:30 2025
+<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
+<!-- % Thu Mar 13 13:10:30 2025 -->
 \begin{table}[ht]
@@
 Children with greater skill in word recognition show an overall positive shift in the growth trajectory of vocabulary development.
-
-```{r}
-# xtable(summary(mod_nlme)$tTable)
-```
 

### 2025-04-04 paper/supplemental_information.Rmd
supplement update

@@
 This model showed a significant effect of initial reaction time on the intercept of the logistic growth curve, but not on its scale (see Table \ref{tab:nlme}).
+```{r}
+# xtable(summary(mod_nlme)$tTable)
+```
 
@@
 Growth Intercept & 2.26 & 2.70 & 1766.00 & 0.84 & 0.40 \\ 
-Growth Intercept $\times$ $t_0$ RT & 3.30 & 0.38 & 1766.00 & 8.59 & 0.00 \\ 
+Growth Intercept $\times$ $t_0$ RT & 3.30 & 0.38 & 1766.00 & 8.59 & < 0.0001 \\ 
 Growth Scale & 2.05 & 2.60 & 1766.00 & 0.79 & 0.43 \\ 
@@
 
-# Latent Factor Growth
-
-```{r}
-fa3_model_long <- "
-# measurement model
-vocab_t1 =~ 1*prod_t1 + s1*comp_t1
-accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
-speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1
-
-vocab_t2 =~ 1*prod_t2 
-accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
-speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2
-
-vocab_t3 =~ 1*prod_t3 
-accuracy_t3 =~ 1*acc_t3 + s2*acc_sd_t3
-speed_t3 =~ 1*log_rt_t3 + s3*log_rt_sd_t3
-
-vocab_t4 =~ 1*prod_t4 
-accuracy_t4 =~ 1*acc_t4 + s2*acc_sd_t4
-speed_t4 =~ 1*log_rt_t4 + s3*log_rt_sd_t4
-
-# means for the latents
-# vocab_t1 ~ 1
-# vocab_t2 ~ 1
-# vocab_t3 ~ 1
-# vocab_t4 ~ 1
-# accuracy_t1 ~ 1
-# accuracy_t2 ~ 1
-# accuracy_t3 ~ 1
-# accuracy_t4 ~ 1
-# speed_t1 ~ 1
-# speed_t2 ~ 1
-# speed_t3 ~ 1
-# speed_t4 ~ 1
-
-# autoregression for the latents
-# vocab_t2 ~ 1 * vocab_t1
-# vocab_t3 ~ 1 * vocab_t2
-# vocab_t4 ~ 1 * vocab_t3
-# accuracy_t2 ~ 1 * accuracy_t1
-# accuracy_t3 ~ 1 * accuracy_t2
-# accuracy_t4 ~ 1 * accuracy_t3
-# speed_t2 ~ 1 * speed_t1
-# speed_t3 ~ 1 * speed_t2
-# speed_t4 ~ 1 * speed_t3
-
-# regressions
-accuracy_intercept =~ 1*accuracy_t1 + 1*accuracy_t2 + 1*accuracy_t3 + 1*accuracy_t4 
-accuracy_slope =~ 1*accuracy_t1 + 2*accuracy_t2 + 3*accuracy_t3 + 4*accuracy_t4
-speed_intercept =~ 1*speed_t1 + 1*speed_t2 + 1*speed_t3 + 1*speed_t4 
-speed_slope =~ 1*speed_t1 + 2*speed_t2 + 3*speed_t3 + 4*speed_t4
-vocab_intercept =~ 1*vocab_t1 + 1*vocab_t2 + 1*vocab_t3 + 1*vocab_t4 
-vocab_slope =~ 1*vocab_t1 + 2*vocab_t2 + 3*vocab_t3 + 4*vocab_t4
-
-# parameters for the regressions
-vocab_intercept ~ 1
-vocab_slope ~ 1
-accuracy_intercept ~ 1
-accuracy_slope ~ 1
-speed_intercept ~ 1
-speed_slope ~ 1
-
-# residual variance for the latents
-vocab_t1 ~~ NA*vocab_t1
-vocab_t2 ~~ NA*vocab_t2
-vocab_t3 ~~ NA*vocab_t3
-vocab_t4 ~~ NA*vocab_t4
-accuracy_t1 ~~ NA*accuracy_t1
-accuracy_t2 ~~ NA*accuracy_t2
-accuracy_t3 ~~ NA*accuracy_t3
-accuracy_t4 ~~ NA*accuracy_t4
-speed_t1 ~~ NA*speed_t1
-speed_t2 ~~ NA*speed_t2
-speed_t3 ~~ NA*speed_t3
-speed_t4 ~~ NA*speed_t4
-
-# residual variance for the other latents
-vocab_intercept ~~ NA*vocab_intercept
-vocab_slope ~~ NA*vocab_slope
-accuracy_intercept ~~ NA*accuracy_intercept
-accuracy_slope ~~ NA*accuracy_slope
-speed_intercept ~~ NA*speed_intercept
-speed_slope ~~ NA*speed_slope
-
-# set observed intercepts to zero - these should go into the latents
-prod_t1 ~ 0
-prod_t2 ~ 0
-prod_t3 ~ 0
-prod_t4 ~ 0
-comp_t1 ~ 0 
-log_rt_t1 ~ 0
-log_rt_t2 ~ 0
-log_rt_t3 ~ 0
-log_rt_t4 ~ 0
-log_rt_sd_t1 ~ 0 
-log_rt_sd_t2 ~ 0
-log_rt_sd_t3 ~ 0 
-log_rt_sd_t4 ~ 0 
-acc_t1 ~ 0 
-acc_t2 ~ 0 
-acc_t3 ~ 0 
-acc_t4 ~ 0 
-acc_sd_t1 ~ 0
-acc_sd_t2 ~ 0
-acc_sd_t3 ~ 0
-acc_sd_t4 ~ 0
-"
-
-fit3_long <- sem(fa3_model_long, d_sub_wide_fine, std.lv=TRUE, missing='fiml')
-
-summary(fit3_long, fit.measures=TRUE, standardize=TRUE)
-
-```
-
-```{r}
-layout <- read.csv(here("misc/layout_dt3.csv"), header = FALSE)
-graph_sem(model = fit3_long, text_size = 3, layout=t(layout))
-
-```
-
-<!-- # Short Window Analysis -->
-
-<!-- Need to redo all figures using short windows ugh.  -->
 # References

### 2025-04-04 paper/supplemental_information.Rmd
editing dataset description

@@
 
-Figure \ref{fig:dataset} gives the age distribution for each separate dataset.
+Figure \ref{fig:dataset} gives the age distribution of unique participants for each separate dataset at different ages. 
+Note that for some datasets, there are multiple administrations (i.e., experimental test sessions) for each participant.
 
 ```{r dataset, fig.cap="\\label{fig:dataset} Age distribution for each dataset, using three-month bins.", fig.pos="h!"}
-ggplot(d_sub, aes(x = age))+ 
-  geom_histogram(binwidth = 3) + 
+
+# summarize participants by age bin
+d_sub_admin_by_age <- d_sub |>
+  mutate(age_bin = age %/% 3 * 3+1.5) |>
+  group_by(dataset_name, age_bin) |>
+  summarise(
+    n_subj = n_distinct(subject_id),
+    n_administration = n_distinct(administration_id)) |>
+  ungroup()
+
+# ggplot(d_sub, aes(x = age))+ 
+#   geom_histogram(binwidth=3) + 
+#   facet_wrap(~dataset_name, scale = "free_y") +
+#   xlab("Age (months)") + 
+#   ylab("Number of administrations")+
+#   theme(strip.text.x = element_text(size = 5))
+
+# plot number of participants across age
+ggplot(d_sub_admin_by_age, aes(x = age_bin,y=n_subj))+ 
+  geom_bar(stat="identity",width=3) + 
   facet_wrap(~dataset_name, scale = "free_y") +
   xlab("Age (months)") + 
-  ylab("Number of participants")
+  ylab("Number of participants")+
+  theme(strip.text.x = element_text(size = 5))
+
 ```
 
-Figure \ref{fig:longitudinal} shows the distribution of measurement intervals for longitudinal studies within the dataset.
+Figure \ref{fig:longitudinal_descriptives} shows the distribution of measurement intervals for longitudinal studies within the dataset.
 
-```{r longitudinal, fig.cap="\\label{fig:longitudinal} Number of administrations for datasets with repeated measurements, plotted by dataset.", fig.pos="h!"}
+```{r longitudinal_descriptives, fig.cap="\\label{fig:longitudinal_descriptives} Number of administrations for datasets with repeated measurements, plotted by dataset.", fig.pos="h!"}
 

### 2025-04-14 paper/paper.Rmd
supplement update
prep for submission

@@
 These samples provide neither the breadth of ages nor the number of participants needed to estimate how word recognition changes developmentally and how it connects with other aspects of early language development (see [@colby2023efficiency; @mcmurray2022slow] for examples of these analyses in school-aged children).
-To overcome this limitation, we created Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank].
+To overcome these limitations, we created Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank].
 This dataset unifies and carefully curates a large amount of eye-tracking data from studies with infants and toddlers, representing cumulatively over 12 million individual samples of children's eye movements during real-time language processing.
-This resource allows us to gain an unprecedented view of the development of word recognition across a large sample of children.
+The Peekbank dataset allows us to gain an unprecedented view of the development of word recognition across a large sample of children.
 
@@
 We began by examining developmental changes in children's word recognition.
-Figure \ref{fig:timecourse} depicts the average timecourse at different ages across all datasets.
+Figure \ref{fig:timecourse} depicts the average timecourse at different ages across all datasets (not controlling for any variation in items and procedures across age groups).
 Intuitively, these timecourses show gradual increases in accuracy (higher overall proportion target looking) and speed (faster looking to the target after hearing a label) as age increases.
-To systematically characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
+To characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms).
@@
 Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables).
-We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant.
+We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study (modeling item and procedural) variation across studies) and random intercepts by participant.
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
@@
 This model revealed a significant effect of $t_0$ RT (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ RT and the quadratic age predictor (`r papaja::apa_print(growth_mod)$full_result$polyage_15_22_rt_t0`).
-This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time, confirming that findings of @fernald2012 were robust across the full set of longitudinal data in our dataset.
+This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time.
 
@@
 We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset.
-(We present a model of coupled growth in the observed variables here for simplicity; see SI: Latent Factor Growth for a comparable model using growth in the latent factors).
 The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}.
@@
 We then excluded participants from the analysis if they contributed fewer than four accuracy measurements or fewer than 2 reaction time measurements.
-At the participant level, these steps together led to `r round(mean(is.na(d_sub$rt)), 2)`% missingness for RTs and `r round(mean(is.na(d_sub$long_window_accuracy)), 2)`% missingness for long window accuracies.
+At the participant level, these steps together led to `r round(mean(is.na(d_sub$rt)), 3)*100`% missingness for RTs and `r round(mean(is.na(d_sub$long_window_accuracy)), 3)*100`% missingness for long window accuracies.
 

### 2025-04-14 paper/supplemental_information.Rmd
prep for submission

@@
 
-```{r}
+Figure 3 shows the result of a parallel analysis supporting the presence of three factors in the exploratory factor analysis.
+Table \ref{tab:loadings} shows the factor loadings for the exploratory three-factor solution using varimax rotation.
+The first factor is primarily driven by vocabulary measures, the second by reaction time, and the third by accuracy measures.
+
+```{r parallel, fig.cap="Parallel analysis scree plot showing the eigenvalues for each factor, for actual, simulated, and resampled data."}
 d_sub_mat <- d_sub |>
@@
                 ~ age_scale(.x, age))) 
-fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
-            use = "pairwise.complete.obs")
+outputs <- capture.output(fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
+            use = "pairwise.complete.obs"))
 ```
 
-```{r}
+```{r eval=FALSE}
 fs <- fa(select(d_sub_mat, -dataset_name), nfactor = 3,
@@
 as.data.frame(unclass(loadings(fs))) |> 
-  kable(caption = "Factor loadings for the exploratory three factor solution using varimax rotation.", digits = 2)
+  xtable(caption = "Factor loadings for the exploratory three factor solution using varimax rotation.", digits = 2)
 ```
+
+<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
+<!-- % Mon Apr 14 12:25:14 2025 -->
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrr}
+  \hline
+ & F1 & F2 & F3 \\ 
+  \hline
+RT & -0.19 & 0.81 & -0.30 \\ 
+  RT var & -0.10 & 0.81 & -0.22 \\ 
+  long window accuracy & 0.33 & -0.31 & 0.55 \\ 
+  long window accuracy var & -0.10 & 0.26 & -0.65 \\ 
+  production vocabulary & 0.95 & -0.04 & 0.30 \\ 
+  comprehension vocabulary & 0.61 & -0.16 & -0.01 \\ 
+  age & 0.63 & -0.14 & 0.37 \\ 
+   \hline
+\end{tabular}
+\caption{\label{tab:loadings}Factor loadings for the exploratory three factor solution using varimax rotation.} 
+\end{table}
 
@@
   Two-factor (speed) & 8 & 46486.96 & 46603.44 & 236.23 & 89.22 & 0.00 & 0 &  \\ 
-  Two-factor (variability) & 8 & 46536.10 & 46652.58 & 285.37 & 49.14 & 0 & 0.00 &  \\ 
+  Two-factor (variability) & 8 & 46536.10 & 46652.58 & 285.37 & 49.14 & 0 & 0 &  \\ 
   One-factor & 9 & 46535.49 & 46645.84 & 286.76 & 1.39 & 0.01 & 1 & 0.24 \\ 

### 2025-04-14 paper/paper.Rmd
prep for submission - of course a typo

@@
 Here, we use a large dataset of eye-tracking data from many different experiments to map out the development of word recognition, finding that this process becomes faster, more accurate, and less variable in the period between one and six years of age, and that children who are better at word recognition tend to have larger and faster-growing vocabularies.
-These data suggest that understanding language in real time is an important part of eearly language learning and that it improves gradually with practice across early childhood.
+These data suggest that understanding language in real time is an important part of early language learning and that it improves gradually with practice across early childhood.
 
@@
   \hline
- & dataset\_name & N subjects & N admins & Mean Age & Min Age & Max Age & Avg Trials & Avg RT Trials & CDIs & longitudinal \\ 
+ & Dataset Name & N subjects & N Admins & Mean Age & Min Age & Max Age & Avg Trials & Avg RT Trials & CDIs & longitudinal \\ 
   \hline

### 2025-08-26 paper/paper.Rmd
various revisons

@@
 Accuracy should change linearly with the logarithm of age, reflecting gradual asymptotic convergence to mature levels of accuracy.
-In addition, we might observe what is known as the "power law of practice," the regularity found in many cases of skill learning that the logarithm of reaction time decreases with the logarithm of experience [@snoddy1926;@kail1991;@anderson1982; cf.
-@heathcote2000].
+In addition, we might observe what is known as the "power law of practice," the regularity found in many cases of skill learning that the logarithm of reaction time decreases with the logarithm of experience across participants [@snoddy1926;@kail1991;@anderson1982; cf.
+@heathcote2000; @heathcote2018].
+Indeed, this pattern is predicted by an influential associative process model of early word learning [@mcmurray2012].
 In our case, we expect that chronological age is a proxy for experience and so the logarithm of reaction time should decrease linearly with the logarithm of age.
@@
 AGEVAR_CUTOFF <- 6
+
+datasets_with_age_variation <- d_trial |>
+  group_by(dataset_name) |>
+  summarise(age_range = max(age) - min(age)) |>
+  filter(age_range >= AGEVAR_CUTOFF)
+
+d_sub_agevar <- d_sub |>
+  filter(dataset_name %in% datasets_with_age_variation$dataset_name)
+
+longitudinal <- d_sub |>
+  group_by(dataset_name, subject_id) |>
+  count() |>
+  filter(n > 1)
+
+d_sub_firstadmin <- d_sub |>
+  group_by(dataset_name, subject_id) |>
+  filter(admin_num == 1) |>
+  ungroup() |>
+  select(-admin_num, -time_since_t0, -delta_t)
 
@@
                arrow = arrow(type = "closed", angle = 20, length = unit(0.05, "inches")), color="black")+
-  scale_color_viridis_d(name="Age Bin\n(mos)")+
-  scale_fill_viridis_d(name="Age Bin\n(mos)")+
+  scale_color_viridis_d(name="Age Bin\n(mos)", guide = guide_legend(reverse = TRUE))+
+  scale_fill_viridis_d(name="Age Bin\n(mos)", guide = guide_legend(reverse = TRUE))+
   #facet_wrap(~age_bin)+
@@
 To characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
+Reaction times were computed only on trials for which the child was fixating the distractor at the point of disambiguation (label onset), and were defined as the time from label onset to the first fixation on the target image (see SI: Reaction Time Computation).
+
 Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms).
@@
 Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables).
-We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study (modeling item and procedural) variation across studies) and random intercepts by participant.
+We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age nested within study (modeling item and procedural variation across studies) and random intercepts by participant.
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
-The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison).
+The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison and SI: Power Law Fits).
 Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analyses.
@@
 In sum, we see continuing improvements in word recognition across the full age range in our dataset that appear roughly linear in the logarithm of age.
-These logarithmic relationships follow theoretical expectations that both speed and accuracy should gradually asymptote to mature levels of performance, as seen in skill learning more generally [@anderson1982,@snoddy1926].
+These logarithmic relationships follow theoretical expectations that both speed and accuracy should gradually asymptote to mature levels of performance, as seen in skill learning more generally [@anderson1982;@snoddy1926].
 
@@
 The dashed line for accuracy shows chance-level looking (.5)", fig.env="figure*", fig.height = 2.5}
-
-datasets_with_age_variation <- d_trial |>
-  group_by(dataset_name) |>
-  summarise(age_range = max(age) - min(age)) |>
-  filter(age_range >= AGEVAR_CUTOFF)
-
-d_sub_agevar <- d_sub |>
-  filter(dataset_name %in% datasets_with_age_variation$dataset_name)
 
@@
 Both within and across datasets, within-individual variation in speed and accuracy decreased across the developmental range we examined (Figure \ref{fig:variance}).
-We fit mixed-effects models predicting the standard deviation of both speed and accuracy for each testing session for each participant, including random slopes of log age by dataset and random intercepts for each participant.
+We fit mixed-effects models predicting the standard deviation of both speed and accuracy for each testing session for each participant, including random slopes of log age nested within dataset and random intercepts for each participant.
 For both speed and accuracy, within-individual variability decreased with age (speed: `r papaja::apa_print(rt_var_mod)$full_result$log_age_s`; accuracy: `r papaja::apa_print(acc_var_mod)$full_result$log_age_s`).
@@
 We fit a series of factor analytic models to explore the dimensionality of the parent report and child LWL data.
+Our goal in these analyses was to understand the underlying relatedness of the various measures of word recognition and vocabulary, and in particular to assess the evidence for 1) whether the speed, accuracy, and variability measures described above all index the same underlying language processing construct and 2) the nature of the relation between this construct (or set of constructs) and early vocabulary.
+We begin developing models using all data, treating each observation as independent even if it comes from a longitudinal study; this assumption is equivalent to asserting an invariant factor structure across development (for a test of this assumption, see SI: Factor Analysis on First Administrations).
+In subsequent models, we relax this assumption and explore longitudinal growth.
+
 Initial exploratory factor analysis using parallel analysis to select the number of factors suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
@@
 
-fit3_age_summary <- suppressMessages(summary(fit3_age, fit.measures=TRUE, standardize=TRUE))
+fit3_age_summary <- suppressMessages(summary(fit3_age, fit.measures=TRUE, 
+                                             standardize=TRUE))
+
 layout_age = matrix(nrow=5, ncol = 6,
@@
 This model supports the idea that variation in speed and accuracy of word recognition is related to individual differences in parent-reported vocabulary beyond the effects of age.
+Further, the broader set of analyses support a factor structure in which speed and accuracy (and their associated variabilities) are related but distinct aspects of word recognition, rather than being measures of one single construct.
 These analyses treat all data as between person, however, rather than modeling change in these factors within individuals.
@@
 ```{r}
-longitudinal <- d_sub |>
-  group_by(dataset_name, subject_id) |>
-  count() |>
-  filter(n > 1)
-
 d_growth <- d_sub |>
@@
 We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in speed during the initial session of the study.
-We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset).
+We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age nested within participant and also age nested within dataset).
 This model revealed a significant effect of $t_0$ RT (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ RT and the quadratic age predictor (`r papaja::apa_print(growth_mod)$full_result$polyage_15_22_rt_t0`).
@@
 
-We confirmed this analysis using a non-linear growth model with a logistic shape, which provides a better fit to vocabulary size than the quadratic model (see SI: Non-Linear Growth Model) [@frank2021].
+We confirmed this analysis using a non-linear growth model with a logistic shape, which provides a better fit to vocabulary size within a fixed-length form than the quadratic model (see SI: Non-Linear Growth Model) [@frank2021].
 Figure \ref{fig:growth} shows predictions from this model, confirming the differentiation of growth curves for children with higher and lower initial reaction time.
@@
              fixed = xmid + scale ~ log_rt_0,
-             random = xmid + scale ~ 1,
+             random = xmid + scale ~ 1 | subject_id,
              groups = ~ dataset_name,
@@
   group_by(dataset_name) |>
-  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0 = c(quantile(d_sub_prod$log_rt, 
+  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0 = c(quantile(d_sub_prod$log_rt_0, 
                                                   c(.25,.75), 
@@
 We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months from the initial measurement (i.e., $t_0, ..., t_4$) and fit individual growth across each of these variables.
-We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset.
+We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset (se SI: SEM Longitudinal Missingness).
 The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}.
@@
 
+Reaction times
+
 We excluded trials entirely if they were missing data on more than 50% of timepoints, and excluded RTs if they were based on fewer than 50% of timepoints in the short analytic window (200 -- 2000 ms).
 We also removed RTs shorter than 367 ms, as these were unlikely to be generated based on the specific linguistic stimulus.
-We then excluded participants from the analysis if they contributed fewer than four accuracy measurements or fewer than 2 reaction time measurements.
+We then excluded participants from the analysis if they contributed fewer than four accuracy measurements or fewer than two reaction time measurements.
 At the participant level, these steps together led to `r round(mean(is.na(d_sub$rt)), 3)*100`% missingness for RTs and `r round(mean(is.na(d_sub$long_window_accuracy)), 3)*100`% missingness for long window accuracies.
@@
 
-We used `lme4` to fit linear mixed effects models, `nlme` to fit non-linear models, and `lavaan` to fit structural equation models. Full model specifications are available in the linked repository.
+We used `lme4` to fit linear mixed effects models, `nlme` to fit non-linear models, and `lavaan` to fit structural equation models.
+Random effects structures for each model are given in text; full model specifications are available in the Supplemental Information and in the reproducible code for this paper, available in the linked repository.
+To aid interpretability, all variables were standardized (z-scored) prior to inclusion in structural equation models.
 

### 2025-08-26 paper/supplemental_information.Rmd
various revisons

@@
   - \usepackage{booktabs}
+  - \renewcommand{\thefigure}{S\arabic{figure}}
+  - \renewcommand{\thetable}{S\arabic{table}}
+
 ---
@@
 
-```{r dataset, fig.cap="\\label{fig:dataset} Age distribution for each dataset, using three-month bins.", fig.pos="h!"}
+```{r dataset, fig.cap="\\label{fig:dataset} Age distribution of unique participants for each dataset, using three-month bins.", fig.pos="h!"}
 
@@
 
-```{r longitudinal_descriptives, fig.cap="\\label{fig:longitudinal_descriptives} Number of administrations for datasets with repeated measurements, plotted by dataset.", fig.pos="h!"}
+```{r longitudinal_descriptives, fig.cap="\\label{fig:longitudinal_descriptives} Distribution of retest administrations across datasets with repeated measurements, colored by dataset.
+Each count indicates a retest administration (initial administrations are excluded).
+Administrations listed with a retest interval of 0 indicate retests within a month of the initial administration.", fig.pos="h!"}
 
@@
   arrange(age) |>
-  mutate(admin_num = 1:n(), 
-         delta_t = age - age[1])
+  mutate(delta_t = age - age[1])
 
-ggplot(d_sub_long, 
+ggplot(filter(d_sub_long, admin_num > 1), 
        aes(x = delta_t, fill = dataset_name)) +
@@
   # facet_wrap(~dataset_name) + 
-  xlab("Age since first test") + 
-  ylab("Number of participants")
+  xlab("Months since first test administration") + 
+  ylab("Number of administrations")
+```
+
+# Reaction Time Computation
+
+Eye-tracking data are stored in Peekbank as a time series of fixations to specific areas of interest (in particular, the target and distractor on each trial).
+Other fixations can be to areas not in the target or distractor as well to off-screen areas.
+This time series has a uniform sample rate of 25ms/sample, based on resampling of the data in Peekbank to 40 Hz during preprocessing [@zettersten2023peekbank].
+Reaction times are computed by filtering trials to only those on which the child is fixating the distractor at the point of disambiguation ($t=0$) and then finding those trials on which the first non-missing fixation is to the target (hence excluding trials without a shift and trials on which a shift is to an off-screen location).
+The reaction time is then the total time from $t=0$ to the first timestep during which the child fixates the target.
+Consistent with standard practice in the literature following Fernald et al.
+(2008), RTs that are shorter than 367 ms are excluded as they are too short to be considered a response to the stimulus.
+
+# Separability of Variability from Other Measures
+
+## Reaction Time
+
+```{r}
+library(fitdistrplus)
+library(statmod)
+
+library(gamlss.dist)
+
+rt_fitting  <- d_sub |>
+  ungroup() |>
+  filter(!is.na(rt)) |>
+  mutate(age_bin = cut(age,
+                       breaks = c(0,12,15,18,21,24,27,30,36,60), 
+                       labels = c("<12", "12-15","15-18", "18-21", 
+                                  "21-24", "24-27","27-30","30-36", ">36"))) |>
+  select(age_bin, rt)
+
+norm <- fitdist(rt_fitting$rt, distr = "norm", method = "mle")
+lnorm <- fitdist(rt_fitting$rt, distr = "lnorm", method = "mle")
+wald <- fitdist(rt_fitting$rt, distr = "invgauss", method = "mle", 
+                start = list(mean = 1, shape = 1, dispersion = 1))
+exgaus <- fitdist(rt_fitting$rt, distr = "exGAUS", method = "mle", 
+                  start = list(mu = 1000, sigma = 500, nu = 1), 
+                  lower = c(0, 0, 0))
+
+BIC(norm)
+BIC(lnorm)
+BIC(wald)
+BIC(exgaus)
+
+rt_fits <- rt_fitting |>
+  group_by(age_bin) |>
+  nest(rt) |>
+  mutate(
+    norm = map(data, \(d) fitdist(d$rt, distr = "norm", method = "mle")),
+    lnorm = map(data, \(d) fitdist(d$rt, distr = "lnorm", method = "mle")),
+    wald = map(data, \(d) fitdist(d$rt, distr = "invgauss", method = "mle", 
+                                  start = list(mean = 1, shape = 1, dispersion = 1))),
+    exgaus = map(data, \(d) fitdist(d$rt, distr = "exGAUS", method = "mle", 
+                                    start = list(mu = 1000, sigma = 500, nu = 1), 
+                                    lower = c(0, 0, 0)))) |>
+  select(-data) |>
+  pivot_longer(norm:exgaus, names_to = "distribution", values_to = "fit") |>
+  mutate(BIC = map_dbl(fit, BIC)) |>
+  group_by(age_bin) |>
+  mutate(min_bic = BIC == min(BIC))
+
+ggplot(rt_fits, 
+       aes(x = distribution, y = BIC, col = min_bic)) + 
+  geom_point(stat = "identity", position = "dodge") + 
+  scale_color_solarized(name = "Lowest BIC") + 
+  xlab("Age bin") + 
+  ylab("BIC") + 
+  coord_flip() + 
+  facet_wrap(~age_bin, scales = "free_y") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+rt_hist_data <- d_sub |>
+  mutate(
+    age_bin = cut(age,
+                  breaks = c(0,12,15,18,21,24,27,30,36,60), 
+                  labels = c("<12", "12-15","15-18", "18-21", 
+                             "21-24", "24-27","27-30","30-36", ">36")))
+
+bw = .2
+
+rt_hist_summary <- rt_hist_data |>
+  group_by(age_bin) |>
+  summarise(m = mean(log_rt, na.rm = TRUE), 
+            sd = sd(log_rt, na.rm = TRUE), 
+            n = n()) |>
+  expand_grid(log_rt = seq(6,8,.01)) |>
+  mutate(norm = dnorm(log_rt, mean = m, sd = sd) * n * bw)
+
+ggplot(rt_hist_data,
+       aes(x = log_rt)) + 
+  geom_histogram(binwidth = bw) + 
+  geom_line(data = rt_hist_summary, aes(x = log_rt, y = norm), col = "blue") +
+  xlab("Log RT (ms)") + 
+  ylab("Number of trials") + 
+  facet_wrap(~age_bin)+ 
+  theme_few()
+```
+
+## Accuracy
+
+```{r}
+acc_hist_data <- d_sub |>
+  mutate(
+    age_bin = cut(age,
+                  breaks = c(0,12,15,18,21,24,27,30,36,60), 
+                  labels = c("<12", "12-15","15-18", "18-21", 
+                             "21-24", "24-27","27-30","30-36", ">36")))
+
+bw = .025
+
+acc_hist_summary <- acc_hist_data |>
+  group_by(age_bin) |>
+  summarise(m = mean(long_window_accuracy, na.rm = TRUE),
+            sd = sd(long_window_accuracy, na.rm = TRUE),
+            n = n()) |>
+  expand_grid(acc = seq(0,1,.01)) |>
+  mutate(norm = dnorm(acc, mean = m, sd = sd) * n * bw)
+
+ggplot(acc_hist_data,
+       aes(x = long_window_accuracy)) + 
+  geom_histogram(binwidth = bw) + 
+  geom_line(data = acc_hist_summary, aes(x = acc, y = norm), col = "blue") +
+  xlab("Long Window Accuracy") + 
+  ylab("Number of trials") + 
+  facet_wrap(~age_bin) + 
+  theme_few()
+```
+
+```{r}
+
+acc_fitting  <- d_sub |>
+  ungroup() |>
+  filter(!is.na(long_window_accuracy)) |>
+  mutate(age_bin = cut(age,
+                       breaks = c(0,12,15,18,21,24,27,30,36,60), 
+                       labels = c("<12", "12-15","15-18", "18-21", 
+                                  "21-24", "24-27","27-30","30-36", ">36"))) |>
+  select(age_bin, long_window_accuracy)
+
+norm <- fitdist(acc_fitting$long_window_accuracy, distr = "norm", method = "mle")
+beta <- fitdist(acc_fitting$long_window_accuracy, distr = "beta", method = "mle")
+
+BIC(norm)
+BIC(beta)
+```
+
+```{r}
+
+acc_fits <- acc_fitting |>
+  group_by(age_bin) |>
+  nest(long_window_accuracy) |>
+  mutate(
+    norm = map(data, \(d) fitdist(d$long_window_accuracy, distr = "norm", method = "mle")),
+    beta = map(data, \(d) fitdist(d$long_window_accuracy, distr = "beta", method = "mle"))) |>
+  select(-data) |>
+  pivot_longer(norm:beta, names_to = "distribution", values_to = "fit") |>
+  mutate(BIC = map_dbl(fit, BIC)) |>
+  group_by(age_bin) |>
+  mutate(min_bic = BIC == min(BIC))
+
+ggplot(acc_fits, 
+       aes(x = distribution, y = BIC, col = min_bic)) + 
+  geom_point(stat = "identity", position = "dodge") + 
+  scale_color_solarized(name = "Lowest BIC") + 
+  xlab("Age bin") + 
+  ylab("BIC") + 
+  coord_flip() + 
+  facet_wrap(~age_bin, scales = "free_y") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
 ```
@@
 We speculate that, on average, more participants looked away from the screen towards the end of trials, leading to a greater number of exclusions of long window trials based on the 50% criterion.
-Note that the total percentage of trials excluded is still small for both measures: `r round(mean(is.na(d_trial$long_window_accuracy)), 2)`% for long window accuracy and `r round(mean(is.na(d_trial$short_window_accuracy)), 2)`% for short window accuracy.
+Note that the total percentage of trials excluded is still small for both measures: `r round(mean(is.na(d_trial$long_window_accuracy)), 3)*100`% for long window accuracy and `r round(mean(is.na(d_trial$short_window_accuracy)), 3)*100`% for short window accuracy.
 
@@
 
-Note that with our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns (Heathcote, Brown, \& Mewhort, 2000).
-Our current results apply to the form of the age gradient as opposed to the form of any individual's pattern of developmental change.
-
 ```{r accmods, results="asis", include = FALSE}
@@
 ```
+
+# Power Law Fits
+
+In the literature on the "law of practice", although the log-log relationship we observed is commonly present in the aggregate across individuals, the situation is substantially more complex when relationships are measured within individuals.
+The best fitting curves for individuals are often exponentials or delayed exponentials  (Evans et al., 2018; Heathcote, Brown, \& Mewhort, 2000).
+With our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns unfortunately, because we have insufficient data about individuals' improvement across time.
+Thus, our current results apply to the form of the age gradient as opposed to the form of any individual's pattern of developmental change.
+
+Nevertheless, here we test for other forms of the aggregate relationship between age and reaction time. In particular, we consider further model comparison between the
+
+```{r rt_mods}
+rt_more_mods_lmer <- 
+  list(log_log = lmer(log_rt ~ log_age_s + 
+                        (log_age_s | dataset_name) + (1 | subject_id), 
+                      data = d_trial),
+       log_log_plus = lmer(log_rt ~ log_age_s + age_s +  
+                             (log_age_s + age_s | dataset_name) + (1 | subject_id), 
+                           data = d_trial), 
+       lin_poly2 = lmer(rt ~ poly(age_s, 2) +  
+                        (age_s | dataset_name) + (1 | subject_id), 
+                      data = d_trial),
+       lin_poly3 = lmer(rt ~ poly(age_s, 3) +  
+                        (age_s | dataset_name) + (1 | subject_id), 
+                      data = d_trial))
+
+rt_more_mods_lmer_summary <- map_df(rt_more_mods_lmer, ~broom.mixed::glance(.x)) |>
+  mutate(model = names(rt_more_mods_lmer), 
+         r2 = map_dbl(rt_more_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
+```
+
+# Mixed effects model specifications
+
+Here we provide specifications for the mixed effects models used in the main text.
+These models are used to estimate the relationship between age and the primary variables of interest, controlling for dataset and subject-level variability.
+
+\verbatim{lmer(long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)}
+
+FIXME
 
@@
 \end{table}
+
+# Factor Analysis: First Administrations
+
+```{r}
+d_sub_mat_firstadmin <- d_sub_firstadmin |>
+  ungroup() |>
+  select(dataset_name, log_rt, log_rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age, log_age)
+
+d_sub_mat_firstadmin_s <- d_sub_mat_firstadmin |>
+  ungroup() |>
+  mutate(across(all_of(c("log_rt", "log_rt_var", "long_window_accuracy", 
+                         "long_window_acc_var", "prod", "comp")), 
+                ~ age_scale(.x, age))) 
+```
+
+```{r}
+fit3_firstadmin <- cfa(fa3_model, d_sub_mat_firstadmin_s, 
+                       std.lv=TRUE, missing='fiml')
+
+fit_stats_firstadmin <- suppressMessages(summary(fit3_firstadmin, 
+                                                 fit.measures=TRUE, 
+                                                 standardize=TRUE))
+```
+
+```{r}
+d_sub_mat_firstadmin_s_renamed <- d_sub_mat_firstadmin_s |>
+  rename(acc = long_window_accuracy, 
+         acc_sd = long_window_acc_var, 
+         log_rt_sd = log_rt_var)
+
+fit3_age_firstadmin <- cfa(fa3_age_model, 
+                           d_sub_mat_firstadmin_s_renamed, 
+                           std.lv=TRUE, missing='fiml')
+
+fit3_age_summary_firstadmin <- 
+  suppressMessages(summary(fit3_age_firstadmin, 
+                           fit.measures=TRUE, 
+                           standardize=TRUE))
+```
+
+```{r}
+library(semoutput)
+semoutput::sem_fitmeasures(fit3)
+semoutput::sem_fitmeasures(fit3_firstadmin)
+semoutput::sem_fitmeasures(fit3_age)
+semoutput::sem_fitmeasures(fit3_age_firstadmin)
+
+```
 
@@
 
-# References
+FIXME
+
+```{r}
+d_sub_prod <- filter(d_sub, !is.na(prod), !is.na(log_rt)) |>
+  ungroup() |>
+  mutate(log_rt_resid = resid(lm(log_rt ~ log_age))) |>
+  group_by(subject_id) |>
+  mutate(log_rt_0 = log_rt[1],
+         log_rt_0_resid = log_rt_resid[1],
+         acc_0 = long_window_accuracy[1]) |> 
+  filter(!is.na(log_rt_0))
+
+mod_nlme <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
+                 data = d_sub_prod,
+                 fixed = xmid + scale ~ log_rt_0_resid,
+                 random = xmid + scale ~ 1 | subject_id,
+                 groups = ~ dataset_name,
+                 start = c(xmid = 20, scale = 3, `xmid:log_rt_0_resid` = 1, `scale:log_rt_0_resid` = 3),
+                 na.action = na.exclude, 
+                 control = list(maxIter = 1000, tolerance = .1))
+
+summary(mod_nlme)
+
+# fixed
+global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
+                                      max(d_sub_prod$age), .1), 
+                            log_rt_0_resid = c(quantile(d_sub_prod$log_rt_0_resid, 
+                                                        c(.25,.75), 
+                                                        na.rm=TRUE)))
+global_preds$pred <- predict(mod_nlme, newdata = global_preds, level = 0)
+
+#random effects
+dataset_preds <- d_sub |>
+  filter(!is.na(prod)) |> # filter on predictor
+  group_by(dataset_name) |>
+  summarise(min_age = min(age),
+            max_age = max(age)) |>
+  group_by(dataset_name) |>
+  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0_resid = c(quantile(d_sub_prod$log_rt_0_resid, 
+                                                                                                     c(.25,.75), 
+                                                                                                     na.rm=TRUE))))) |>
+  select(-min_age, -max_age) |>
+  unnest(col = "df") |>
+  ungroup()
+
+dataset_preds$pred <- predict(mod_nlme,
+                              newdata = dataset_preds,
+                              level = c(0,1))$predict.dataset_name
+
+ggplot(d_sub_prod, aes(x = age, y = prod, col = as.factor(log_rt_0_resid))) + 
+  geom_point(alpha = .03, pch = ".", col = "black") + 
+  geom_line(aes(group = subject_id), alpha = .02, col = "black") + 
+  geom_line(data = dataset_preds |>   
+              filter(dataset_name %in% datasets_with_age_variation$dataset_name), 
+            aes(y = pred, group = interaction(log_rt_0_resid, dataset_name)), lty="dashed", alpha=.5)+
+  geom_line(data = global_preds, aes(y = pred, group = log_rt_0_resid, col = as.factor(log_rt_0_resid)), size=1)+
+  scale_color_viridis(discrete=T, option = "H")+
+  theme(legend.position = "none")+
+  labs(y="Production", x="Age in months")
+
+```
+
+```{r}
+library(brms)
+
+d_sub_prod$age_c <- d_sub_prod$age - mean(d_sub_prod$age, na.rm = TRUE)
+d_sub_prod$log_rt_0_c <- d_sub_prod$log_rt_0 - mean(d_sub_prod$log_rt_0, na.rm = TRUE)
+
+# Define the nonlinear formula
+nlform <- brms::bf(
+  prod ~ 1 / (1 + exp((xmid - age_c) / exp(logscale))),
+  xmid ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
+  logscale ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
+  # scale ~ 1 + log_rt_0_c,
+  nl = TRUE
+)
+
+# xmid ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
+# scale ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
+
+priors <- c(
+  prior(normal(0, 5), nlpar = "xmid", coef = "Intercept"),  # xmid near center of age_c
+  prior(normal(1, 1), nlpar = "logscale", coef = "Intercept"), # scale > 0, mildly steep curve
+  prior(normal(0, 1), nlpar = "logscale", coef = "log_rt_0_c"),
+  prior(exponential(1), class = "sigma"),                # residual error
+  prior(normal(0, 2), nlpar = "xmid", coef = "log_rt_0_c"),
+  # prior(normal(0, 1), nlpar = "scale", lb = 0)
+
+  # Random effects for xmid
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name:subject_id"),
+
+  # Random effects for scale
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name:subject_id")
+)
+#
+
+# Fit the model
+mod_brms <- brm(
+  formula = nlform,
+  data = d_sub_prod,
+  prior = priors,
+  family = gaussian(),
+  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
+  chains = 4, cores = 4,
+  control = list(adapt_delta = 0.95),
+  # backend = "cmdstanr",  # optional but faster
+  save_pars = save_pars(all = TRUE),
+  sample_prior = "yes"
+)
+
+new_data <- expand.grid(
+  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
+  log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
+)
+
+fitted_vals <- fitted(mod_brms, newdata = new_data, re_formula = NA, summary = TRUE)
+
+new_data$fit <- fitted_vals[, "Estimate"]
+new_data$lower <- fitted_vals[, "Q2.5"]
+new_data$upper <- fitted_vals[, "Q97.5"]
+
+ggplot(new_data, aes(x = age_c, y = fit, col = as.factor(log_rt_0_c))) +
+  geom_line(linewidth = 2) +
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = as.factor(log_rt_0_c), group = as.factor(log_rt_0_c)), 
+              alpha = 0.2) +
+  labs(x = "Centered Age", y = "Predicted prod (fixed effects only)") +
+  theme_minimal() + 
+  geom_point(data = d_sub_prod, aes(x = age_c, y = prod), col = "black") + 
+  scale_color_solarized(name = "log RT at t0 (SD)") + 
+  scale_fill_solarized(guide = FALSE)
+
+```
+
+```{r}
+library(brms)
+
+d_sub_prod$age_c <- d_sub_prod$age - mean(d_sub_prod$age, na.rm = TRUE)
+d_sub_prod$log_rt_0_c <- d_sub_prod$log_rt_0 - mean(d_sub_prod$log_rt_0, na.rm = TRUE)
+
+d_sub_prod$resid_log_rt_0_c <- resid(lm(log_rt_0_c ~ log(age), data = d_sub_prod))
+
+# Define the nonlinear formula
+nlform_resid <- brms::bf(
+  prod ~ 1 / (1 + exp((xmid - age_c) / exp(logscale))),
+  xmid ~ 1 + resid_log_rt_0_c  + (1 | dataset_name/subject_id),
+  logscale ~ 1 + resid_log_rt_0_c  + (1 | dataset_name/subject_id),
+  # scale ~ 1 + log_rt_0_c,
+  nl = TRUE
+)
+
+# xmid ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
+# scale ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
+
+priors_resid <- c(
+  prior(normal(0, 5), nlpar = "xmid", coef = "Intercept"),  # xmid near center of age_c
+  prior(normal(1, 1), nlpar = "logscale", coef = "Intercept"), # scale > 0, mildly steep curve
+  prior(normal(0, 1), nlpar = "logscale", coef = "resid_log_rt_0_c"),
+  prior(exponential(1), class = "sigma"),                # residual error
+  prior(normal(0, 2), nlpar = "xmid", coef = "resid_log_rt_0_c"),
+  # prior(normal(0, 1), nlpar = "scale", lb = 0)
+
+  # Random effects for xmid
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name:subject_id"),
+
+  # Random effects for scale
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name:subject_id")
+)
+#
+
+# Fit the model
+resid_mod_brms <- brm(
+  formula = nlform_resid,
+  data = d_sub_prod,
+  prior = priors_resid,
+  family = gaussian(),
+  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
+  chains = 4, cores = 4,
+  control = list(adapt_delta = 0.95),
+  # backend = "cmdstanr",  # optional but faster
+  save_pars = save_pars(all = TRUE),
+  sample_prior = "yes"
+)
+
+new_data_resid <- expand.grid(
+  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
+  resid_log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
+)
+
+fitted_vals <- fitted(mod_brms, newdata = new_data, re_formula = NA, summary = TRUE)
+
+new_data_resid$fit <- fitted_vals[, "Estimate"]
+new_data_resid$lower <- fitted_vals[, "Q2.5"]
+new_data_resid$upper <- fitted_vals[, "Q97.5"]
+
+ggplot(new_data_resid, aes(x = age_c, y = fit, col = as.factor(resid_log_rt_0_c))) +
+  geom_line(linewidth = 2) +
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = as.factor(resid_log_rt_0_c), group = as.factor(resid_log_rt_0_c)), 
+              alpha = 0.2) +
+  labs(x = "Centered Age", y = "Predicted prod (fixed effects only)") +
+  theme_minimal() + 
+  scale_color_viridis() + 
+  geom_point(data = d_sub_prod, aes(x = age_c, y = prod), col = "black")
+
+```
+
+# SEM Longitudinal Missingness
+
+FIXME
+
+```{r}
+colMeans(!is.na(d_sub_wide_fine))
+```
+
+# Additional References
 

### 2025-08-28 paper/supplemental_information.Rmd
add creel re-analysis to supplement

@@
   - \renewcommand{\thetable}{S\arabic{table}}
-
 ---
@@
 (2008), RTs that are shorter than 367 ms are excluded as they are too short to be considered a response to the stimulus.
+
+# Comparison of Reaction Times for Correct and Incorrect Trials: Re-analysis of Creel (2024)
+
+```{r}
+#load data using Creel (2024) analysis scripts
+# To run this script, you need to first pull in two datasets from the Creel (2024) OSF repository (https://osf.io/wszd5/):
+#allfamiliars.csv and allnovels.csv
+library(osfr)
+creel_proj <- osf_retrieve_node("wszd5")
+#donwload familiar trials
+#TO DO: separate this out so files are downloaded only once
+creel_proj %>%
+  osf_ls_files(pattern = "allfamiliars.csv") %>% # Replace "data.csv" with your file's name
+  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")
+creel_proj %>%
+  osf_ls_files(pattern = "allnovels.csv") %>% # Replace "data.csv" with your file's name
+  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")
+
+#run Creel prep script
+source(here("supplemental_data","creel_2024","Step2aCorrsNDataPrep.R"))
+# run Peekbank-style rt computation over Creel data
+source(here("supplemental_data","creel_2024","creel_peekbank_acc_rt_analysis.R"))
+
+ creel_plot <- ggplot(filter(subj_rt_acc,n_correct_trials>1&n_trials>1),aes(mean_rt,mean_rt_correct))+
+  geom_point(aes(size=n_correct_trials),alpha=0.3,fill="black",stroke=NA)+
+  geom_smooth(method="lm")+
+  xlab("Mean Participant Reaction Time (ms)\nAll Trials")+
+  ylab("Mean Participant Reaction Time (ms)\nTrials w/ Correct Pointing Response")+
+  theme_cowplot(font_size=20)+
+  #add correlation and p-value as text box
+  annotate("text",x=500,y=1500,label=paste0("r = ",round(cor(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs"),2)),size=9)+
+  theme(legend.position=c(0.8,0.2))+
+  scale_size_continuous(name="# Trials",breaks=c(2,5,10,15,20))
+ggsave(here("supplemental_data","creel_2024","creel_rt_correlation_correcttrials_vs_alltrials.png"),width=9,height=6)
+
+creel_correlation <- cor.test(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs")
+
+```
+
+The Peekbank dataset only includes measurements of infants' looking behavior, with no measure of a final target selection.
+This contrasts with work in the visual-world paradigm with older children and adults, in which participants make a final explicit choice about which image matches the target label (e.g.
+Colby & McMurray, 2023).
+Having this additional response allows a clearer separation of accuracy and reaction times, because researchers can compute reaction times specifically on those trials in which participants responded correctly.
+This strategy helps avoid a possible mixing of reaction times for incorrect and correct responses, which might be generated by different underlying cognitive processes.
+A possible concern with the Peekbank datasets --- and reaction times in infant looking-while-listening studies more generally --- is that it is difficult to separate reaction times for correct vs.
+incorrect responses in the absence of an independent final choice response.
+
+To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024).
+This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image.
+Using this dataset, we investigated the correlation between reaction times (following the same procedure as in our main analyses, i.e.
+focusing specifically on distractor-to-target shifts) computed over all trials and reaction times computed only for those trials in which children selected the correct referent.
+The results are visualized below.
+Reaction times (i.e., distractor to target shifts) for correct trials only were highly correlated with reaction times across all trials (`r apa_print(creel_correlation)$full_result`).
+This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children.
+While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.
+
+```{r}
+creel_plot
+```
 
@@
 
+```{=tex}
 \begin{table}[tbp]
@@
 \end{table}
-
+```
 ```{r corrs, results = "asis", include = FALSE}
@@
 \end{table}
-
+```
+```{=tex}
 \begin{table}[tbp]
@@
 \end{table}
-
+```
 ```{r accmods, results="asis", include = FALSE}
@@
 In the literature on the "law of practice", although the log-log relationship we observed is commonly present in the aggregate across individuals, the situation is substantially more complex when relationships are measured within individuals.
-The best fitting curves for individuals are often exponentials or delayed exponentials  (Evans et al., 2018; Heathcote, Brown, \& Mewhort, 2000).
+The best fitting curves for individuals are often exponentials or delayed exponentials (Evans et al., 2018; Heathcote, Brown, & Mewhort, 2000).
 With our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns unfortunately, because we have insufficient data about individuals' improvement across time.
@@
 <!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
+
 <!-- % Mon Apr 14 12:25:14 2025 -->
+
+```{=tex}
 \begin{table}[ht]
@@
 \end{table}
-
+```
 # Factor Analysis: First Administrations
@@
 
-* a one-factor model;
-* a two-factor model with vocabulary separated from speed and accuracy;
-* a two-factor model with speed separated from accuracy and vocabulary; and 
-* a two-factor model with variability terms separated from speed, accuracy, and vocabulary.
+-   a one-factor model;
+-   a two-factor model with vocabulary separated from speed and accuracy;
+-   a two-factor model with speed separated from accuracy and vocabulary; and
+-   a two-factor model with variability terms separated from speed, accuracy, and vocabulary.
 
@@
 ```
+
 <!-- xtable 1.8-4 package -->
+
 <!-- % Fri Apr  4 15:15:52 2025 -->
+
+```{=tex}
 \begin{table}[ht]
@@
 \end{table}
-
+```
 # Non-linear Growth Models
@@
 This model showed a significant effect of initial reaction time on the intercept of the logistic growth curve, but not on its scale (see Table \ref{tab:nlme}).
+
 ```{r}
@@
 <!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
+
 <!-- % Thu Mar 13 13:10:30 2025 -->
+
+```{=tex}
 \begin{table}[ht]
@@
 \end{table}
-
+```
 Interpretation of growth in both this model and the linear growth model in the main text is complicated by the fact that the CDI form puts a ceiling on the total number of words that can be recorded; both the quadratic growth functions and the logistic functions come together at the form ceiling.
@@
 
-Heathcote, A., Brown, S., \& Mewhort, D. J. K. (2000). The power law repealed: The case for an exponential law of practice. Psychonomic Bulletin & Review, 7(2), 185–207.
+Heathcote, A., Brown, S., & Mewhort, D. J. K. (2000). The power law repealed: The case for an exponential law of practice. Psychonomic Bulletin & Review, 7(2), 185–207.
+
+Creel, S. (2024). Connecting the tots: Strong looking-pointing correlations in preschoolers' word learning and implications for continuity in language development. Child Development, 96, 87-103.

### 2025-09-17 paper/paper.Rmd
minor tweaks

@@
 
-output: rticles::pnas_article
+output: papaja::apa6_pdf
 
@@
   ) |>
-  #downsample t_norm to 50ms bins
+  #downsample t_norm to 50ms binsv
   #mutate(t_norm_downsampled = round(t_norm/50)*50) |>

### 2025-09-18 paper.R
push missing .R scripts

@@ -1,0 +1,606 @@
+suppressPackageStartupMessages(library(here))
+suppressPackageStartupMessages(library(psych))
+suppressPackageStartupMessages(library(lmerTest))
+suppressPackageStartupMessages(library(lavaan))
+suppressPackageStartupMessages(library(tidySEM))
+suppressPackageStartupMessages(library(xtable))
+suppressPackageStartupMessages(library(papaja))
+suppressPackageStartupMessages(library(viridis))
+suppressPackageStartupMessages(library(nlme))
+suppressPackageStartupMessages(library(cowplot))
+
+source(here("helper","common.R"))
+
+source(here("helper/modified_sem_plot.R"))
+
+#load d_aoi for timecourse plot
+d_aoi <- readRDS(here("cached_intermediates","0_d_aoi.Rds"))
+d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
+  mutate(log_age = log(age))
+d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds")) |>
+  group_by(subject_id) |>
+  arrange(age) |>
+  mutate(admin_num = 1:n(), 
+         time_since_t0 = age - age[1],
+         delta_t = c(0, diff(age)))
+
+d_trial$age_s <- scale(d_trial$age)[,1]
+d_trial$log_age_s <- scale(d_trial$log_age)[,1]   
+d_sub$age_s <- scale(d_sub$age)[,1]
+d_sub$log_age_s <- scale(d_sub$log_age)[,1]   
+
+AGEVAR_CUTOFF <- 6
+
+datasets_with_age_variation <- d_trial |>
+  group_by(dataset_name) |>
+  summarise(age_range = max(age) - min(age)) |>
+  filter(age_range >= AGEVAR_CUTOFF)
+
+d_sub_agevar <- d_sub |>
+  filter(dataset_name %in% datasets_with_age_variation$dataset_name)
+
+longitudinal <- d_sub |>
+  group_by(dataset_name, subject_id) |>
+  count() |>
+  filter(n > 1)
+
+d_sub_firstadmin <- d_sub |>
+  group_by(dataset_name, subject_id) |>
+  filter(admin_num == 1) |>
+  ungroup() |>
+  select(-admin_num, -time_since_t0, -delta_t)
+
+knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
+
+dataset_characteristics <- d_sub |>
+  group_by(dataset_name) |>
[…]

### 2025-09-18 supplemental_data/creel_2024/Step1README.R

@@ -1,0 +1,244 @@
+# Step1README.R
+# README file for eyemagic
+# AUTHOR
+# 8/15/23
+# Last updated 5/2/24
+
+############################################################
+############################################################
+############## Throughout, really important ################
+##############    text appears in # boxes.  ################
+############################################################
+############################################################
+
+### Not all important text may appear in boxes. ###
+
+### If something is commented out and far-indented, it's   ###
+### something not very useful that I was afraid to delete. ###
+
+
+### You need to run Step2a and Step2b scripts to run most ###
+### any of the remaining scripts. Step2a needs to be run  ###
+### with different values of whichCorrsToGenerate, which  ###
+### will give you different values of all resulting       ###
+### variables. Also writes some to file.                  ###
+
+
+### STUFF:
+
+### Data file: allnovels.csv
+  # NOVEL WORD, MELODY LEARNING, TALKER LEARNING trials but not familiar stuff
+  #
+  # Contains averaged look locations every 50 ms of every trial of every child.
+  # Experiment ID, experimental condition, age, and sex data are noted.
+  # Age outliers, dropped Ss from old studies, etc. are already eliminated.
+  # Children whose data are eliminated for too few onscreen looks still included,
+  #   since this can vary depending on your analysis presets.
+
+
+### Data file: allfamiliar.csv
+  # FAMILIAR WORD trials, all 2AFC, that were interspersed in KL studies and MN3.
+  #
+  # Contains averaged look locations every 50 ms of every trial of every child.
+  # Experiment ID, experimental condition, age, and sex data are noted.
+  # Age outliers, dropped Ss from old studies, etc. are already eliminated.
+  # Children whose data are eliminated for too few onscreen looks still included,
+  #   since this can vary depending on your analysis presets.
+
+
+
+### SCRIPTS:
+
+### Step2aCorrsNDataPrep.R *****
+  # Basically NO scripts will work if you haven't run this
+  # CALCULATES several different sets of looking-pointing correlations
+  # Preprocessing of gaze data to be used in following analyses
+  # 
+  # Select data to be processed from all learning studies; familiar words; 
[…]

### 2025-09-18 supplemental_data/creel_2024/Step2aCorrsNDataPrep.R

@@ -1,0 +1,530 @@
+# Step 2 looped version to extract point and look avgs and then correlate
+# Sarah C Creel
+# March 7, 2023
+# Last edited 4/23/24 to accommodate processing familiar word trials
+#
+# Previously in Step 1:
+# - read in buttload of data
+# - dropped bad data files as indicated in records
+#
+# Here in Step 2:
+# - extract accuracy values from ET data
+# - figure out bad ET trials and subjects to drop
+# - compute correlations between point acc and look proportion
+# 
+#  IMPORTANT: set the variable whichCorrsToGenerate to one of 3 different values (see below)
+#    default is set to 1 (overall Pearson correlations, and input data to linear mixed fx model)
+#
+# Default look measure:
+# Target look proportion minus Competitor look proportion in a window
+# Can also do 
+#   - proportion of total looks that are to Target (Target / (Target + Comp))
+#   - elogit (without weighting)
+
+library(dplyr)
+library(lme4)
+library(ggplot2) # underutilized right now
+library(readr)
+library(here)
+read_path <- here("supplemental_data","creel_2024")
+
+
+#setwd("~/Documents/current studies/A0 priority/EyeMagic/newRscripts")
+
+# included this in set of possible functions
+elogit <- function(proportion,trialN) {log((trialN*proportion + .5) / (trialN*(1-proportion) + .5))}  # log() defaults to ln
+# I'm not currently doing weighting...4/10/23
+
+
+###################################################
+# Make some selections about your analysis
+# I could not get readline() to run smoothly
+# Oh bother
+###################################################
+
+# FIRST: What dataset do you want to read in? n = all novel learning studies, f = familiar words only, w = familiar and novel words, a = all
+
+readthisone='w'
+
+#  NEXT: WHICH correlations (and cleanlookacc2 etc) do you want to generate?
+#  WARNING: 1=pretty quick, 2=less so, 3=if you have a lot of parameters to loop through can take hours. ##
+#  1: Overall correlations (Suppl Table 1) and input data, and main linear mixed effects model
+#  2: varied window lengths in Figure 4
+#  3: kid age windows in suppl figs
[…]

### 2025-09-18 supplemental_data/creel_2024/creel_peekbank_acc_rt_analysis.R

@@ -1,0 +1,80 @@
+# run after running Step2a in Creel code
+
+library(tidyverse)
+source(here("supplemental_data","creel_2024","rt_helper.R")) #loading local helper, this is just to adjust the sampling rate for rt computation to match creel data
+library(cowplot)
+library(ggpubr)
+library(here)
+#data <- read_csv(here("supplemental_data","creel_2024","all_data.csv"))
+##dat is all data
+
+target_window_min <- 200
+target_window_max <- 4000
+#compute RT
+binarize_dat <- dat %>%
+  mutate(prop = Targ/(Targ+Comp)) %>%
+  mutate(
+    target_look_binary = case_when(
+      prop>.5 ~ 1,
+      prop<=.5 ~ 0,
+    )
+  ) %>%
+  mutate(
+    aoi = case_when(
+      target_look_binary == 1 ~ "target",
+      target_look_binary == 0 ~ "distractor",
+      TRUE ~ NA
+    )
+  ) 
+
+# #reformat data
+rt_data <- binarize_dat %>%
+  filter(any(WOT == 50), # must have data at starting point
+         WOT >= 50) %>% # only pass data after 0
+  select(Subject,Age,AgeGroups,Tloc,Trialcount,Response,aoi,) %>%
+  group_by(Subject,Age,AgeGroups,Tloc,Trialcount,Response) %>%
+  summarise(lengths = rle(aoi)$lengths, 
+            values = rle(aoi)$values) #run-length-encoded format expected
+
+# compute RTs
+rts <- rt_data %>%
+  group_by(Subject,Age,AgeGroups,Tloc,Trialcount,Response) %>%
+  nest() %>%
+  mutate(data = lapply(data, get_rt)) %>% 
+  unnest(cols = c(data)) 
+
+#summarize
+summarized_trial_data <- binarize_dat %>%
+  filter(WOT>=target_window_min) %>%
+  filter(WOT<=target_window_max) %>%
+  group_by(Subject,Age,AgeGroups,Tloc,Trialcount) %>%
+  summarize(
+    pointing_acc = mean(Response=="target", na.rm=TRUE),
+    looking_acc_prop = mean(prop,na.rm=TRUE),
+    looking_acc_bin = mean(target_look_binary,na.rm=TRUE)
+  )
+
[…]

### 2025-09-18 supplemental_data/creel_2024/rt_helper.R

@@ -1,0 +1,63 @@
+## RT function tested in peekbank_rts.Rmd
+# takes rle_data dataframe (already rle'd)
+get_rt <- function (rle_data, SAMPLING_RATE = 20) {
+  
+  # end if no data
+  if (is.null(rle_data$values) | is.null(rle_data$lengths)) {
+    return(tibble(rt = NA, 
+                  shift_type = NA))
+  }
+  
+  onset_aoi <- rle_data$values[1] # zero point AOI
+  
+  # end if missing for start
+  if (!(onset_aoi %in% c("target","distractor"))) {
+    return(tibble(rt = NA, 
+                  shift_type = "other"))
+  }
+  
+  first_landing <- rle_data$values[rle_data$values != onset_aoi &
+                                     rle_data$values %in% c("target","distractor")][1]
+  
+  # end if no shift
+  if (is.na(first_landing)) {
+    return(tibble(rt = NA, 
+                  shift_type = "no shift"))
+  }
+  
+  shift_type <- case_when(onset_aoi == "distractor" &
+                            first_landing == "target" ~ "D-T",
+                          onset_aoi == "target" &
+                            first_landing == "distractor" ~ "T-D",
+                          TRUE ~ "other")
+  
+  first_landing_idx <- which(rle_data$values == first_landing)[1]
+  
+  values_before_first_landing <- rle_data$lengths[1:(first_landing_idx-1)]
+  value_before_first_shift <- rle_data$lengths[1]
+  
+  # if (first_landing_idx>2 & length(rle_data$lengths)>1) {
+  #   shift_values <- rle_data$lengths[2:(first_landing_idx-1)]
+  # } else {
+  #   shift_values <- c()
+  # }
+  
+  # rt is the number of samples happening before arrival + 1 
+  # (first sample of arrival)
+  # times the length of a sample
+  landing_time_rt <- (sum(values_before_first_landing) + 1) * (1000/SAMPLING_RATE)
+  
+  # rt is the number of samples happening before a shift is initiated + 1
+  # first sample off of onset AOI prior to moving to the next location
+  # times the length of a sample
+  shift_start_rt <- (value_before_first_shift + 1) * (1000/SAMPLING_RATE)
+  
+  # shift length: how long is the time between shift initiation and first landing
+  shift_length <- landing_time_rt - shift_start_rt
+    
+  
+  return(tibble(rt = landing_time_rt, # define rt as landing time RT for now to avoid breaking things downstream
[…]

### 2025-09-18 timecourse_plot.R

@@ -1,0 +1,120 @@
+library(cowplot)
+
+#summarize by subject
+d_timecourse_by_dataset_age_subj <- d_aoi |>
+  filter(t_norm>-1000) |> 
+  #create age bins cutoffs
+  # mutate(
+  #   age_bin = cut(age, breaks = c(0,12,15,18,21,24,30,36,60), 
+  #                 labels = c("<12", "12-15","15-18", "18-21", "21-24", "24-30","30-36", ">36"))
+  # ) |> 
+  mutate(
+    age_bin = cut(age, breaks = c(0,12,15,18,21,24,27,30,36,60), 
+                  labels = c("<12", "12-15","15-18", "18-21", "21-24", "24-27","27-30","30-36", ">36"))
+  ) |>
+  #downsample t_norm to 50ms bins
+  #mutate(t_norm_downsampled = round(t_norm/50)*50) |>
+  group_by(dataset_name,dataset_id,subject_id,age_bin,t_norm) |> 
+  #calculate proportion time spent looking at target
+  summarize(
+    n=n(),
+    accuracy = mean(correct, na.rm=TRUE))
+
+#by dataset and age bin
+d_timecourse_by_dataset_age_overall <- d_timecourse_by_dataset_age_subj |>
+  filter(n>=5) |> 
+  group_by(dataset_name,dataset_id,age_bin,t_norm) |>
+  summarize(
+    n_subj=n(),
+    proportion_target_looking = mean(accuracy,na.rm=TRUE)
+  ) |> 
+  ungroup() |> 
+  #create joined dataset_name and age_bin column
+  tidyr::unite(dataset_age_bin,dataset_name, age_bin,remove=F)
+#across dataset
+d_timecourse_across_dataset_by_age_overall <- d_timecourse_by_dataset_age_overall |>
+  #filter(n_subj>10) |> 
+  group_by(age_bin,t_norm) |>
+  summarize(
+    N = n(),
+    accuracy = mean(proportion_target_looking,na.rm=TRUE),
+    #standard error of the mean
+    sem = sd(proportion_target_looking,na.rm=TRUE)/sqrt(N)
+  ) 
+
+#by age bin, not representing dataset structure
+d_timecourse_by_age_overall <- d_timecourse_by_dataset_age_subj |>
+  filter(n>=5) |> 
+  group_by(age_bin,t_norm) |>
+  summarize(
[…]

### 2025-09-19 paper/paper.Rmd
attempts to knit paper and supp

@@
 
+shorttitle: Continuous developmental changes
 # Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
@@
 lineno: false
-
+floatsintext: yes
 output: papaja::apa6_pdf
@@
 
-\dropcap{C}hildren acquiring a language are learning a body of knowledge -- a set of words and the ways they are combined -- but they are also learning to deploy this knowledge in the myriad complex, noisy, and fast-moving environments in which language is used.
+Children acquiring a language are learning a body of knowledge -- a set of words and the ways they are combined -- but they are also learning to deploy this knowledge in the myriad complex, noisy, and fast-moving environments in which language is used.
 As children enter their second year, language explodes onto the scene; both vocabulary and grammatical abilities grow rapidly and in tandem [@bates1994developmental;@frank2021].
@@
 
-```{r age_regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age.
+```{r age-regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age.
 Observed variables are notated as squares and latent variables are notated as circles Factor loadings and regression coefficients are shown with straight, solid lines; covariances are shown with dashed lines; residual variances are shown as solid circular connections.
@@
 
-Figure \ref{fig:age_regression} shows a regression model fit to this confirmatory factor analysis, with log age predicting each latent variable.
+Figure \ref{fig:age-regression} shows a regression model fit to this confirmatory factor analysis, with log age predicting each latent variable.
 This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age).
@@
 <!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
-\showmatmethods
+<!--\showmatmethods
 \showacknow
@@
 
+\newpage -->
 \newpage
+
+# References
+
+::: {#refs custom-style="Bibliography"}
+:::

### 2025-09-19 paper/supplemental_information.Rmd
attempts to knit paper and supp

@@
 
-```{r}
+```{r, eval=F}
 #load data using Creel (2024) analysis scripts
@@
 library(osfr)
+library(here)
 creel_proj <- osf_retrieve_node("wszd5")
@@
 
+creel_proj %>%
+  osf_ls_files(pattern = "Step2aCorrsNDataPrep.R") %>% # Replace "data.csv" with your file's name
+  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")
+```
+
+```{r, eval=F}
 #run Creel prep script
-source(here("supplemental_data","creel_2024","Step2aCorrsNDataPrep.R"))
+
+#creel prep script expects files in root directory
+setwd(here("supplemental_data","creel_2024"))
+source("Step2aCorrsNDataPrep.R")
+setwd(here())
 # run Peekbank-style rt computation over Creel data
@@
 
-To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024).
+```{r}
+# <!--To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024).
 This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image.
@@
 This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children.
-While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.
+While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.-->
 
-```{r}
+```
+
+```{r, eval=F}
 creel_plot
@@
 library(fitdistrplus)
+
 library(statmod)
@@
 library(gamlss.dist)
+
+#note that fitdistrplus attaches MASS which also has a select function, and that's annoying
+detach("package:dplyr", character.only = TRUE)
+library("dplyr", character.only = TRUE)
 
@@
 
-```{r rt_mods}
+```{r rt_more_mods}
 rt_more_mods_lmer <- 
@@
 ```{r}
+#remotes::install_github("dr-JT/semoutput")
 library(semoutput)

### 2025-09-19 paper/supplemental_information.Rmd
work on implementing SI changes

@@
   - \renewcommand{\thetable}{S\arabic{table}}
+  - \usepackage{float}
+  - \renewcommand{\textfraction}{0.00}
+  - \renewcommand{\topfraction}{1}
+  - \renewcommand{\bottomfraction}{1}
+  - \renewcommand{\floatpagefraction}{1}
+  - \setcounter{topnumber}{5}
+  - \setcounter{bottomnumber}{5}
+  - \setcounter{totalnumber}{6}
+
 ---
@@
 source(here("helper","common.R"))
+library(papaja)
 ```
 
-```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
+```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, include=F}
 paper <- suppressMessages(knitr::purl(here("paper","paper.Rmd"), documentation = 0))
@@
 
-```{r, eval=F}
+```{r, include=F}
 #run Creel prep script
@@
 #creel prep script expects files in root directory
+detach("package:dplyr", character.only = TRUE)
+library("dplyr", character.only = TRUE)
 setwd(here("supplemental_data","creel_2024"))
@@
 source(here("supplemental_data","creel_2024","creel_peekbank_acc_rt_analysis.R"))
-
- creel_plot <- ggplot(filter(subj_rt_acc,n_correct_trials>1&n_trials>1),aes(mean_rt,mean_rt_correct))+
-  geom_point(aes(size=n_correct_trials),alpha=0.3,fill="black",stroke=NA)+
-  geom_smooth(method="lm")+
-  xlab("Mean Participant Reaction Time (ms)\nAll Trials")+
-  ylab("Mean Participant Reaction Time (ms)\nTrials w/ Correct Pointing Response")+
-  theme_cowplot(font_size=20)+
-  #add correlation and p-value as text box
-  annotate("text",x=500,y=1500,label=paste0("r = ",round(cor(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs"),2)),size=9)+
-  theme(legend.position=c(0.8,0.2))+
-  scale_size_continuous(name="# Trials",breaks=c(2,5,10,15,20))
-ggsave(here("supplemental_data","creel_2024","creel_rt_correlation_correcttrials_vs_alltrials.png"),width=9,height=6)
 
@@
 
-```{r}
-# <!--To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024).
+To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024).
 This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image.
@@
 focusing specifically on distractor-to-target shifts) computed over all trials and reaction times computed only for those trials in which children selected the correct referent.
-The results are visualized below.
+The results are shown in Figure \ref{fig:creelplot}.
 Reaction times (i.e., distractor to target shifts) for correct trials only were highly correlated with reaction times across all trials (`r apa_print(creel_correlation)$full_result`).
 This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children.
-While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.-->
+While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.
 
-```
+```{r creel-plot, fig.cap="\\label{fig:creelplot} Correlation between reaction times on all trials and reaction times on trials where the child pointed to the correct target.
+Data from Creel (2024).", fig.pos="h!", fig.width=8, fig.height=6}
 
-```{r, eval=F}
-creel_plot
+ggplot(filter(subj_rt_acc,n_correct_trials>1&n_trials>1),aes(mean_rt,mean_rt_correct))+
+  geom_point(aes(size=n_correct_trials),alpha=0.3,fill="black",stroke=NA)+
+  geom_smooth(method="lm")+
+  xlab("Mean Participant Reaction Time (ms)\nAll Trials")+
+  ylab("Mean Participant Reaction Time (ms)\nTrials w/ Correct Pointing Response")+
+  theme_cowplot(font_size=20)+
+  #add correlation and p-value as text box
+  annotate("text",x=500,y=1500,label=paste0("r = ",round(cor(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs"),2)),size=7)+
+  theme(legend.position=c(0.8,0.2))+
+  scale_size_continuous(name="# Trials",breaks=c(2,5,10,15,20))
 ```
@@
 
-## Reaction Time
+(TODO someone should edit this text)
+A potential concern with multiple measures deriving from the same underlying data (as with RT/accuracy in the looking-while-listening paradigm) is that distributional assumptions about each variable could implicitly assume some connections between RT/accuracy and variability that might not be warranted.
+
+To begin, we note that RT/accuracy and variability could be connected due to true age-related covariation - that is, some common cause leading to improvement in both, namely improvement in speed of response and improvement in consistency of response (e.g., skill learning).
+But they could be necessarily connected by virtue of their distributional form, and this is the concern we address here.
 
@@
 
-BIC(norm)
-BIC(lnorm)
-BIC(wald)
-BIC(exgaus)
+#BIC(norm)
+#BIC(lnorm)
+#BIC(wald)
+#BIC(exgaus)
 
@@
   mutate(min_bic = BIC == min(BIC))
+```
 
+## Reaction Time
+
+We’ll look at reaction time first.
+Under certain distributional forms for reaction time, there would be a necessary relationship between RT and its variation.
+However, the literature focuses on the use of the Exponential-Gaussian (ex-Gaussian) distribution as well as Wald, Weibull, gamma, and log normal distributions (see for example Luce, 1986; Ratcliff, 1993; Van Zandt, 2002).
+All of these are 2- or 3-parameter distributions meaning that there is no necessary relationship between mean and variance.
+
+The problem of fitting RT distributions is complex and a substantial literature exists (e.g., Ratcliff, 1979; Luce, 1986; Van Zandt, 2000; Baayen & Milin, 2010).
+One of the big challenges in our dataset as well as elsewhere is that distributions are conditional on factors such as participant and task, so it is challenging to draw inferences about the underlying distribution when looking at average data.
+
+That said, we have attempted some fitting and we find that overall the data are best fit by either an ex-Gaussian or a log normal distribution, again consistent with prior literature, giving us confidence in this conclusion.
+Across the full dataset, the BIC values for ex-Gaussian (`r BIC(exgaus)`) and log normal (`r BIC(lnorm)`) are quite close to one another, and better than the Wald (`r BIC(wald)`) and normal (`r BIC(norm)`) fits.
+When binned by age (Fig \ref{fig:rt-fits}), younger children seem better fit for a log normal distribution and older children seem better fit by ex-Gaussian (models with lowest BICs are shown in red since significant differences can be obscured by the large scale).
+Figure \ref{fig:rt-hist} shows the RT data distribution overlaid with the corresponding log-normal distributions.
+
+Overall, we think this result generally vitiates our decision to use log-transformed RTs as our primary dependent measure, and further suggests that reaction time means and variabilities are, in principle, theoretically independent.
+
+```{r rt-fits, fig.cap="\\label{fig:rt-fits} Goodness of fit for different distributional models for RT, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
 ggplot(rt_fits, 
@@
 
-```{r}
+```{r rt-hist, fig.cap="\\label{fig:rt-hist} Distribution of RT overlaid with log normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
+
 rt_hist_data <- d_sub |>
@@
 ```{r}
+
+acc_fitting  <- d_sub |>
+  ungroup() |>
+  filter(!is.na(long_window_accuracy)) |>
+  mutate(age_bin = cut(age,
+                       breaks = c(0,12,15,18,21,24,27,30,36,60), 
+                       labels = c("<12", "12-15","15-18", "18-21", 
+                                  "21-24", "24-27","27-30","30-36", ">36"))) |>
+  select(age_bin, long_window_accuracy)
+
+norm_acc <- fitdist(acc_fitting$long_window_accuracy, distr = "norm", method = "mle")
+beta_acc <- fitdist(acc_fitting$long_window_accuracy, distr = "beta", method = "mle")
+
+#BIC(norm)
+#BIC(beta)
+```
+
+Individual trial-level accuracies are not binomial because they are an average probability of fixation over a viewing window, as described in the manuscript.
+They are bounded at 0 and 1, but in general they tend towards the range .5 - .8 in most studies of this population.
+Figure \ref{acc-hist} shows the data binned by age group with fitted gaussian distributions.
+
+These distributions seem well fit by standard gaussians, but they are in principle bounded and so we asked whether this made a difference, using a Beta distribution (a two-parameter continuous distribution bounded at 0 and 1) for fitting.
+Surprisingly, across all data, the BIC values for the two distributions were very similar (`r BIC(norm_acc)` for normal versus `r BIC(beta_acc)` for Beta), though the Beta was slightly favored.
+Across age groups, there was heterogeneity with some groups better fit by a gaussian and others better fit by a Beta (Figure \ref{fig:acc-fits}).
+
+Again, we feel that this result generally vitiates our approach of modeling accuracies via standard linear mixed effects models: their distributional form is quite close to normal.
+Thus, again there is no necessary connection between mean and variance: they could in principle be independent.
+As above, we will plan on including this information in the Supplement.
+
+```{r acc-hist, fig.cap="\\label{fig:acc-hist} Goodness of fit for different distributional models of accuracy, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
 acc_hist_data <- d_sub |>
@@
 
-```{r}
-
-acc_fitting  <- d_sub |>
-  ungroup() |>
-  filter(!is.na(long_window_accuracy)) |>
-  mutate(age_bin = cut(age,
-                       breaks = c(0,12,15,18,21,24,27,30,36,60), 
-                       labels = c("<12", "12-15","15-18", "18-21", 
-                                  "21-24", "24-27","27-30","30-36", ">36"))) |>
-  select(age_bin, long_window_accuracy)
-
-norm <- fitdist(acc_fitting$long_window_accuracy, distr = "norm", method = "mle")
-beta <- fitdist(acc_fitting$long_window_accuracy, distr = "beta", method = "mle")
-
-BIC(norm)
-BIC(beta)
-```
-
-```{r}
-
+```{r acc-fits, fig.cap="\\label{fig:acc-fits} Distribution of accuracies overlaid with normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
 acc_fits <- acc_fitting |>
@@
 
-\verbatim{lmer(long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)}
+<!--\verbatim{lmer(long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)}-->
 
-FIXME
+FIXME because I cause things to crash!
 
@@
 library(semoutput)
-semoutput::sem_fitmeasures(fit3)
-semoutput::sem_fitmeasures(fit3_firstadmin)
-semoutput::sem_fitmeasures(fit3_age)
-semoutput::sem_fitmeasures(fit3_age_firstadmin)
+semoutput::sem_fitmeasures(fit3, print=F) |> kable()
+semoutput::sem_fitmeasures(fit3_firstadmin, print=F) |> kable()
+semoutput::sem_fitmeasures(fit3_age, print=F) |> kable()
+semoutput::sem_fitmeasures(fit3_age_firstadmin, print=F) |> kable()
 
@@
 
-summary(mod_nlme)
+#summary(mod_nlme)
 
@@
                                                                                                      na.rm=TRUE))))) |>
-  select(-min_age, -max_age) |>
+  dplyr::select(-min_age, -max_age) |>
   unnest(col = "df") |>
@@
 
-```{r}
+```{r, eval=F}
 library(brms)
@@
   save_pars = save_pars(all = TRUE),
-  sample_prior = "yes"
+  sample_prior = "yes",
+  file="brms1.rds"
 )
@@
 fitted_vals <- fitted(mod_brms, newdata = new_data, re_formula = NA, summary = TRUE)
+
+saveRDS(fitted_vals, "brms1_fitted.rds")
+```
+
+```{r}
+fitted_vals <- readRDS("brms1_fitted.rds")
+
+d_sub_prod$age_c <- d_sub_prod$age - mean(d_sub_prod$age, na.rm = TRUE)
+d_sub_prod$log_rt_0_c <- d_sub_prod$log_rt_0 - mean(d_sub_prod$log_rt_0, na.rm = TRUE)
+
+new_data <- expand.grid(
+  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
+  log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
+)
 
@@
   save_pars = save_pars(all = TRUE),
-  sample_prior = "yes"
+  sample_prior = "yes",
+  file="brms2.rds"
 )
@@
 fitted_vals <- fitted(mod_brms, newdata = new_data, re_formula = NA, summary = TRUE)
+
+saveRDS(fitted_vals, "brms2_fitted.rds")
+```
+
+```{r}
+
+d_sub_prod$age_c <- d_sub_prod$age - mean(d_sub_prod$age, na.rm = TRUE)
+d_sub_prod$log_rt_0_c <- d_sub_prod$log_rt_0 - mean(d_sub_prod$log_rt_0, na.rm = TRUE)
+
+d_sub_prod$resid_log_rt_0_c <- resid(lm(log_rt_0_c ~ log(age), data = d_sub_prod))
+
+fitted_vals <- readRDS("brms2_fitted.rds")
+
+new_data_resid <- expand.grid(
+  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
+  resid_log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
+)
 
@@
   theme_minimal() + 
-  scale_color_viridis() + 
+  scale_color_viridis(discrete=T) + 
+  scale_fill_viridis(discrete=T, guide=F)+
   geom_point(data = d_sub_prod, aes(x = age_c, y = prod), col = "black")
@@
 
-```{r}
+```{r, eval=F}
 colMeans(!is.na(d_sub_wide_fine))

### 2025-09-24 paper/paper.Rmd
assorted table futzing, refs

@@
 output: papaja::apa6_pdf
+header-includes:
+  - \usepackage{threeparttablex}
+  - \usepackage{booktabs}
+  - \usepackage{placeins}
+  - \usepackage{float}
+  - \renewcommand{\textfraction}{0.00}
+  - \renewcommand{\topfraction}{1}
+  - \renewcommand{\bottomfraction}{1}
+  - \renewcommand{\floatpagefraction}{1}
+  - \setcounter{topnumber}{5}
+  - \setcounter{bottomnumber}{5}
+  - \setcounter{totalnumber}{6}
 
@@
 
-\begin{table*}[ht]
+\begin{table}[ht]
 \centering
+\setlength\tabcolsep{2pt}
+\begin{scriptsize}
 \begin{tabular}{rlrrrrrrrll}
@@
 \end{tabular}
+\end{scriptsize}
 \caption{Characteristics of included datasets from Peekbank.
@@
 \label{tab:datasets}}
-\end{table*}
+\end{table}
 
@@
 In addition, we might observe what is known as the "power law of practice," the regularity found in many cases of skill learning that the logarithm of reaction time decreases with the logarithm of experience across participants [@snoddy1926;@kail1991;@anderson1982; cf.
-@heathcote2000; @heathcote2018].
+@heathcote2000; @heathcote2018, @evans2018].
 Indeed, this pattern is predicted by an influential associative process model of early word learning [@mcmurray2012].
@@
 suppressPackageStartupMessages(library(viridis))
-suppressPackageStartupMessages(library(nlme))
+#suppressPackageStartupMessages(library(nlme))
 suppressPackageStartupMessages(library(cowplot))
@@
 Thin colored lines show linear fits for those datasets spanning six or more months of age.
-The dashed line for accuracy shows chance-level looking (.5)", fig.env="figure*", fig.height = 2.5}
+The dashed line for accuracy shows chance-level looking (.5)", fig.env="figure", fig.height = 2.5}
 
@@
 ```{r variance, fig.cap="Participant-level variability in target looking and reaction time (log RT), plotted by age (log).
-Plotting conventions are as in Figure 1.", fig.env="figure*", fig.height = 2.5}
+Plotting conventions are as in Figure 1.", fig.env="figure", fig.height = 2.5}
   mycolors = c("#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E", "#1B9E77", "#D95F02", "#E7298A", "#66A61E","#1B9E77", "#D95F02")
@@
 
-```{r age-regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age.
+```{r age-regression, fig.env="figure", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age.
 Observed variables are notated as squares and latent variables are notated as circles Factor loadings and regression coefficients are shown with straight, solid lines; covariances are shown with dashed lines; residual variances are shown as solid circular connections.
@@
 
-```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for the 25th (red) and 75th (blue) percentile on initial reaction times.
-Individual longitudinal trajectories are shown in light gray.
-Solid lines show global model estimates and dashed lines show estimates for individual datasets with sufficient age variation as above.", fig.env="figure", fig.height = 2.5, fig.width=4, out.width=".5\\textwidth", include=TRUE}
-
-d_sub_prod <- filter(d_sub, !is.na(prod)) |>
+```{r}
+d_sub_prod <- filter(d_sub, !is.na(prod), !is.na(log_rt)) |>
+  ungroup() |>
+  mutate(log_rt_resid = resid(lm(log_rt ~ log_age))) |>
   group_by(subject_id) |>
   mutate(log_rt_0 = log_rt[1],
+         log_rt_0_resid = log_rt_resid[1],
          acc_0 = long_window_accuracy[1]) |> 
@@
 
-mod_nlme <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
+d_sub_prod$age_c <- d_sub_prod$age - mean(d_sub_prod$age, na.rm = TRUE)
+d_sub_prod$log_rt_0_c <- d_sub_prod$log_rt_0 - mean(d_sub_prod$log_rt_0, na.rm = TRUE)
+```
+
+```{r, eval=F}
+library(brms)
+
+# Define the nonlinear formula
+nlform <- brms::bf(
+  prod ~ 1 / (1 + exp((xmid - age_c) / exp(logscale))),
+  xmid ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
+  logscale ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
+  # scale ~ 1 + log_rt_0_c,
+  nl = TRUE
+)
+
+# xmid ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
+# scale ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
+
+priors <- c(
+  prior(normal(0, 5), nlpar = "xmid", coef = "Intercept"),  # xmid near center of age_c
+  prior(normal(1, 1), nlpar = "logscale", coef = "Intercept"), # scale > 0, mildly steep curve
+  prior(normal(0, 1), nlpar = "logscale", coef = "log_rt_0_c"),
+  prior(exponential(1), class = "sigma"),                # residual error
+  prior(normal(0, 2), nlpar = "xmid", coef = "log_rt_0_c"),
+  # prior(normal(0, 1), nlpar = "scale", lb = 0)
+
+  # Random effects for xmid
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name:subject_id"),
+
+  # Random effects for scale
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name:subject_id")
+)
+#
+
+# Fit the model
+mod_brms <- brm(
+  formula = nlform,
   data = d_sub_prod,
-             fixed = xmid + scale ~ log_rt_0,
-             random = xmid + scale ~ 1 | subject_id,
-             groups = ~ dataset_name,
-             start = c(xmid = 20, scale = 3, `xmid:log_rt_0` = 1, `scale:log_rt_0` = 3),
-             na.action = na.exclude, 
-             control = list(maxIter = 1000, tolerance = 10))
-
-# fixed effects (no CIs)
-global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
-                                      max(d_sub_prod$age), .1), 
-                            log_rt_0 = c(quantile(d_sub_prod$log_rt, 
-                                                  c(.25,.75), 
-                                                  na.rm=TRUE)))
-global_preds$pred <- predict(mod_nlme, newdata = global_preds, level = 0)
-
-#random effects
-dataset_preds <- d_sub |>
-  filter(!is.na(prod)) |> # filter on predictor
-  group_by(dataset_name) |>
-  summarise(min_age = min(age),
-            max_age = max(age)) |>
-  group_by(dataset_name) |>
-  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0 = c(quantile(d_sub_prod$log_rt_0, 
-                                                  c(.25,.75), 
-                                                  na.rm=TRUE))))) |>
-  select(-min_age, -max_age) |>
-  unnest(col = "df") |>
-  ungroup()
-
-dataset_preds$pred <- predict(mod_nlme,
-                              newdata = dataset_preds,
-                              level = c(0,1))$predict.dataset_name
-
-ggplot(d_sub_prod, aes(x = age, y = prod, col = as.factor(log_rt_0))) + 
-  geom_point(alpha = .03, pch = ".", col = "black") + 
-  geom_line(aes(group = subject_id), alpha = .02, col = "black") + 
-  geom_line(data = dataset_preds |>   
-              filter(dataset_name %in% datasets_with_age_variation$dataset_name), 
-            aes(y = pred, group = interaction(log_rt_0, dataset_name)), lty="dashed", alpha=.5)+
-  geom_line(data = global_preds, aes(y = pred, group = log_rt_0, col = as.factor(log_rt_0)), size=1)+
-  scale_color_viridis(discrete=T, option = "H")+
-  theme(legend.position = "none")+
-  labs(y="Production", x="Age in months")
-# +
-#   geom_line(data = dataset_preds, aes(y = pred))
+  prior = priors,
+  family = gaussian(),
+  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
+  chains = 4, cores = 4,
+  control = list(adapt_delta = 0.95),
+  # backend = "cmdstanr",  # optional but faster
+  save_pars = save_pars(all = TRUE),
+  sample_prior = "yes",
+  file="brms1.rds"
+)
+
+summary(mod_brms)$fixed |> saveRDS("brms1_coef.rds")
+
+new_data <- expand.grid(
+  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
+  log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
+)
+
+fitted_vals <- fitted(mod_brms, newdata = new_data, re_formula = NA, summary = TRUE)
+
+saveRDS(fitted_vals, "brms1_fitted.rds")
+
+```
+
+```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for children with initial reaction times one SD faster than the mean (blue), at the mean (red), and one SD slower than the mean (green).
+Individual longitudinal trajectories are shown in light gray.
+Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, fig.width=6, out.width=".6\\textwidth", include=TRUE}
+
+fitted_vals <- readRDS("brms1_fitted.rds")
+
+new_data <- expand.grid(
+  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
+  log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
+)
+
+new_data$fit <- fitted_vals[, "Estimate"]
+new_data$lower <- fitted_vals[, "Q2.5"]
+new_data$upper <- fitted_vals[, "Q97.5"]
+
+ggplot(new_data, aes(x = age_c, y = fit, col = as.factor(log_rt_0_c))) +
+    geom_line(data = d_sub_prod, aes(x = age_c, y = prod, group=subject_id), col = "black", alpha=.1) + 
+  geom_line(linewidth = 2) +
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = as.factor(log_rt_0_c), group = as.factor(log_rt_0_c)), 
+              alpha = 0.3) +
+  labs(x = "Centered Age", y = "Predicted prod (fixed effects only)") +
+  theme_minimal() + 
+  scale_color_solarized(name = "log RT at t0 (SD)") + 
+  scale_fill_solarized(guide = FALSE)
+
 ```
@@
 
-```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between growth parameters. "}
+```{r longitudinal, fig.env="figure", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between growth parameters. "}
 layout <- read.csv(here("misc/layout_slopes_observed_growth.csv"), header = FALSE)
@@
 
-Reaction times
-
 We excluded trials entirely if they were missing data on more than 50% of timepoints, and excluded RTs if they were based on fewer than 50% of timepoints in the short analytic window (200 -- 2000 ms).
@@
 
-We used `lme4` to fit linear mixed effects models, `nlme` to fit non-linear models, and `lavaan` to fit structural equation models.
+We used `lme4` to fit linear mixed effects models, `brms` to fit non-linear growth models, and `lavaan` to fit structural equation models.
 Random effects structures for each model are given in text; full model specifications are available in the Supplemental Information and in the reproducible code for this paper, available in the linked repository.

### 2025-09-24 paper/supplemental_information.Rmd
assorted table futzing, refs

@@
   - \usepackage{booktabs}
+  - \usepackage{placeins}
   - \renewcommand{\thefigure}{S\arabic{figure}}
@@
 source(here("helper","common.R"))
+#library(kableExtra)
 library(papaja)
@@
 
+\FloatBarrier
+
 # Reaction Time Computation
@@
 Other fixations can be to areas not in the target or distractor as well to off-screen areas.
-This time series has a uniform sample rate of 25ms/sample, based on resampling of the data in Peekbank to 40 Hz during preprocessing [@zettersten2023peekbank].
+This time series has a uniform sample rate of 25ms/sample, based on resampling of the data in Peekbank to 40 Hz during preprocessing (Zettersten et al.
+2023).
 Reaction times are computed by filtering trials to only those on which the child is fixating the distractor at the point of disambiguation ($t=0$) and then finding those trials on which the first non-missing fixation is to the target (hence excluding trials without a shift and trials on which a shift is to an off-screen location).
@@
 (2008), RTs that are shorter than 367 ms are excluded as they are too short to be considered a response to the stimulus.
+
+\FloatBarrier
 
@@
 
-# Separability of Variability from Other Measures
-
-(TODO someone should edit this text)
-A potential concern with multiple measures deriving from the same underlying data (as with RT/accuracy in the looking-while-listening paradigm) is that distributional assumptions about each variable could implicitly assume some connections between RT/accuracy and variability that might not be warranted.
-
-To begin, we note that RT/accuracy and variability could be connected due to true age-related covariation - that is, some common cause leading to improvement in both, namely improvement in speed of response and improvement in consistency of response (e.g., skill learning).
-But they could be necessarily connected by virtue of their distributional form, and this is the concern we address here.
+\FloatBarrier
+
+# Checks on data distributional assumptions
+
+Here, we check whether the distributional forms that are assumed for the distributions of RT and accuracy are a reasonable empirical fit to the data, and compare against other commonly used distributional forms.
+
+We confirm that, across the age range, the choice to use a log-normal distribution for RT and a normal distribution for accuracy is justified.
 
@@
 #note that fitdistrplus attaches MASS which also has a select function, and that's annoying
-detach("package:dplyr", character.only = TRUE)
-library("dplyr", character.only = TRUE)
 
@@
                                   "21-24", "24-27","27-30","30-36", ">36"))) |>
-  select(age_bin, rt)
+  dplyr::select(age_bin, rt)
 
@@
 
-#BIC(norm)
-#BIC(lnorm)
-#BIC(wald)
-#BIC(exgaus)
+norm_bic <- BIC(norm)
+lnorm_bic <- BIC(lnorm)
+wald_bic <- BIC(wald)
+exgaus_bic <- BIC(exgaus)
 
@@
                                     lower = c(0, 0, 0)))) |>
-  select(-data) |>
+  dplyr::select(-data) |>
   pivot_longer(norm:exgaus, names_to = "distribution", values_to = "fit") |>
@@
 
-We’ll look at reaction time first.
-Under certain distributional forms for reaction time, there would be a necessary relationship between RT and its variation.
-However, the literature focuses on the use of the Exponential-Gaussian (ex-Gaussian) distribution as well as Wald, Weibull, gamma, and log normal distributions (see for example Luce, 1986; Ratcliff, 1993; Van Zandt, 2002).
+The literature focuses on the use of the Exponential-Gaussian (ex-Gaussian) distribution as well as Wald, Weibull, gamma, and log normal distributions (see for example Luce, 1986; Ratcliff, 1993; Van Zandt, 2002).
 All of these are 2- or 3-parameter distributions meaning that there is no necessary relationship between mean and variance.
@@
 
-That said, we have attempted some fitting and we find that overall the data are best fit by either an ex-Gaussian or a log normal distribution, again consistent with prior literature, giving us confidence in this conclusion.
-Across the full dataset, the BIC values for ex-Gaussian (`r BIC(exgaus)`) and log normal (`r BIC(lnorm)`) are quite close to one another, and better than the Wald (`r BIC(wald)`) and normal (`r BIC(norm)`) fits.
+That said, we find that overall the data are best fit by either an ex-Gaussian or a log normal distribution, again consistent with prior literature, giving us confidence in this conclusion.
+Across the full dataset, the BIC values for ex-Gaussian (`r round(exgaus_bic)|> as.character()`) and log normal (`r round(lnorm_bic)|> as.character()`) are quite close to one another, and better than the Wald (`r round(wald_bic)|> as.character()`) and normal (`r round(norm_bic)|> as.character()`) fits.
 When binned by age (Fig \ref{fig:rt-fits}), younger children seem better fit for a log normal distribution and older children seem better fit by ex-Gaussian (models with lowest BICs are shown in red since significant differences can be obscured by the large scale).
@@
 
-Overall, we think this result generally vitiates our decision to use log-transformed RTs as our primary dependent measure, and further suggests that reaction time means and variabilities are, in principle, theoretically independent.
-
-```{r rt-fits, fig.cap="\\label{fig:rt-fits} Goodness of fit for different distributional models for RT, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
+Overall, we think this result generally vitiates our decision to use log-transformed RTs as our primary dependent measure.
+
+```{r rt-fits, fig.cap="\\label{fig:rt-fits} Goodness of fit for different distributional models for RT, split by age.", fig.pos="h!", fig.width=8, fig.height=5}
 ggplot(rt_fits, 
@@
 
-```{r rt-hist, fig.cap="\\label{fig:rt-hist} Distribution of RT overlaid with log normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
+```{r rt-hist, fig.cap="\\label{fig:rt-hist} Distribution of RT overlaid with log normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=5}
 
@@
                                   "21-24", "24-27","27-30","30-36", ">36"))) |>
-  select(age_bin, long_window_accuracy)
+  dplyr::select(age_bin, long_window_accuracy)
 
@@
 
-#BIC(norm)
-#BIC(beta)
-```
-
-Individual trial-level accuracies are not binomial because they are an average probability of fixation over a viewing window, as described in the manuscript.
+norm_acc_bic <- BIC(norm_acc)
+beta_acc_bic <- BIC(beta_acc)
+```
+
+Individual trial-level accuracies are not binomial because they are an average probability of fixation over a viewing window.
 They are bounded at 0 and 1, but in general they tend towards the range .5 - .8 in most studies of this population.
-Figure \ref{acc-hist} shows the data binned by age group with fitted gaussian distributions.
+Figure \ref{fig:acc-hist} shows the data binned by age group with fitted gaussian distributions.
 
 These distributions seem well fit by standard gaussians, but they are in principle bounded and so we asked whether this made a difference, using a Beta distribution (a two-parameter continuous distribution bounded at 0 and 1) for fitting.
-Surprisingly, across all data, the BIC values for the two distributions were very similar (`r BIC(norm_acc)` for normal versus `r BIC(beta_acc)` for Beta), though the Beta was slightly favored.
+Surprisingly, across all data, the BIC values for the two distributions were very similar (`r round(norm_acc_bic) |> as.character()` for normal versus `r round(beta_acc_bic)` for Beta), though the normal distribution was slightly favored.
 Across age groups, there was heterogeneity with some groups better fit by a gaussian and others better fit by a Beta (Figure \ref{fig:acc-fits}).
@@
 Again, we feel that this result generally vitiates our approach of modeling accuracies via standard linear mixed effects models: their distributional form is quite close to normal.
-Thus, again there is no necessary connection between mean and variance: they could in principle be independent.
-As above, we will plan on including this information in the Supplement.
-
-```{r acc-hist, fig.cap="\\label{fig:acc-hist} Goodness of fit for different distributional models of accuracy, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
+
+```{r acc-hist, fig.cap="\\label{fig:acc-hist} Goodness of fit for different distributional models of accuracy, split by age.", fig.pos="h!", fig.width=8, fig.height=5}
 acc_hist_data <- d_sub |>
@@
 
-```{r acc-fits, fig.cap="\\label{fig:acc-fits} Distribution of accuracies overlaid with normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
+```{r acc-fits, fig.cap="\\label{fig:acc-fits} Distribution of accuracies overlaid with normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=5}
 acc_fits <- acc_fitting |>
@@
     beta = map(data, \(d) fitdist(d$long_window_accuracy, distr = "beta", method = "mle"))) |>
-  select(-data) |>
+  dplyr::select(-data) |>
   pivot_longer(norm:beta, names_to = "distribution", values_to = "fit") |>
@@
 
+\FloatBarrier
+
 # Test-Retest Reliability
@@
 
+\FloatBarrier
 # Correlations
@@
 ```{=tex}
-\begin{table}[tbp]
+\begin{table}[H]
+\caption{Pairwise correlations between primary variables of interest. \label{tab:corrs}}
 \begin{center}
@@
 \end{center}
-\caption{Pairwise correlations between primary variables of interest. \label{tab:corrs}}
+
 \end{table}
 ```
-```{r corrs, results = "asis", include = FALSE}
-
-cor(select(d_sub, age, log_age, rt, log_rt, 
+
+```{r corrs, results = "asis", include = FALSE, eval=F}
+
+cor(dplyr::select(d_sub, age, log_age, rt, log_rt, 
            long_window_accuracy, short_window_accuracy, prod, comp), 
@@
 ```
+
+\FloatBarrier
 
@@
 
-```{=tex}
-\begin{table}[tbp]
-\begin{center}
-\begin{threeparttable}
-\begin{tabular}{lllllllll}
-\toprule
-n.
-obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
-\midrule
-48354 & 0.28 & -8,338.83 & 16,691.65 & 16,753.16 & 16,677.65 & 48347 & Long window, linear age & 0.14\\
-48354 & 0.28 & -8,305.39 & 16,624.78 & 16,686.28 & 16,610.78 & 48347 & Long window, log age & 0.12\\
-50244 & 0.31 & -12,486.60 & 24,987.21 & 25,048.98 & 24,973.21 & 50237 & Short window, linear age & 0.09\\
-50244 & 0.31 & -12,458.77 & 24,931.54 & 24,993.31 & 24,917.54 & 50237 & Short window, log age & 0.08\\
-\bottomrule
-\end{tabular}
-\end{threeparttable}
-\end{center}
-\caption{Model comparison metrics for different functional forms of the relationship between accuracy and age. \label{tab:accmods}}
-\end{table}
-```
-```{=tex}
-\begin{table}[tbp]
-\begin{center}
-\begin{threeparttable}
-\begin{tabular}{lllllllll}
-\toprule
-n.
-obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
-\midrule
-18940 & 0.45 & -12,414.69 & 24,843.38 & 24,898.32 & 24,829.38 & 18933 & Log RT, linear age & 0.21\\
-18940 & 0.45 & -12,393.27 & 24,800.53 & 24,855.48 & 24,786.53 & 18933 & Log RT, log age & 0.21\\
-18940 & 570.49 & -147,644.43 & 295,302.87 & 295,357.81 & 295,288.87 & 18933 & Linear RT, linear age & 0.18\\
-18940 & 570.02 & -147,622.67 & 295,259.34 & 295,314.29 & 295,245.34 & 18933 & Linear RT, log age & 0.18\\
-\bottomrule
-\end{tabular}
-\end{threeparttable}
-\end{center}
-\caption{Model comparison metrics for different functional forms of the relationship between RT and age. \label{tab:rtmods}}
-\end{table}
-```
-```{r accmods, results="asis", include = FALSE}
-papaja::apa_table(acc_mods_lmer_summary)
-```
-
-```{r results="asis", include = FALSE}
-papaja::apa_table(rt_mods_lmer_summary)
-```
+```{r}
+acc_mods_lmer_summary |>
+  mutate(BIC=round(BIC),
+         AIC=round(AIC),
+         logLik=round(logLik),
+                  REMLcrit=round(REMLcrit),
+         model=str_replace(model, "lwa", "Long window") |> str_replace("swa", "Short window") |> str_replace("lin", "linear age") |> str_replace("log", "log age") |> str_replace("_", ", ")
+         ) |>rename(`n. obs`=nobs) |>  dplyr::select(model, everything()) |> 
+kable(digits=3, caption="Model comparison metrics for different functional forms of the relationship between accuracy and age. \\label{tab:accmods}", latex_options = "HOLD_position")
+```
+
+```{r}
+rt_mods_lmer_summary |>
+  mutate(BIC=round(BIC),
+         AIC=round(AIC),
+         logLik=round(logLik),
+         REMLcrit=round(REMLcrit),
+         model=str_replace(model, "log_", "Log RT, ") |> str_replace("lin_", "Linear RT, ") |> str_replace("lin", "linear age") |> str_replace("log", "log age") |> str_replace("_", ", ")
+         ) |> rename(`n.  obs`=nobs) |> dplyr::select(model, everything()) |> 
+kable(digits=3, caption="Model comparison metrics for different functional forms of the relationship between RT and age.
+\\label{tab:rtmods}", format="latex", booktabs=T, latex_options = "HOLD_position")
+```
+
+\FloatBarrier
 
@@
 The best fitting curves for individuals are often exponentials or delayed exponentials (Evans et al., 2018; Heathcote, Brown, & Mewhort, 2000).
+
 With our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns unfortunately, because we have insufficient data about individuals' improvement across time.
@@
 
-Nevertheless, here we test for other forms of the aggregate relationship between age and reaction time. In particular, we consider further model comparison between the
+We believe that, unlike the skills being studied in the prior adult literature (e.g., Anderson, 1982; Heathcote, Brown, & Mewhort, 2000; Logan, 1988), language processing is being learned over the course of a child’s lifetime.
+Thus, we do not expect to see within-paradigm changes in learning in what is a narrow period of time compared to the duration over which language processing skills are refined.
+
+Nevertheless, here we test for other forms of the aggregate relationship between age and reaction time.
+In particular, we consider 1) a log~log relationship between RT and age (presented in the main text), 2) using both a log age and a linear age to predict log RT, 3) a quadratic relationship between age and RT, and 4) a cubic relationship between age and RT.
+As shown in Table \ref{tab:age-rt}, the model with a linear age term in addition to a log age term has the best fit, although the linear age term coefficient is only marginally significant (coefficients in Table \ref{tab:loglin}).
+
+These models reveal a small but significant additional linear age term over and above log age, but – because individual participant-level fits are not possible – this term can’t really be used to weigh in on the debate about the precise nature of the learning pattern.
 
@@
 
-rt_more_mods_lmer_summary <- map_df(rt_more_mods_lmer, ~broom.mixed::glance(.x)) |>
+ map_df(rt_more_mods_lmer, ~broom.mixed::glance(.x)) |>
   mutate(model = names(rt_more_mods_lmer), 
-         r2 = map_dbl(rt_more_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional))
-```
+         model = case_when(
+           model=="log_log" ~ "Log RT, log age",
+           model=="log_log_plus"~ "Log RT, log+linear age",
+           model=="lin_poly2" ~ "Linear RT, poly(age,2)",
+           model=="lin_poly3" ~ "Linear RT, poly(age,3)"
+         ),
+         r2 = map_dbl(rt_more_mods_lmer, ~performance::r2_nakagawa(.x)$R2_conditional)) |> mutate(AIC=round(AIC), BIC=round(BIC), logLik=round(logLik),          REMLcrit=round(REMLcrit),
+) |> dplyr::select(model, everything()) |>  rename(`n.
+obs`=nobs) |> kable(digits=3, caption="Goodness of fit comparison between different models of the relationship between age and RT.
+\\label{tab:age-rt}", padding=0, latex_options = "HOLD_position")
+```
+
+```{r}
+ log_log_plus = lmer(log_rt ~ log_age_s + age_s +  
+                             (log_age_s + age_s | dataset_name) + (1 | subject_id), 
+                           data = d_trial)
+
+summary(log_log_plus)$coef |> as_tibble(rownames="Term") |> 
+  mutate(Term=case_when(
+  Term=="(Intercept)"~"Intercept",
+  Term=="log_age_s"~ "log age",
+  Term=="age_s"~"linear age",)) |> rename(`$p$-value`=`Pr(>|t|)`)|> kable(digits=3, caption="Fixed effects coefficients for a model predicting log RT from both log age and linear age.
+\\label{tab:loglin}", format="latex", escape=F, booktabs=T, latex_options = "HOLD_position") |>   kableExtra::kable_styling(full_width = F)
+```
+
+\FloatBarrier
 
@@
 
-Here we provide specifications for the mixed effects models used in the main text.
+Here we provide specifications for the lmer mixed effects models used in the main text.
 These models are used to estimate the relationship between age and the primary variables of interest, controlling for dataset and subject-level variability.
 
-<!--\verbatim{lmer(long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)}-->
-
-FIXME because I cause things to crash!
+For accuracy, 4 models were run, crossing long and short windows as the dependent variable with age or log age as the predictor.
+
+\begin{verbatim}
+long_window_accuracy ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)
+long_window_accuracy ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)
+short_window_accuracy ~ age_s  + (age_s | dataset_name) + (1 | subject_id)
+short_window_accuracy ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)
+\end{verbatim}
+
+For reaction time, 4 models were run, crossing rt and log rt as the dependent variable  with age or log age as the predictor.
+
+\begin{verbatim}
+log_rt ~ age_s + (age_s | dataset_name) + (1 | subject_id)
+log_rt ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)
+rt ~ age_s + (age_s | dataset_name) + (1 | subject_id)
+rt ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)
+\end{verbatim}
+
+To look at the relationship between variance in the accuracy and reaction time measures and children's age, we ran two models.
+
+\begin{verbatim}
+long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) +  (1 | subject_id)
+log_rt_var ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)
+
+\end{verbatim}
+
+In the growth curve analysis, fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at
+study initiation (t0), and their interaction, using the formula below
+
+\begin{verbatim}
+prod ~ poly(age_15,2) *rt_t0 + (age  | subject_id) + (1 | dataset_name)
+\end{verbatim}
+
+\FloatBarrier
 
@@
   ungroup() |>
-  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)
+  dplyr::select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)
 
@@
                 ~ age_scale(.x, age))) 
-outputs <- capture.output(fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
+outputs <- capture.output(fa.parallel(dplyr::select(d_sub_mat, -dataset_name, -age), fa = "fa", 
                                       use = "pairwise.complete.obs"))
@@
 ```{r eval=FALSE}
-fs <- fa(select(d_sub_mat, -dataset_name), nfactor = 3,
+fs <- fa(dplyr::select(d_sub_mat, -dataset_name), nfactor = 3,
          use = "pairwise", rotate = "varimax")
@@
 ```{=tex}
-\begin{table}[ht]
+\begin{table}[H]
 \centering
@@
 ```
+
+The confirmatory factor analysis of this three-factor solution was fit using the following specification
+
+\begin{verbatim}
+vocab =~ prod + comp
+accuracy =~ long_window_accuracy + long_window_acc_var
+speed =~ log_rt + log_rt_var
+\end{verbatim}
+
+The confirmatory factor analysis of the three-factor solution  with a relation to age was fit using the following specification
+
+\begin{verbatim}
+vocab =~ prod + comp
+accuracy =~ acc + acc_sd
+speed =~ log_rt + log_rt_sd
+
+vocab ~ log_age
+accuracy ~ log_age
+speed ~ log_age
+\end{verbatim}
+
+The SEM with a linear growth curve used the following specification
+
+\begin{verbatim}
+accuracy_intercept =~ 1*acc_t0 + 1*acc_t1 + 1*acc_t2 + 1*acc_t3 + 1*acc_t4
+accuracy_slope =~ 1*acc_t0 + 2*acc_t1 + 3*acc_t2 + 4*acc_t3 + 5*acc_t4
+speed_intercept =~ 1*log_rt_t0 + 1*log_rt_t1 + 1*log_rt_t2 + 1*log_rt_t3 + 1*log_rt_t4
+
+speed_slope =~ 1*log_rt_t0 + 2*log_rt_t1 + 3*log_rt_t2 + 4*log_rt_t3 + 5*log_rt_t4
+vocab_intercept =~ 1*prod_t0 + 1*prod_t1 + 1*prod_t2 + 1*prod_t3 + 1*prod_t4
+vocab_slope =~ 1*prod_t0 + 2*prod_t1 + 3*prod_t2 + 4*prod_t3 + 5*prod_t4
+
+accuracy_intercept ~~ NA*accuracy_intercept
+accuracy_slope ~~ NA*accuracy_slope
+speed_intercept ~~ NA*speed_intercept
+
+speed_slope ~~ NA*speed_slope
+vocab_intercept ~~ NA*vocab_intercept
+vocab_slope ~~ NA*vocab_slope
+\end{verbatim}
+
+\FloatBarrier
+
 # Factor Analysis: First Administrations
+
+As a robustness check, we tested our best factor analytic models using only cross-sectional data (filtering to the first test session in longitudinal datasets; N=1963 instead of N=3553).
+A comparison of all 4 models is shown in Table \ref{tab:factor-first}.
+For the three-factor CFA, the first administration model shows increased CFI (.992 instead of .972) and decreased RMSEA (.030 instead of .065).
+The same is true for the age-regressed three-factor CFA, which shows very good statistics on both first administrations and longitudinal data (CFI = .999 and .991, respectively and RMSEA = .009 and .037 respectively).
 
@@
   ungroup() |>
-  select(dataset_name, log_rt, log_rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age, log_age)
+  dplyr::select(dataset_name, log_rt, log_rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age, log_age)
 
@@
 library(semoutput)
-semoutput::sem_fitmeasures(fit3, print=F) |> kable()
-semoutput::sem_fitmeasures(fit3_firstadmin, print=F) |> kable()
-semoutput::sem_fitmeasures(fit3_age, print=F) |> kable()
-semoutput::sem_fitmeasures(fit3_age_firstadmin, print=F) |> kable()
-
-```
-
+
+output <- sem_fitmeasures(fit3, print=F) |> as_tibble() |> mutate(Model="No Age, longitudinal") |>
+  bind_rows(sem_fitmeasures(fit3_firstadmin, print=F) |> as_tibble() |> mutate(Model="No Age, first admin")) |> 
+  bind_rows(sem_fitmeasures(fit3_age, print=F) |> as_tibble() |> mutate(Model="Age, longitudinal")) |> 
+  bind_rows(sem_fitmeasures(fit3_age_firstadmin, print=F) |> as_tibble() |> mutate(Model="Age, first admin")) |>
+  dplyr::select(Model, everything()) |> mutate(AIC=round(AIC), BIC=round(BIC)) |> rename(`RMSEA$_{Lower}$`=`RMSEA_Lower`, `RMSEA$_{Upper}$`=`RMSEA_Upper`)
+
+kable(output, digits=3, caption="Comparison of confirmatory factor analysis models on longitudinal data or first administrations only. \\label{tab:factor-first}", padding=0)
+```
+
+\FloatBarrier
 # Alternative Factor Structures
@@
 
-\noindent Table \ref{tab:modcomp} shows the result of these comparisons. The three-factor model shows the lowest AIC and BIC, as well as being significantly better fittign than the next-best model.
+\noindent Table \ref{tab:modcomp} shows the result of these comparisons. The three-factor model shows the lowest AIC and BIC, as well as being significantly better fitting than the next-best model.
 
@@
 ```
+
+\FloatBarrier
 # Non-linear Growth Models
-
-To test for the differentiation of vocabulary growth based on initial reaction time, we used the package `nlme` to fit a logistic growth model to the production data.
-This model has two parameters for the logistic curve, a scale and an intercept.
-Both were allowed to interact with initial reaction time.
-We also included random effects of logistic intercept and scale by participant and a grouping term across datasets.
-This model showed a significant effect of initial reaction time on the intercept of the logistic growth curve, but not on its scale (see Table \ref{tab:nlme}).
-
-```{r}
-# xtable(summary(mod_nlme)$tTable)
-```
-
-<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
-
-<!-- % Thu Mar 13 13:10:30 2025 -->
-
-```{=tex}
-\begin{table}[ht]
-\centering
-\begin{tabular}{rrrrrr}
-\hline
-& Value & Std Error & DF & $t$-value & $p$-value \\ 
-\hline
-Growth Intercept & 2.26 & 2.70 & 1766.00 & 0.84 & 0.40 \\ 
-Growth Intercept $\times$ $t_0$ RT & 3.30 & 0.38 & 1766.00 & 8.59 & < 0.0001 \\ 
-Growth Scale & 2.05 & 2.60 & 1766.00 & 0.79 & 0.43 \\ 
-Growth Scale $\times$ $t_0$ RT & 0.28 & 0.38 & 1766.00 & 0.75 & 0.46 \\ 
-\hline
-\end{tabular}
-\caption{Fixed effects estimates from logistic growth model. \label{tab:nlme}}
-\end{table}
-```
-Interpretation of growth in both this model and the linear growth model in the main text is complicated by the fact that the CDI form puts a ceiling on the total number of words that can be recorded; both the quadratic growth functions and the logistic functions come together at the form ceiling.
-Thus, a shift in quadratic growth in the linear model and a shift in intercept in the logistic model both point to the same overall effect, which is faster growth at the point of maximal sensitivity of the CDI.
-Neither model can estimate whether the overall growth trajectory is different beyond the range of the CDI.
-Thus, although these models might initially seem to be in conflict, we believe that they actually point to the same phenomenon, which is perhaps better described by the longitudinal SEM model reported in the main text.
-Children with greater skill in word recognition show an overall positive shift in the growth trajectory of vocabulary development.
-
-FIXME
-
-```{r}
-d_sub_prod <- filter(d_sub, !is.na(prod), !is.na(log_rt)) |>
-  ungroup() |>
-  mutate(log_rt_resid = resid(lm(log_rt ~ log_age))) |>
-  group_by(subject_id) |>
-  mutate(log_rt_0 = log_rt[1],
-         log_rt_0_resid = log_rt_resid[1],
-         acc_0 = long_window_accuracy[1]) |> 
-  filter(!is.na(log_rt_0))
-
-mod_nlme <- nlme(model = prod ~ SSlogis(age, Asym = 1, xmid, scale),
-                 data = d_sub_prod,
-                 fixed = xmid + scale ~ log_rt_0_resid,
-                 random = xmid + scale ~ 1 | subject_id,
-                 groups = ~ dataset_name,
-                 start = c(xmid = 20, scale = 3, `xmid:log_rt_0_resid` = 1, `scale:log_rt_0_resid` = 3),
-                 na.action = na.exclude, 
-                 control = list(maxIter = 1000, tolerance = .1))
-
-#summary(mod_nlme)
-
-# fixed
-global_preds <- expand_grid(age = seq(min(d_sub_prod$age), 
-                                      max(d_sub_prod$age), .1), 
-                            log_rt_0_resid = c(quantile(d_sub_prod$log_rt_0_resid, 
-                                                        c(.25,.75), 
-                                                        na.rm=TRUE)))
-global_preds$pred <- predict(mod_nlme, newdata = global_preds, level = 0)
-
-#random effects
-dataset_preds <- d_sub |>
-  filter(!is.na(prod)) |> # filter on predictor
-  group_by(dataset_name) |>
-  summarise(min_age = min(age),
-            max_age = max(age)) |>
-  group_by(dataset_name) |>
-  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0_resid = c(quantile(d_sub_prod$log_rt_0_resid, 
-                                                                                                     c(.25,.75), 
-                                                                                                     na.rm=TRUE))))) |>
-  dplyr::select(-min_age, -max_age) |>
-  unnest(col = "df") |>
-  ungroup()
-
-dataset_preds$pred <- predict(mod_nlme,
-                              newdata = dataset_preds,
-                              level = c(0,1))$predict.dataset_name
-
-ggplot(d_sub_prod, aes(x = age, y = prod, col = as.factor(log_rt_0_resid))) + 
-  geom_point(alpha = .03, pch = ".", col = "black") + 
-  geom_line(aes(group = subject_id), alpha = .02, col = "black") + 
-  geom_line(data = dataset_preds |>   
-              filter(dataset_name %in% datasets_with_age_variation$dataset_name), 
-            aes(y = pred, group = interaction(log_rt_0_resid, dataset_name)), lty="dashed", alpha=.5)+
-  geom_line(data = global_preds, aes(y = pred, group = log_rt_0_resid, col = as.factor(log_rt_0_resid)), size=1)+
-  scale_color_viridis(discrete=T, option = "H")+
-  theme(legend.position = "none")+
-  labs(y="Production", x="Age in months")
-
-```
 
@@
 library(brms)
-
-d_sub_prod$age_c <- d_sub_prod$age - mean(d_sub_prod$age, na.rm = TRUE)
-d_sub_prod$log_rt_0_c <- d_sub_prod$log_rt_0 - mean(d_sub_prod$log_rt_0, na.rm = TRUE)
-
-# Define the nonlinear formula
-nlform <- brms::bf(
-  prod ~ 1 / (1 + exp((xmid - age_c) / exp(logscale))),
-  xmid ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
-  logscale ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
-  # scale ~ 1 + log_rt_0_c,
-  nl = TRUE
-)
-
-# xmid ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
-# scale ~ 1 + log_rt_0_c + (1 | dataset_name/subject_id),
-
-priors <- c(
-  prior(normal(0, 5), nlpar = "xmid", coef = "Intercept"),  # xmid near center of age_c
-  prior(normal(1, 1), nlpar = "logscale", coef = "Intercept"), # scale > 0, mildly steep curve
-  prior(normal(0, 1), nlpar = "logscale", coef = "log_rt_0_c"),
-  prior(exponential(1), class = "sigma"),                # residual error
-  prior(normal(0, 2), nlpar = "xmid", coef = "log_rt_0_c"),
-  # prior(normal(0, 1), nlpar = "scale", lb = 0)
-
-  # Random effects for xmid
-  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name"),
-  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name:subject_id"),
-
-  # Random effects for scale
-  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name"),
-  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name:subject_id")
-)
-#
-
-# Fit the model
-mod_brms <- brm(
-  formula = nlform,
-  data = d_sub_prod,
-  prior = priors,
-  family = gaussian(),
-  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
-  chains = 4, cores = 4,
-  control = list(adapt_delta = 0.95),
-  # backend = "cmdstanr",  # optional but faster
-  save_pars = save_pars(all = TRUE),
-  sample_prior = "yes",
-  file="brms1.rds"
-)
-
-new_data <- expand.grid(
-  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
-  log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
-)
-
-fitted_vals <- fitted(mod_brms, newdata = new_data, re_formula = NA, summary = TRUE)
-
-saveRDS(fitted_vals, "brms1_fitted.rds")
-```
-
-```{r}
-fitted_vals <- readRDS("brms1_fitted.rds")
-
-d_sub_prod$age_c <- d_sub_prod$age - mean(d_sub_prod$age, na.rm = TRUE)
-d_sub_prod$log_rt_0_c <- d_sub_prod$log_rt_0 - mean(d_sub_prod$log_rt_0, na.rm = TRUE)
-
-new_data <- expand.grid(
-  age_c = seq(min(d_sub_prod$age_c), max(d_sub_prod$age_c), length.out = 100),
-  log_rt_0_c = c(-1, 0, 1)  # e.g., low, mean, high values
-)
-
-new_data$fit <- fitted_vals[, "Estimate"]
-new_data$lower <- fitted_vals[, "Q2.5"]
-new_data$upper <- fitted_vals[, "Q97.5"]
-
-ggplot(new_data, aes(x = age_c, y = fit, col = as.factor(log_rt_0_c))) +
-  geom_line(linewidth = 2) +
-  geom_ribbon(aes(ymin = lower, ymax = upper, fill = as.factor(log_rt_0_c), group = as.factor(log_rt_0_c)), 
-              alpha = 0.2) +
-  labs(x = "Centered Age", y = "Predicted prod (fixed effects only)") +
-  theme_minimal() + 
-  geom_point(data = d_sub_prod, aes(x = age_c, y = prod), col = "black") + 
-  scale_color_solarized(name = "log RT at t0 (SD)") + 
-  scale_fill_solarized(guide = FALSE)
-
-```
-
-```{r, eval=F}
-library(brms)
+library(tidybayes)
 
@@
 
+summary(resid_mod_brms)$fixed |> saveRDS("brms2_coef.rds")
+
 new_data_resid <- expand.grid(
@@
 ```{r}
+readRDS("brms1_coef.rds") |> as_tibble(rownames="type") |> 
+  mutate(Type=case_when(
+    type=="xmid_Intercept" ~ "Constant component of growth intercept",
+    type=="xmid_log_rt_0_c" ~ "Effect of $t_0$ RT on growth intercept",
+    type=="logscale_Intercept" ~ "Constant compontent of growth scale",
+    type=="logscale_log_rt_0_c" ~ "Effect of $t_0$ RT on growth scale"
+  )) |> 
+  dplyr::select(Type, Estimate, `Est.Error`, `l-95\\% CI`=`l-95% CI`, `u-95\\% CI`=`u-95% CI`) |> kable(digits=3, caption="Fixed effects estimates from logistic growth model.
+\\label{tab:nlme}", format="latex", escape=F, booktabs=T, latex_options = "HOLD_position") |>   kableExtra::kable_styling(full_width = F)
+
+```
+
+```{r}
+readRDS("brms2_coef.rds") |> as_tibble(rownames="type") |> 
+  mutate(Type=case_when(
+    type=="xmid_Intercept" ~ "Constant component of growth intercept",
+    type=="xmid_resid_log_rt_0_c" ~ "Effect of residualized $t_0$ RT on growth intercept",
+    type=="logscale_Intercept" ~ "Constant component of growth scale",
+    type=="logscale_resid_log_rt_0_c" ~ "Effect of residualized $t_0$ RT on growth scale"
+  )) |> 
+  dplyr::select(Type, Estimate, `Est.Error`, `l-95\\% CI`=`l-95% CI`, `u-95\\% CI`=`u-95% CI`) |> kable(digits=3, caption="Fixed effects estimates from logistic growth model using RT residualized on age as the predictor.
+\\label{tab:nlme-resid}", format="latex", escape=F, booktabs=T, latex_options = "HOLD_position") |>  kableExtra:: kable_styling(full_width = F)
+
+```
+
+To test for the differentiation of vocabulary growth based on initial reaction time, we used the package `brms` to fit a (Bayesian) logistic growth model to the production data.
+This model has two parameters for the logistic curve, a scale and an intercept.
+Both were allowed to interact with initial reaction time.
+We also included random effects of logistic intercept and scale by participant and a grouping term across datasets.
+This model showed a significant effect of initial reaction time on the intercept of the logistic growth curve, but not on its scale (see Table \ref{tab:nlme}).
+
+Age and initial reaction time were both mean-centered.
+
+```{r nlme-residual, fig.cap="\\label{fig:nlme-resid} Growth curves from a logistic growth model showing predicted vocabulary growth for children based on their age-residualized initial reaction times.
+Predictions are shown for children with initial reaction times one SD faster than the mean for their age (blue), at the mean  for their age (red), and one SD slower than the mean for their age (green).
+Individual longitudinal trajectories are shown in light gray.
+Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, fig.width=6, out.width=".6\\textwidth", fig.pos="h!"}
 
@@
 ggplot(new_data_resid, aes(x = age_c, y = fit, col = as.factor(resid_log_rt_0_c))) +
+ geom_line(data = d_sub_prod, aes(x = age_c, y = prod, group=subject_id), col = "black", alpha=.1) + 
   geom_line(linewidth = 2) +
   geom_ribbon(aes(ymin = lower, ymax = upper, fill = as.factor(resid_log_rt_0_c), group = as.factor(resid_log_rt_0_c)), 
-              alpha = 0.2) +
+              alpha = 0.3) +
   labs(x = "Centered Age", y = "Predicted prod (fixed effects only)") +
   theme_minimal() + 
-  scale_color_viridis(discrete=T) + 
-  scale_fill_viridis(discrete=T, guide=F)+
-  geom_point(data = d_sub_prod, aes(x = age_c, y = prod), col = "black")
-
-```
-
+  scale_color_solarized(name = "residualized log RT at t0 (SD)") + 
+  scale_fill_solarized(guide = FALSE)
+
+```
+
+The formula specification was
+\begin{verbatim}
+nlform <- brms::bf(
+  prod ~ 1 / (1 + exp((xmid - age_c) / exp(logscale))),
+  xmid ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
+  logscale ~ 1 + log_rt_0_c  + (1 | dataset_name/subject_id),
+  # scale ~ 1 + log_rt_0_c,
+  nl = TRUE
+)
+\end{verbatim}
+
+And the priors were
+
+\begin{verbatim}
+priors <- c(
+  prior(normal(0, 5), nlpar = "xmid", coef = "Intercept"), 
+  prior(normal(1, 1), nlpar = "logscale", coef = "Intercept"), 
+  prior(normal(0, 1), nlpar = "logscale", coef = "log_rt_0_c"),
+  prior(exponential(1), class = "sigma"),              
+  prior(normal(0, 2), nlpar = "xmid", coef = "log_rt_0_c"),
+
+  # Random effects for xmid
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "xmid", group = "dataset_name:subject_id"),
+
+  # Random effects for scale
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name"),
+  prior(exponential(1), class = "sd", nlpar = "logscale", group = "dataset_name:subject_id")
+)
+\end{verbatim}
+
+Age and reaction time are correlated, so to check that the effects of initial reaction time were not due to age effects, we reran the model using residualized reaction time to remove effects of age.
+As seen in Table \ref{tab:nlme-resid} and Figure \ref{fig:nlme-resid}, the pattern of effects is similar for residualized reaction time as for reaction time.
+
+Interpretation of growth in both this model and the linear growth model in the main text is complicated by the fact that the CDI form puts a ceiling on the total number of words that can be recorded; both the quadratic growth functions and the logistic functions come together at the form ceiling.
+Thus, a shift in quadratic growth in the linear model and a shift in intercept in the logistic model both point to the same overall effect, which is faster growth at the point of maximal sensitivity of the CDI.
+Neither model can estimate whether the overall growth trajectory is different beyond the range of the CDI.
+Thus, although these models might initially seem to be in conflict, we believe that they actually point to the same phenomenon, which is perhaps better described by the longitudinal SEM model reported in the main text.
+Children with greater skill in word recognition show an overall positive shift in the growth trajectory of vocabulary development.
+
+\FloatBarrier
 # SEM Longitudinal Missingness
 
-FIXME
-
-```{r, eval=F}
-colMeans(!is.na(d_sub_wide_fine))
-```
-
+The SEM model was fit to the entire dataset, including the large mass of cross-sectional data (to anchor the estimates of t0 coefficients) and the sparse longitudinal data for each time point.
+We have 3%-12% of the total t0 datapoints for any given time point (see Table \ref{tab:sem-missing}), given the sparsity of longitudinal sampling (only 6/24 of the datasets are longitudinal).
+
+Our data are MAR (missing at random) rather than MCAR (missing completely at random).
+This is because their missingness is due to which dataset they are part of – if they are from a cross-sectional dataset, they are by definition missing all longitudinal observations.
+For our analyses to be appropriate given this structure, we have to assume that the general developmental patterns we are studying are replicated across datasets.
+We believe that they are, and we show this statistically using our mixed effects and non-linear mixed effects models, which control for dataset-related variation.
+(We also show dataset-level effects in a number of our visualizations for this same reason).
+The same degree of random effect specification that we can do in the mixed effects models is not possible in the SEM model, however, purely for technical reasons.
+Again, this point highlights the importance of convergence across analyses.
+
+```{r}
+colMeans(!is.na(d_sub_wide_fine)) |> as_tibble(rownames="type") |> filter(!type%in%c("subject_id", "dataset_name")) |>  separate_wider_regex(type, c(type = ".*", "_", timepoint = ".*")) |> pivot_wider(names_from=type, values_from=value) |> arrange(timepoint) |> dplyr::select(-log_rt_sd, -acc_sd) |>
+  rename("Log RT"=log_rt, Accuracy=acc, Production=prod, Comprehension=comp) |> kable(digits=3, caption="\\label{tab:sem-missing} Fraction of data present for each measure at each time point for the longitudinal SEM.
+", format="latex",  booktabs = TRUE, linesep = "", latex_options = "HOLD_position") |>   kableExtra::kable_styling(full_width = F)
+
+```
+\FloatBarrier
 # Additional References
 
-Heathcote, A., Brown, S., & Mewhort, D. J. K. (2000). The power law repealed: The case for an exponential law of practice. Psychonomic Bulletin & Review, 7(2), 185–207.
+Baayen, R. H., & Milin, P. (2010). Analyzing reaction times. International journal of psychological research, 3(2), 12-28.
 
 Creel, S. (2024). Connecting the tots: Strong looking-pointing correlations in preschoolers' word learning and implications for continuity in language development. Child Development, 96, 87-103.
+
+Logan, G. D. (1988). Toward an instance theory of automatization. Psychological review, 95(4), 492.
+
+Luce, P. A. (1986). A computational analysis of uniqueness points in auditory word recognition. Perception & Psychophysics, 39(3), 155-158.
+
+Ratcliff, R. (1979). Group reaction time distributions and an analysis of distribution statistics. Psychological bulletin, 86(3), 446.
+
+Ratcliff, R. (1993). Methods for dealing with reaction time outliers. Psychological bulletin, 114(3), 510.
+
+Van Zandt, T. (2000). How to fit a response time distribution. Psychonomic bulletin & review, 7(3), 424-465.
+
+Van Zandt, T. (2002). Analysis of response time distributions. Stevens’ handbook of experimental psychology, 4, 461-516.

### 2025-10-13 paper/supplemental_information.Rmd
fix a couple typos and allow for "first time" eval

@@
 library(papaja)
-```
-
-```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, include=F}
+
+#set to TRUE to fit the Bayesian models
+FIT_BAYES = FALSE
+```
+
+```{r, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE, include=F}
 paper <- suppressMessages(knitr::purl(here("paper","paper.Rmd"), documentation = 0))
@@
 
-```{r, eval=F}
+```{r, eval=FIT_BAYES}
 library(brms)
@@
 
-fitted_vals <- fitted(mod_brms, newdata = new_data, re_formula = NA, summary = TRUE)
+fitted_vals <- fitted(resid_mod_brms, newdata = new_data_resid, re_formula = NA, summary = TRUE)
 

### 2025-10-13 paper.R

@@ -6,7 +6,7 @@ suppressPackageStartupMessages(library(tidySEM))
 suppressPackageStartupMessages(library(xtable))
 suppressPackageStartupMessages(library(papaja))
 suppressPackageStartupMessages(library(viridis))
-suppressPackageStartupMessages(library(nlme))
+#suppressPackageStartupMessages(library(nlme))
 suppressPackageStartupMessages(library(cowplot))
 
 source(here("helper","common.R"))
@@ -52,6 +52,9 @@ d_sub_firstadmin <- d_sub |>
 
 knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
 
+#set to TRUE the first time running the document to ensure all cached elements are created (FALSE thereafter)
+FIRST_TIME = FALSE
+
 dataset_characteristics <- d_sub |>
   group_by(dataset_name) |>
   summarise(`N subjects` = length(unique(subject_id)),
@@ -89,7 +92,7 @@ d_timecourse_by_dataset_age_subj <- d_aoi |>
     age_bin = cut(age, breaks = c(0,12,15,18,21,24,27,30,36,60), 
                   labels = c("<12", "12-15","15-18", "18-21", "21-24", "24-27","27-30","30-36", ">36"))
   ) |>
-  #downsample t_norm to 50ms bins
+  #downsample t_norm to 50ms binsv
   #mutate(t_norm_downsampled = round(t_norm/50)*50) |>
   group_by(dataset_name,dataset_id,subject_id,age_bin,t_norm) |> 
   #calculate proportion time spent looking at target
@@ -362,62 +365,45 @@ growth_mod <- lmer(prod ~ poly(age_15,2) *rt_t0 + (age  | subject_id) + (1 | dat
 
 # summary(growth_mod)
 
-
-d_sub_prod <- filter(d_sub, !is.na(prod)) |>
+d_sub_prod <- filter(d_sub, !is.na(prod), !is.na(log_rt)) |>
+  ungroup() |>
+  mutate(log_rt_resid = resid(lm(log_rt ~ log_age))) |>
   group_by(subject_id) |>
   mutate(log_rt_0 = log_rt[1],
+         log_rt_0_resid = log_rt_resid[1],
          acc_0 = long_window_accuracy[1]) |> 
-  filter(!is.na(log_rt_0))
+  filter(!is.na(log_rt_0)) 
 
[…]

### 2025-10-13 paper/paper.Rmd
resolve citation
fix a couple typos and allow for "first time" eval
tweak Fig 1 sizing

@@
 In addition, we might observe what is known as the "power law of practice," the regularity found in many cases of skill learning that the logarithm of reaction time decreases with the logarithm of experience across participants [@snoddy1926;@kail1991;@anderson1982; cf.
-@heathcote2000; @heathcote2018, @evans2018].
+@heathcote2000; @evans2018].
 Indeed, this pattern is predicted by an influential associative process model of early word learning [@mcmurray2012].
@@
 knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
+
+#set to TRUE the first time running the document to ensure all cached elements are created (FALSE thereafter)
+FIRST_TIME = FALSE
 ```
@@
 Error bands represent standard errors of the mean.
-Grey backgrounds highlight the short and long time windows used in subsequent analyses.", fig.env="figure", fig.height = 3, fig.width=4, out.width=".5\\textwidth", include=TRUE}
+Grey backgrounds highlight the short and long time windows used in subsequent analyses.", fig.env="figure", fig.height = 4, include=TRUE}
 
@@
 
-```{r, eval=F}
+```{r, eval=FIRST_TIME}
 library(brms)

### 2025-10-14 paper/paper.Rmd
fix a couple typos and add TOC to supplement

@@
 
-Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables).
-We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age nested within study (modeling item and procedural variation across studies) and random intercepts by participant.
+Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw pairwise correlations between variables).
+We began by fitting linear mixed-effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age nested within study (modeling item and procedural variation across studies) and random intercepts by participant.
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
@@
 We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in speed during the initial session of the study.
-We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age nested within participant and also age nested within dataset).
+We fit a mixed-effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age nested within participant and also age nested within dataset).
 This model revealed a significant effect of $t_0$ RT (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ RT and the quadratic age predictor (`r papaja::apa_print(growth_mod)$full_result$polyage_15_22_rt_t0`).
@@
 
-We used `lme4` to fit linear mixed effects models, `brms` to fit non-linear growth models, and `lavaan` to fit structural equation models.
+We used `lme4` to fit linear mixed-effects models, `brms` to fit non-linear growth models, and `lavaan` to fit structural equation models.
 Random effects structures for each model are given in text; full model specifications are available in the Supplemental Information and in the reproducible code for this paper, available in the linked repository.

### 2025-10-14 paper/supplemental_information.Rmd
fix a couple typos and add TOC to supplement

@@
     keep_tex: yes
+    toc: true
 header-includes:
@@
 ```
-
-# Dataset Description
+\newpage
+# S1. Dataset Description
 
@@
 
-# Reaction Time Computation
+# S2. Reaction Times
+
+## S2.1. Reaction Time Computation
 
@@
 
-# Comparison of Reaction Times for Correct and Incorrect Trials: Re-analysis of Creel (2024)
+## S2.2. Comparison of Reaction Times for Correct and Incorrect Trials: Re-analysis of Creel (2024)
 
@@
 
-# Checks on data distributional assumptions
+# S3. Checks on Data Distributional Assumptions
 
@@
 
-## Reaction Time
+## S3.1. Reaction Time
 
@@
 
-## Accuracy
+## S3.2. Accuracy
 
@@
 
-Again, we feel that this result generally vitiates our approach of modeling accuracies via standard linear mixed effects models: their distributional form is quite close to normal.
+Again, we feel that this result generally vitiates our approach of modeling accuracies via standard linear mixed-effects models: their distributional form is quite close to normal.
 
@@
 
-# Test-Retest Reliability
+# S4. Test-Retest Reliability
 
@@
 \FloatBarrier
-# Correlations
+# S5. Pairwise Correlations of Main Measures
 
@@
 \toprule
-& \multicolumn{1}{l}{age} & \multicolumn{1}{l}{log age} & \multicolumn{1}{l}{RT} & \multicolumn{1}{l}{log RT} & \multicolumn{1}{l}{long acc} & \multicolumn{1}{l}{short acc } & \multicolumn{1}{l}{prod} & \multicolumn{1}{l}{comp}\\
+& \multicolumn{1}{l}{age} & \multicolumn{1}{l}{log age} & \multicolumn{1}{l}{rt} & \multicolumn{1}{l}{log rt} & \multicolumn{1}{l}{long acc} & \multicolumn{1}{l}{short acc } & \multicolumn{1}{l}{prod} & \multicolumn{1}{l}{comp}\\
 \midrule
@@
 
-# Functional Form Model Comparison
+# S6. Functional Form Model Comparison
 
@@
 
-# Power Law Fits
+# S7. Power Law Fits
 
@@
 
-# Mixed effects model specifications
-
-Here we provide specifications for the lmer mixed effects models used in the main text.
+# S8. Mixed-effects model specifications
+
+Here we provide specifications for the lmer mixed-effects models used in the main text.
 These models are used to estimate the relationship between age and the primary variables of interest, controlling for dataset and subject-level variability.
@@
 
-In the growth curve analysis, fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at
+In the growth curve analysis, fit a mixed-effects model predicting growth in vocabulary as a quadratic function of age, RT at
 study initiation (t0), and their interaction, using the formula below
@@
 
-# Factor Analysis
+# S9. Factor Analysis
 
@@
 
-# Factor Analysis: First Administrations
+# S10. Factor Analysis on First Administrations
 
@@
 \FloatBarrier
-# Alternative Factor Structures
+# S11. Alternative Factor Structures
 
@@
 \FloatBarrier
-# Non-linear Growth Models
+# S12. Non-linear Growth Models
 
@@
 \FloatBarrier
-# SEM Longitudinal Missingness
+# S13. SEM Longitudinal Missingness
 
@@
 For our analyses to be appropriate given this structure, we have to assume that the general developmental patterns we are studying are replicated across datasets.
-We believe that they are, and we show this statistically using our mixed effects and non-linear mixed effects models, which control for dataset-related variation.
+We believe that they are, and we show this statistically using our mixed-effects and non-linear mixed-effects models, which control for dataset-related variation.
 (We also show dataset-level effects in a number of our visualizations for this same reason).
-The same degree of random effect specification that we can do in the mixed effects models is not possible in the SEM model, however, purely for technical reasons.
+The same degree of random effect specification that we can do in the mixed-effects models is not possible in the SEM model, however, purely for technical reasons.
 Again, this point highlights the importance of convergence across analyses.

### 2025-10-14 paper/paper.Rmd
update SI section naming and numbering in paper

@@
 
-Table \ref{tab:datasets} shows the characteristics of individual datasets (see also SI: Dataset Description).
+Table \ref{tab:datasets} shows the characteristics of individual datasets (see also S1 Dataset Description in the Supplementary Information).
 The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm all allowed us to make a more detailed study of the development of word recognition than has previously been possible.
@@
 To characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
-Reaction times were computed only on trials for which the child was fixating the distractor at the point of disambiguation (label onset), and were defined as the time from label onset to the first fixation on the target image (see SI: Reaction Time Computation).
+Reaction times were computed only on trials for which the child was fixating the distractor at the point of disambiguation (label onset), and were defined as the time from label onset to the first fixation on the target image (see S2.1 Reaction Time Computation).
 
@@
 For each window, we averaged all fixations within the window to compute a continuous proportion of target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial.
-In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy (see SI: Test-Retest Reliability); in subsequent analyses we estimate latent variables representing these constructs.
-
-Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw pairwise correlations between variables).
+In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy (see S4 Test-Retest Reliability); in subsequent analyses we estimate latent variables representing these constructs.
+
+Our first question was about the functional form of the relationships between age, speed, and accuracy (see S5 Pairwise Correlations of Main Measures for raw pairwise correlations between variables).
 We began by fitting linear mixed-effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age nested within study (modeling item and procedural variation across studies) and random intercepts by participant.
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
-The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison and SI: Power Law Fits).
+The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see S6 Functional Form Model Comparison and S7 Power Law Fits).
 Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analyses.
@@
 Our goal in these analyses was to understand the underlying relatedness of the various measures of word recognition and vocabulary, and in particular to assess the evidence for 1) whether the speed, accuracy, and variability measures described above all index the same underlying language processing construct and 2) the nature of the relation between this construct (or set of constructs) and early vocabulary.
-We begin developing models using all data, treating each observation as independent even if it comes from a longitudinal study; this assumption is equivalent to asserting an invariant factor structure across development (for a test of this assumption, see SI: Factor Analysis on First Administrations).
+We begin developing models using all data, treating each observation as independent even if it comes from a longitudinal study; this assumption is equivalent to asserting an invariant factor structure across development (for a test of this assumption, see S10 Factor Analysis on First Administrations).
 In subsequent models, we relax this assumption and explore longitudinal growth.
 
-Initial exploratory factor analysis using parallel analysis to select the number of factors suggested that three factors explained substantial variance in the data (see SI: Factor Analysis).
+Initial exploratory factor analysis using parallel analysis to select the number of factors suggested that three factors explained substantial variance in the data (see S9 Factor Analysis).
 To better accommodate missing data under the assumption of data missing at random (e.g., missingness due to the age sampling schemes of the various datasets), we used confirmatory factor analysis with full information maximum likelihood to find the best set of loadings.
 The best fitting model was a three-factor model with factors for speed (RT and RT variability), accuracy (proportion looking to target on each trial and associated variability of this measure), and vocabulary (comprehension and production from the CDI).
-Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`); see SI: Alternative Factor Structures).
+Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`); see S11 Alternative Factor Structures).
 
@@
 
-We confirmed this analysis using a non-linear growth model with a logistic shape, which provides a better fit to vocabulary size within a fixed-length form than the quadratic model (see SI: Non-Linear Growth Model) [@frank2021].
+We confirmed this analysis using a non-linear growth model with a logistic shape, which provides a better fit to vocabulary size within a fixed-length form than the quadratic model (see S12 Non-Linear Growth Model) [@frank2021].
 Figure \ref{fig:growth} shows predictions from this model, confirming the differentiation of growth curves for children with higher and lower initial reaction time.
@@
 We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months from the initial measurement (i.e., $t_0, ..., t_4$) and fit individual growth across each of these variables.
-We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset (se SI: SEM Longitudinal Missingness).
+We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset (see S13 SEM Longitudinal Missingness).
 The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}.
@@
 We retrieved all data from Peekbank release 2025.1 using the `peekbankr` package.
-All code and data necessary to reproduce this manuscript are available at [http://github.com/peekbank/peekbank-development]().
+All code and data necessary to reproduce this manuscript are available at [https://github.com/peekbank/peekbank-development]().
 

### 2025-10-14 paper/paper.Rmd
making sure each SI section is referenced in main

@@
 To characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
-Reaction times were computed only on trials for which the child was fixating the distractor at the point of disambiguation (label onset), and were defined as the time from label onset to the first fixation on the target image (see S2.1 Reaction Time Computation).
+Reaction times were computed only on trials for which the child was fixating the distractor at the point of disambiguation (label onset), and were defined as the time from label onset to the first fixation on the target image (see S2 Reaction Times for further details on how reaction times were computed in S2.1 and discussion of issues surrounding distinguishing "correct" vs.
+"incorrect" trials when computing looking-based reaction times in S2.2).
 
@@
 We began by fitting linear mixed-effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age nested within study (modeling item and procedural variation across studies) and random intercepts by participant.
-We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT.
+We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT (see S3 Checks on Data Distributional Assumptions for further analyses and discussion of these modeling choices).
 The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see S6 Functional Form Model Comparison and S7 Power Law Fits).

### 2025-10-14 paper/paper.Rmd
correct typos in model specifications (and a couple other small things)

@@
 To characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008].
-Reaction times were computed only on trials for which the child was fixating the distractor at the point of disambiguation (label onset), and were defined as the time from label onset to the first fixation on the target image (see S2 Reaction Times for further details on how reaction times were computed in S2.1 and discussion of issues surrounding distinguishing "correct" vs.
+Reaction times were computed only on trials for which the child was fixating the distractor at the point of disambiguation (label onset), and were defined as the time from label onset to the first fixation on the target image (see S2 Reaction Times, including further details on how reaction times were computed in S2.1 and discussion of issues surrounding distinguishing "correct" vs.
 "incorrect" trials when computing looking-based reaction times in S2.2).
@@
 Our first question was about the functional form of the relationships between age, speed, and accuracy (see S5 Pairwise Correlations of Main Measures for raw pairwise correlations between variables).
-We began by fitting linear mixed-effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age nested within study (modeling item and procedural variation across studies) and random intercepts by participant.
+We began by fitting linear mixed-effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age nested within study (modeling item and procedural variation across studies) and random intercepts by participant (see S8 Mixed-effects model specifications for further details on these specifications).
 We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT (see S3 Checks on Data Distributional Assumptions for further analyses and discussion of these modeling choices).

### 2025-10-14 paper/supplemental_information.Rmd
correct typos in model specifications (and a couple other small things)

@@
 
-```{r rt-hist, fig.cap="\\label{fig:rt-hist} Distribution of RT overlaid with log normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=5}
+```{r rt-hist, fig.cap="\\label{fig:rt-hist} Distribution of RT overlaid with a log normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=5}
 
@@
 
-Table \ref{tab:accmods} shows model comparison measures between for different models of the functional form of the relationship between accuracy and age and Table \ref{tab:rtmods} shows the same for reaction time.
+Table \ref{tab:accmods} shows model comparison measures for different models of the functional form of the relationship between accuracy and age and Table \ref{tab:rtmods} shows the same for reaction time.
 Age gradients are estimated substantially better with long window accuracies.
@@
 
-With our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns unfortunately, because we have insufficient data about individuals' improvement across time.
+With our current dataset, we unfortunately cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns, because we have insufficient data about individuals' improvement across time.
 Thus, our current results apply to the form of the age gradient as opposed to the form of any individual's pattern of developmental change.
@@
 \begin{verbatim}
-long_window_accuracy ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)
+long_window_accuracy ~ age_s + (age_s | dataset_name) + (1 | subject_id)
 long_window_accuracy ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)

### 2025-10-14 paper/paper.Rmd
add detail on vanilla trials, fix figure formatting in main and SI references

@@
 Individual longitudinal trajectories are shown in light gray.
-Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, fig.width=6, out.width=".6\\textwidth", include=TRUE}
+Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, include=TRUE}
 
@@
 Dataset information is given in Table \ref{tab:datasets}.
-Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed only data from simple word recognition trials; these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations.
+Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed only data from standard, simple word recognition trials; these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations.
+Requirements for being considered a standard word recognition trial included that (a) the target word was familiar (also no part-words); (b) the target word was the first point of disambiguation; (c) the target word was embedded in a well-formed, grammatical carrier phrase; (d) there was no informative language presented prenominally (e.g.
+semantically informative verbs, adjectives); (e) there were no nonsense words presented anywhere during the trial (including the carrier phrase); (f) there was no language-, speaker-, or accent-switching within trial; (g) the distractor image was unrelated to both the target label and the target image; and (h) there was no phonological overlap between the distractor label and target label.
 We focus here on English purely for practical reasons -- the Peekbank dataset at present contains limited data from other languages.
@@
 We used `lme4` to fit linear mixed-effects models, `brms` to fit non-linear growth models, and `lavaan` to fit structural equation models.
-Random effects structures for each model are given in text; full model specifications are available in the Supplemental Information and in the reproducible code for this paper, available in the linked repository.
+Random effects structures for each model are given in text; full model specifications are available in the Supplemental Information (S8, S9, and S12) and in the reproducible code for this paper, available in the linked repository.
 To aid interpretability, all variables were standardized (z-scored) prior to inclusion in structural equation models.

### 2025-10-14 paper/supplemental_information.Rmd
add detail on vanilla trials, fix figure formatting in main and SI references

@@
 Individual longitudinal trajectories are shown in light gray.
-Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, fig.width=6, out.width=".6\\textwidth", fig.pos="h!"}
+Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4}
 
@@
 We believe that they are, and we show this statistically using our mixed-effects and non-linear mixed-effects models, which control for dataset-related variation.
-(We also show dataset-level effects in a number of our visualizations for this same reason).
+We also show dataset-level effects in a number of our visualizations for this same reason.
 The same degree of random effect specification that we can do in the mixed-effects models is not possible in the SEM model, however, purely for technical reasons.
@@
 
-Baayen, R. H., & Milin, P. (2010). Analyzing reaction times. International journal of psychological research, 3(2), 12-28.
-
-Creel, S. (2024). Connecting the tots: Strong looking-pointing correlations in preschoolers' word learning and implications for continuity in language development. Child Development, 96, 87-103.
-
-Logan, G. D. (1988). Toward an instance theory of automatization. Psychological review, 95(4), 492.
-
-Luce, P. A. (1986). A computational analysis of uniqueness points in auditory word recognition. Perception & Psychophysics, 39(3), 155-158.
-
-Ratcliff, R. (1979). Group reaction time distributions and an analysis of distribution statistics. Psychological bulletin, 86(3), 446.
-
-Ratcliff, R. (1993). Methods for dealing with reaction time outliers. Psychological bulletin, 114(3), 510.
-
-Van Zandt, T. (2000). How to fit a response time distribution. Psychonomic bulletin & review, 7(3), 424-465.
-
-Van Zandt, T. (2002). Analysis of response time distributions. Stevens’ handbook of experimental psychology, 4, 461-516.
+Baayen, R. H., & Milin, P. (2010). Analyzing reaction times. *International Journal of Psychological Research, 3*(2), 12-28.
+
+Creel, S. (2024). Connecting the tots: Strong looking-pointing correlations in preschoolers' word learning and implications for continuity in language development. *Child Development, 96*, 87-103.
+
+Logan, G. D. (1988). Toward an instance theory of automatization. *Psychological Review, 95*(4), 492–527.
+
+Luce, P. A. (1986). A computational analysis of uniqueness points in auditory word recognition. *Perception & Psychophysics, 39*(3), 155-158.
+
+Ratcliff, R. (1979). Group reaction time distributions and an analysis of distribution statistics. *Psychological bulletin, 86*(3), 446–461.
+
+Ratcliff, R. (1993). Methods for dealing with reaction time outliers. *Psychological Bulletin, 114*(3), 510-532.
+
+Van Zandt, T. (2000). How to fit a response time distribution. *Psychonomic Bulletin & Review, 7*(3), 424-465.
+
+Van Zandt, T. (2002). Analysis of response time distributions. *Stevens’ Handbook of Experimental Psychology, 4*, 461-516.

### 2025-10-16 paper/supplemental_information.Rmd
minor changes to allow for rendering

@@
 library(papaja)
+# library(fitdistrplus)
+# library(statmod)
+# library(gamlss.dist)
+# library(semoutput)
 
 #set to TRUE to fit the Bayesian models
-FIT_BAYES = FALSE
+FIT_BAYES = TRUE
 ```

### 2025-10-16 supplemental_data/creel_2024/Step2aCorrsNDataPrep.R

@@ -25,8 +25,6 @@ library(dplyr)
 library(lme4)
 library(ggplot2) # underutilized right now
 library(readr)
-library(here)
-read_path <- here("supplemental_data","creel_2024")
 
 
 #setwd("~/Documents/current studies/A0 priority/EyeMagic/newRscripts")
@@ -60,15 +58,15 @@ whichCorrsToGenerate=1
 # Read in selected dataset
 ##################################
 if (readthisone=="n"){
-  dat=read.csv(here(read_path,'allnovels.csv')) ## allnovels is generated by a long ugly script that is omitted in shared version.
+  dat=read.csv('allnovels.csv') ## allnovels is generated by a long ugly script that is omitted in shared version.
 } else if (readthisone=="f"){
-  dat=read.csv(here(read_path,'allfamiliars.csv')) ## allfamiliars is similarly generated by a long ugly script that is omitted in shared version. 
+  dat=read.csv('allfamiliars.csv') ## allfamiliars is similarly generated by a long ugly script that is omitted in shared version. 
 } else if (readthisone=="w"){
-  dat=read.csv(here(read_path,'allnovels.csv'));dat2=read.csv(here(read_path,'allfamiliars.csv'));
+  dat=read.csv('allnovels.csv');dat2=read.csv('allfamiliars.csv');
   dat=rbind(select(dat,-Subsub),dat2) # this gives you novels and fam-word trials 
   dat= dat %>% filter(ExptSubcat %in% c("wordlearning","familiarwords"))
 } else if (readthisone=="a"){
-  dat=read.csv(here(read_path,'allnovels.csv'));dat2=read.csv(here(read_path,'allfamiliars.csv'));
+  dat=read.csv('allnovels.csv');dat2=read.csv('allfamiliars.csv');
   dat=rbind(select(dat,-Subsub),dat2) # this gives you novels and fam-word trials
 } else {print("Did not recognize selection, restart.")
 }
@@ -453,8 +451,8 @@ windfToSave=windf %>% mutate(groupis=reftype,lowerlimit=lowerlimit,upperlimit=up
                              useacc=useacc,accmin=accmin,accmax=accmax,
                              lookaccfilter=lookaccfilter,lookfiltermode=lookfiltermode,
                              movwinsize=movwinsize,stepsize=stepsize,totalsubs=totalsubs,date=date())
-#MZ commented out
[…]
